---
title: "NM_Envr_Data"
output: pdf_document
date: "2024-03-21"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Set working directory
```{r}
#set working directory to Rmd file location
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```


```{r}
#read in appropriate packages
library(dplyr)
library(readr)
library(forcats)
library(RColorBrewer)
library(circlize)
library(BiocManager)
library(ComplexHeatmap)
library(vegan)#for Mantel test
```

All the data for each site are in separate csv files. There is one temp and one salinity csv for each site. We need to combine those files into one salinity data frame and one temperature data frame, and then join the salinity and temperature data frames into one to conduct the distance analysis.

## Combining temp and salinity csvs respectively from each site
```{r}
#set working directory to temperature files location
setwd("../../data/envr_of_origin/Temp")
# read in all temp files as csvs and combine into one file named "temp"
temp <- list.files(path=".") %>% 
  lapply(read_csv) %>% 
  bind_rows

#View temp file
View(temp)
#set working directory to Salinity folder
setwd("../Salinity")
# read in all salinity files as csvs and combine into one file named "sal"
sal <- list.files(path=".") %>% 
  lapply(read_csv) %>% 
  bind_rows

View(sal)

#set working directory to location where code is saved
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))

#remove extra column at the end of each data frame
ncol(temp) #ncol = 15
temp <- temp[ , -c(15)]

ncol(sal) #ncol = 15
sal <- sal[ , -c(15)]
```

## Join temp and sal data frames
```{r}
#left join temp and salinity dfs and save into an object called "envr"
envr <-temp %>%
          left_join(sal, by = "site_name")

#remove rows unnecessary for distance analysis and save into an object called "envr_sub"
envr_sub <- subset(envr, select = -c(download_date.x, firstyear.x, finalyear.x, source_description.x, lat.x, lon.x, Temperature_n, Temperature_years, collection_type.x, download_date.y, source_description.y, lat.y, lon.y, firstyear.y, finalyear.y, Salinity_n, Salinity_years, collection_type.y))

# make envr_sub into a data frame and save as "envr_df"
envr.df <- data.frame(envr_sub)

#update site names with labeling scheme
envr.df["site_name"][envr.df["site_name"] == "TX2"] <- "W1-TX"
envr.df["site_name"][envr.df["site_name"] == "LA4"] <- "W2-LA"
envr.df["site_name"][envr.df["site_name"] == "FL3"] <- "W3-FL"
envr.df["site_name"][envr.df["site_name"] == "LOLA"] <- "S1-LOLA"
envr.df["site_name"][envr.df["site_name"] == "DEBY"] <- "S2-DEBY"
envr.df["site_name"][envr.df["site_name"] == "JR"] <- "W4-VA"
envr.df["site_name"][envr.df["site_name"] == "NH1"] <- "W5-NH"
envr.df["site_name"][envr.df["site_name"] == "ME2"] <- "W6-ME"

#reorder rows of data frame
envr.df <- envr.df %>% 
  arrange(factor(site_name, levels = c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME")))

#save envr.df as a csv for future analyses
write.csv(envr.df, "../../data/envr_of_origin/envr_summary_stats.csv")
          
#rename rows with MVP population names
rownames(envr.df) <- c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME")
#remove site names from data frame and save as object called "envr.df.dist"
envr.df.dist <- subset(envr.df, select = -c(site_name))
```

## Run distance analysis and visualize with a heat map
```{r}
#make run distance analysis on envr.df.dist and save the results as a matrix named "envr.dist.mat"
envr.dist.mat <- as.matrix(dist(envr.df.dist))

#save envr distance matrix as csv
write.csv(envr.dist.mat, "../../data/envr_of_origin/envr_dist.csv")

#View distance matrix
View(envr.dist.mat)
```
