---
title: "envr_quantiles"
author: "Nicole Mongillo"
output: pdf_document
date: "2024-10-08"
---

This file compiles temperature and salinity data from all environment-of-origin (EoO) sites and common garden field sites (LEW, YRK), computes quantile summaries (10th and 90th percentiles), and produces visualization panels for Figure 1 of H2F: violin plots with boxplots for salinity distributions (Fig 1B, 1C) and single-row heatmaps for temperature and salinity quantiles at both EoO and field sites. The main data product is `envr_quantiles_allsites.csv`, used by downstream modeling scripts (Q2_envr_quantile_models_field.Rmd, Q2_condition_index_models_field.Rmd, Q2_models_hatchery_nursery.Rmd).

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory relative to project root
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("tidyverse", "here")
pacman::p_load(char = packages)

# log versions for reproducibility
r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

## Load data

First, all temperature and salinity summary files need to be combined into one large summary file for all sites. It is important to run this AFTER all individual site EnvrData.Rmd scripts have been run. This replaces the append = TRUE approach and avoids duplicate rows.

```{r}
# define all site temperature CSV paths
temp_files <- c(
  "data/environment/W1-TX/W1-TX_temperature.csv",
  "data/environment/W2-LA/W2-LA_temperature.csv",
  "data/environment/W3-FL/W3-FL_temperature.csv",
  "data/environment/W4-VA/W4-VA_temperature.csv",
  "data/environment/W5-NH/W5-NH_temperature.csv",
  "data/environment/W6-ME/W6-ME_temperature.csv",
  "data/environment/S1-LOLA/S1-LOLA_temperature.csv",
  "data/environment/S2-DEBY/S2-DEBY_temperature.csv",
  "data/environment/Lewisetta/LEW_temperature.csv",
  "data/environment/York-River/YRK_temperature.csv"
)

# define all site salinity CSV paths
sal_files <- c(
  "data/environment/W1-TX/W1-TX_salinity.csv",
  "data/environment/W2-LA/W2-LA_salinity.csv",
  "data/environment/W3-FL/W3-FL_salinity.csv",
  "data/environment/W4-VA/W4-VA_salinity.csv",
  "data/environment/W5-NH/W5-NH_salinity.csv",
  "data/environment/W6-ME/W6-ME_salinity.csv",
  "data/environment/S1-LOLA/S1-LOLA_salinity.csv",
  "data/environment/S2-DEBY/S2-DEBY_salinity.csv",
  "data/environment/Lewisetta/LEW_salinity.csv",
  "data/environment/York-River/YRK_salinity.csv"
)

# read and combine all temperature summaries
all_temperature <- temp_files %>%
  lapply(function(f) {
    if (file.exists(f)) {
      read.csv(f, stringsAsFactors = FALSE)
    } else {
      warning(paste("File not found:", f))
      return(NULL)
    }
  }) %>%
  bind_rows()

# read and combine all salinity summaries
all_salinity <- sal_files %>%
  lapply(function(f) {
    if (file.exists(f)) {
      read.csv(f, stringsAsFactors = FALSE)
    } else {
      warning(paste("File not found:", f))
      return(NULL)
    }
  }) %>%
  bind_rows()

# write combined files (overwrites, no append)
write.csv(all_temperature, "data/environment/all_temperature.csv", row.names = FALSE)
write.csv(all_salinity, "data/environment/all_salinity.csv", row.names = FALSE)

cat("Combined", nrow(all_temperature), "temperature rows and", nrow(all_salinity), "salinity rows.\n")
cat("Expected: 10 rows each (one per site).\n")
```

All `_full` files now have a standardized structure: `site_name`, `datetime`, and either `temp` or `salinity`. File paths point to the `data/environment/` directory with site-specific subdirectories.

```{r}
# read in salinity _full files from standardized locations
TX_sal <- read.csv("data/environment/W1-TX/W1-TX_sal_full.csv")
LA_sal <- read.csv("data/environment/W2-LA/W2-LA_sal_full.csv")
FL_sal <- read.csv("data/environment/W3-FL/W3-FL_sal_full.csv")
VA_sal <- read.csv("data/environment/W4-VA/W4-VA_sal_full.csv")
LOLA_sal <- read.csv("data/environment/S1-LOLA/S1-LOLA_sal_full.csv")
DEBY_sal <- read.csv("data/environment/S2-DEBY/S2-DEBY_sal_full.csv")
NH_sal <- read.csv("data/environment/W5-NH/W5-NH_sal_full.csv")
ME_sal <- read.csv("data/environment/W6-ME/W6-ME_sal_full.csv")
LEW_sal <- read.csv("data/environment/Lewisetta/LEW_sal_full.csv")
YRK_sal <- read.csv("data/environment/York-River/YRK_sal_full.csv")

# read in data frames with salinity and temp summary stats (including quantiles) for each eoo and field site
temp_all <- read.csv("data/environment/all_temperature.csv")
sal_all <- read.csv("data/environment/all_salinity.csv")
```

## Data preparation

### POSIXct date time
```{r}
# all _full files write datetime in the same format from write.csv, so we can use a single conversion
# write.csv outputs POSIXct as "%Y-%m-%d %H:%M:%S" or "%Y-%m-%d" for date-only
all_sal_dfs <- list(TX_sal, LA_sal, FL_sal, VA_sal, LOLA_sal, DEBY_sal, NH_sal, ME_sal, LEW_sal, YRK_sal)
all_sal_names <- c("TX_sal", "LA_sal", "FL_sal", "VA_sal", "LOLA_sal", "DEBY_sal", "NH_sal", "ME_sal", "LEW_sal", "YRK_sal")

for (i in seq_along(all_sal_dfs)) {
  all_sal_dfs[[i]]$datetime <- as.POSIXct(all_sal_dfs[[i]]$datetime, tz = "UTC")
  assign(all_sal_names[i], all_sal_dfs[[i]])
}
```

### Merge and clean data
```{r}
# merge all salinity data frames — all have the same columns (site_name, datetime, salinity)
full_sal <- bind_rows(TX_sal, LA_sal, FL_sal, VA_sal, LOLA_sal, DEBY_sal, NH_sal, ME_sal, LEW_sal, YRK_sal)

# make all salinity variables numeric
full_sal$salinity <- as.numeric(full_sal$salinity)

# check for NAs
na_full_sal <- is.na(full_sal)
summary(na_full_sal)
# remove NAs
full_sal <- na.omit(full_sal) 

# filter out salinity values below 0 and above 44 (this already happens in each separate file - most are above 40 but W1-TX is high salinity so it gets filtered with 44 as the upper bound)
filtered_sal <- full_sal %>%
    filter(between(salinity, 0, 44)) 

# convert site_name to factor with desired level order (names already hyphenated from EnvrData scripts)
filtered_sal$site_name <- factor(filtered_sal$site_name, 
  levels = c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME", "LEW", "YRK"))
```

### Data product: Quantiles
```{r}
# combine quantile data from the separate temp and salinity data frames
# site names are already in hyphenated format from the EnvrData scripts

# select relevant columns and join
temp_quant <- temp_all %>% 
  select(site_name, temp_quantile_10, temp_quantile_90, Temperature_st_dev)
sal_quant <- sal_all %>% 
  select(site_name, salinity_quantile_10, salinity_quantile_90, Salinity_st_dev)
quantiles <- left_join(sal_quant, temp_quant, by = "site_name")

# calculate mean salinity and save in quantiles
mean_sal <- filtered_sal %>% 
  aggregate(salinity ~ site_name, FUN = mean)
quantiles <- left_join(quantiles, mean_sal, by = "site_name")

# reorder levels
quantiles$site_name <-
  factor(quantiles$site_name, levels = c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME", "LEW", "YRK"))

quantiles <- quantiles[order(quantiles$site_name),]

# write to csv
write.csv(quantiles, file = "results/environment/envr_quantiles_allsites.csv", row.names = FALSE)
```

### Subset quantiles for EOO and field sites
```{r}
# filter out YRK and LEW from quantiles
quant_eoo <- quantiles %>% 
  filter(site_name != "YRK" & site_name != "LEW")
# quantiles for only YRK and LEW
quant_field <- quantiles %>% 
  filter(site_name == "YRK" | site_name == "LEW")
```

## Figures

### Environment of origin salinity violin plot
```{r}
eoo_sal_violin <- filtered_sal %>% 
  filter(site_name != "YRK" & site_name != "LEW") %>% 
  ggplot(aes(site_name, salinity)) +
  geom_violin() +
  geom_boxplot(width = .1) +
  geom_errorbar(data = quant_eoo, aes(ymin = salinity_quantile_10, ymax = salinity_quantile_90), width=.3, color = "blue")+
  ggtitle("Environments-of-origin")+
  ylab("Salinity (ppt)")+
  ylim(c(0,44))+
  scale_x_discrete(name = "Group", limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.ticks.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.title.x = element_blank(),
        axis.text.y = element_text(color = "black", size = 10),
        axis.title.y = element_text(size = 12))

eoo_sal_violin

ggsave("Fig1B_eoo_salinity_violin.pdf",
       plot = eoo_sal_violin,
       device = "pdf",
       path = "figures/envr_of_origin",
       height = 4,
       width = 9.25)
```

### Common gardens salinity violin plot
```{r}
field_sal_violin <- filtered_sal %>% 
  filter(site_name == "YRK" | site_name == "LEW") %>% 
  ggplot(aes(site_name, salinity)) +
  geom_violin() +
  geom_boxplot(width = .1) +
  geom_errorbar(data = quant_field, aes(ymin = salinity_quantile_10, ymax = salinity_quantile_90), width=.3, color = "blue")+
  ggtitle("Common garden sites")+
  ylab("Salinity (ppt)")+
  ylim(c(0, 44))+
  scale_x_discrete(limits = c("LEW", "YRK"))+
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank())

field_sal_violin

ggsave("Fig_1C_field_salinity_violin.pdf",
       plot = field_sal_violin,
       device = "pdf",
       path = "figures/envr_of_origin",
       height = 4,
       width = 2.5)
```

### EoO 0.1 salinity quantile heatmap
```{r}
eoo_sal_10_heatmap <- ggplot(quant_eoo, aes(site_name, 1, fill = salinity_quantile_10)) +
  geom_tile()+
  geom_text(aes(label = round(salinity_quantile_10, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "G", limits = c(0,38), direction=-1)+
  scale_x_discrete(limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  labs(fill = "Salinity (ppt)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = "none")

eoo_sal_10_heatmap
```

### EoO 0.9 salinity quantile heatmap
```{r}
eoo_sal_90_heatmap <- ggplot(quant_eoo, aes(site_name, 1, fill = salinity_quantile_90)) +
  geom_tile()+
  geom_text(aes(label = round(salinity_quantile_90, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "G", limits = c(0,38), direction = -1)+
  scale_x_discrete(limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  labs(fill = "Salinity (ppt)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = "none")

eoo_sal_90_heatmap
```

### EoO 0.1 temperature quantile heatmap
```{r}
eoo_temp_10_heatmap <- ggplot(quant_eoo, aes(site_name, 1, fill = temp_quantile_10)) +
  geom_tile()+
  geom_text(aes(label = round(temp_quantile_10, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "C", limits = c(5,32))+
  scale_x_discrete(limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  labs(fill = "Temperature °C ") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = "none")

eoo_temp_10_heatmap

ggsave("Fig1B_eoo_temp10_heatmap.pdf",
       plot = eoo_temp_10_heatmap,
       device = "pdf",
       path = "figures/envr_of_origin",
       height = 1.9,
       width = 9)
```

### EoO 0.9 temperature quantile heatmap
```{r}
eoo_temp_90_heatmap <- ggplot(quant_eoo, aes(site_name, 1, fill = temp_quantile_90), height = 2, width = 7) +
  geom_tile()+
  geom_text(aes(label = round(temp_quantile_90, digits = 1)), size = 5)+
  scale_fill_viridis_c(option = "C", limits = c(5,32)) +
  scale_x_discrete(limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  labs(fill = "Temperature °C ") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black"), 
       axis.ticks.y = element_blank(), 
       axis.text.y = element_blank(), 
       axis.title.y = element_blank(), 
       panel.grid = element_blank(), 
       legend.position = "none")

eoo_temp_90_heatmap

ggsave("Fig1B_eoo_temp_90_heatmap.pdf",
       plot = eoo_temp_90_heatmap,
       device = "pdf",
       path = "figures/envr_of_origin",
       height = 1.9,
       width = 9)
```

### Common gardens 0.1 salinity quantile heatmap
```{r}
field_sal_10_heatmap <- ggplot(quant_field, aes(site_name, 1, fill = salinity_quantile_10)) +
  geom_tile()+
  geom_text(aes(label = round(salinity_quantile_10, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "G", limits = c(0,38), direction = -1) +
  scale_x_discrete(limits = c("LEW", "YRK"))+
  labs(fill = "Salinity (ppt)") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1), 
         axis.ticks.y = element_blank(), 
         axis.text.y = element_blank(), 
         axis.title.y = element_blank(), 
         panel.grid = element_blank(), 
         legend.key.size = unit(.4, 'cm'), 
         legend.title = element_text(size = 10), 
         legend.text = element_text(size = 10))
field_sal_10_heatmap
```

### Common gardens 0.9 salinity quantile heatmap
```{r}
field_sal_90_heatmap <- ggplot(quant_field, aes(site_name, 1, fill = salinity_quantile_90)) +
  geom_tile()+
  geom_text(aes(label = round(salinity_quantile_90, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "G", limits = c(0,38), direction = -1) +
  scale_x_discrete(limits = c("LEW", "YRK"))+
  labs(fill = "Salinity (ppt)") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1), 
         axis.ticks.y = element_blank(), 
         axis.text.y = element_blank(), 
         axis.title.y = element_blank(), 
         panel.grid = element_blank(), 
         legend.key.size = unit(.4, 'cm'), 
         legend.title = element_text(size = 10), 
         legend.text = element_text(size = 10))
field_sal_90_heatmap
```

### Common gardens 0.1 temperature quantile heatmap
```{r}
field_temp_10_heatmap <- ggplot(quant_field, aes(site_name, 1, fill = temp_quantile_10)) +
  geom_tile()+
  geom_text(aes(label = round(temp_quantile_10, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "C", limits = c(5,32)) +
  scale_x_discrete(limits = c("LEW", "YRK"))+
  labs(fill = "Temp °C ") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1), 
         axis.ticks.y = element_blank(), 
         axis.text.y = element_blank(), 
         axis.title.y = element_blank(), 
         panel.grid = element_blank(), 
         legend.key.size = unit(.4, 'cm'), 
         legend.title = element_text(size = 10), 
         legend.text = element_text(size = 10))
field_temp_10_heatmap

ggsave("Fig1C_field_temp10_heatmap.pdf",
       plot = field_temp_10_heatmap,
       device = "pdf",
       path = "figures/envr_of_origin",
       height = 1.7,
       width = 3.4)
```

### Common gardens 0.9 temperature quantile heatmap
```{r}
field_temp_90_heatmap <- ggplot(quant_field, aes(site_name, 1, fill = temp_quantile_90), height = 2, width = 7) +
  geom_tile()+
  geom_text(aes(label = round(temp_quantile_90, digits = 1)), size = 5)+
  scale_fill_viridis_c(option = "C", limits = c(5,32)) +
  scale_x_discrete(limits = c("LEW", "YRK"))+
  labs(fill = "Temp °C ") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.key.size = unit(.4, 'cm'), 
        legend.title = element_text(size = 10), 
        legend.text = element_text(size = 10))

field_temp_90_heatmap

ggsave("Fig1C_field_temp_90_heatmap.pdf",
       plot = field_temp_90_heatmap,
       device = "pdf",
       path = "figures/envr_of_origin",
       height = 1.7,
       width = 3.4)
```