---
title: "envr_dist_quantiles"
author: "Kiran Bajaj"
output: pdf_document
date: "25-10-19"
---

This analysis calculates pairwise environmental distances between sites based on scaled 
temperature and salinity quantiles (10th and 90th quantiles).
This produces two distance matrices: one with all sites + one excluding field sites.

## Edit history:
 - 2025-11-10: NM updated environmental quantiles

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory relative to project root
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("tidyverse", "here")
pacman::p_load(char = packages)

# log versions for reproducibility
r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

## Global variables
```{r config}
# site ordering
POP_ORDER <- c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME")
FIELD_SITES <- c("LEW", "YRK")
ALL_SITES <- c(POP_ORDER, FIELD_SITES)

# columns used for distance calculation (10th/90th quantiles for temp and salinity)
DIST_COLS <- c("salinity_quantile_10", "salinity_quantile_90", "temp_quantile_10", "temp_quantile_90")
```

## Helper functions
```{r helpers}
# calculate scaled Euclidean distance matrix from environmental data
# scaling is done via z-score normalization (scale()) so that temperature and salinity contribute equally regardless of raw units/ranges. Note that scaling is computed independently for each input, so the 8-site and 10-site matrices use slightly different z-scores.

#' @param df data frame with site_name and environmental variables
#' @param cols columns to use for distance calculation
#' @param sites vector of site names for row/column labels
calc_envr_dist <- function(df, cols, sites) {
  df %>%
    select(all_of(cols)) %>%
    scale() %>%
    dist() %>%
    as.matrix() %>%
    `rownames<-`(sites) %>%
    `colnames<-`(sites)
}
```

## Data preparation
```{r load-data}
# read quantiles produced by envr_quantiles.Rmd
# includes SD, 10th, and 90th quantiles for temp/salinity
# covers 8 environment-of-origin sites + 2 field experiment sites
envr_quants_raw <- read_csv("results/environment/envr_quantiles_allsites.csv",
                            show_col_types = FALSE)

# create ordered versions
envr_all <- envr_quants_raw %>% arrange(factor(site_name, levels = ALL_SITES))
envr_origin <- envr_all %>% filter(!site_name %in% FIELD_SITES)
```

## Run distance analysis
```{r distance-matrices}
# all sites (including field experiment locations)
envr_dist_all <- calc_envr_dist(envr_all, DIST_COLS, ALL_SITES)

# environment-of-origin sites only (excluding field sites)
envr_dist_origin <- calc_envr_dist(envr_origin, DIST_COLS, POP_ORDER)
```

## Save results
```{r export}
# distance matrix with all sites
write.csv(envr_dist_all, "results/environment/envr_quantiles_dist_all.csv")

# distance matrix for environment-of-origin sites only for downstream analysis
saveRDS(envr_dist_origin, "results/environment/envr_quantiles_dist_origin.rds")

# display matrices
envr_dist_all
envr_dist_origin
```