---
title: "EoO_Field_Envr_Figures"
output: pdf_document
date: "2024-10-08"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
#set working directory
setwd("~/Desktop/GitHub/MVP-H2F-HatcheryField/src/envr_data")
```

### Packages
```{r packages}
library(dplyr) #Used for working with data frames
library(lubridate) #Used for time-date conversions
library(readr) #Used to read the CSV file
library(stringr) #for relpacing parts of a string
library(ggplot2) #plot with ggplot
library(cowplot) #arrange ggplots
library(ggpmisc)
library(car) #Anova function
```

### Load Data
```{r}
#Read in salinity files
TX_sal <- read.csv("../../data/envr_of_origin/full_sal/TX_sal_full.csv")
#remove extra empty columns
TX_sal <- subset(TX_sal, select = c(site_name, datetime, salinity))

LA_sal <- read.csv("../../data/envr_of_origin/full_sal/LA_sal_full.csv")

FL_sal <- read.csv("../../data/envr_of_origin/full_sal/FL_sal_full.csv")

LOLA_sal <- read.csv("../../data/envr_of_origin/full_sal/LOLA_sal_full.csv")

DEBY_sal <- read.csv("../../data/envr_of_origin/full_sal/DEBY_sal_full.csv")

VA_sal <- read.csv("../../data/envr_of_origin/full_sal/VA_sal_full.csv")
#remove uncorrected salinity column and rename corrected_salinity to salinity
VA_sal <- subset(VA_sal, select = -c(salinity))
colnames(VA_sal) <- c("site_name", "datetime", "salinity")

NH_sal <- read.csv("../../data/envr_of_origin/full_sal/NH_sal_full.csv")
#remove extra empty columns
NH_sal <- subset(NH_sal, select = c(site_name, datetime, salinity))

ME_sal <- read.csv("../../data/envr_of_origin/full_sal/ME_sal_full.csv")

YRK_sal <- read.csv("../../data/envr_of_origin/full_sal/YRK_sal_full.csv")[, 2:4]

#Read in data frames with salinity and temp summary stats (including quantiles) for each eoo and field site
temp_all <- read.csv("../../data/environment/all_temperature.csv")
sal_all <- read.csv("../../data/environment/all_salinity.csv")
```

### Data preparation

#### POSIXct date time
```{r}
#convert all date times to POSIXct
TX_sal$datetime <- as.POSIXct(TX_sal$datetime, "%m/%d/%y %H:%M", tz = "")

LA_sal$datetime <- as.POSIXct(LA_sal$datetime, "%Y-%m-%d %H:%M:%S", tz = "")

FL_sal$datetime <- as.POSIXct(FL_sal$datetime, "%m/%d/%y %H:%M", tz = "")

LOLA_sal$datetime <- as.POSIXct(LOLA_sal$datetime, "%Y-%m-%d %H:%M:%S", tz = "")

DEBY_sal$datetime <- as.POSIXct(DEBY_sal$datetime, "%Y-%m-%d %H:%M:%S", tz = "")

VA_sal$datetime <- as.POSIXct(VA_sal$datetime, "%Y-%m-%d %H:%M:%S", tz = "")

NH_sal$datetime <- as.POSIXct(NH_sal$datetime,"%m/%d/%y %H:%M", tz = "")

ME_sal$datetime <- as.POSIXct(ME_sal$datetime, "%m/%d/%y %H:%M", tz = "")
```

#### Field environmental data
```{r}
#S1-LOLA and the Lewisetta field site use the same environmental data, but the field site data should only include the time frame of the field experiment (August 2023 - November 2024). Make a copy of the LOLA_sal df, rename it LEW_sal, and filter the time frame.
#currently subsetting dates as 08-01-2023- 11-20-2024
LEW_sal <- LOLA_sal %>% 
  filter(datetime>= 1690862400 & datetime <= 1732078800)
LEW_sal["site_name"][LEW_sal["site_name"] == "LOLA"] <- "LEW"
```

#### Merge and clean data
```{r}
#merge data frames
full_sal <- rbind(TX_sal, LA_sal,  FL_sal, LOLA_sal, DEBY_sal, VA_sal, NH_sal, ME_sal, LEW_sal, YRK_sal)

#make all salinity variables numeric
full_sal$salinity <- as.numeric(full_sal$salinity)

#check for NAs
na_full_sal <- is.na(full_sal) # store our NAs in a variable
summary(na_full_sal) # we have 553 NAs in datetime and 192380 NAs in salinity that are stored as "TRUE" 
#remove NAs using na.omit
full_sal <- na.omit(full_sal) 

#filter out salinity values below 0 and above 40
filtered_sal <- full_sal %>%
    filter(between(salinity, 0, 40)) 

filtered_sal <- filtered_sal %>%
  mutate(site_name = recode_factor(site_name,
                                   DEBY = "S2-DEBY",
                                   FL = "W3-FL",
                                   LA = "W2-LA",
                                   LOLA = "S1-LOLA",
                                   ME = "W6-ME",
                                   NH = "W5-NH",
                                   TX = "W1-TX",
                                   W4_VA = "W4-VA"))
```

#### Data product: Quantiles
```{r}
#combine quantile data from the separate temp and salinity data frames
#replace underscores in site names with hyphens
temp_all$site_name <- str_replace(temp_all$site_name, "_", "-")
sal_all$site_name <- str_replace(sal_all$site_name, "_", "-")

#select relevant columns and join
temp_quant <- subset(temp_all, select = c(site_name, temp_quantile_10, temp_quantile_90, Temperature_st_dev))
sal_quant <-subset(sal_all, select = c(site_name, salinity_quantile_10, salinity_quantile_90, Salinity_st_dev))
quantiles <- left_join(sal_quant, temp_quant, by = "site_name")

#calculate mean salinity and save in quantiles
mean_sal <- filtered_sal %>% 
  aggregate(salinity ~ site_name, FUN = mean)
quantiles <- left_join(quantiles, mean_sal, by = "site_name")

#reorder levels
quantiles$site_name <-
  factor(quantiles$site_name, levels = c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME", "LEW", "YRK"))

quantiles <- quantiles[order(quantiles$site_name),]

#write to csv
write.csv(quantiles, file = "../../data/environment/quantiles_05232025.csv")

```

### Subset quantiles for EOO and field sites
```{r}
#filter out YRK and LEW from quantiles
quant_eoo <- quantiles %>% 
  filter(site_name != "YRK" & site_name != "LEW")
#quantiles for only YRK and LEW
quant_field <- quantiles %>% 
  filter(site_name == "YRK" | site_name == "LEW")
```

### Figure: Environment of origin salinity violin plot
```{r}
eoo_sal_violin <- filtered_sal %>% 
  filter(site_name != "YRK" & site_name != "LEW") %>% 
  ggplot(aes(site_name, salinity)) +
  geom_violin() +
  geom_boxplot(width = .1) +
  geom_errorbar(data = quant_eoo, aes(ymin = salinity_quantile_10, ymax = salinity_quantile_90), width=.3, color = "blue")+
  ggtitle("Environments-of-Origin")+
  ylab("Salinity (ppt)")+
  ylim(c(0,40))+
  scale_x_discrete(name = "Group", limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.ticks.x = element_blank(), 
        axis.text.x = element_blank(), 
        axis.title.x = element_blank(),
        axis.text.y = element_text(color = "black", size = 10),
        axis.title.y = element_text(size = 12))

eoo_sal_violin

ggsave("eoo_salinity_violin.pdf",
       plot = eoo_sal_violin,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 4,
       width = 9.25)
```

### Figure: Common gardens salinity violin plot
```{r}
yrk <- filtered_sal %>% 
  filter(site_name == "YRK")
field_sal_violin <- filtered_sal %>% 
  filter(site_name == "YRK" | site_name == "LEW") %>% 
  ggplot(aes(site_name, salinity)) +
  geom_violin() +
  geom_boxplot(width = .1) +
  geom_errorbar(data = quant_field, aes(ymin = salinity_quantile_10, ymax = salinity_quantile_90), width=.3, color = "blue")+
  ggtitle("Common Gardens")+
  ylab("Salinity (ppt)")+
  ylim(c(0, 40))+
  scale_x_discrete(limits = c("LEW", "YRK"))+
  theme_bw()+
  theme(panel.grid.major = element_blank(), 
        panel.grid.minor = element_blank(), 
        axis.ticks.x = element_blank(),
        axis.text.x = element_blank(),
        axis.title.x = element_blank(),
        axis.text.y = element_blank(),
        axis.title.y = element_blank(),
        axis.ticks.y = element_blank())

field_sal_violin

ggsave("field_salinity_violin.pdf",
       plot = field_sal_violin,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 4,
       width = 2.5)
```

###Figure: EoO 0.1 salinity quantile heatmap
```{r temp_heatmap}
eoo_sal_10_heatmap <- ggplot(quant_eoo, aes(site_name, 1, fill = salinity_quantile_10)) +
  geom_tile()+
  geom_text(aes(label = round(salinity_quantile_10, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "G", limits = c(0,38), direction=-1)+  # use viridis color scale
  scale_x_discrete(limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  labs(fill = "Salinity (ppt)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = "none")

eoo_sal_10_heatmap

ggsave("eoo_sal10_heat.pdf",
       plot = eoo_sal_10_heatmap,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.9,
       width = 9)
```

###Figure: EoO 0.9 salinity quantile heatmap
```{r temp_heatmap}
eoo_sal_90_heatmap <- ggplot(quant_eoo, aes(site_name, 1, fill = salinity_quantile_90)) +
  geom_tile()+
  geom_text(aes(label = round(salinity_quantile_90, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "G", limits = c(0,38), direction = -1)+  # use viridis color scale
  scale_x_discrete(limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  labs(fill = "Salinity (ppt)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = "none")

eoo_sal_90_heatmap

ggsave("eoo_sal90_heat.pdf",
       plot = eoo_sal_90_heatmap,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.9,
       width = 9)
```

###Figure: EoO 0.1 temperature quantile heatmap
```{r temp_heatmap}
eoo_temp_10_heatmap <- ggplot(quant_eoo, aes(site_name, 1, fill = temp_quantile_10)) +
  geom_tile()+
  geom_text(aes(label = round(temp_quantile_10, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "C", limits = c(5,32))+  # use viridis color scale
  scale_x_discrete(limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  labs(fill = "Temp 째C ") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = "none")

eoo_temp_10_heatmap

ggsave("eoo_temp10_heat.pdf",
       plot = eoo_temp_10_heatmap,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.9,
       width = 9)
```

###Figure: EoO 0.9 temperature quantile heatmap
```{r}
eoo_temp_90_heatmap <- ggplot(quant_eoo, aes(site_name, 1, fill = temp_quantile_90), height = 2, width = 7) +
  geom_tile()+
  geom_text(aes(label = round(temp_quantile_90, digits = 1)), size = 5)+
  scale_fill_viridis_c(option = "C", limits = c(5,32)) +  # use viridis color scale
  scale_x_discrete(limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  labs(fill = "Temp 째C ") +
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black"), 
       axis.ticks.y = element_blank(), 
       axis.text.y = element_blank(), 
       axis.title.y = element_blank(), 
       panel.grid = element_blank(), 
       legend.position = "none")

eoo_temp_90_heatmap

ggsave("eoo_temp_90_heat.pdf",
       plot = eoo_temp_90_heatmap,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.9,
       width = 9)
```

###Figure: Common gardens 0.1 salinity quantile heatmap
```{r}
field_sal_10_heatmap <- ggplot(quant_field, aes(site_name, 1, fill = salinity_quantile_10)) +
  geom_tile()+
  geom_text(aes(label = round(salinity_quantile_10, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "G", limits = c(0,38), direction = -1) +  # use viridis color scale
  scale_x_discrete(limits = c("LEW", "YRK"))+
  labs(fill = "Salinity (ppt)") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1), 
         axis.ticks.y = element_blank(), 
         axis.text.y = element_blank(), 
         axis.title.y = element_blank(), 
         panel.grid = element_blank(), 
         legend.key.size = unit(.4, 'cm'), 
         legend.title = element_text(size = 10), 
         legend.text = element_text(size = 10))
field_sal_10_heatmap

ggsave("field_sal10_heat.pdf",
       plot = field_sal_10_heatmap,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.7,
       width = 3.4)
```

###Figure: Common gardens 0.9 salinity quantile heatmap
```{r}
field_sal_90_heatmap <- ggplot(quant_field, aes(site_name, 1, fill = salinity_quantile_90)) +
  geom_tile()+
  geom_text(aes(label = round(salinity_quantile_90, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "G", limits = c(0,38), direction = -1) +  # use viridis color scale
  scale_x_discrete(limits = c("LEW", "YRK"))+
  labs(fill = "Salinity (ppt)") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1), 
         axis.ticks.y = element_blank(), 
         axis.text.y = element_blank(), 
         axis.title.y = element_blank(), 
         panel.grid = element_blank(), 
         legend.key.size = unit(.4, 'cm'), 
         legend.title = element_text(size = 10), 
         legend.text = element_text(size = 10))
field_sal_90_heatmap

ggsave("field_sal90_heat.pdf",
       plot = field_sal_90_heatmap,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.7,
       width = 3.4)
```

###Figure: Common gardens 0.1 temperature quantile heatmap
```{r}
field_temp_10_heatmap <- ggplot(quant_field, aes(site_name, 1, fill = temp_quantile_10)) +
  geom_tile()+
  geom_text(aes(label = round(temp_quantile_10, digits = 1)), color = "white", size = 5)+
  scale_fill_viridis_c(option = "C", limits = c(5,32)) +  # use viridis color scale
  scale_x_discrete(limits = c("LEW", "YRK"))+
  labs(fill = "Temp 째C ") +
  theme_minimal() +
   theme(axis.text.x = element_text(angle = 45, hjust = 1), 
         axis.ticks.y = element_blank(), 
         axis.text.y = element_blank(), 
         axis.title.y = element_blank(), 
         panel.grid = element_blank(), 
         legend.key.size = unit(.4, 'cm'), 
         legend.title = element_text(size = 10), 
         legend.text = element_text(size = 10))
field_temp_10_heatmap

ggsave("field_temp10_heat.pdf",
       plot = field_temp_10_heatmap,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.7,
       width = 3.4)
```

###Figure: Common gardens 0.9 temperature quantile heatmap
```{r}
field_temp_90_heatmap <- ggplot(quant_field, aes(site_name, 1, fill = temp_quantile_90), height = 2, width = 7) +
  geom_tile()+
  geom_text(aes(label = round(temp_quantile_90, digits = 1)), size = 5)+
  scale_fill_viridis_c(option = "C", limits = c(5,32)) +  # use viridis color scale
  scale_x_discrete(limits = c("LEW", "YRK"))+
  labs(fill = "Temp 째C ") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1),
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.key.size = unit(.4, 'cm'), 
        legend.title = element_text(size = 10), 
        legend.text = element_text(size = 10))

field_temp_90_heatmap

ggsave("field_temp_90_heat.pdf",
       plot = field_temp_90_heatmap,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.7,
       width = 3.4)
```

### Disease data
Disease data are classified as the following:
- 0: no disease pressure
- 1: lower disease pressure
- 2: higher disease pressure

#### Disease data preparation
```{r}
#Make disease dataframe
site_name <- c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME", "LEW", "YRK")
dermo<- c(2, 1, 2, 1, 1, 2, 2, 2, 1, 2)
MSX <- c(0, 0, 2, 1, 1, 1, 2, 2, 1, 2)
disease <- as.data.frame(cbind(site_name, MSX, dermo))
summary(disease)
```

###Figure: EoO Dermo Pressure
```{r}
#Make heatmap of environment-of-origin Dermo disease pressure
eoo_dermo <- disease %>% 
  filter(site_name != "YRK" | site_name != "LEW") %>% 
  ggplot(aes(site_name, 1, fill = factor(dermo)), height = 2, width = 7)+
  geom_tile(show.legend = T)+
  geom_text(aes(label = dermo), size = 5)+
  scale_fill_manual(name="Pressure",
                    limits=c("0", "1", "2"),
                    values=c("0"="#FFF2F2", "1"="#FF7C7C", "2"="#DA0707"))+
  scale_x_discrete(limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  theme_minimal() +
 theme(axis.text.x = element_text(angle = 45, hjust = 1), 
       axis.ticks.y = element_blank(), 
       axis.text.y = element_blank(), 
       axis.title.y = element_blank(), 
       panel.grid = element_blank(), 
       legend.position = "none")

eoo_dermo

ggsave("eoo_dermo.pdf",
       plot = eoo_dermo,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.9,
       width = 9)
```

###Figure: EoO MSX Pressure
```{r}
#Make heatmap of environment-of-origin MSX disease pressure
eoo_msx <- disease %>% 
  filter(site_name != "YRK" | site_name != "LEW") %>% 
    ggplot(aes(site_name, 1, fill = factor(MSX)), height = 2, width = 7)+
  geom_tile(show.legend = T)+
  geom_text(aes(label = MSX), size = 5)+
  scale_fill_manual(name="Pressure",
                    limits=c("0", "1", "2"),
                    values=c("0"="#FFF2F2", "1"="#FF7C7C", "2"="#DA0707"))+
  scale_x_discrete(name = "Environment-of-Origin", limits = c("W1-TX","W2-LA","W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME"))+
  labs(fill = "MSX Pressure") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black"), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.position = "none")

eoo_msx

ggsave("eoo_msx.pdf",
       plot = eoo_msx,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.9,
       width = 9)
```

###Figure: Common Gardens Dermo Pressure
```{r}
#Make heatmap of field site Dermo disease pressure
field_dermo <- disease %>% 
  filter(site_name == "YRK" | site_name == "LEW") %>% 
    ggplot(aes(site_name, 1, fill = factor(dermo)), height = 2, width = 7)+
  geom_tile(show.legend = T)+
  geom_text(aes(label = dermo), size = 5)+
  scale_fill_manual(name="Pressure",
                    limits=c("0", "1", "2"),
                    values=c("0"="#FFF2F2", "1"="#FF7C7C", "2"="#DA0707"))+
  scale_x_discrete(name = "Group", limits = c("LEW", "YRK"))+
  labs( fill = "Pressure") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.key.size = unit(.4, 'cm'), 
        legend.title = element_text(size = 10), 
        legend.text = element_text(size = 10))

field_dermo

ggsave("field_dermo.pdf",
       plot = field_dermo,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.7,
       width = 3.4)
```

###Figure: Common Gardens MSX Pressure
```{r}
#Make heatmap of field site MSX disease pressure
field_msx <- disease %>% 
  filter(site_name == "YRK" | site_name == "LEW") %>% 
    ggplot(aes(site_name, 1, fill = factor(MSX)), color = colors_df$colors, height = 2, width = 7) +
  geom_tile(show.legend = T)+
  geom_text(aes(label = MSX), size = 5)+
   scale_fill_manual(name="Pressure",
                    limits=c("0", "1", "2"),
                    values=c("0"="#FFF2F2", "1"="#FF7C7C", "2"="#DA0707"))+
  scale_x_discrete(name = "Common Garden Site", limits = c("LEW", "YRK"))+
  labs(fill = "Pressure") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black"), 
        axis.ticks.y = element_blank(), 
        axis.text.y = element_blank(), 
        axis.title.y = element_blank(), 
        panel.grid = element_blank(), 
        legend.key.size = unit(.4, 'cm'), 
        legend.title = element_text(size = 10), 
        legend.text = element_text(size = 10))

field_msx

ggsave("field_msx.pdf",
       plot = field_msx,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 1.7,
       width = 3.4)
```