---
title: "LEW_Data_Comparisons"
output: pdf_document
date: "2025-11-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

This code examines the differences between environmental data for LOLA/Lewisetta. The data sources are:

1. Willy Reay's data (VIMS data) from the Coan River
2. NOAA National Buoy Data Center (NDBC), Station 44042 (Potomac, MD) for salinity
3. NOAA National Buoy Data Center (NDBC), Station 8635750 (Lewisetta, VA) for temp

```{r setwd}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

```{r packages}
library("dplyr") #Used for working with data frames
library("lubridate") #Used for time-date conversions
library("readr") #Used to read the CSV file
library("ggplot2") 
library("stringr")
```

#NDBC Data Upload and Cleaning
```{r NDBC}
# directory for files
lew_sonde <- "data/EnvDat/exp/lew_raw/lew_sonde" 
pot_sonde <- "data/EnvDat/exp/lew_raw/potomac_sonde"

# Get a list of all .txt files in the directory
files_lew <- list.files(path = lew_sonde, pattern = "\\.txt$", full.names = T)
files_pot <- list.files(path = pot_sonde, pattern = "\\.txt$", full.names = T)

# Read each file into a list of data frames
lew_sonde_df <- lapply(files_lew, function(file) {
  read.table(file, sep = " ")
})
pot_sonde_df <- lapply(files_pot, function(file) {
  read.table(file, sep = " ")
})

# Combine all data frames in the list into a single data frame
lew_check <- bind_rows(lew_sonde_df)
pot_check <- bind_rows(pot_sonde_df)

# colnames
colnames(lew_check) <- c("YY", "MM", "DD", "hh", "mm", "WDIR", "WSPD", "GST", "WVHT", "DPD", "APD", "MWD", "PRES", "ATMP", "temp", "DEWP", "VIS", "TIDE")
colnames(pot_check) <- c("YY", "MM", "DD", "hh", "mm", "DEPTH", "OTMP", "COND", "sal", "O2%", "O2PPM", "CLCON", "TURB", "PH", "EH")

# just relevant data
lew_check_red <- lew_check[,c("YY","MM","DD", "hh", "mm","temp")]
pot_check_red <- pot_check[,c("YY","MM","DD", "hh", "mm", "sal")]

# put together datetime
lew_check_red$datetime <- paste0(lew_check_red$YY, "-", lew_check_red$MM, "-", lew_check_red$DD, " ", lew_check_red$hh, ":", lew_check_red$mm, ":00 EST")
pot_check_red$datetime <- paste0(pot_check_red$YY, "-", pot_check_red$MM, "-", pot_check_red$DD, " ", pot_check_red$hh, ":", pot_check_red$mm, ":00 EST")

lew_check_red$datetime <- as.POSIXct(lew_check_red$datetime, "%Y-%m-%d %H:%M:%S", tz = "")
pot_check_red$datetime <- as.POSIXct(pot_check_red$datetime, "%Y-%m-%d %H:%M:%S", tz = "")

# merge these datesets
check_lewisetta <- merge(lew_check_red[,c("datetime","temp")], pot_check_red[,c("datetime","sal")], by = "datetime", all = T)

# check data
check_lewisetta$temp[check_lewisetta$temp == 999] <- NA
check_lewisetta$sal[check_lewisetta$sal == 99] <- NA
summary(check_lewisetta)

# qc filter
filtered_check_lewisetta <- check_lewisetta %>%
    filter(between(sal, 0, 40)) 
filtered_check_lewisetta <- filtered_check_lewisetta %>%
    filter(between(temp, 0, 40))
```

#VIMS Data Upload and Cleaning

```{r VIMS_cleaning}
# read in data
lew <- read.csv("data/EnvDat/exp/lew_raw/lewisetta_coandat.csv")[,-1]

# vims data is every 15 mins vs more every hour at noaa buoys -- use daily averages to calculate the other summary stats
filtered_check_lewisetta$date <- as.Date(filtered_check_lewisetta$datetime, tz = "EST")
daily_check_lewisetta <- filtered_check_lewisetta %>% 
  group_by(date) %>% 
  mutate(mean_daily_temp = mean(temp), mean_daily_sal = mean(sal))

# do the same with other lewisetta data
lew_daily <- lew
lew_daily$date <- as.Date(lew$datetime, tz = "EST")
daily_lewisetta <- lew_daily %>% 
  group_by(date) %>% 
  mutate(mean_daily_temp = mean(temp), mean_daily_sal = mean(salinity))

#filter so each day is represented once
day_check_lewisetta <- daily_check_lewisetta[match(unique(daily_check_lewisetta$date), daily_check_lewisetta$date),]
noaa_day_filtered <- day_check_lewisetta[day_check_lewisetta$date %in% as.Date(daily_lewisetta$datetime, tz = "EST"),]
lola_day_filtered <- daily_lewisetta[as.Date(daily_lewisetta$datetime, tz = "EST") %in% day_check_lewisetta$date,]

# put together full dataset
colnames(noaa_day_filtered) <- c("datetime", "temp_noaa", "sal_noaa", "date", "mean_daily_temp_noaa", "mean_daily_sal_noaa")
lew_mg <- merge(noaa_day_filtered, lola_day_filtered, by = c("datetime", "date"), all = T)
```


Plot to visualize differences in salinity & temperature:

```{r env plot}
# plot to look at differences
filter_tempplot <- ggplot(lew_mg, aes(x=datetime, y = mean_daily_temp, color = "VIMS")) +
    geom_line()+
    geom_line(data = lew_mg, aes(x=datetime, y = mean_daily_temp_noaa, color = "NOAA"))+
  scale_color_manual(values=c("black", "red"))+
  labs(x = "Time", y = "Temperature C", title = "Temperature Plot for NOAA and VIMS Data Sets") 
filter_tempplot # basically right on top of each other

filter_salplot <- ggplot(lew_mg, aes(x=datetime, y = mean_daily_sal, color = "VIMS")) +
    geom_line()+
    geom_line(data = lew_mg, aes(x=datetime, y = mean_daily_sal_noaa, color = "NOAA"))+
  scale_color_manual(values=c("black", "blue"))+
  labs(x = "Time", y = "Salinity psu", title = "Salinity Plot for NOAA and VIMS Data Sets") 
filter_salplot # trends follow, but need a large correction
```

Now we correct the NDBC data

```{r correction_value}
# corrections
lew_mg <- lew_mg %>% 
  mutate(temp_diff = mean_daily_temp-mean_daily_temp_noaa, sal_diff = mean_daily_sal -mean_daily_sal_noaa)

temp_correction <- mean(lew_mg$temp_diff)
# VIMS data (from Willy Coan) is 0.09287575ºC above NOAA data, apply this correction to past NOAA data (2014-2020)

sal_correction <- mean(lew_mg$sal_diff)
# VIMS data (from Willy Coan) is -1.192284 below NOAA data, apply this correction to past NOAA data (2014-2020)

# apply correction
day_check_lewisetta$temp_corrected <- day_check_lewisetta$mean_daily_temp + 0.09287575
day_check_lewisetta$sal_corrected <- day_check_lewisetta$mean_daily_sal - 1.192284

# write it out
lew_correct <- day_check_lewisetta
lew_correct <- lew_correct[!(is.na(lew_correct$datetime)),]
write.csv(lew_correct, "../../../../data/envr_of_origin/raw_envr_data/lewisetta_corrected.csv", row.names = F)

# summary of these data
lew_correct_yr <- lew_correct %>%
    dplyr::mutate(year = year(datetime)) %>%
    dplyr::group_by(year, .drop = FALSE) %>%
    dplyr::summarise(
      min_salinity = min(sal_corrected),
      max_salinity = max(sal_corrected),
      mean_salinity = mean(sal_corrected),
      salinity_quantile10 = quantile(sal_corrected, 0.1),
      salinity_quantile90 = quantile(sal_corrected, 0.9),
      min_temp = min(temp_corrected),
      max_temp = max(temp_corrected),
      mean_temp = mean(temp_corrected),
      temp_quantile10 = quantile(temp_corrected, 0.1),
      temp_quantile90 = quantile(temp_corrected, 0.9))
```

