---
title: "LOLA_raw_env"
output: pdf_document
date: "2024-04-03"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/src/NM")
```

### Load required packages. 
```{r}
library("dplyr") #Used for working with data frames
library("lubridate") #Used for time-date conversions
library("readr") #Used to read the CSV file
library("ggplot2") 
```

### Note the date of data download and source. All available data should be used for each site regardless of year. Note from the CSV file how often the site was sampled, and if there are replicates in the data. Also describe if the sampling occurred at only low tide, only high tide, or continuously.  
```{r}
#Data was downloaded on 05/10/2024
#Source - Dr. William "Willy" Reay of VIMS shared the file with Jess Small via email. Jess emailed the file to KEL who shared it with me (NM).
#The site was sampled continuously every 15 min from May 2023 to April 2024

#Create text strings with metadata information that we want to include in the final data frame. 
download_date <- ("05-10-2024")
source_description <- ("William Reay, VIMS")
site_name <- ("LOLA") 
collection_type <- ("continuous")
```

### Use the file path name in your working directory or desktop, see example below. Or, import data set through the "Files" window in R studio. Store the file in a variable with the "raw_ID_Site" format. If salinity and temperature data are in separate files, read in both and store them with "_sal" or "_temp" in the variable names. 
```{r}
#The files we will be working with are from the Coan River, VA, which is for the LOLA selection line. The ID_Site for this site is LOLA. 

#set working directory to files location
setwd("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/LOLA_Coan")

#merge files into one
raw_LOLA_env <- read.csv("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/LOLA_Coan/LOLA_Coan_raw_VIMS.csv")

#The data include separate columns for date and time, which I want to merge into one.
raw_LOLA_env$datetime <- paste(raw_LOLA_env$Date, raw_LOLA_env$EST.Time)

#Remove Date and EST.Time columns, as well as depth column.
raw_LOLA_env <- subset(raw_LOLA_env, select = -c(Date, EST.Time, Depth))

#Reorder columns so datetime is first
raw_LOLA_env <- raw_LOLA_env[, c(3,1,2)]


# View how the data are stored. Note the variable names and the format and units that the data are stored in.  
summary(raw_LOLA_env)

#rename columns. "datetime" = date and time of data collection, "temp" = water temperature in degrees C, "salinity" = water salinity in parts per thousand (ppt)

colnames(raw_LOLA_env) <- c("datetime", "temp", "salinity")
```

### Start with the date and time of collection. We will use the lubridate package to standardize all values into the date-time format called POSIXct. This format stores the date and time in number of seconds since a past point (1/1/1970). This makes comparisons easy and helps to standardizes values. 
```{r}

#Convert to POSIXct format. Tell R what the current date/time format is so it knows how to convert it. Store it into a column named datetime in the data frame.
raw_LOLA_env$datetime <- as.POSIXct(raw_LOLA_env$datetime, "%m/%d/%y %H:%M:%S", tz = "")



#Print the new data frame and examine to make sure the new datetime column is in the correct format. 
head(raw_LOLA_env)
```

## Set up data frames for violin plots
```{r}
#add site name and create new data frame with full envr data set
LOLA_env_full <- raw_LOLA_env %>% 
  mutate(site_name, site_name = "LOLA")

#reorder columns with site_name first
LOLA_env_full <- LOLA_env_full[, c(4, 1, 2, 3)]

#separate out salinity and temp data into new frames for other analyses
LOLA_temp_full <- LOLA_env_full[, c(1,2,3)]

LOLA_sal_full <- LOLA_env_full[, c(1,2,4)]

#save LOLA  as csvs for future analyses
write.csv(LOLA_temp_full, "/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/full_datasets/full_temp/LOLA_temp_full.csv", row.names = FALSE)

write.csv(LOLA_sal_full, "/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/full_datasets/full_sal/LOLA_sal_full.csv", row.names = FALSE)
```

### #Standardize column and variable names. We will use "lat" for latitude in degrees, and "lon" for longitude in degrees. 

```{r}


#Store variables that we will include in the final data frame
lat <- 37.9803
lon <- -76.4619
#Lat and long data are from this site(https://www.vims.edu/cbnerr/monitoring/water_quality/). Data came from the KCB hatchery in the Coan as part of the CBNERR water quality monitoring initiative.
firstyear <- 2023
finalyear <- 2024
```


### Filter any of the variables that have data points outside of normal range. We will use 0-40 as the accepted range for salinity (ppt) and temperature (C) values. Note, in the summer, salinity values can sometimes exceed 40. Check to see if there are values above 40. In this case, adjust the range or notify someone that the site has particularly high salinity values. 
```{r}
#Filter the data between the values of 0 and 40 for both salinity and temperature. 
filtered_LOLA_sal <- raw_LOLA_env %>%
    filter(between(salinity, 0, 40)) 
           
filtered_LOLA_env <- filtered_LOLA_sal %>%
    filter(between(temp, 0, 40))

# Sanity check - print the ranges to ensure values are filtered properly. We can see that the ranges for both are now in the appropriate range.  
print(summary(filtered_LOLA_env$salinity))
print(summary(filtered_LOLA_env$temp))

#Store our data into a variable name with just the site name. 
LOLA_env <- filtered_LOLA_env

# we have NAs in the our data frame in the datetime column - need to remove these
count.nas_LOLA_env <- is.na(LOLA_env$datetime) # store our NAs in a variable
summary(count.nas_LOLA_env) # we have 4 NAs that are stored as "TRUE" in our count.nas

nrow(LOLA_env) # figure out how many rows we have in the original df: 34082
which(count.nas_LOLA_env == TRUE) # find the number of NA rows that we need to remove: 4
LOLA_env <- na.omit(LOLA_env) #remove NAs using na.omit
nrow(LOLA_env) #there are 34078 rows in the new data frame. 34082-34078 = 4, meaning we removed the right number of rows

# check for NAs in our temperature column
count.nas_LOLA_temp <- is.na(LOLA_env$temp) # store our NAs in temp in a variable
summary(count.nas_LOLA_temp) # we have 0 NAs

#check for NAs in our salinity column
count.nas_LOLA_sal <- is.na(LOLA_env$salinity) #store our NAs in salinity in a variable
summary(count.nas_LOLA_sal) #we have no NAs

```

### Visualize the salinity, temperature, and date ranges over time. This can help us see if there are any anomalies or gaps in the data and make sure the filtering was done correctly. Sanity check - do the temperature and salinity ranges look appropriate for the geography of the site (ex. near full ocean salinity for coastal sites, lower salinity for estuaries or near rivers)?

```{r salinity-plot}

salplot <- ggplot(LOLA_env, aes(x = datetime)) +
    geom_line(aes(y = salinity, color = "Salinity (ppt)")) +
    ylim(0,40) +
    labs(x = "Time", y = "Salinity ppt", title = "Salinity Plot for LOLA - Coan River, Virginia") +
    scale_color_manual(values = c("Salinity (ppt)" = "blue")) +
    theme_minimal()


salplot
```


```{r temperature-plot}
tempplot <- ggplot(LOLA_env, aes(x = datetime)) +
    geom_line(aes(y = temp, color = "Temperature (C)")) +
    ylim(0, 40) +
    labs(x = "Time", y = "Temperature C", title = "Temperature Plot for LOLA - Coan River, Virginia") +
    scale_color_manual(values = c( "Temperature (C)" = "red")) +
    theme_minimal()

tempplot
```

### We need to calculate the mean, maximum, and minimum values for salinity and temperature per month and year. First make two data frames to contain each of the annual and monthly averages.
```{r}
#Calculate the mean, maximum, and minimum values for salinity and temperature for each month. 
LOLA_envrmonth_sal <- LOLA_env %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity))
      
LOLA_envrmonth_temp <- LOLA_env %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(      
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp))
  
print(LOLA_envrmonth_sal)
print(LOLA_envrmonth_temp)


#Calculate the mean, maximum, and minimum values for salinity and temperature for each year. 
LOLA_envryear_sal <- LOLA_env %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity))

LOLA_envryear_temp <- LOLA_env %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp))

print(LOLA_envryear_sal)
print(LOLA_envryear_temp)

```
### Plot the months and years of data collection to check if there are any collection gaps in the data.
```{r timeplot - salinity}
timeplot_sal_LOLA <- ggplot(LOLA_envrmonth_sal, aes(x = year)) +
    geom_point(aes(y = month, color = length_salinity), size = 4) +
    labs(x = "Time", y = "Month", title = "Salinity Timeplot for LOLA - Coan River, Virginia") +
    ylim(1,12) +
    theme_minimal()

timeplot_sal_LOLA
```

### Plot the months and years of data collection to check if there are any collection gaps in the data.
```{r timeplot - temperature}
timeplot_temp_LOLA <- ggplot(LOLA_envrmonth_temp, aes(x = year)) +
    geom_point(aes(y = month, color = length_temp), size = 4) +
    labs(x = "Time", y = "Month", title = "Temperature Timeplot for LOLA - Coan River, Virginia") +
    ylim(1,12) +
    theme_minimal()

timeplot_temp_LOLA
```

### We can now calculate a list of variables that we will have collected for all sites. This will allow us to compare sites easily. We will calculate the number of observations from each site, the mean annual, maximum annual, and minimum annual value for all variables. 

Our list of variables includes: 

- Mean_Annual_Temperature_C: average of all available data
- Mean_max_temperature_C: average of maximums for each year
- Mean_min_temperature_C: average of minimums for each year
- Temperature_st_dev: standard deviation of all available data
- Temperature_n: total number of data points
- Temperature_years: number of years in data set

- Mean_Annual_Salinity_ppt: average of all available data
- Mean_min_Salinity_ppt: average of minimums for each year
- Mean_max_Salinity_ppt: average of maximums for each year
- Salinity_st_dev: standard deviation of all available data
- Salinity_n: total number of data points
- Salinity_years: number of years in data set

```{r}
#Calculate temperature variables. 
#Calculate temperature variables. 
Mean_Annual_Temperature_C <- mean(LOLA_env$temp)
Mean_max_temperature_C <- mean(LOLA_envryear_temp$max_temp)
Mean_min_temperature_C <- mean(LOLA_envryear_temp$min_temp)
Temperature_st_dev <- sd(LOLA_env$temp)
Temperature_n <- nrow(LOLA_env)
Temperature_years <- nrow(LOLA_envryear_temp)

#Create a data frame to store the temperature results
LOLA_temp <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Temperature_C, Mean_max_temperature_C, Mean_min_temperature_C, Temperature_st_dev, Temperature_n, Temperature_years, collection_type)
print(LOLA_temp)

# Write to a unique new CSV file
write.csv(LOLA_temp, "/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/Temp/LOLA_temperature.csv", row.names = FALSE)

```

```{r}
#Calculate the salinity variables
Mean_Annual_Salinity_ppt <- mean(LOLA_env$salinity)
Mean_max_Salinity_ppt <- mean(LOLA_envryear_sal$max_salinity)
Mean_min_Salinity_ppt <- mean(LOLA_envryear_sal$min_salinity)
Salinity_st_dev <- sd(LOLA_env$salinity)
Salinity_n <- nrow(LOLA_env)
Salinity_years <- nrow(LOLA_envryear_sal)


#Create a data frame to store the temperature results
LOLA_salinity <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Salinity_ppt, Mean_max_Salinity_ppt, Mean_min_Salinity_ppt, Salinity_st_dev, Salinity_n, Salinity_years, collection_type)
print(LOLA_salinity)

# Write to the combined file with all sites 
# Write to a unique new CSV file
write.csv(LOLA_salinity, "/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/Salinity/LOLA_salinity.csv", row.names = FALSE)
```