---
title: "S1-LOLA-EnvrData"
author: "Nicole Mongillo"
date: "2025-11-09"
output: pdf_document
---

All code based on Madeline Eppley's code for environmental data processing.

Long-term LOLA data uses multiple data sources. These sources are compiled in LEW_corrections.Rmd.

1. Willy Reay's data (VIMS data) from the Coan River
2. NOAA National Buoy Data Center (NDBC), Station 44042 (Potomac, MD) for salinity
3. NOAA National Buoy Data Center (NDBC), Station 8635750 (Lewisetta, VA) for temp

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory relative to project root
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("purrr", "dplyr", "lubridate", "readr", "here", "ggplot2", "stringr")
pacman::p_load(char = packages)

# log versions for reproducibility
r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

Note the date of data download and source. All available data should be used for each site regardless of year. Note from the CSV file how often the site was sampled, and if there are replicates in the data. Also describe if the sampling occurred at only low tide, only high tide, or continuously.  

## Download info

Data was downloaded on 05/10/2024 for 2023 data, 2/26/2025 for 2024 data
Source - Dr. William "Willy" Reay of VIMS shared the file with Jess Small, KEL, MGE, and NM via email. 
The site was sampled continuously every 15 min from May 2023 to April 2024. 
For longer term data, we used data from two NOAA NBDC buoys: Station 44042 (Potomac, MD) for salinity and Station 8635750 (Lewisetta, VA) for temp.

```{r}
# create text strings with metadata information that we want to include in the final data frame. 
download_date <- "05-10-2024"
source_description <- "VIMS - Willy Reay, NOAA NBDC"
site_name <- "S1-LOLA"
collection_type <- "daily"
```

## Data preparation

Use the file path name in your working directory or desktop, see example below. Or, import data set through the "Files" window in R studio. Store the file in a variable with the "raw_ID_Site" format. If salinity and temperature data are in separate files, read in both and store them with "_sal" or "_temp" in the variable names.

The file we will be working with is from Lewisetta, Coan River, VA. The ID_Site for this site is S1-LOLA. Since we did not have long-term environmental data at the exact location where S1-LOLA oysters were bred, we used a nearby buoy with longer-term data to fill in the gaps. Camille Rumberger calculated the mean differences in temperature and salinity values for overlapping datetimes between the two data sources, and applied the results as correction values to the long-term, less local data.

```{r}
corr_S1_LOLA <- read_csv("data/environment/Lewisetta/lewisetta_corrected.csv")

# select date, corrected sal, and corrected temp columns
raw_S1_LOLA <- corr_S1_LOLA %>%
  select(date, sal_corrected, temp_corrected)
```

Start with the date and time of collection. We will use the lubridate package to standardize all values into the date-time format called POSIXct. This format stores the date and time in number of seconds since a past point (1/1/1970). This makes comparisons easy and helps to standardizes values. 

```{r}
# we will be using the date column in the file, not the datetime column. Only some datetime column entries have a time, but each date only has 1 data entry.
str(raw_S1_LOLA)
raw_S1_LOLA$date <- as.character(raw_S1_LOLA$date)

# convert to POSIXct format. Store it into a column named datetime in the data frame.
raw_S1_LOLA$datetime <- as.POSIXct(raw_S1_LOLA$date, format = "%Y-%m-%d", tz = "UTC")

# print the new data frame and examine to make sure the new datetime column is in the correct format. 
summary(raw_S1_LOLA$datetime)

# remove any NAs
raw_S1_LOLA <- raw_S1_LOLA %>%
    filter(!is.na(datetime))
summary(raw_S1_LOLA$datetime)
```

Analyze the ranges of all of our variables of interest - time, salinity, and temperature. Make sure that the latitude and longitude values are consistent for a static collection site. This is a quick check so we can determine how to conduct the next filtering step.

```{r}
# standardize column and variable names. We will use "temp" for temperature in degrees C, "salinity" for salinity in parts per thousand (ppt), "lat" for latitude in degrees, and "lon" for longitude in degrees. 
raw_S1_LOLA <- raw_S1_LOLA %>% rename(temp = temp_corrected, salinity = sal_corrected) #No lat and long data in this file - check metadata files

# print the range (minimum and maximum) of dates of data collection. 
print(range(raw_S1_LOLA$datetime))

# print the range (minimum and maximum) of the salinity values. 
print(summary(raw_S1_LOLA$salinity))

# print the range (minimum and maximum) of the temperature values.
print(summary(raw_S1_LOLA$temp))

# store variables that we will include in the final data frame
lat <- 37.980
lon <- -76.462
firstyear <- 2014
finalyear <- 2024
```

Filter any of the variables that have data points outside of normal range. We will use 0-40 as the accepted range for salinity (ppt) and temperature (C) values. Note, in the summer, salinity values can sometimes exceed 40. Check to see if there are values above 40. In this case, adjust the range or notify someone that the site has particularly high salinity values. 

```{r}
# filter the data between the values of 0 and 40 for both salinity and temperature. 
S1_LOLA <- raw_S1_LOLA %>%
    filter(between(salinity, 0, 40),
           between(temp, 0, 40))

# sanity check - print the ranges to ensure values are filtered properly. We can see that the ranges for both are now in the appropriate range.  
print(summary(S1_LOLA$salinity))
print(summary(S1_LOLA$temp))
```

## Datasets for violin plots
```{r}
# add site name
S1_LOLA <- S1_LOLA %>% 
  mutate(site_name = "S1-LOLA") %>%
  select(site_name, everything())

# create temp subset: site_name, datetime, temp
S1_LOLA_temp_full <- S1_LOLA %>% 
  select(site_name, datetime, temp)

# create salinity subset: site_name, datetime, salinity
S1_LOLA_sal_full <- S1_LOLA %>% 
  select(site_name, datetime, salinity)

# save as csv for future analyses
write.csv(S1_LOLA_temp_full, "data/environment/S1-LOLA/S1-LOLA_temp_full.csv", row.names = FALSE)
write.csv(S1_LOLA_sal_full, "data/environment/S1-LOLA/S1-LOLA_sal_full.csv", row.names = FALSE)
```

Visualize the salinity, temperature, and date ranges over time. This can help us see if there are any anomalies or gaps in the data and make sure the filtering was done correctly. Sanity check - do the temperature and salinity ranges look appropriate for the geography of the site (ex. near full ocean salinity for coastal sites, lower salinity for estuaries or near rivers)?

```{r salinity-plot}
salplot <- ggplot(S1_LOLA, aes(x = datetime)) +
    geom_line(aes(y = salinity, color = "Salinity (ppt)")) +
    ylim(0, 45) +
    labs(x = "Time", y = "Salinity ppt", title = "Salinity Plot for S1_LOLA") +
    scale_color_manual(values = c("Salinity (ppt)" = "blue")) +
    theme_minimal()

salplot
```

```{r temperature-plot}
tempplot <- ggplot(S1_LOLA, aes(x = datetime)) +
    geom_line(aes(y = temp, color = "Temperature (C)")) +
    ylim(0, 45) +
    labs(x = "Time", y = "Temperature C", title = "Temperature Plot for S1_LOLA") +
    scale_color_manual(values = c("Temperature (C)" = "red")) +
    theme_minimal()

tempplot
```

## Monthly and yearly means

We need to calculate the mean, maximum, and minimum values for salinity and temperature per month and year. First make two data frames to contain each of the annual and monthly averages.

```{r}
# calculate the mean, maximum, and minimum values for salinity and temperature for each month. 
S1_LOLA_envrmonth <- S1_LOLA %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp),
      .groups = "drop")
  
print(S1_LOLA_envrmonth)

# calculate the mean, maximum, and minimum values for salinity and temperature for each year. 
S1_LOLA_envryear <- S1_LOLA %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      .groups = "drop")

print(S1_LOLA_envryear)

# calculate the mean, maximum, and minimum values for salinity and temperature for each day. 
S1_LOLA_envrday <- S1_LOLA %>%
    mutate(year = year(datetime), month = month(datetime), day = day(datetime)) %>%
    group_by(year, month, day) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp),
      .groups = "drop")
  
print(S1_LOLA_envrday)
```

Plot the months and years of data collection to check if there are any collection gaps in the data.
```{r timeplot}
timeplot <- ggplot(S1_LOLA_envrmonth, aes(x = year)) +
    geom_point(aes(y = month, color = length_salinity), size = 4) +
    labs(x = "Time", y = "Month", title = "Timeplot for S1_LOLA") +
    ylim(1, 12) +
    theme_minimal()

timeplot
```

## Stress calculations

Calculate days above and below thresholds and plot

```{r}
# open up a blank data frame that spans all 12 months for all years of data collection for this site
# we will merge this with the observations so that we can plot over time 
complete_year_month <- expand.grid(
  year = firstyear:finalyear,
  month = 1:12
)
```

### Low salinity stress
```{r low-sal-stress}
# calculate stress inline so we don't add extra columns to S1_LOLA
low_sal_stress_count <- S1_LOLA %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         low_sal_stress = salinity < 12) %>% 
  filter(low_sal_stress) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(low_sal_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(low_sal_stress_count)

# this plots low sal stress days only, years with no observations of low salinity stress will not appear in the graph 
ggplot(low_sal_stress_count, aes(x = factor(month), y = low_sal_stress, fill = factor(month))) +
  geom_bar(stat = "identity", fill = "#1f77b4") +
  facet_wrap(~ year) +
  labs(title = "S1-LOLA: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

# merge with complete grid, filling missing values with 0
low_sal_complete_data <- complete_year_month %>%
  left_join(low_sal_stress_count, by = c("year", "month")) %>%
  mutate(low_sal_stress = replace_na(low_sal_stress, 0),
         month = as.numeric(month))

# bars
ggplot(low_sal_complete_data, aes(x = factor(month), y = low_sal_stress)) +
  geom_bar(stat = "identity", fill = "#1f77b4") + 
  facet_wrap(~ year) +
  labs(title = "S1-LOLA: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of low salinity stress days") +
  theme_minimal()

# points
ggplot(low_sal_complete_data, aes(x = factor(month), y = low_sal_stress)) +
  geom_point(color = "#1f77b4", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "S1-LOLA: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of low salinity stress days") +
  theme_minimal()
```

### High salinity stress
```{r high-sal-stress}
high_sal_stress_count <- S1_LOLA %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         high_sal_stress = salinity > 35) %>% 
  filter(high_sal_stress) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(high_sal_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(high_sal_stress_count) #no high sal stress days

high_sal_complete_data <- complete_year_month %>%
  left_join(high_sal_stress_count, by = c("year", "month")) %>%
  mutate(high_sal_stress = replace_na(high_sal_stress, 0),
         month = as.numeric(month))

# bars
ggplot(high_sal_complete_data, aes(x = factor(month), y = high_sal_stress)) +
  geom_bar(stat = "identity", fill = "#1B9AAA") + 
  facet_wrap(~ year) +
  labs(title = "S1-LOLA: count of high salinity (> 35 ppt) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

# points
ggplot(high_sal_complete_data, aes(x = factor(month), y = high_sal_stress)) +
  geom_point(color = "#1B9AAA", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "S1-LOLA: count of high salinity (> 35 ppt) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()
```

```{r combined-salinity-stress}
ggplot() +
  geom_bar(data = low_sal_complete_data, aes(x = factor(month), y = low_sal_stress), 
           stat = "identity", fill = "#1f77b4") +
  geom_bar(data = high_sal_complete_data, aes(x = factor(month), y = high_sal_stress), 
           stat = "identity", fill = "#1B9AAA") + 
  facet_wrap(~ year) +
  labs(title = "S1-LOLA: count of stressful salinity days per month", 
       x = "month", 
       y = "count of days") + 
  theme_minimal()

ggplot() +
  geom_point(data = low_sal_complete_data, aes(x = factor(month), y = low_sal_stress), 
           color = "#1f77b4", size = 2) +
  geom_point(data = high_sal_complete_data, aes(x = factor(month), y = high_sal_stress), 
           color = "#1B9AAA", size = 2) + 
  facet_wrap(~ year) +
  labs(title = "S1-LOLA: count of stressful salinity days per month", 
       x = "month", 
       y = "count of days") + 
  theme_minimal()
```

### High temp stress 
```{r high-temp-stress}
high_temp_stress_count <- S1_LOLA %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         high_temp_stress = temp > 28) %>% 
  filter(high_temp_stress) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(high_temp_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(high_temp_stress_count)

ggplot(high_temp_stress_count, aes(x = factor(month), y = high_temp_stress, fill = factor(month))) +
  geom_bar(stat = "identity", fill = "#DD4124FF") +
  facet_wrap(~ year) +
  labs(title = "S1-LOLA: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

high_temp_complete_data <- complete_year_month %>%
  left_join(high_temp_stress_count, by = c("year", "month")) %>%
  mutate(high_temp_stress = replace_na(high_temp_stress, 0),
         month = as.numeric(month))

# bars
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_bar(stat = "identity", fill = "#DD4124FF") + 
  facet_wrap(~ year) +
  labs(title = "S1-LOLA: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

# points
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_point(color = "#DD4124FF", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "S1-LOLA: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()
```

## Summary stats

We can now calculate a list of variables that we will have collected for all sites. This will allow us to compare sites easily. We will calculate the number of observations from each site, the mean annual, maximum annual, and minimum annual value for all variables.

Our list of variables includes: 

- Mean_Annual_Temperature_C: average of all available data
- Mean_max_temperature_C: average of maximums for each year
- Mean_min_temperature_C: average of minimums for each year
- Temperature_st_dev: standard deviation of all available data
- Temperature_n: total number of data points
- Temperature_years: number of years in data set
- temp_quantile_10: 10th quantile value
- temp_quantile_90: 90th quantile value

- Mean_Annual_Salinity_ppt: average of all available data
- Mean_min_Salinity_ppt: average of minimums for each year
- Mean_max_Salinity_ppt: average of maximums for each year
- Salinity_st_dev: standard deviation of all available data
- Salinity_n: total number of data points
- Salinity_years: number of years in data set
- salinity_quantile_10: 10th quantile value
- salinity_quantile_90: 90th quantile value

```{r}
# calculate temperature variables. 
Mean_Annual_Temperature_C <- mean(S1_LOLA$temp)
Mean_max_temperature_C <- mean(S1_LOLA_envryear$max_temp)
Mean_min_temperature_C <- mean(S1_LOLA_envryear$min_temp)
Temperature_st_dev <- sd(S1_LOLA$temp)
Temperature_n <- nrow(S1_LOLA)
Temperature_years <- nrow(S1_LOLA_envryear)

# safe sum with fallback for empty data frames
high_temp_stress_days <- if (nrow(high_temp_stress_count) > 0) sum(high_temp_stress_count$high_temp_stress) else 0
frac_high_temp_stress_days <- high_temp_stress_days / nrow(S1_LOLA_envrday)
temp_quantile_10 <- unname(quantile(S1_LOLA$temp, 0.1))
temp_quantile_90 <- unname(quantile(S1_LOLA$temp, 0.9))

# use left_join instead of cbind to avoid duplicate month columns
Mean_Monthly_Temperature_C <- S1_LOLA_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_Temperature = mean(mean_temp), .groups = "drop")

Mean_min_Monthly_Temperature_C <- S1_LOLA_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_min_Temperature = mean(min_temp), .groups = "drop")

Mean_max_Monthly_Temperature_C <- S1_LOLA_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_max_Temperature = mean(max_temp), .groups = "drop")

# create a data frame to store the temperature results
S1_LOLA_temp <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Temperature_C, Mean_max_temperature_C, Mean_min_temperature_C, temp_quantile_10, temp_quantile_90, Temperature_st_dev, high_temp_stress_days, frac_high_temp_stress_days, Temperature_n, Temperature_years, collection_type)
print(S1_LOLA_temp)

# join monthly temp data instead of cbind + dedup
S1_LOLA_monthly_temp <- Mean_Monthly_Temperature_C %>%
  left_join(Mean_min_Monthly_Temperature_C, by = "month") %>%
  left_join(Mean_max_Monthly_Temperature_C, by = "month")
print(S1_LOLA_monthly_temp)

# write to the combined file with all sites 
# write.table(S1_LOLA_temp, "data/environment/all_temperature.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE)

# write to a unique new CSV file
write.csv(S1_LOLA_temp, "data/environment/S1-LOLA/S1-LOLA_temperature.csv", row.names = FALSE)

# write all monthly data to a unique new CSV file
write.csv(S1_LOLA_monthly_temp, "data/environment/S1-LOLA/S1-LOLA_monthly_temp.csv", row.names = FALSE)
```

```{r}
# calculate the salinity variables
Mean_Annual_Salinity_ppt <- mean(S1_LOLA$salinity)
Mean_max_Salinity_ppt <- mean(S1_LOLA_envryear$max_salinity)
Mean_min_Salinity_ppt <- mean(S1_LOLA_envryear$min_salinity)
Salinity_st_dev <- sd(S1_LOLA$salinity)
Salinity_n <- nrow(S1_LOLA)
Salinity_years <- nrow(S1_LOLA_envryear)

# safe sums with fallback
high_sal_stress_days <- if (nrow(high_sal_stress_count) > 0) sum(high_sal_stress_count$high_sal_stress) else 0
low_sal_stress_days <- if (nrow(low_sal_stress_count) > 0) sum(low_sal_stress_count$low_sal_stress) else 0
frac_high_sal_stress_days <- high_sal_stress_days / nrow(S1_LOLA_envrday)
frac_low_sal_stress_days <- low_sal_stress_days / nrow(S1_LOLA_envrday)
salinity_quantile_10 <- unname(quantile(S1_LOLA$salinity, 0.1))
salinity_quantile_90 <- unname(quantile(S1_LOLA$salinity, 0.9))

# join monthly salinity data instead of cbind + dedup
Mean_Monthly_Salinity <- S1_LOLA_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Mean_Salinity = mean(mean_salinity), .groups = "drop")

Min_Monthly_Salinity <- S1_LOLA_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Min_Salinity = mean(min_salinity), .groups = "drop")

Max_Monthly_Salinity <- S1_LOLA_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Max_Salinity = mean(max_salinity), .groups = "drop")

# create a data frame to store the salinity results
S1_LOLA_salinity <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Salinity_ppt, Mean_max_Salinity_ppt, Mean_min_Salinity_ppt, salinity_quantile_10, salinity_quantile_90, high_sal_stress_days, low_sal_stress_days, frac_high_sal_stress_days, frac_low_sal_stress_days, Salinity_st_dev, Salinity_n, Salinity_years, collection_type)
print(S1_LOLA_salinity)

S1_LOLA_monthly_sal <- Mean_Monthly_Salinity %>%
  left_join(Min_Monthly_Salinity, by = "month") %>%
  left_join(Max_Monthly_Salinity, by = "month")
print(S1_LOLA_monthly_sal)

# write to the combined file with all sites 
# write.table(S1_LOLA_salinity, "data/environment/all_salinity.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE)

# write all year data to a unique new CSV file
write.csv(S1_LOLA_salinity, "data/environment/S1-LOLA/S1-LOLA_salinity.csv", row.names = FALSE)
# write all monthly data to a unique new CSV file
write.csv(S1_LOLA_monthly_sal, "data/environment/S1-LOLA/S1-LOLA_monthly_sal.csv", row.names = FALSE)
```