---
title: "S2-DEBY-EnvrData"
author: "Madeline Eppley"
date: "3/17/2025"
output: github_document
---

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory relative to project root
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("purrr", "dplyr", "lubridate", "readr", "here", "ggplot2", "stringr")
pacman::p_load(char = packages)

# log versions for reproducibility
r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

Note the date of data download and source. All available data should be used for each site regardless of year. Note from the CSV file how often the site was sampled, and if there are replicates in the data. Also describe if the sampling occurred at only low tide, only high tide, or continuously.  

## Download info

Data was downloaded on 04/02/2024
Source - https://vecos.vims.edu/dashboard?station=YRK005.40
The site was sampled continuously every 15 min from 2003.

```{r}
# create text strings with metadata information that we want to include in the final data frame. 
download_date <- "04-02-2024"
source_description <- "Virginia Estuarine and Coastal Observing System, VIMS"
site_name <- "S2-DEBY"
collection_type <- "continuous"
```

## Data preparation

Use the file path name in your working directory or desktop, see example below. Or, import data set through the "Files" window in R studio. Store the file in a variable with the "raw_ID_Site" format. If salinity and temperature data are in separate files, read in both and store them with "_sal" or "_temp" in the variable names.

```{r}
# the files we will be working with are from Gloucester Point, VA, which is for the DEBY selection line. The ID_Site for this site is DEBY. 

# environmental data could only be downloaded by year, so first we need to merge the yearly data sets.
combine_york_data <- function(directory) {
  years <- 2003:2024
  file_names <- paste0(directory, "/York_", years, ".csv")
  data_list <- lapply(file_names, function(f) {
    if (file.exists(f)) {
      read.csv(f, stringsAsFactors = FALSE, fill = TRUE)
    } else {
      warning(paste("File not found:", f))
      return(NULL)
    }
  })
  
  combined_data <- do.call(rbind, data_list)
  return(combined_data)
}

raw_DEBY_env <- combine_york_data("data/environment/S2-DEBY") 
```

The metadata for these data (located at the bottom of this site: http://vecos.vims.edu/Content.aspx?idContent=44) explain the various error codes listed alongside observations that are unusable or questionable. Error codes for a certain water quality observation are listed in a column titled WATERQUALITY_A, where WATERQUALITY is whatever aspect of water quality is being measured. For example, next to the SALINITY column is SALINITY_A, which lists error codes for salinity observations if applicable. Blank or NA WATERQUALITY_A cells indicate no error code is applicable. We need to remove observations marked with these codes by keeping only data with blank cells or NA in these columns.

```{r}
# subset data to include sample collection day and time, water temperature, water temperature error codes, salinity, and salinity error codes
subset_DEBY_env <- raw_DEBY_env %>%
  select(SAMPLE_DATETIME, WTEMP, WTEMP_A, SALINITY, SALINITY_A) %>%
  filter(WTEMP_A %in% c("", NA),
         SALINITY_A %in% c("", NA)) %>%
  select(SAMPLE_DATETIME, WTEMP, SALINITY) %>%
  rename(datetime = SAMPLE_DATETIME,
         temp = WTEMP,
         salinity = SALINITY)

# view how the data are stored. Note the variable names and the format and units that the data are stored in.  
summary(subset_DEBY_env)
```

Start with the date and time of collection. We will use the lubridate package to standardize all values into the date-time format called POSIXct. This format stores the date and time in number of seconds since a past point (1/1/1970). This makes comparisons easy and helps to standardizes values. 

```{r}
# convert to POSIXct format. Tell R what the current date/time format is so it knows how to convert it.
subset_DEBY_env$datetime <- as.POSIXct(subset_DEBY_env$datetime, format = "%m/%d/%Y %H:%M:%S %p", tz = "UTC")

# print the new data frame and examine to make sure the new datetime column is in the correct format. 
#head(subset_DEBY_env)
```

Standardize column and variable names. We will use "lat" for latitude in degrees, and "lon" for longitude in degrees. 

```{r}
# store variables that we will include in the final data frame
lat <- 37.247284
lon <- -76.499369
firstyear <- 2003
finalyear <- 2024
```

Filter any of the variables that have data points outside of normal range. We will use 0-40 as the accepted range for salinity (ppt) and temperature (C) values. Note, in the summer, salinity values can sometimes exceed 40. Check to see if there are values above 40. In this case, adjust the range or notify someone that the site has particularly high salinity values.

```{r}
# filter the data between the values of 0 and 40 for both salinity and temperature. 
DEBY_env <- subset_DEBY_env %>%
    filter(between(salinity, 0, 40),
           between(temp, 0, 40))

# sanity check - print the ranges to ensure values are filtered properly. We can see that the ranges for both are now in the appropriate range.  
print(summary(DEBY_env$salinity)) 
print(summary(DEBY_env$temp))

# remove NAs in the datetime column
DEBY_env <- DEBY_env %>% filter(!is.na(datetime))
nrow(DEBY_env)

# check for NAs in our temperature and salinity columns
summary(is.na(DEBY_env$temp)) # we have 0 NAs
summary(is.na(DEBY_env$salinity)) # we have no NAs
```

## Datasets for violin plots
```{r}
# add site name
DEBY_env <- DEBY_env %>% 
  mutate(site_name = "S2-DEBY") %>%
  select(site_name, everything())

# create temp subset: site_name, datetime, temp
DEBY_temp_full <- DEBY_env %>% 
  select(site_name, datetime, temp)

# create salinity subset: site_name, datetime, salinity
DEBY_sal_full <- DEBY_env %>% 
  select(site_name, datetime, salinity)

# save as csv for future analyses
write.csv(DEBY_temp_full, "data/environment/S2-DEBY/S2-DEBY_temp_full.csv", row.names = FALSE)
write.csv(DEBY_sal_full, "data/environment/S2-DEBY/S2-DEBY_sal_full.csv", row.names = FALSE)
```

Visualize the salinity, temperature, and date ranges over time. This can help us see if there are any anomalies or gaps in the data and make sure the filtering was done correctly. Sanity check - do the temperature and salinity ranges look appropriate for the geography of the site (ex. near full ocean salinity for coastal sites, lower salinity for estuaries or near rivers)?

```{r salinity-plot}
salplot <- ggplot(DEBY_env, aes(x = datetime)) +
    geom_line(aes(y = salinity, color = "Salinity (ppt)")) +
    ylim(0, 40) +
    labs(x = "Time", y = "Salinity ppt", title = "Salinity Plot for DEBY - Gloucester Point, Virginia") +
    scale_color_manual(values = c("Salinity (ppt)" = "blue")) +
    theme_minimal()

salplot
```

```{r temperature-plot}
tempplot <- ggplot(DEBY_env, aes(x = datetime)) +
    geom_line(aes(y = temp, color = "Temperature (C)")) +
    ylim(0, 40) +
    labs(x = "Time", y = "Temperature C", title = "Temperature Plot for DEBY - Gloucester Point, Virginia") +
    scale_color_manual(values = c("Temperature (C)" = "red")) +
    theme_minimal()

tempplot
```

## Monthly and yearly means

We need to calculate the mean, maximum, and minimum values for salinity and temperature per month and year. First make two data frames to contain each of the annual and monthly averages.

```{r}
# calculate the mean, maximum, and minimum values for salinity and temperature for each month. 
DEBY_envrmonth_sal <- DEBY_env %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      .groups = "drop")
      
DEBY_envrmonth_temp <- DEBY_env %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(      
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp),
      .groups = "drop")
  
print(DEBY_envrmonth_sal)
print(DEBY_envrmonth_temp)

# calculate the mean, maximum, and minimum values for salinity and temperature for each year. 
DEBY_envryear_sal <- DEBY_env %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      .groups = "drop")

DEBY_envryear_temp <- DEBY_env %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      .groups = "drop")

print(DEBY_envryear_sal)
print(DEBY_envryear_temp)

# calculate the mean, maximum, and minimum values for salinity and temperature for each day. 
DEBY_envrday <- DEBY_env %>%
    mutate(year = year(datetime), month = month(datetime), day = day(datetime)) %>%
    group_by(year, month, day) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp),
      .groups = "drop")
  
print(DEBY_envrday)
```

Plot the months and years of data collection to check if there are any collection gaps in the data.
```{r timeplot-salinity}
timeplot <- ggplot(DEBY_envrmonth_sal, aes(x = year)) +
    geom_point(aes(y = month, color = length_salinity), size = 4) +
    labs(x = "Time", y = "Month", title = "Salinity Timeplot for DEBY - Gloucester Point, Virginia") +
    ylim(1, 12) +
    theme_minimal()

timeplot
```

Plot the months and years of data collection to check if there are any collection gaps in the data.
```{r timeplot-temperature}
timeplot_temp <- ggplot(DEBY_envrmonth_temp, aes(x = year)) +
    geom_point(aes(y = month, color = length_temp), size = 4) +
    labs(x = "Time", y = "Month", title = "Temperature Timeplot for DEBY - Gloucester Point, Virginia") +
    ylim(1, 12) +
    theme_minimal()

timeplot_temp
```

## Stress calculations

Calculate days above and below thresholds and plot

```{r}
# open up a blank data frame that spans all 12 months for all years of data collection for this site
# we will merge this with the observations so that we can plot over time 
complete_year_month <- expand.grid(
  year = firstyear:finalyear,
  month = 1:12
)
```

### Low salinity stress
```{r low-sal-stress}
# calculate stress inline so we don't add extra columns to DEBY_env
low_sal_stress_count <- DEBY_env %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         low_sal_stress = salinity < 12) %>% 
  filter(low_sal_stress) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(low_sal_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(low_sal_stress_count) # 0 days of low salinity stress
```

### High salinity stress
```{r high-sal-stress}
high_sal_stress_count <- DEBY_env %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         high_sal_stress = salinity > 35) %>% 
  filter(high_sal_stress) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(high_sal_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(high_sal_stress_count) # no high sal stress
```

### High temp stress 
```{r high-temp-stress}
high_temp_stress_count <- DEBY_env %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         high_temp_stress = temp > 28) %>% 
  filter(high_temp_stress) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(high_temp_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(high_temp_stress_count)

ggplot(high_temp_stress_count, aes(x = factor(month), y = high_temp_stress, fill = factor(month))) +
  geom_bar(stat = "identity", fill = "#DD4124FF") +
  facet_wrap(~ year) +
  labs(title = "DEBY: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

high_temp_complete_data <- complete_year_month %>%
  left_join(high_temp_stress_count, by = c("year", "month")) %>%
  mutate(high_temp_stress = replace_na(high_temp_stress, 0),
         month = as.numeric(month))

# bars
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_bar(stat = "identity", fill = "#DD4124FF") + 
  facet_wrap(~ year) +
  labs(title = "DEBY: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

# points
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_point(color = "#DD4124FF", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "DEBY: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()
```

## Summary stats

We can now calculate a list of variables that we will have collected for all sites. This will allow us to compare sites easily. We will calculate the number of observations from each site, the mean annual, maximum annual, and minimum annual value for all variables. 

Our list of variables includes: 

- Mean_Annual_Temperature_C: average of all available data
- Mean_max_temperature_C: average of maximums for each year
- Mean_min_temperature_C: average of minimums for each year
- Temperature_st_dev: standard deviation of all available data
- Temperature_n: total number of data points
- Temperature_years: number of years in data set

- Mean_Annual_Salinity_ppt: average of all available data
- Mean_min_Salinity_ppt: average of minimums for each year
- Mean_max_Salinity_ppt: average of maximums for each year
- Salinity_st_dev: standard deviation of all available data
- Salinity_n: total number of data points
- Salinity_years: number of years in data set

```{r}
# calculate temperature variables. 
Mean_Annual_Temperature_C <- mean(DEBY_env$temp)
Mean_max_temperature_C <- mean(DEBY_envryear_temp$max_temp)
Mean_min_temperature_C <- mean(DEBY_envryear_temp$min_temp)
Temperature_st_dev <- sd(DEBY_env$temp)
Temperature_n <- nrow(DEBY_env)
Temperature_years <- nrow(DEBY_envryear_temp)

# safe sum with fallback for empty data frames
high_temp_stress_days <- if (nrow(high_temp_stress_count) > 0) sum(high_temp_stress_count$high_temp_stress) else 0
frac_high_temp_stress_days <- high_temp_stress_days / nrow(DEBY_envrday)
temp_quantile_10 <- unname(quantile(DEBY_env$temp, 0.1))
temp_quantile_90 <- unname(quantile(DEBY_env$temp, 0.9))

# use left_join instead of cbind to avoid duplicate month columns
Mean_Monthly_Temperature_C <- DEBY_envrmonth_temp %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_Temperature = mean(mean_temp), .groups = "drop")

Mean_min_Monthly_Temperature_C <- DEBY_envrmonth_temp %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_min_Temperature = mean(min_temp), .groups = "drop")

Mean_max_Monthly_Temperature_C <- DEBY_envrmonth_temp %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_max_Temperature = mean(max_temp), .groups = "drop")

# create a data frame to store the temperature results
DEBY_temp <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Temperature_C, Mean_max_temperature_C, Mean_min_temperature_C, temp_quantile_10, temp_quantile_90, Temperature_st_dev, high_temp_stress_days, frac_high_temp_stress_days, Temperature_n, Temperature_years, collection_type)
print(DEBY_temp)

# join monthly temp data instead of cbind + dedup
DEBY_monthly_temp <- Mean_Monthly_Temperature_C %>%
  left_join(Mean_min_Monthly_Temperature_C, by = "month") %>%
  left_join(Mean_max_Monthly_Temperature_C, by = "month")
print(DEBY_monthly_temp)

# write to the combined file with all sites 
# write.table(DEBY_temp, "data/environment/all_temperature.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE)

# write to a unique new CSV file
write.csv(DEBY_temp, "data/environment/S2-DEBY/S2-DEBY_temperature.csv", row.names = FALSE)

# write all monthly data to a unique new CSV file
write.csv(DEBY_monthly_temp, "data/environment/S2-DEBY/S2-DEBY_monthly_temp.csv", row.names = FALSE)
```

```{r}
# calculate the salinity variables
Mean_Annual_Salinity_ppt <- mean(DEBY_env$salinity)
Mean_max_Salinity_ppt <- mean(DEBY_envryear_sal$max_salinity)
Mean_min_Salinity_ppt <- mean(DEBY_envryear_sal$min_salinity)
Salinity_st_dev <- sd(DEBY_env$salinity)
Salinity_n <- nrow(DEBY_env)
Salinity_years <- nrow(DEBY_envryear_sal)

# safe sums with fallback
high_sal_stress_days <- if (nrow(high_sal_stress_count) > 0) sum(high_sal_stress_count$high_sal_stress) else 0
low_sal_stress_days <- if (nrow(low_sal_stress_count) > 0) sum(low_sal_stress_count$low_sal_stress) else 0
frac_high_sal_stress_days <- high_sal_stress_days / nrow(DEBY_envrday)
frac_low_sal_stress_days <- low_sal_stress_days / nrow(DEBY_envrday)
salinity_quantile_10 <- unname(quantile(DEBY_env$salinity, 0.1))
salinity_quantile_90 <- unname(quantile(DEBY_env$salinity, 0.9))

# join monthly salinity data instead of cbind + dedup
Mean_Monthly_Salinity <- DEBY_envrmonth_sal %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Mean_Salinity = mean(mean_salinity), .groups = "drop")

Min_Monthly_Salinity <- DEBY_envrmonth_sal %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Min_Salinity = mean(min_salinity), .groups = "drop")

Max_Monthly_Salinity <- DEBY_envrmonth_sal %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Max_Salinity = mean(max_salinity), .groups = "drop")

# create a data frame to store the salinity results
DEBY_salinity <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Salinity_ppt, Mean_max_Salinity_ppt, Mean_min_Salinity_ppt, salinity_quantile_10, salinity_quantile_90, high_sal_stress_days, low_sal_stress_days, frac_high_sal_stress_days, frac_low_sal_stress_days, Salinity_st_dev, Salinity_n, Salinity_years, collection_type)
print(DEBY_salinity)

DEBY_monthly_sal <- Mean_Monthly_Salinity %>%
  left_join(Min_Monthly_Salinity, by = "month") %>%
  left_join(Max_Monthly_Salinity, by = "month")
print(DEBY_monthly_sal)

# write to the combined file with all sites 
# write.table(DEBY_salinity, "data/environment/all_salinity.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE)

# write all year data to a unique new CSV file
write.csv(DEBY_salinity, "data/environment/S2-DEBY/S2-DEBY_salinity.csv", row.names = FALSE)
# write all monthly data to a unique new CSV file
write.csv(DEBY_monthly_sal, "data/environment/S2-DEBY/S2-DEBY_monthly_sal.csv", row.names = FALSE)
```