---
title: "W2-LA-EnvrData"
author: "Madeline Eppley"
date: "8/15/2023"
output: github_document
---

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory relative to project root
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("purrr", "dplyr", "lubridate", "readr", "here", "ggplot2", "stringr", "dataRetrieval")
pacman::p_load(char = packages)

# log versions for reproducibility
r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

Note the date of data download and source. All available data should be used for each site regardless of year. Note from the CSV file how often the site was sampled, and if there are replicates in the data. Also describe if the sampling occurred at only low tide, only high tide, or continuously.  

## Download info

Data was downloaded on 8/15/2023
Source - https://waterdata.usgs.gov/monitoring-location/07381349/#parameterCode=00065&period=P7D - Caillou Lake (Sister Lake) SW of Dulac, LA - Site Number 07381349

```{r}
# create text strings with metadata information that we want to include in the final data frame. 
download_date <- "08-15-2023"
source_description <- "USGS Water Data Sister Lake - 07381349"
site_name <- "W2-LA"
collection_type <- "continuous"
```

Read in the data using the USGS Data Retrieval Package in R. This will skip the step of downloading a .csv file or similar and importing that file from the desktop. We will import the salinity and temperature data separately and store them with "_sal" or "_temp" in the variable names. Then we will combine them into one file together.

```{r}
siteNumber <- "07381349" # USGS Site Code

# import our site info and read the associated metadata.
W2_LAInfo <- readNWISsite(siteNumber)
comment(W2_LAInfo)

# store the parameter codes that we want to collect data for. The USGS codes salinity as 00480 and temperature (C) as 00010. 
parameterCd_sal <- "00480"
parameterCd_temp <- "00010"
 
# we will retrieve the unit values, or the data values collected at regular intervals. Note the regularity of collection, for this site it is every 15 minutes. 
rawUnitValues_sal <- readNWISuv(siteNumber, parameterCd_sal, "2007-10-01", "2022-08-09")
rawUnitValues_temp <- readNWISuv(siteNumber, parameterCd_temp, "2007-10-01", "2022-08-09")

# view how the data is stored. Note the variable names and the format and units that the data are stored in.  
#View(rawUnitValues_sal)
#View(rawUnitValues_temp)
```

## Data preparation

Combine the salinity and temperature data into one common data frame and name it with the appropriate site code using the "raw_" format. Filter the combined data frame to include no duplicate columns and rename column headers. 

```{r}
# join the data frames by common time using the dateTime column
raw_W2_LA <- rawUnitValues_sal %>%
  inner_join(rawUnitValues_temp, by = "dateTime")

# we now have "double" columns for site code, agency, time zone, and other parameters. Remove those columns and rename.
raw_W2_LA <- raw_W2_LA %>%
  select(agency_cd.x, site_no.x, dateTime, X_00480_00000, X_00010_00000) %>%
  rename(agency = agency_cd.x,
         site = site_no.x,
         temp = X_00010_00000,
         salinity = X_00480_00000)
```

Start with the date and time of collection. We will use the lubridate package to standardize all values into the date-time format called POSIXct. This format stores the date and time in number of seconds since a past point (1/1/1970). This makes comparisons easy and helps to standardizes values. 

```{r}
# SKIP combining, date and time of collection is already in a column together 

# convert to POSIXct format. Store it into a column named datetime in the data frame.
raw_W2_LA$datetime <- as.POSIXct(raw_W2_LA$dateTime, format = "%Y/%m/%d %H:%M:%S", tz = "UTC")

# drop the old date-time column
raw_W2_LA <- raw_W2_LA %>% select(-dateTime)

# print the new data frame and examine to make sure the new datetime column is in the correct format. 
summary(raw_W2_LA)

# we have NAs, so remove those
raw_W2_LA <- raw_W2_LA %>%
    filter(!is.na(datetime))
summary(raw_W2_LA$datetime)
```

Analyze the ranges of all of our variables of interest - time, salinity, and temperature. Make sure that the latitude and longitude values are consistent for a static collection site. This is a quick check so we can determine how to conduct the next filtering step. 

```{r}
# print the range (minimum and maximum) of dates of data collection. 
print(summary(raw_W2_LA$datetime))

# print the range (minimum and maximum) of the salinity values. 
print(summary(raw_W2_LA$salinity))

# print the range (minimum and maximum) of the temperature values.
print(summary(raw_W2_LA$temp))

# store variables that we will include in the final data frame. Pull metadata from the USGS website link located above or the site description from the comment function above. 
lat <- 29.24916667
lon <- -90.9211111
firstyear <- 2007
finalyear <- 2022
```

Filter any of the variables that have data points outside of normal range. We will use 0-40 as the accepted range for salinity (ppt) and temperature (C) values. Note, in the summer, salinity values can sometimes exceed 40. Check to see if there are values above 40. In this case, adjust the range or notify someone that the site has particularly high salinity values. 

```{r}
# filter the data between the values of 0 and 40 for salinity and -1 to 40 for temperature. 
W2_LA <- raw_W2_LA %>%
    filter(between(salinity, 0, 40),
           between(temp, -1, 40))

# sanity check - print the ranges to ensure values are filtered properly. We can see that the ranges for both are now in the appropriate range.  
print(summary(W2_LA$salinity))
print(summary(W2_LA$temp))
```

Write the final processed data frame to a .csv file to create a reproducible "raw" file
```{r}
write.table(W2_LA, "data/environment/W2-LA/W2-LA-raw.csv", sep = ",", append = TRUE, col.names = TRUE, row.names = FALSE)
```

## Datasets for violin plots
```{r}
# add site name
W2_LA <- W2_LA %>% 
  mutate(site_name = "W2-LA") %>%
  select(site_name, everything())

# create temp subset: site_name, datetime, temp
W2_LA_temp_full <- W2_LA %>% 
  select(site_name, datetime, temp)

# create salinity subset: site_name, datetime, salinity
W2_LA_sal_full <- W2_LA %>% 
  select(site_name, datetime, salinity)

# save as csv for future analyses
write.csv(W2_LA_temp_full, "data/environment/W2-LA/W2-LA_temp_full.csv", row.names = FALSE)
write.csv(W2_LA_sal_full, "data/environment/W2-LA/W2-LA_sal_full.csv", row.names = FALSE)
```

Visualize the salinity, temperature, and date ranges over time. This can help us see if there are any anomalies or gaps in the data and make sure the filtering was done correctly. Sanity check - do the temperature and salinity ranges look appropriate for the geography of the site (ex. near full ocean salinity for coastal sites, lower salinity for estuaries or near rivers)?

```{r salinity-plot}
salplot <- ggplot(W2_LA, aes(x = datetime)) +
    geom_line(aes(y = salinity, color = "Salinity (ppt)")) +
    ylim(0, 45) +
    labs(x = "Time", y = "Salinity ppt", title = "Salinity Plot for W2_LA") +
    scale_color_manual(values = c("Salinity (ppt)" = "blue")) +
    theme_minimal()

salplot
```

```{r temperature-plot}
tempplot <- ggplot(W2_LA, aes(x = datetime)) +
    geom_line(aes(y = temp, color = "Temperature (C)")) +
    ylim(-10, 45) +
    labs(x = "Time", y = "Temperature C", title = "Temperature Plot for W2_LA") +
    scale_color_manual(values = c("Temperature (C)" = "red")) +
    theme_minimal()

tempplot
```

## Monthly and yearly means

We need to calculate the mean, maximum, and minimum values for salinity and temperature per month and year. First make two data frames to contain each of the annual and monthly averages.

```{r}
# calculate the mean, maximum, and minimum values for salinity and temperature for each month. 
W2_LA_envrmonth <- W2_LA %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp),
      .groups = "drop")
  
print(W2_LA_envrmonth)

# calculate the mean, maximum, and minimum values for salinity and temperature for each year. 
W2_LA_envryear <- W2_LA %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      .groups = "drop")

print(W2_LA_envryear)

# calculate the mean, maximum, and minimum values for salinity and temperature for each day. 
W2_LA_envrday <- W2_LA %>%
    mutate(year = year(datetime), month = month(datetime), day = day(datetime)) %>%
    group_by(year, month, day) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp),
      .groups = "drop")
  
print(W2_LA_envrday)
```

Plot the months and years of data collection to check if there are any collection gaps in the data.
```{r timeplot}
timeplot <- ggplot(W2_LA_envrmonth, aes(x = year)) +
    geom_point(aes(y = month, color = length_salinity), size = 4) +
    labs(x = "Time", y = "Month", title = "Timeplot for W2_LA") +
    ylim(1, 12) +
    theme_minimal()

timeplot
```

## Stress calculations

Calculate days above and below thresholds and plot

```{r}
# open up a blank data frame that spans all 12 months for all years of data collection for this site
# we will merge this with the observations so that we can plot over time 
complete_year_month <- expand.grid(
  year = firstyear:finalyear,
  month = 1:12
)
```

### Low salinity stress 
```{r low-sal-stress}
# calculate stress inline so we don't add extra columns to W2_LA
low_sal_stress_count <- W2_LA %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         low_sal_stress = salinity < 12) %>% 
  filter(low_sal_stress) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(low_sal_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(low_sal_stress_count)

# this plots low sal stress days only, years with no observations of low salinity stress will not appear in the graph 
ggplot(low_sal_stress_count, aes(x = factor(month), y = low_sal_stress, fill = factor(month))) +
  geom_bar(stat = "identity", fill = "#1f77b4") +
  facet_wrap(~ year) +
  labs(title = "W2-LA: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

# merge with complete grid, filling missing values with 0
low_sal_complete_data <- complete_year_month %>%
  left_join(low_sal_stress_count, by = c("year", "month")) %>%
  mutate(low_sal_stress = replace_na(low_sal_stress, 0),
         month = as.numeric(month))

# bars
ggplot(low_sal_complete_data, aes(x = factor(month), y = low_sal_stress)) +
  geom_bar(stat = "identity", fill = "#1f77b4") + 
  facet_wrap(~ year) +
  labs(title = "W2-LA: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of low salinity stress days") +
  theme_minimal()

# points
ggplot(low_sal_complete_data, aes(x = factor(month), y = low_sal_stress)) +
  geom_point(color = "#1f77b4", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "W2-LA: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of low salinity stress days") +
  theme_minimal()
```

### High salinity stress
```{r high-sal-stress}
high_sal_stress_count <- W2_LA %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         high_sal_stress = salinity > 35) %>% 
  filter(high_sal_stress) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(high_sal_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(high_sal_stress_count) # we don't have any high-stress salinity days at this site
```

### High temp stress
```{r high-temp-stress}
high_temp_stress_count <- W2_LA %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         high_temp_stress = temp > 28) %>% 
  filter(high_temp_stress) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(high_temp_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(high_temp_stress_count)

ggplot(high_temp_stress_count, aes(x = factor(month), y = high_temp_stress, fill = factor(month))) +
  geom_bar(stat = "identity", fill = "#DD4124FF") +
  facet_wrap(~ year) +
  labs(title = "W2-LA: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

high_temp_complete_data <- complete_year_month %>%
  left_join(high_temp_stress_count, by = c("year", "month")) %>%
  mutate(high_temp_stress = replace_na(high_temp_stress, 0),
         month = as.numeric(month))

# bars
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_bar(stat = "identity", fill = "#DD4124FF") + 
  facet_wrap(~ year) +
  labs(title = "W2-LA: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

# points
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_point(color = "#DD4124FF", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "W2-LA: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()
```

## Summary stats

We can now calculate a list of variables that we will have collected for all sites. This will allow us to compare sites easily. We will calculate the number of observations from each site, the mean annual, maximum annual, and minimum annual value for all variables. 

Our list of variables includes: 

- Mean_Annual_Temperature_C: average of all available data
- Mean_max_temperature_C: average of maximums for each year
- Mean_min_temperature_C: average of minimums for each year
- Temperature_st_dev: standard deviation of all available data
- Temperature_n: total number of data points
- Temperature_years: number of years in data set

- Mean_Annual_Salinity_ppt: average of all available data
- Mean_min_Salinity_ppt: average of minimums for each year
- Mean_max_Salinity_ppt: average of maximums for each year
- Salinity_st_dev: standard deviation of all available data
- Salinity_n: total number of data points
- Salinity_years: number of years in data set

```{r}
# calculate temperature variables. 
Mean_Annual_Temperature_C <- mean(W2_LA$temp)
Mean_max_temperature_C <- mean(W2_LA_envryear$max_temp)
Mean_min_temperature_C <- mean(W2_LA_envryear$min_temp)
Temperature_st_dev <- sd(W2_LA$temp)
Temperature_n <- nrow(W2_LA)
Temperature_years <- nrow(W2_LA_envryear)

# safe sum with fallback for empty data frames
high_temp_stress_days <- if (nrow(high_temp_stress_count) > 0) sum(high_temp_stress_count$high_temp_stress) else 0
frac_high_temp_stress_days <- high_temp_stress_days / nrow(W2_LA_envrday)
temp_quantile_10 <- unname(quantile(W2_LA$temp, 0.1))
temp_quantile_90 <- unname(quantile(W2_LA$temp, 0.9))

# use left_join instead of cbind to avoid duplicate month columns
Mean_Monthly_Temperature_C <- W2_LA_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_Temperature = mean(mean_temp), .groups = "drop")

Mean_min_Monthly_Temperature_C <- W2_LA_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_min_Temperature = mean(min_temp), .groups = "drop")

Mean_max_Monthly_Temperature_C <- W2_LA_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_max_Temperature = mean(max_temp), .groups = "drop")

# create a data frame to store the temperature results
W2_LA_temp <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Temperature_C, Mean_max_temperature_C, Mean_min_temperature_C, temp_quantile_10, temp_quantile_90, Temperature_st_dev, high_temp_stress_days, frac_high_temp_stress_days, Temperature_n, Temperature_years, collection_type)
print(W2_LA_temp)

# join monthly temp data instead of cbind + dedup
W2_LA_monthly_temp <- Mean_Monthly_Temperature_C %>%
  left_join(Mean_min_Monthly_Temperature_C, by = "month") %>%
  left_join(Mean_max_Monthly_Temperature_C, by = "month")
print(W2_LA_monthly_temp)

# write to the combined file with all sites 
# write.table(W2_LA_temp, "data/environment/all_temperature.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE)

# write to a unique new CSV file
write.csv(W2_LA_temp, "data/environment/W2-LA/W2-LA_temperature.csv", row.names = FALSE)

# write all monthly data to a unique new CSV file
write.csv(W2_LA_monthly_temp, "data/environment/W2-LA/W2-LA_monthly_temp.csv", row.names = FALSE)
```

```{r}
# calculate the salinity variables
Mean_Annual_Salinity_ppt <- mean(W2_LA$salinity)
Mean_max_Salinity_ppt <- mean(W2_LA_envryear$max_salinity)
Mean_min_Salinity_ppt <- mean(W2_LA_envryear$min_salinity)
Salinity_st_dev <- sd(W2_LA$salinity)
Salinity_n <- nrow(W2_LA)
Salinity_years <- nrow(W2_LA_envryear)

# safe sums with fallback
high_sal_stress_days <- if (nrow(high_sal_stress_count) > 0) sum(high_sal_stress_count$high_sal_stress) else 0
low_sal_stress_days <- if (nrow(low_sal_stress_count) > 0) sum(low_sal_stress_count$low_sal_stress) else 0
frac_high_sal_stress_days <- high_sal_stress_days / nrow(W2_LA_envrday)
frac_low_sal_stress_days <- low_sal_stress_days / nrow(W2_LA_envrday)
salinity_quantile_10 <- unname(quantile(W2_LA$salinity, 0.1))
salinity_quantile_90 <- unname(quantile(W2_LA$salinity, 0.9))

# join monthly salinity data instead of cbind + dedup
Mean_Monthly_Salinity <- W2_LA_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Mean_Salinity = mean(mean_salinity), .groups = "drop")

Min_Monthly_Salinity <- W2_LA_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Min_Salinity = mean(min_salinity), .groups = "drop")

Max_Monthly_Salinity <- W2_LA_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Max_Salinity = mean(max_salinity), .groups = "drop")

# create a data frame to store the salinity results
W2_LA_salinity <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Salinity_ppt, Mean_max_Salinity_ppt, Mean_min_Salinity_ppt, salinity_quantile_10, salinity_quantile_90, high_sal_stress_days, low_sal_stress_days, frac_high_sal_stress_days, frac_low_sal_stress_days, Salinity_st_dev, Salinity_n, Salinity_years, collection_type)
print(W2_LA_salinity)

W2_LA_monthly_sal <- Mean_Monthly_Salinity %>%
  left_join(Min_Monthly_Salinity, by = "month") %>%
  left_join(Max_Monthly_Salinity, by = "month")
print(W2_LA_monthly_sal)

# write to the combined file with all sites 
# write.table(W2_LA_salinity, "data/environment/all_salinity.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE)

# write all year data to a unique new CSV file
write.csv(W2_LA_salinity, "data/environment/W2-LA/W2-LA_salinity.csv", row.names = FALSE)
# write all monthly data to a unique new CSV file
write.csv(W2_LA_monthly_sal, "data/environment/W2-LA/W2-LA_monthly_sal.csv", row.names = FALSE)
```