---
title: "W4-VA-EnvrData"
author: "Madeline Eppley"
date: "3/14/2025"
output: github_document
---

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory relative to project root
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("purrr", "dplyr", "lubridate", "readr", "here", "ggplot2", "stringr")
pacman::p_load(char = packages)

# log versions for reproducibility
r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

Note the date of data download and source. All available data should be used for each site regardless of year. Note from the CSV file how often the site was sampled, and if there are replicates in the data. Also describe if the sampling occurred at only low tide, only high tide, or continuously.  

** Note that this analysis combines two data sources. The data sources were corrected against each other in VA_corrections.Rmd.

## Download info

Data was downloaded on 10/01/2024
Source: https://www.ndbc.noaa.gov/historical_data.shtml#ocean
The site was sampled continuously every hour from 2008-2019. When compared to data from VIMS water quality monitoring in the York River, NDBC temperature data were on average 1.4°C higher than those recorded by VIMS and NDBC salinity data were on average 12.1 ppt lower than those recorded by VIMS. The data used here are the NDBC data with corrected temperature and salinity values, with temperature values 1.4°C lower and salinity values 12.1 ppt higher than the raw file. Datetime is already in POSIXct format.

```{r}
# create text strings with metadata information that we want to include in the final data frame. 
download_date <- "10/01/24"
source_description <- "NOAA National Buoy Data Center (NDBC), Chesapeake Bay Interpretive Buoy System with values corrected based on data from VIMS Water Quality Data"
site_name <- "W4-VA"
collection_type <- "continuous"
```

## Data preparation

Use the file path name in your working directory or desktop, see example below. Or, import data set through the "Files" window in R studio. Store the file in a variable with the "raw_ID_Site" format. If salinity and temperature data are in separate files, read in both and store them with "_sal" or "_temp" in the variable names. 

```{r}
# the files we will be working with are from the James River, Jamestown, W4_VA. The ID_Site for this site W4_VA. 

# read in file
raw_W4_VA <- read.csv("data/environment/W4-VA/VA-corrected.csv")

# view how the data are stored. Note the variable names and the format and units that the data are stored in.  
summary(raw_W4_VA)

# remove extra column X
raw_W4_VA <- raw_W4_VA %>% select(-X)

# remove row with no time in the datetime column
raw_W4_VA <- raw_W4_VA[-c(1), ]

# explicitly name the format argument and set a consistent timezone
raw_W4_VA$datetime <- as.POSIXct(raw_W4_VA$datetime, format = "%Y-%m-%d %H:%M:%S", tz = "UTC")

# we have NAs, so remove those
raw_W4_VA <- raw_W4_VA %>%
    filter(!is.na(datetime))
summary(raw_W4_VA$datetime)
```

Standardize column and variable names. We will use "lat" for latitude in degrees, and "lon" for longitude in degrees. 

```{r}
# store variables that we will include in the final data frame. Lat and lon data from this site: https://www.ndbc.noaa.gov/station_history.php?station=44041
lat <- 37.211
lon <- -76.787
firstyear <- 2008
finalyear <- 2019
```

Filter any of the variables that have data points outside of normal range. We will use 0-40 as the accepted range for salinity (ppt) and temperature (C) values. Note, in the summer, salinity values can sometimes exceed 40. Check to see if there are values above 40. In this case, adjust the range or notify someone that the site has particularly high salinity values. 

```{r}
# filter the data between the values of 0 and 40 for both corrected salinity and corrected temperature
W4_VA <- raw_W4_VA %>%
    filter(between(salinity_corrected, 0, 40),
           between(temp_corrected, 0, 40))

# drop raw columns and rename corrected columns to match other sites
# from this point forward, "temp" = corrected_temp and "salinity" = corrected_salinity
W4_VA <- W4_VA %>%
  select(datetime, temp_corrected, salinity_corrected) %>%
  rename(temp = temp_corrected, salinity = salinity_corrected)

# sanity check - print the ranges to ensure values are filtered properly.
print(summary(W4_VA$salinity)) 
print(summary(W4_VA$temp))

# re-check for NAs
count.nas_env <- is.na(W4_VA)
summary(count.nas_env)
```

## Datasets for violin plots
```{r}
# add site name
W4_VA <- W4_VA %>% 
  mutate(site_name = "W4-VA") %>%
  select(site_name, everything())

# create temp subset: site_name, datetime, temp
W4_VA_temp_full <- W4_VA %>% 
  select(site_name, datetime, temp)

# create salinity subset: site_name, datetime, salinity
W4_VA_sal_full <- W4_VA %>% 
  select(site_name, datetime, salinity)

# save as csv for future analyses
write.csv(W4_VA_temp_full, "data/environment/W4-VA/W4-VA_temp_full.csv", row.names = FALSE)
write.csv(W4_VA_sal_full, "data/environment/W4-VA/W4-VA_sal_full.csv", row.names = FALSE)
```

Visualize the salinity, temperature, and date ranges over time.

```{r salinity-plot}
salplot <- ggplot(W4_VA, aes(x = datetime)) +
    geom_line(aes(y = salinity, color = "Salinity (ppt)")) +
    ylim(0, 40) +
    labs(x = "Time", y = "Salinity ppt", title = "Salinity Plot for W4_VA - James River, Jamestown Virginia") +
    scale_color_manual(values = c("Salinity (ppt)" = "blue")) +
    theme_minimal()

salplot
```

```{r temperature-plot}
tempplot <- ggplot(W4_VA, aes(x = datetime)) +
    geom_line(aes(y = temp, color = "Temperature (C)")) +
    ylim(0, 40) +
    labs(x = "Time", y = "Temperature C", title = "Temperature Plot for W4_VA - James River, Jamestown, Virginia") +
    scale_color_manual(values = c("Temperature (C)" = "red")) +
    theme_minimal()

tempplot
```

## Monthly and yearly means

```{r}
# calculate the mean, maximum, and minimum values for salinity and temperature for each month. 
W4_VA_envrmonth <- W4_VA %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp),
      .groups = "drop")
  
print(W4_VA_envrmonth)

# calculate the mean, maximum, and minimum values for salinity and temperature for each year. 
W4_VA_envryear <- W4_VA %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      .groups = "drop")

print(W4_VA_envryear)

# calculate the mean, maximum, and minimum values for salinity and temperature for each day. 
W4_VA_envrday <- W4_VA %>%
    mutate(year = year(datetime), month = month(datetime), day = day(datetime)) %>%
    group_by(year, month, day) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp),
      .groups = "drop")
  
print(W4_VA_envrday)
```

```{r timeplot}
timeplot <- ggplot(W4_VA_envrmonth, aes(x = year)) +
    geom_point(aes(y = month, color = length_salinity), size = 4) +
    labs(x = "Time", y = "Month", title = "Timeplot for W4_VA") +
    ylim(1, 12) +
    theme_minimal()

timeplot
```

## Stress calculations

```{r}
# open up a blank data frame that spans all 12 months for all years of data collection
complete_year_month <- expand.grid(
  year = firstyear:finalyear,
  month = 1:12
)
```

### Low salinity stress 
```{r low-sal-stress}
# calculate stress inline so we don't add extra columns to W4_VA
low_sal_stress_count <- W4_VA %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         low_sal_stress = salinity < 12) %>% 
  filter(low_sal_stress) %>%  
  distinct(year, month, day) %>%
  group_by(year, month) %>%
  summarise(low_sal_stress = n(), .groups = "drop")

print(low_sal_stress_count)

# merge with complete grid, filling missing values with 0
low_sal_complete_data <- complete_year_month %>%
  left_join(low_sal_stress_count, by = c("year", "month")) %>%
  mutate(low_sal_stress = replace_na(low_sal_stress, 0),
         month = as.numeric(month))

# bars
ggplot(low_sal_complete_data, aes(x = factor(month), y = low_sal_stress)) +
  geom_bar(stat = "identity", fill = "#1f77b4") + 
  facet_wrap(~ year) +
  labs(title = "W4-VA: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of low salinity stress days") +
  theme_minimal()

# points
ggplot(low_sal_complete_data, aes(x = factor(month), y = low_sal_stress)) +
  geom_point(color = "#1f77b4", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "W4-VA: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of low salinity stress days") +
  theme_minimal()
```

### High salinity stress
```{r high-sal-stress}
high_sal_stress_count <- W4_VA %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         high_sal_stress = salinity > 35) %>% 
  filter(high_sal_stress) %>%  
  distinct(year, month, day) %>%
  group_by(year, month) %>%
  summarise(high_sal_stress = n(), .groups = "drop")

print(high_sal_stress_count)
```

### High temp stress 
```{r high-temp-stress}
high_temp_stress_count <- W4_VA %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime),
         high_temp_stress = temp > 28) %>% 
  filter(high_temp_stress) %>%  
  distinct(year, month, day) %>%
  group_by(year, month) %>%
  summarise(high_temp_stress = n(), .groups = "drop")

print(high_temp_stress_count)

high_temp_complete_data <- complete_year_month %>%
  left_join(high_temp_stress_count, by = c("year", "month")) %>%
  mutate(high_temp_stress = replace_na(high_temp_stress, 0),
         month = as.numeric(month))

# bars
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_bar(stat = "identity", fill = "#DD4124FF") + 
  facet_wrap(~ year) +
  labs(title = "W4-VA: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

# points
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_point(color = "#DD4124FF", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "W4-VA: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()
```

## Summary stats

```{r}
# calculate temperature variables. 
Mean_Annual_Temperature_C <- mean(W4_VA$temp)
Mean_max_temperature_C <- mean(W4_VA_envryear$max_temp)
Mean_min_temperature_C <- mean(W4_VA_envryear$min_temp)
Temperature_st_dev <- sd(W4_VA$temp)
Temperature_n <- nrow(W4_VA)
Temperature_years <- nrow(W4_VA_envryear)

# safe sum with fallback for empty data frames
high_temp_stress_days <- if (nrow(high_temp_stress_count) > 0) sum(high_temp_stress_count$high_temp_stress) else 0
frac_high_temp_stress_days <- high_temp_stress_days / nrow(W4_VA_envrday)
temp_quantile_10 <- unname(quantile(W4_VA$temp, 0.1))
temp_quantile_90 <- unname(quantile(W4_VA$temp, 0.9))

# use left_join instead of cbind to avoid duplicate month columns
Mean_Monthly_Temperature_C <- W4_VA_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_Temperature = mean(mean_temp), .groups = "drop")

Mean_min_Monthly_Temperature_C <- W4_VA_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_min_Temperature = mean(min_temp), .groups = "drop")

Mean_max_Monthly_Temperature_C <- W4_VA_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_max_Temperature = mean(max_temp), .groups = "drop")

# create a data frame to store the temperature results
W4_VA_temp <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Temperature_C, Mean_max_temperature_C, Mean_min_temperature_C, temp_quantile_10, temp_quantile_90, Temperature_st_dev, high_temp_stress_days, frac_high_temp_stress_days, Temperature_n, Temperature_years, collection_type)
print(W4_VA_temp)

# join monthly temp data instead of cbind + dedup
W4_VA_monthly_temp <- Mean_Monthly_Temperature_C %>%
  left_join(Mean_min_Monthly_Temperature_C, by = "month") %>%
  left_join(Mean_max_Monthly_Temperature_C, by = "month")
print(W4_VA_monthly_temp)

# write to the combined file with all sites 
# write.table(W4_VA_temp, "data/environment/all_temperature.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE)

# write to a unique new CSV file
write.csv(W4_VA_temp, "data/environment/W4-VA/W4-VA_temperature.csv", row.names = FALSE)

# write all monthly data to a unique new CSV file
write.csv(W4_VA_monthly_temp, "data/environment/W4-VA/W4-VA_monthly_temp.csv", row.names = FALSE)
```

```{r}
# calculate the salinity variables
Mean_Annual_Salinity_ppt <- mean(W4_VA$salinity)
Mean_max_Salinity_ppt <- mean(W4_VA_envryear$max_salinity)
Mean_min_Salinity_ppt <- mean(W4_VA_envryear$min_salinity)
Salinity_st_dev <- sd(W4_VA$salinity)
Salinity_n <- nrow(W4_VA)
Salinity_years <- nrow(W4_VA_envryear)

# safe sums with fallback
high_sal_stress_days <- if (nrow(high_sal_stress_count) > 0) sum(high_sal_stress_count$high_sal_stress) else 0
low_sal_stress_days <- if (nrow(low_sal_stress_count) > 0) sum(low_sal_stress_count$low_sal_stress) else 0
frac_high_sal_stress_days <- high_sal_stress_days / nrow(W4_VA_envrday)
frac_low_sal_stress_days <- low_sal_stress_days / nrow(W4_VA_envrday)
salinity_quantile_10 <- unname(quantile(W4_VA$salinity, 0.1))
salinity_quantile_90 <- unname(quantile(W4_VA$salinity, 0.9))

# join monthly salinity data instead of cbind + dedup
Mean_Monthly_Salinity <- W4_VA_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Mean_Salinity = mean(mean_salinity), .groups = "drop")

Min_Monthly_Salinity <- W4_VA_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Min_Salinity = mean(min_salinity), .groups = "drop")

Max_Monthly_Salinity <- W4_VA_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Max_Salinity = mean(max_salinity), .groups = "drop")

# create a data frame to store the salinity results
W4_VA_salinity <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Salinity_ppt, Mean_max_Salinity_ppt, Mean_min_Salinity_ppt, salinity_quantile_10, salinity_quantile_90, high_sal_stress_days, low_sal_stress_days, frac_high_sal_stress_days, frac_low_sal_stress_days, Salinity_st_dev, Salinity_n, Salinity_years, collection_type)
print(W4_VA_salinity)

W4_VA_monthly_sal <- Mean_Monthly_Salinity %>%
  left_join(Min_Monthly_Salinity, by = "month") %>%
  left_join(Max_Monthly_Salinity, by = "month")
print(W4_VA_monthly_sal)

# write to the combined file with all sites 
# write.table(W4_VA_salinity, "data/environment/all_salinity.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE)

# write all year data to a unique new CSV file
write.csv(W4_VA_salinity, "data/environment/W4-VA/W4-VA_salinity.csv", row.names = FALSE)
# write all monthly data to a unique new CSV file
write.csv(W4_VA_monthly_sal, "data/environment/W4-VA/W4-VA_monthly_sal.csv", row.names = FALSE)
```