---
title: "YRK_EnvrData"
author: "Nicole Mongillo"
date: "2025-02-28"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

### Load required packages. 
```{r}
library("dplyr") #Used for working with data frames
library("lubridate") #Used for time-date conversions
library("readr") #Used to read the CSV file
library("ggplot2") 
```

### Note the date of data download and source. All available data should be used for each site regardless of year. Note from the CSV file how often the site was sampled, and if there are replicates in the data. Also describe if the sampling occurred at only low tide, only high tide, or continuously.  
```{r}
#Data was downloaded on 04/02/2024
#Source - http://vecos.vims.edu/StationDetail.aspx?param=YRK005.40&program=CMON

#Chesapeake Bay National Estuarine Research Reserve in Virginia, Virginia Institute of Marine Science (CBNERR-VA VIMS), 2025. Virginia Estuarine and Coastal Observing System (VECOS). Data accessed from VECOS website: http://vecos.vims.edu; accessed 28 February 2025.
#The site was sampled continuously every 15 min from 2003

#Create text strings with metadata information that we want to include in the final data frame. 
download_date <- ("02-28-2025")
source_description <- ("Virginia Estuarine and Coastal Observing System, VIMS")
site_name <- ("YRK") 
collection_type <- ("continuous")
```

### Use the file path name in your working directory or desktop, see example below. Or, import data set through the "Files" window in R studio. Store the file in a variable with the "raw_ID_Site" format. If salinity and temperature data are in separate files, read in both and store them with "_sal" or "_temp" in the variable names. 
```{r}
#The files we will be working with are from Gloucester Point, VA, which is for the DEBY selection line. The ID_Site for this site is DEBY. 

#Environmental data could only be downloaded by year, so first we need to merge the yearly data sets.

#set working directory to location of files
setwd("../../data/envr_of_origin/raw_envr_data/YRK-EnvrData_files")

#merge files into one
raw_YRK <- list.files(path=".") %>% 
  lapply(read.csv) %>% 
  bind_rows 

#set working directory back to rmd file location
setwd("../../../../src/envr_data")

#The metadata for these data (located at the bottom of this site: http://vecos.vims.edu/Content.aspx?idContent=44) explain the various error codes listed alongside observations that are unusable or questionable. Error codes for a certain water quality observation are listed in a column titled WATERQUALITY_A, where WATERQUALITY is whatever aspect of water quality is being measured. For example, next to the SALINITY column is SALINITY_A, which lists error codes for salinity observations if applicable. Blank or NA WATERQUALITY_A cells indicate no error code is applicable. We need to remove observations marked with these codes by keeping only data with blank cells or NA in these columns.

YRK_error <- subset(raw_YRK, select = c(SAMPLE_DATETIME, WTEMP, WTEMP_A, SALINITY, SALINITY_A)) #subset data to include sample collection day and time, water temperature, water temperature error codes, salinity, and salinity error codes

subset1_YRK_error <- YRK_error[(YRK_error$WTEMP_A %in% c("", NA)), ] #keep only rows where the WTEMP_A columns are blank or NA, i.e. where there were no error codes recorded for the temperature data, and save it to a new data frame

subset2_YRK_error <- subset1_YRK_error[(subset1_YRK_error$SALINITY_A %in% c("", NA)),] #keep only rows where the SALINITY_A columns are blank or NA, i.e. where there were no error codes recorded for the salinity data, and save it to a new data frame

View(subset2_YRK_error)
#sorting the data frame by ascending and descending values for WTEMP_A and SALINITY_A shows only blank cells or NA for each of those columns. We have removed the data marked with error codes

raw_YRK <- subset(subset2_YRK_error, select = c(SAMPLE_DATETIME, WTEMP, SALINITY)) #remove WTEMP_A and SALINITY_A from the dataframe now

# View how the data are stored. Note the variable names and the format and units that the data are stored in.  
summary(raw_YRK)

#rename columns. "datetime" = date and time of data collection, "temp" = water temperature in degrees C, "salinity" = water salinity in parts per thousand (ppt)

colnames(raw_YRK) <- c("datetime", "temp", "salinity")
```

### Start with the date and time of collection. We will use the lubridate package to standardize all values into the date-time format called POSIXct. This format stores the date and time in number of seconds since a past point (01/01/1970). This makes comparisons easy and helps to standardizes values. 
```{r}

#Convert to POSIXct format. Tell R what the current date/time format is so it knows how to convert it. Store it into a column named datetime in the data frame.
raw_YRK$datetime <- as.POSIXct(raw_YRK$datetime, "%Y-%m-%d %H:%M:%S", tz = "")


#Print the new data frame and examine to make sure the new datetime column is in the correct format. 
head(raw_YRK)

#exclude time before August 1 (approximate start of field data)
raw_YRK <- raw_YRK %>% 
  filter(datetime >= 1690848000)

#Print the new data frame and examine to make sure the new datetime column is in the correct format. 
summary(raw_YRK$datetime)

# we have NAs, so remove those
raw_YRK <- raw_YRK %>%
    filter(!is.na(datetime))

# double check that the NA removal worked
summary(raw_YRK$datetime)
```


### #Standardize column and variable names. We will use "lat" for latitude in degrees, and "lon" for longitude in degrees. 

```{r}
#Store variables that we will include in the final data frame
lat <- 37.247284
lon <- -76.499369
firstyear <- 2023
finalyear <- 2025
```


### Filter any of the variables that have data points outside of normal range. We will use 0-40 as the accepted range for salinity (ppt) and temperature (C) values. Note, in the summer, salinity values can sometimes exceed 40. Check to see if there are values above 40. In this case, adjust the range or notify someone that the site has particularly high salinity values. 
```{r}
#Filter the data between the values of 0 and 40 for both salinity and temperature. 
filtered_YRK_sal <- raw_YRK %>%
    filter(between(salinity, 0, 40)) 
           
filtered_YRK_env <- filtered_YRK_sal %>%
    filter(between(temp, 0, 40))

# Sanity check - print the ranges to ensure values are filtered properly. We can see that the ranges for both are now in the appropriate range.  
print(summary(filtered_YRK_env$salinity)) 
print(summary(filtered_YRK_env$temp))

#Store our data into a variable name with just the site name. 
YRK <- filtered_YRK_env
```

### Visualize the salinity, temperature, and date ranges over time. This can help us see if there are any anomalies or gaps in the data and make sure the filtering was done correctly. Sanity check - do the temperature and salinity ranges look appropriate for the geography of the site (ex. near full ocean salinity for coastal sites, lower salinity for estuaries or near rivers)?

```{r salinity-plot}

salplot <- ggplot(YRK, aes(x = datetime)) +
    geom_line(aes(y = salinity, color = "Salinity (ppt)")) +
    ylim(0,40) +
    labs(x = "Time", y = "Salinity ppt", title = "Salinity Plot for YRK - Gloucester Point, Virginia") +
    scale_color_manual(values = c("Salinity (ppt)" = "blue")) +
    theme_minimal()


salplot
```


```{r temperature-plot}
tempplot <- ggplot(YRK, aes(x = datetime)) +
    geom_line(aes(y = temp, color = "Temperature (C)")) +
    ylim(0, 40) +
    labs(x = "Time", y = "Temperature C", title = "Temperature Plot for YRK - Gloucester Point, Virginia") +
    scale_color_manual(values = c( "Temperature (C)" = "red")) +
    theme_minimal()

tempplot
```

### We need to calculate the mean, maximum, and minimum values for salinity and temperature per month and year. First make two data frames to contain each of the annual and monthly averages.
```{r}
#Calculate the mean, maximum, and minimum values for salinity and temperature for each month. 
YRK_envrmonth <- YRK %>%
    mutate(year = year(datetime), month = month(datetime)) %>%
    group_by(year, month) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp))
  
print(YRK_envrmonth)


#Calculate the mean, maximum, and minimum values for salinity and temperature for each year. 
YRK_envryear <- YRK %>%
    mutate(year = year(datetime)) %>%
    group_by(year) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp))

print(YRK_envryear)

#Calculate the mean, maximum, and minimum values for salinity and temperature for each day. 
YRK_envrday <- YRK %>%
    mutate(year = year(datetime), month = month(datetime), day = day(datetime)) %>%
    group_by(year, month, day) %>%
    summarise(
      min_salinity = min(salinity),
      max_salinity = max(salinity),
      mean_salinity = mean(salinity),
      length_salinity = length(salinity),
      min_temp = min(temp),
      max_temp = max(temp),
      mean_temp = mean(temp),
      length_temp = length(temp))
  
print(YRK_envrday)

```
### Plot the months and years of data collection to check if there are any collection gaps in the data.
```{r timeplot - salinity}
timeplot <- ggplot(YRK_envrmonth, aes(x = year)) +
    geom_point(aes(y = month, color = length_salinity), size = 4) +
    labs(x = "Time", y = "Month", title = "Salinity Timeplot for YRK - Gloucester Point, Virginia") +
    ylim(1,12) +
    theme_minimal()

timeplot
```

# Calculate days above and below thresholds and plot
```{r}
# open up a blank data frame that spans all 12 months for all years of data collection for this site
# we will merge this with the observations so that we can plot over time 

complete_year_month <- expand.grid(
  year = unique(firstyear:finalyear),
  month = 1:12
)

```

# start with low salinity stress 
```{r low-sal-stress}
# first calculate for low salinity
YRK$low_sal_stress <- YRK$salinity < 12

low_sal_stress_count <- YRK %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime)) %>% 
  filter(low_sal_stress == 1) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(low_sal_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(low_sal_stress_count)

# this plots low sal stress days only, years with no observations of low salinity stress will not appear in the graph 
ggplot(low_sal_stress_count, aes(x = factor(month), y = low_sal_stress, fill = factor(month))) +
  geom_bar(stat = "identity", fill = "#1f77b4") +
  facet_wrap(~ year) +
  labs(title = "YRK: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

# Merge with the original dataset, filling missing values with 0 for low_sal_stress
low_sal_complete_data <- complete_year_month %>%
  left_join(low_sal_stress_count, by = c("year", "month")) %>%
  mutate(low_sal_stress = ifelse(is.na(low_sal_stress), 0, low_sal_stress))

low_sal_complete_data$month <- as.numeric(low_sal_complete_data$month)

# bars
ggplot(low_sal_complete_data, aes(x = factor(month), y = low_sal_stress)) +
  geom_bar(stat = "identity", fill = "#1f77b4") + 
  facet_wrap(~ year) +
  labs(title = "YRK: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of low salinity stress days") +
  theme_minimal()

# points
ggplot(low_sal_complete_data, aes(x = factor(month), y = low_sal_stress)) +
  geom_point(color = "#1f77b4", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "YRK: count of low salinity (< 12 ppt) days per month",
       x = "month",
       y = "count of low salinity stress days") +
  theme_minimal()
```

# now calculate for high salinity stress
```{r high-sal-stress}
YRK$high_sal_stress <- YRK$salinity > 35

high_sal_stress_count <- YRK %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime)) %>% 
  filter(high_sal_stress == 1) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(high_sal_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(high_sal_stress_count) # no high sal stress 

```


# high temp stress calculations
```{r high-temp-stress}
YRK$high_temp_stress <- YRK$temp > 28

high_temp_stress_count <- YRK %>%
  mutate(year = year(datetime), 
         month = month(datetime), 
         day = day(datetime)) %>% 
  filter(high_temp_stress == 1) %>%  
  distinct(year, month, day) %>%  # remove dups
  group_by(year, month) %>%
  summarise(high_temp_stress = n(), .groups = "drop") # group all occurrences together by month rather than by numeric day

print(high_temp_stress_count)

ggplot(high_temp_stress_count, aes(x = factor(month), y = high_temp_stress, fill = factor(month))) +
  geom_bar(stat = "identity", fill = "#DD4124FF") +
  facet_wrap(~ year) +
  labs(title = "YRK: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()


high_temp_complete_data <- complete_year_month %>%
  left_join(high_temp_stress_count, by = c("year", "month")) %>%
  mutate(high_temp_stress = ifelse(is.na(high_temp_stress), 0, high_temp_stress))

high_temp_complete_data$month <- as.numeric(high_temp_complete_data$month)

# bars
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_bar(stat = "identity", fill = "#DD4124FF") + 
  facet_wrap(~ year) +
  labs(title = "YRK: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

# points
ggplot(high_temp_complete_data, aes(x = factor(month), y = high_temp_stress)) +
  geom_point(color = "#DD4124FF", size = 2) + 
  facet_wrap(~ year) +  
  labs(title = "YRK: count of high temperature (> 28 C) days per month",
       x = "month",
       y = "count of days") +
  theme_minimal()

```

### We can now calculate a list of variables that we will have collected for all sites. This will allow us to compare sites easily. We will calculate the number of observations from each site, the mean annual, maximum annual, and minimum annual value for all variables. 

Our list of variables includes: 

- Mean_Annual_Temperature_C: average of all available data
- Mean_max_temperature_C: average of maximums for each year
- Mean_min_temperature_C: average of minimums for each year
- Temperature_st_dev: standard deviation of all available data
- Temperature_n: total number of data points
- Temperature_years: number of years in data set

- Mean_Annual_Salinity_ppt: average of all available data
- Mean_min_Salinity_ppt: average of minimums for each year
- Mean_max_Salinity_ppt: average of maximums for each year
- Salinity_st_dev: standard deviation of all available data
- Salinity_n: total number of data points
- Salinity_years: number of years in data set

```{r}
#Calculate temperature variables. 
Mean_Annual_Temperature_C <- mean(YRK$temp)
Mean_max_temperature_C <- mean(YRK_envryear$max_temp)
Mean_min_temperature_C <- mean(YRK_envryear$min_temp)
Temperature_st_dev <- sd(YRK$temp)
Temperature_n <- nrow(YRK)
Temperature_years <- nrow(YRK_envryear)
high_temp_stress_days <- sum(high_temp_stress_count$high_temp_stress)
frac_high_temp_stress_days <- high_temp_stress_days/nrow(YRK_envrday)

Mean_Monthly_Temperature_C <- YRK_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_Temperature = mean(mean_temp))

Mean_min_Monthly_Temperature_C <- YRK_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_min_Temperature = mean(min_temp))

Mean_max_Monthly_Temperature_C <- YRK_envrmonth %>%
  filter(!is.na(month)) %>% 
  group_by(month) %>%
  summarise(Mean_max_Temperature = mean(max_temp))

#Create a data frame to store the temperature results
YRK_temp <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Temperature_C, Mean_max_temperature_C, Mean_min_temperature_C, Temperature_st_dev, high_temp_stress_days, frac_high_temp_stress_days, Temperature_n, Temperature_years, collection_type)
print(YRK_temp)

YRK_monthly_temp <- cbind(Mean_Monthly_Temperature_C, Mean_min_Monthly_Temperature_C, Mean_max_Monthly_Temperature_C)
YRK_monthly_temp <- YRK_monthly_temp[, !duplicated(names(YRK_monthly_temp))]
print(YRK_monthly_temp)

# Write to the combined file with all sites 
write.table(YRK_temp, "../../data/environment/all_temperature.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE) # The column names should be changed to FALSE after 1st row is added to the data frame

# Write to a unique new CSV file
write.csv(YRK_temp, "../../data/environment/YRK_temperature.csv")

# Write all montly data to a unique new CSV file
write.csv(YRK_monthly_temp, "../../data/environment/YRK_monthly_temp.csv")

```

```{r}
#Calculate the salinity variables
Mean_Annual_Salinity_ppt <- mean(YRK$salinity)
Mean_max_Salinity_ppt <- mean(YRK_envryear$max_salinity)
Mean_min_Salinity_ppt <- mean(YRK_envryear$min_salinity)
Salinity_st_dev <- sd(YRK$salinity)
Salinity_n <- nrow(YRK)
Salinity_years <- nrow(YRK_envryear)
high_sal_stress_days <- sum(high_sal_stress_count$high_sal_stress)
low_sal_stress_days <- sum(low_sal_stress_count$low_sal_stress)
frac_high_sal_stress_days <- high_sal_stress_days/nrow(YRK_envrday)
frac_low_sal_stress_days <- low_sal_stress_days/nrow(YRK_envrday)

Mean_Monthly_Salinity <- YRK_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Mean_Salinity = mean(mean_salinity))

Min_Monthly_Salinity <- YRK_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Min_Salinity = mean(min_salinity))

Max_Monthly_Salinity <- YRK_envrmonth %>%
  filter(!is.na(month)) %>%
  group_by(month) %>%
  summarise(Max_Salinity = mean(max_salinity))

#Create a data frame to store the temperature results
YRK_salinity <- cbind(site_name, download_date, source_description, lat, lon, firstyear, finalyear, Mean_Annual_Salinity_ppt, Mean_max_Salinity_ppt, Mean_min_Salinity_ppt, high_sal_stress_days,low_sal_stress_days, frac_high_sal_stress_days, frac_low_sal_stress_days, Salinity_st_dev, Salinity_n, Salinity_years, collection_type)
print(YRK_salinity)

YRK_monthly_sal <- cbind(Mean_Monthly_Salinity, Min_Monthly_Salinity, Max_Monthly_Salinity)
YRK_monthly_sal <- YRK_monthly_sal[, !duplicated(names(YRK_monthly_sal))]
print(YRK_monthly_sal)

# Write to the combined file with all sites 
write.table(YRK_salinity, "../../data/environment/all_salinity.csv", sep = ",", append = TRUE, col.names = FALSE, row.names = FALSE) # The column names should be changed to FALSE after 1st row is added to the data frame

# Write all year data to a unique new CSV file
write.csv(YRK_salinity, "../../data/environment/YRK_salinity.csv")
# Write all montly data to a unique new CSV file
write.csv(YRK_monthly_sal, "../../data/environment/YRK_monthly_sal.csv")
```
