---
title: "Q2c_JUV_genetic_diversity_models_field"
author: "Kiran Bajaj"
date: "2026-08-23"
output: pdf_document
---

This file addresses Q2c: the direct effect of offspring genetic diversity (heterozygosity, allelic richness) on oyster survival, length, and condition index at each field timepoint, **including both monoculture and polyculture treatments**. The mixed-origin groups (H1-HYBRIDMIX, H2-SEEDMIX) are included here because this approach uses only genetic diversity (which they have) and no environment-of-origin (which they lack).

Model structure matches Q2a/Q2b: full models, a population random effect (survival) or nested population/bag effect (length, CI) so genetic predictors are tested against the source groups, with Satterthwaite inference. Because predictors vary only across the 10 source groups, population-level df is low; non-significant effects mean "undetectable at this replication," not "absent."

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")
packages <- c("tidyverse", "here", "ggplot2", "lme4", "lmerTest", "car", "broom", "broom.mixed", "corrplot", "ggstance", "stringr")
pacman::p_load(char = packages)

r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>% map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>% paste(collapse = ", ")
```

```{r}
getwd()
```

## Read in data
```{r}
surv_mod_ord_rename     <- read.csv("data/performance_H2F/field_surv_all_times.csv")[, -1]
lengths_mod_narm_nodead <- read.csv("data/performance_H2F/field_length_all_times.csv")[ ,-1]
ci_df                   <- read_csv("data/performance_H2F/MVP_phenotyping_final.csv", show_col_types = FALSE)

# juvenile genetics — 10 groups, INCLUDING SEEDMIX and HYBRIDMIX
juv_genetics <- read.csv("results/juvenile_effects_H2F/juvenile_genetic_effects_full.csv")
```

## Data preparation
```{r}
colnames(juv_genetics) <- c("pop", "ho", "exp_hs", "ar")

# genetics file labels the mix groups SEEDMIX/HYBRIDMIX; performance data uses
# H2-SEEDMIX/H1-HYBRIDMIX. Recode genetics to match. dplyr:: guards against car::recode.
juv_genetics$pop <- dplyr::recode(juv_genetics$pop,
  "SEEDMIX" = "H2-SEEDMIX", "HYBRIDMIX" = "H1-HYBRIDMIX")

# survival + genetics (mix groups retained)
surv_mod_gen <- merge(surv_mod_ord_rename, juv_genetics, by = "pop")

# length + genetics
len_mod_gen  <- merge(lengths_mod_narm_nodead, juv_genetics, by = "pop")

# condition index: build frame, THEN recode hybrid label (CI data calls it H1-LARMIX), THEN merge
ci_sub <- ci_df %>%
  mutate(population = as.character(population)) %>%
  filter(!is.na(population), population != "NA") %>%
  transmute(
    ci         = CI,
    t          = fct_recode(as.factor(samplingTimePoint), t1 = "2023-11", t2 = "2024-05", t3 = "2024-11"),
    bag_site   = fct_recode(as.factor(site), Lewisetta = "LEW", YorkRiver = "YR"),
    pop        = population,
    bags_label = bagID
  )
ci_sub$pop <- dplyr::recode(ci_sub$pop, "H1-LARMIX" = "H1-HYBRIDMIX")
ci_gen <- merge(ci_sub, juv_genetics, by = "pop")

# GUARD: all three frames must carry all 10 groups, or a label drift silently dropped some
stopifnot(
  length(unique(surv_mod_gen$pop)) == 10,
  length(unique(len_mod_gen$pop))  == 10,
  length(unique(ci_gen$pop))       == 10
)
```

## Survival ~ ho + ar

```{r}
surv_lew_t1 <- surv_mod_gen[surv_mod_gen$t == 1 & surv_mod_gen$bag_site == "Lewisetta", ]
surv_G1.lew <- lmer(surv ~ ho + ar + (1 | pop), data = surv_lew_t1)
isSingular(surv_G1.lew); summary(surv_G1.lew)
surv_t1_lew_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + (1 | pop), data = surv_lew_t1)

surv_lew_t2 <- surv_mod_gen[surv_mod_gen$t == 2 & surv_mod_gen$bag_site == "Lewisetta", ]
surv_G2.lew <- lmer(surv ~ ho + ar + (1 | pop), data = surv_lew_t2)
isSingular(surv_G2.lew); summary(surv_G2.lew)
surv_t2_lew_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + (1 | pop), data = surv_lew_t2)

surv_lew_t3 <- surv_mod_gen[surv_mod_gen$t == 3 & surv_mod_gen$bag_site == "Lewisetta", ]
surv_G3.lew <- lmer(surv ~ ho + ar + (1 | pop), data = surv_lew_t3)
isSingular(surv_G3.lew); summary(surv_G3.lew)
surv_t3_lew_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + (1 | pop), data = surv_lew_t3)

surv_yrk_t1 <- surv_mod_gen[surv_mod_gen$t == 1 & surv_mod_gen$bag_site == "YorkRiver", ]
surv_G1.york <- lmer(surv ~ ho + ar + (1 | pop), data = surv_yrk_t1)
isSingular(surv_G1.york); summary(surv_G1.york)
surv_t1_yrk_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + (1 | pop), data = surv_yrk_t1)

surv_yrk_t2 <- surv_mod_gen[surv_mod_gen$t == 2 & surv_mod_gen$bag_site == "YorkRiver", ]
surv_G2.york <- lmer(surv ~ ho + ar + (1 | pop), data = surv_yrk_t2)
isSingular(surv_G2.york); summary(surv_G2.york)
surv_t2_yrk_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + (1 | pop), data = surv_yrk_t2)

surv_yrk_t3 <- surv_mod_gen[surv_mod_gen$t == 3 & surv_mod_gen$bag_site == "YorkRiver", ]
surv_G3.york <- lmer(surv ~ ho + ar + (1 | pop), data = surv_yrk_t3)
isSingular(surv_G3.york); summary(surv_G3.york)
surv_t3_yrk_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + (1 | pop), data = surv_yrk_t3)

# save survival results
sink(paste0("results/juvenile_effects_H2F/models/survival_Q2c_genetics_results_", Sys.Date(), ".txt"))
for (nm in c("surv_G1.lew","surv_G2.lew","surv_G3.lew","surv_G1.york","surv_G2.york","surv_G3.york")) {
  m <- get(nm); cat("\n", nm, "  singular fit:", isSingular(m), "\n"); print(summary(m))
}
for (nm in c("surv_t1_lew_scaled","surv_t2_lew_scaled","surv_t3_lew_scaled","surv_t1_yrk_scaled","surv_t2_yrk_scaled","surv_t3_yrk_scaled")) {
  cat("\n", nm, "(scaled)\n"); print(summary(get(nm)))
}
sink()
```

## Length ~ ho + ar + (1|pop/bags_label)

```{r}
len_lew_t1 <- len_mod_gen[len_mod_gen$t == 1 & len_mod_gen$bag_site == "Lewisetta", ]
len_G1.lew <- lmer(length ~ ho + ar + (1 | pop / bags_label), data = len_lew_t1)
isSingular(len_G1.lew); VarCorr(len_G1.lew); summary(len_G1.lew)
len_t1_lew_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = len_lew_t1)

len_lew_t2 <- len_mod_gen[len_mod_gen$t == 2 & len_mod_gen$bag_site == "Lewisetta", ]
len_G2.lew <- lmer(length ~ ho + ar + (1 | pop / bags_label), data = len_lew_t2)
isSingular(len_G2.lew); VarCorr(len_G2.lew); summary(len_G2.lew)
len_t2_lew_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = len_lew_t2)

len_lew_t3 <- len_mod_gen[len_mod_gen$t == 3 & len_mod_gen$bag_site == "Lewisetta", ]
len_G3.lew <- lmer(length ~ ho + ar + (1 | pop / bags_label), data = len_lew_t3)
isSingular(len_G3.lew); VarCorr(len_G3.lew); summary(len_G3.lew)
len_t3_lew_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = len_lew_t3)

len_yrk_t1 <- len_mod_gen[len_mod_gen$t == 1 & len_mod_gen$bag_site == "YorkRiver", ]
len_G1.york <- lmer(length ~ ho + ar + (1 | pop / bags_label), data = len_yrk_t1)
isSingular(len_G1.york); VarCorr(len_G1.york); summary(len_G1.york)
len_t1_yrk_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = len_yrk_t1)

len_yrk_t2 <- len_mod_gen[len_mod_gen$t == 2 & len_mod_gen$bag_site == "YorkRiver", ]
len_G2.york <- lmer(length ~ ho + ar + (1 | pop / bags_label), data = len_yrk_t2)
isSingular(len_G2.york); VarCorr(len_G2.york); summary(len_G2.york)
len_t2_yrk_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = len_yrk_t2)

len_yrk_t3 <- len_mod_gen[len_mod_gen$t == 3 & len_mod_gen$bag_site == "YorkRiver", ]
len_G3.york <- lmer(length ~ ho + ar + (1 | pop / bags_label), data = len_yrk_t3)
isSingular(len_G3.york); VarCorr(len_G3.york); summary(len_G3.york)
len_t3_yrk_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = len_yrk_t3)

# save length results
sink(paste0("results/juvenile_effects_H2F/models/length_Q2c_genetics_results_", Sys.Date(), ".txt"))
for (nm in c("len_G1.lew","len_G2.lew","len_G3.lew","len_G1.york","len_G2.york","len_G3.york")) {
  m <- get(nm); cat("\n", nm, "  singular fit:", isSingular(m), "\n"); print(VarCorr(m)); print(summary(m))
}
for (nm in c("len_t1_lew_scaled","len_t2_lew_scaled","len_t3_lew_scaled","len_t1_yrk_scaled","len_t2_yrk_scaled","len_t3_yrk_scaled")) {
  cat("\n", nm, "(scaled)\n"); print(summary(get(nm)))
}
sink()
```

## Condition index ~ ho + ar + (1|pop/bags_label)

```{r}
ci_lew_t1 <- ci_gen[ci_gen$t == "t1" & ci_gen$bag_site == "Lewisetta", ]
ci_G1.lew <- lmer(ci ~ ho + ar + (1 | pop / bags_label), data = ci_lew_t1)
isSingular(ci_G1.lew); VarCorr(ci_G1.lew); summary(ci_G1.lew)
ci_t1_lew_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = ci_lew_t1)

ci_lew_t2 <- ci_gen[ci_gen$t == "t2" & ci_gen$bag_site == "Lewisetta", ]
ci_G2.lew <- lmer(ci ~ ho + ar + (1 | pop / bags_label), data = ci_lew_t2)
isSingular(ci_G2.lew); VarCorr(ci_G2.lew); summary(ci_G2.lew)
ci_t2_lew_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = ci_lew_t2)

ci_lew_t3 <- ci_gen[ci_gen$t == "t3" & ci_gen$bag_site == "Lewisetta", ]
ci_G3.lew <- lmer(ci ~ ho + ar + (1 | pop / bags_label), data = ci_lew_t3)
isSingular(ci_G3.lew); VarCorr(ci_G3.lew); summary(ci_G3.lew)
ci_t3_lew_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = ci_lew_t3)

ci_yrk_t1 <- ci_gen[ci_gen$t == "t1" & ci_gen$bag_site == "YorkRiver", ]
ci_G1.york <- lmer(ci ~ ho + ar + (1 | pop / bags_label), data = ci_yrk_t1)
isSingular(ci_G1.york); VarCorr(ci_G1.york); summary(ci_G1.york)
ci_t1_yrk_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = ci_yrk_t1)

ci_yrk_t2 <- ci_gen[ci_gen$t == "t2" & ci_gen$bag_site == "YorkRiver", ]
ci_G2.york <- lmer(ci ~ ho + ar + (1 | pop / bags_label), data = ci_yrk_t2)
isSingular(ci_G2.york); VarCorr(ci_G2.york); summary(ci_G2.york)
ci_t2_yrk_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = ci_yrk_t2)

ci_yrk_t3 <- ci_gen[ci_gen$t == "t3" & ci_gen$bag_site == "YorkRiver", ]
ci_G3.york <- lmer(ci ~ ho + ar + (1 | pop / bags_label), data = ci_yrk_t3)
isSingular(ci_G3.york); VarCorr(ci_G3.york); summary(ci_G3.york)
ci_t3_yrk_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + (1 | pop / bags_label), data = ci_yrk_t3)

# save CI results
sink(paste0("results/juvenile_effects_H2F/models/condition_Q2c_genetics_results_", Sys.Date(), ".txt"))
for (nm in c("ci_G1.lew","ci_G2.lew","ci_G3.lew","ci_G1.york","ci_G2.york","ci_G3.york")) {
  m <- get(nm); cat("\n", nm, "  singular fit:", isSingular(m), "\n"); print(VarCorr(m)); print(summary(m))
}
for (nm in c("ci_t1_lew_scaled","ci_t2_lew_scaled","ci_t3_lew_scaled","ci_t1_yrk_scaled","ci_t2_yrk_scaled","ci_t3_yrk_scaled")) {
  cat("\n", nm, "(scaled)\n"); print(summary(get(nm)))
}
sink()
```

## Figures
```{r}
build_slope_df <- function(m_t1, m_t2, m_t3) {
  bind_rows(
    tidy(m_t1, effects = "fixed") %>% mutate(t = "t1"),
    tidy(m_t2, effects = "fixed") %>% mutate(t = "t2"),
    tidy(m_t3, effects = "fixed") %>% mutate(t = "t3")
  ) %>%
    filter(term != "(Intercept)") %>%
    mutate(tmult = qt(0.975, df),
           lower = estimate - tmult * std.error,
           upper = estimate + tmult * std.error) %>%
    select(explanatory = term, slope = estimate, lower, upper, t)
}

time_labels <- c("t1" = "November 2023", "t2" = "May 2024", "t3" = "November 2024")
y_limits <- c("scale(ho)", "scale(ar)")
y_labels <- c("Heterozygosity", "Allelic richness")

plot_slopes <- function(df, point_color) {
  ggplot(df, aes(x = slope, y = explanatory)) +
    geom_point(aes(alpha = t, shape = t), color = point_color, size = 3.5,
               position = position_dodge2v(height = .85, reverse = T, preserve = "total")) +
    geom_errorbarh(aes(xmax = upper, xmin = lower, alpha = t), color = point_color,
                   height = 0, size = 1.2, position = position_dodgev(height = -.85, preserve = "total")) +
    scale_alpha_manual(name = "Time", na.translate = F,
                       values = c("t1" = .45, "t2" = .7, "t3" = 1), labels = time_labels) +
    scale_shape_manual(name = "Time", na.translate = F,
                       values = c("t1" = 15, "t2" = 16, "t3" = 17), labels = time_labels) +
    coord_cartesian(xlim = c(-1.6, 2)) +
    xlab("Slope on scaled data") + ylab("Variable") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_y_discrete(name = "Explanatory variable", limits = y_limits, labels = y_labels) +
    theme_classic() +
    theme(axis.text = element_text(color = "black", size = 14),
          axis.title = element_text(size = 15),
          legend.text = element_text(size = 14),
          legend.title = element_text(size = 15))
}

# survival
plot_slopes(build_slope_df(surv_t1_lew_scaled, surv_t2_lew_scaled, surv_t3_lew_scaled), "#F06292") -> p; p
ggsave("FigSx_lew_surv_genetics.pdf", p, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)
plot_slopes(build_slope_df(surv_t1_yrk_scaled, surv_t2_yrk_scaled, surv_t3_yrk_scaled), "#880E4F") -> p; p
ggsave("FigSx_york_surv_genetics.pdf", p, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)

# length
plot_slopes(build_slope_df(len_t1_lew_scaled, len_t2_lew_scaled, len_t3_lew_scaled), "#F06292") -> p; p
ggsave("FigSx_lew_length_genetics.pdf", p, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)
plot_slopes(build_slope_df(len_t1_yrk_scaled, len_t2_yrk_scaled, len_t3_yrk_scaled), "#880E4F") -> p; p
ggsave("FigSx_york_length_genetics.pdf", p, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)

# condition index
plot_slopes(build_slope_df(ci_t1_lew_scaled, ci_t2_lew_scaled, ci_t3_lew_scaled), "#F06292") -> p; p
ggsave("FigSx_lew_ci_genetics.pdf", p, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)
plot_slopes(build_slope_df(ci_t1_yrk_scaled, ci_t2_yrk_scaled, ci_t3_yrk_scaled), "#880E4F") -> p; p
ggsave("FigSx_york_ci_genetics.pdf", p, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)
```

## Check assumptions
```{r}
all_models <- list(
  "surv Lew t1"=surv_G1.lew,"surv Lew t2"=surv_G2.lew,"surv Lew t3"=surv_G3.lew,
  "surv Yrk t1"=surv_G1.york,"surv Yrk t2"=surv_G2.york,"surv Yrk t3"=surv_G3.york,
  "len Lew t1"=len_G1.lew,"len Lew t2"=len_G2.lew,"len Lew t3"=len_G3.lew,
  "len Yrk t1"=len_G1.york,"len Yrk t2"=len_G2.york,"len Yrk t3"=len_G3.york,
  "CI Lew t1"=ci_G1.lew,"CI Lew t2"=ci_G2.lew,"CI Lew t3"=ci_G3.lew,
  "CI Yrk t1"=ci_G1.york,"CI Yrk t2"=ci_G2.york,"CI Yrk t3"=ci_G3.york
)

op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (nm in names(all_models)) {
  m <- all_models[[nm]]
  plot(fitted(m), resid(m), main = paste(nm, "resid vs fitted"), xlab = "Fitted", ylab = "Residual", pch = 19, col = rgb(0,0,0,0.3)); abline(h = 0, lty = 2)
  qqnorm(resid(m), main = paste(nm, "QQ")); qqline(resid(m))
}
par(op)

sapply(all_models, function(m) c(singular = isSingular(m),
                                 shapiro_p = tryCatch(shapiro.test(resid(m))$p.value, error=function(e) NA),
                                 n = length(resid(m))))
```