---
title: "Q2_JUV_condition_index_models_field"
author: "Kiran Bajaj"
date: "2026-08-23"
output: pdf_document
---

This file models the effects of offspring genetic diversity (heterozygosity, allelic richness) and environment-of-origin — as both absolute environmental quantiles (Q2a) and environmental distance to the field site (Q2b) — on oyster condition index (monocultures only).

CI data and calculations are from ZS. Modeling structure mirrors the Q2a/Q2b notebooks: full models (no stepwise selection), a nested population/bag random effect so population-level predictors are tested against the 8 source populations, and Satterthwaite-based inference.

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("tidyverse", "here", "ggplot2", "lme4", "lmerTest", "car", "broom", "broom.mixed", "corrplot", "ggstance", "stringr")
pacman::p_load(char = packages)

r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

## Read in data
```{r}
# condition index
ci_df <- read_csv("data/performance_H2F/MVP_phenotyping_final.csv", show_col_types = FALSE)

# juvenile heterozygosity and allelic richness (recoded genotype matrix)
juv_genetics <- read.csv("results/juvenile_effects_H2F/juvenile_genetic_effects_full.csv")

# environment of origin quantiles
env_parents <- read.csv("results/environment/envr_quantiles_allsites.csv")

# pairwise environmental distance matrix; LEW/YRK cols = distance to each field site
env_dist <- read.csv("results/environment/envr_quantiles_dist_all.csv")
```

## Data preparation

### Clean CI data
```{r}
# recode to match the other notebooks: pop / t / bag_site / bags_label
ci_sub <- ci_df %>%
  mutate(population = as.character(population)) %>%
  # drop mixed-origin groups (CI data labels the hybrid as LARMIX) and missing pops
  filter(!population %in% c("H1-LARMIX", "H1-HYBRIDMIX", "H2-SEEDMIX"),
         !is.na(population)) %>%
  transmute(
    ci         = CI,
    t          = fct_recode(as.factor(samplingTimePoint),
                            t1 = "2023-11", t2 = "2024-05", t3 = "2024-11"),
    bag_site   = fct_recode(as.factor(site), Lewisetta = "LEW", YorkRiver = "YR"),
    pop        = population,
    bags_label = bagID
  )
```

### Combine CI + genetics + environment (quantiles, Q2a)
```{r}
colnames(juv_genetics) <- c("pop", "ho", "exp_hs", "ar")
ci_gen <- merge(ci_sub, juv_genetics, by = "pop")

colnames(env_parents) <- c("pop","salinity_quantile_10","salinity_quantile_90","Salinity_st_dev","temp_quantile_10","temp_quantile_90","Temperature_st_dev", "salinity")
ci_gen_env <- merge(ci_gen, env_parents, by = "pop")
```

### Combine CI + genetics + distance (Q2b)
```{r}
# melt distance matrix to one dist column, site-matched (same approach as Q2b)
env_dist_long <- env_dist %>%
  select(pop = X, Lewisetta = LEW, YorkRiver = YRK) %>%
  pivot_longer(cols = c(Lewisetta, YorkRiver),
               names_to = "bag_site", values_to = "dist")

ci_gen_dist <- merge(ci_gen, env_dist_long, by = c("pop", "bag_site"))

# confirm mixed groups gone and check structure
unique(ci_gen_env$pop)
table(ci_gen_env$pop, ci_gen_env$bag_site)
# confirm bag labels are unique per population (justifies the nested random effect)
# with(ci_gen_env, table(bags_label, pop))
```

## Assess collinearity
```{r}
# --- corrplots: quantile predictors + genetics + CI ---
cor_vars_q   <- c("temp_quantile_90", "salinity_quantile_10", "salinity_quantile_90", "ho", "ar", "ci")
cor_labels_q <- c("Temperature Q90", "Salinity Q10", "Salinity Q90", "Heterozygosity", "Allelic richness", "Condition index")

q_mat <- cor(ci_gen_env[, cor_vars_q], use = "complete.obs")
dimnames(q_mat) <- list(cor_labels_q, cor_labels_q)

pdf(file = "figures/juvenile_effects_H2F/FigS7_corrplot_ci_quantiles.pdf")
corrplot(q_mat, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         title = "Pairwise correlations — CI quantiles", addCoef.col = "black", mar = c(0,0,2,0))
dev.off()
corrplot(q_mat, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         title = "Pairwise correlations — CI quantiles", addCoef.col = "black", mar = c(0,0,2,0))

# --- corrplots: distance + genetics + CI, per site (dist is site-specific) ---
cor_vars_d   <- c("dist", "ho", "ar", "ci")
cor_labels_d <- c("Environmental distance", "Heterozygosity", "Allelic richness", "Condition index")

for (s in c("Lewisetta", "YorkRiver")) {
  m <- cor(ci_gen_dist[ci_gen_dist$bag_site == s, cor_vars_d], use = "complete.obs")
  dimnames(m) <- list(cor_labels_d, cor_labels_d)
  corrplot(m, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
           title = paste("Pairwise correlations — CI distance,", s),
           addCoef.col = "black", mar = c(0,0,2,0))
}
```

```{r}
# brief VIF. NOTE: predictors vary across only 8 source populations, so indicative.
VIF5 <- function(a, b, c, d, e) 1 / (1 - summary(lm(a ~ b + c + d + e))$r.squared)
d <- ci_gen_env
VIF5(d$temp_quantile_90,     d$salinity_quantile_10, d$salinity_quantile_90, d$ho, d$ar)  # temp90
VIF5(d$salinity_quantile_10, d$salinity_quantile_90, d$temp_quantile_90,     d$ho, d$ar)  # sal10
VIF5(d$salinity_quantile_90, d$salinity_quantile_10, d$temp_quantile_90,     d$ho, d$ar)  # sal90
VIF5(d$ho, d$salinity_quantile_10, d$salinity_quantile_90, d$temp_quantile_90, d$ar)      # ho
VIF5(d$ar, d$salinity_quantile_10, d$salinity_quantile_90, d$temp_quantile_90, d$ho)      # ar
```

## Model 1: genetic diversity + environmental quantiles (Q2a)

### t1 CI, Lewisetta
```{r}
ci_lew_t1 <- ci_gen_env[ci_gen_env$t == "t1" & ci_gen_env$bag_site == "Lewisetta", ]
ci_GE1.lew <- lmer(ci ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 + (1 | pop / bags_label), data = ci_lew_t1)
isSingular(ci_GE1.lew); VarCorr(ci_GE1.lew); summary(ci_GE1.lew)
ci_t1_lew_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) + scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label), data = ci_lew_t1)
isSingular(ci_t1_lew_scaled)
```

### t2 CI, Lewisetta
```{r}
ci_lew_t2 <- ci_gen_env[ci_gen_env$t == "t2" & ci_gen_env$bag_site == "Lewisetta", ]
ci_GE2.lew <- lmer(ci ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 + (1 | pop / bags_label), data = ci_lew_t2)
isSingular(ci_GE2.lew); VarCorr(ci_GE2.lew); summary(ci_GE2.lew)
ci_t2_lew_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) + scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label), data = ci_lew_t2)
isSingular(ci_t2_lew_scaled)
```

### t3 CI, Lewisetta
```{r}
ci_lew_t3 <- ci_gen_env[ci_gen_env$t == "t3" & ci_gen_env$bag_site == "Lewisetta", ]
ci_GE3.lew <- lmer(ci ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 + (1 | pop / bags_label), data = ci_lew_t3)
isSingular(ci_GE3.lew); VarCorr(ci_GE3.lew); summary(ci_GE3.lew)
ci_t3_lew_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) + scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label), data = ci_lew_t3)
isSingular(ci_t3_lew_scaled)
```

### t1 CI, York River
```{r}
ci_yrk_t1 <- ci_gen_env[ci_gen_env$t == "t1" & ci_gen_env$bag_site == "YorkRiver", ]
ci_GE1.york <- lmer(ci ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 + (1 | pop / bags_label), data = ci_yrk_t1)
isSingular(ci_GE1.york); VarCorr(ci_GE1.york); summary(ci_GE1.york)
ci_t1_yrk_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) + scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label), data = ci_yrk_t1)
isSingular(ci_t1_yrk_scaled)
```

### t2 CI, York River
```{r}
ci_yrk_t2 <- ci_gen_env[ci_gen_env$t == "t2" & ci_gen_env$bag_site == "YorkRiver", ]
ci_GE2.york <- lmer(ci ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 + (1 | pop / bags_label), data = ci_yrk_t2)
isSingular(ci_GE2.york); VarCorr(ci_GE2.york); summary(ci_GE2.york)
ci_t2_yrk_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) + scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label), data = ci_yrk_t2)
isSingular(ci_t2_yrk_scaled)
```

### t3 CI, York River
```{r}
ci_yrk_t3 <- ci_gen_env[ci_gen_env$t == "t3" & ci_gen_env$bag_site == "YorkRiver", ]
ci_GE3.york <- lmer(ci ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 + (1 | pop / bags_label), data = ci_yrk_t3)
isSingular(ci_GE3.york); VarCorr(ci_GE3.york); summary(ci_GE3.york)
ci_t3_yrk_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) + scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label), data = ci_yrk_t3)
isSingular(ci_t3_yrk_scaled)

# save quantile results
sink(paste0("results/juvenile_effects_H2F/models/condition_Q2a_quantiles_results_", Sys.Date(), ".txt"))
for (nm in c("ci_GE1.lew","ci_GE2.lew","ci_GE3.lew","ci_GE1.york","ci_GE2.york","ci_GE3.york")) {
  m <- get(nm)
  cat("\n", nm, "  singular fit:", isSingular(m), "\n"); print(VarCorr(m)); print(summary(m))
}
for (nm in c("ci_t1_lew_scaled","ci_t2_lew_scaled","ci_t3_lew_scaled","ci_t1_yrk_scaled","ci_t2_yrk_scaled","ci_t3_yrk_scaled")) {
  cat("\n", nm, "(scaled)\n"); print(summary(get(nm)))
}
sink()
```

## Model 2: genetic diversity + environmental distance (Q2b)

### t1 CI, Lewisetta
```{r}
cid_lew_t1 <- ci_gen_dist[ci_gen_dist$t == "t1" & ci_gen_dist$bag_site == "Lewisetta", ]
ci_D1.lew <- lmer(ci ~ ho + ar + dist + (1 | pop / bags_label), data = cid_lew_t1)
isSingular(ci_D1.lew); VarCorr(ci_D1.lew); summary(ci_D1.lew)
cid_t1_lew_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = cid_lew_t1)
isSingular(cid_t1_lew_scaled)
```

### t2 CI, Lewisetta
```{r}
cid_lew_t2 <- ci_gen_dist[ci_gen_dist$t == "t2" & ci_gen_dist$bag_site == "Lewisetta", ]
ci_D2.lew <- lmer(ci ~ ho + ar + dist + (1 | pop / bags_label), data = cid_lew_t2)
isSingular(ci_D2.lew); VarCorr(ci_D2.lew); summary(ci_D2.lew)
cid_t2_lew_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = cid_lew_t2)
isSingular(cid_t2_lew_scaled)
```

### t3 CI, Lewisetta
```{r}
cid_lew_t3 <- ci_gen_dist[ci_gen_dist$t == "t3" & ci_gen_dist$bag_site == "Lewisetta", ]
ci_D3.lew <- lmer(ci ~ ho + ar + dist + (1 | pop / bags_label), data = cid_lew_t3)
isSingular(ci_D3.lew); VarCorr(ci_D3.lew); summary(ci_D3.lew)
cid_t3_lew_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = cid_lew_t3)
isSingular(cid_t3_lew_scaled)
```

### t1 CI, York River
```{r}
cid_yrk_t1 <- ci_gen_dist[ci_gen_dist$t == "t1" & ci_gen_dist$bag_site == "YorkRiver", ]
ci_D1.york <- lmer(ci ~ ho + ar + dist + (1 | pop / bags_label), data = cid_yrk_t1)
isSingular(ci_D1.york); VarCorr(ci_D1.york); summary(ci_D1.york)
cid_t1_yrk_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = cid_yrk_t1)
isSingular(cid_t1_yrk_scaled)
```

### t2 CI, York River
```{r}
cid_yrk_t2 <- ci_gen_dist[ci_gen_dist$t == "t2" & ci_gen_dist$bag_site == "YorkRiver", ]
ci_D2.york <- lmer(ci ~ ho + ar + dist + (1 | pop / bags_label), data = cid_yrk_t2)
isSingular(ci_D2.york); VarCorr(ci_D2.york); summary(ci_D2.york)
cid_t2_yrk_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = cid_yrk_t2)
isSingular(cid_t2_yrk_scaled)
```

### t3 CI, York River
```{r}
cid_yrk_t3 <- ci_gen_dist[ci_gen_dist$t == "t3" & ci_gen_dist$bag_site == "YorkRiver", ]
ci_D3.york <- lmer(ci ~ ho + ar + dist + (1 | pop / bags_label), data = cid_yrk_t3)
isSingular(ci_D3.york); VarCorr(ci_D3.york); summary(ci_D3.york)
cid_t3_yrk_scaled <- lmer(scale(ci) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = cid_yrk_t3)
isSingular(cid_t3_yrk_scaled)

# save distance results
sink(paste0("results/juvenile_effects_H2F/models/condition_Q2b_envrdist_results_", Sys.Date(), ".txt"))
for (nm in c("ci_D1.lew","ci_D2.lew","ci_D3.lew","ci_D1.york","ci_D2.york","ci_D3.york")) {
  m <- get(nm)
  cat("\n", nm, "  singular fit:", isSingular(m), "\n"); print(VarCorr(m)); print(summary(m))
}
for (nm in c("cid_t1_lew_scaled","cid_t2_lew_scaled","cid_t3_lew_scaled","cid_t1_yrk_scaled","cid_t2_yrk_scaled","cid_t3_yrk_scaled")) {
  cat("\n", nm, "(scaled)\n"); print(summary(get(nm)))
}
sink()
```

## Figures
```{r}
build_slope_df <- function(m_t1, m_t2, m_t3) {
  bind_rows(
    tidy(m_t1, effects = "fixed") %>% mutate(t = "t1"),
    tidy(m_t2, effects = "fixed") %>% mutate(t = "t2"),
    tidy(m_t3, effects = "fixed") %>% mutate(t = "t3")
  ) %>%
    filter(term != "(Intercept)") %>%
    mutate(tmult = qt(0.975, df),
           lower = estimate - tmult * std.error,
           upper = estimate + tmult * std.error) %>%
    select(explanatory = term, slope = estimate, lower, upper, t)
}

time_labels <- c("t1" = "November 2023", "t2" = "May 2024", "t3" = "November 2024")

plot_slopes <- function(df, point_color, y_limits, y_labels) {
  ggplot(df, aes(x = slope, y = explanatory)) +
    geom_point(aes(alpha = t, shape = t), color = point_color, size = 3.5,
               position = position_dodge2v(height = .85, reverse = T, preserve = "total")) +
    geom_errorbarh(aes(xmax = upper, xmin = lower, alpha = t), color = point_color,
                   height = 0, size = 1.2,
                   position = position_dodgev(height = -.85, preserve = "total")) +
    scale_alpha_manual(name = "Time", na.translate = F,
                       values = c("t1" = .45, "t2" = .7, "t3" = 1), labels = time_labels) +
    scale_shape_manual(name = "Time", na.translate = F,
                       values = c("t1" = 15, "t2" = 16, "t3" = 17), labels = time_labels) +
    coord_cartesian(xlim = c(-1.6, 2)) +
    xlab("Slope on scaled data") + ylab("Variable") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_y_discrete(name = "Explanatory variable", limits = y_limits, labels = y_labels) +
    theme_classic() +
    theme(axis.text = element_text(color = "black", size = 14),
          axis.title = element_text(size = 15),
          legend.text = element_text(size = 14),
          legend.title = element_text(size = 15))
}

# quantile label set
yq_limits <- c("scale(salinity_quantile_10)", "scale(salinity_quantile_90)", "scale(temp_quantile_90)", "scale(ho)", "scale(ar)")
yq_labels <- c("Salinity Q10", "Salinity Q90", "Temperature Q90", "Heterozygosity", "Allelic richness")

# distance label set
yd_limits <- c("scale(dist)", "scale(ho)", "scale(ar)")
yd_labels <- c(str_wrap("Environmental distance (EoO to field site)", 20), "Heterozygosity", "Allelic richness")

# --- CI quantiles ---
lew_ci1 <- plot_slopes(build_slope_df(ci_t1_lew_scaled, ci_t2_lew_scaled, ci_t3_lew_scaled), "#F06292", yq_limits, yq_labels)
lew_ci1; ggsave("FigS7A_lew_ci_quantiles.pdf", lew_ci1, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)

yrk_ci1 <- plot_slopes(build_slope_df(ci_t1_yrk_scaled, ci_t2_yrk_scaled, ci_t3_yrk_scaled), "#880E4F", yq_limits, yq_labels)
yrk_ci1; ggsave("FigS7C_york_ci_quantiles.pdf", yrk_ci1, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)

# --- CI distance ---
lew_ci2 <- plot_slopes(build_slope_df(cid_t1_lew_scaled, cid_t2_lew_scaled, cid_t3_lew_scaled), "#F06292", yd_limits, yd_labels)
lew_ci2; ggsave("FigS7B_lew_ci_envrdist.pdf", lew_ci2, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)

yrk_ci2 <- plot_slopes(build_slope_df(cid_t1_yrk_scaled, cid_t2_yrk_scaled, cid_t3_yrk_scaled), "#880E4F", yd_limits, yd_labels)
yrk_ci2; ggsave("FigS7D_york_ci_envrdist.pdf", yrk_ci2, device = "pdf", path = "figures/juvenile_effects_H2F", height = 4, width = 7)
```

## Check assumptions
```{r}
ci_models <- list(
  "CI-q Lew t1"=ci_GE1.lew, "CI-q Lew t2"=ci_GE2.lew, "CI-q Lew t3"=ci_GE3.lew,
  "CI-q Yrk t1"=ci_GE1.york,"CI-q Yrk t2"=ci_GE2.york,"CI-q Yrk t3"=ci_GE3.york,
  "CI-d Lew t1"=ci_D1.lew,  "CI-d Lew t2"=ci_D2.lew,  "CI-d Lew t3"=ci_D3.lew,
  "CI-d Yrk t1"=ci_D1.york, "CI-d Yrk t2"=ci_D2.york, "CI-d Yrk t3"=ci_D3.york
)
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (nm in names(ci_models)) {
  m <- ci_models[[nm]]
  plot(fitted(m), resid(m), main = paste(nm, "- resid vs fitted"), xlab = "Fitted", ylab = "Residual", pch = 19, col = rgb(0,0,0,0.3)); abline(h = 0, lty = 2)
  qqnorm(resid(m), main = paste(nm, "- QQ")); qqline(resid(m))
}
par(op)

sapply(ci_models, function(m) c(shapiro_p = tryCatch(shapiro.test(resid(m))$p.value, error=function(e) NA), n = length(resid(m))))
```