---
title: "Q2a_JUV_envr_quantile_models_field"
author: "Kiran Bajaj"
date: "2026-08-23"
output: pdf_document
---

This file contains analyses that address Q2a: effects of offspring genetic diversity (heterozygosity, allelic richness) and absolute environmental variables (temperature and salinity quantiles of environments-of-origin) on oyster survival and length at a given timepoint (monocultures only).

Modeling is based on code from CR.

260823 KEB: updated these models to include juvenile heterozygosity and allelic richness data which is calculated correctly based on the recoded genotype matrix.

## Load packages

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("tidyverse", "here", "ggplot2", "lme4", "lmerTest", "car", "broom", "broom.mixed", "corrplot", "ggstance", "stringr")
pacman::p_load(char = packages)

r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")

```

```{r}
getwd()
```

## Read in data

```{r}

# survival
surv_mod_ord_rename <- read.csv("data/performance_H2F/field_surv_all_times.csv")[, -1]

# length
lengths_mod_narm_nodead <- read.csv("data/performance_H2F/field_length_all_times.csv")[ ,-1]

# this file contains data on mean heterozygosity and allelic richness in the juvenile populations
juv_genetics <- read.csv("results/juvenile_effects_H2F/juvenile_genetic_effects_full.csv")

# environment of origin quantiles
env_parents <- read.csv("results/environment/envr_quantiles_allsites.csv")

```

For the H2F paper, we want to know how environment and genetics are impacting differential survival. I'll start by putting together the complete dataframe.

Because the H1-HYBRIDMIX and H2-SEEDMIX groups do not have a single environment-of-origin, those groups are excluded from these analyses.

## Modeling drivers of differences in survival

### Data preparation

Combine environmental, genetic, and survival data into a data frame for the downstream models.

```{r}

# genetic data
colnames(juv_genetics) <- c("pop","ho","exp_hs","ar")
surv_mod_gen <- merge(surv_mod_ord_rename, juv_genetics, by = "pop")

# clean env_parents
colnames(env_parents) <- c("pop","salinity_quantile_10","salinity_quantile_90","Salinity_st_dev","temp_quantile_10","temp_quantile_90","Temperature_st_dev", "salinity")

# survival + genetics + environment (used for the survival models)
surv_mod_gen_env <- merge(surv_mod_gen, env_parents, by = "pop")

# also build a survival + length frame for the correlation matrix only
# summarize length to mean per pop/bag_site/bags_label/t so it matches survival
length_summarized <- lengths_mod_narm_nodead %>%
  group_by(bag_site, pop, t, bags_label) %>%
  summarise(mean_len = mean(length, na.rm = TRUE), .groups = "drop")

surv_len_mod_gen     <- merge(length_summarized, surv_mod_gen,
                              by = c("pop", "bag_site", "bags_label", "t"))
surv_len_mod_gen_env <- merge(surv_len_mod_gen, env_parents, by = "pop")

```

### Assess multicollinearity and calculate pairwise correlations

KEB 260823: Using the recoded juvenile genetic data to calculate heterozygosity and allelic richness, nothing is correlated.

```{r}

# full matrix incl. SD terms, survival, and mean length
cor_matrix_all <- cor(surv_len_mod_gen_env[,c("salinity_quantile_10","salinity_quantile_90","Salinity_st_dev","temp_quantile_10","temp_quantile_90","Temperature_st_dev", "ho", "ar", "surv", "mean_len")])
colnames(cor_matrix_all) <- c("Salinity 0.1 quantile", "Salinity 0.9 quantile", "Salinity SD", "Temperature 0.1 quantile", "Temperature 0.9 quantile", "Temperature SD", "Heterozygosity", "Allelic richness", "Survival", "Mean length")
rownames(cor_matrix_all) <- colnames(cor_matrix_all)

corrplot_all <- corrplot(cor_matrix_all,
         method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         title = "Pairwise correlations Q2a",
         addCoef.col = 'black', mar = c(0,0,2,0))

# reduced matrix: quantiles only (no SD), plus survival and mean length
cor_matrix_quant <- cor(surv_len_mod_gen_env[,c("salinity_quantile_10","salinity_quantile_90","temp_quantile_10","temp_quantile_90","ho", "ar", "surv", "mean_len")])
colnames(cor_matrix_quant) <- c("Salinity 0.1 quantile", "Salinity 0.9 quantile", "Temperature 0.1 quantile", "Temperature 0.9 quantile", "Heterozygosity", "Allelic richness", "Survival", "Mean length")
rownames(cor_matrix_quant) <- colnames(cor_matrix_quant)

pdf(file = "figures/juvenile_effects_H2F/FigS2_corrplot_envr_quantiles_H0_Ar.pdf")
corrplot(cor_matrix_quant,
         method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         title = "Pairwise correlations Q2a",
         addCoef.col = 'black', mar = c(0,0,2,0))
dev.off() 

```

### Assess VIF

```{r}

# Collinearity check. Predictors vary across only 8 source populations
VIF <- function(a, b, c, d, e){
  1/(1-summary(lm(a ~ b + c + d + e))$r.squared)
}

d <- surv_mod_gen_env

# predictor set retaining temp Q90 (excluding temp Q10)
VIF(d$temp_quantile_90,     d$salinity_quantile_10, d$salinity_quantile_90, d$ho, d$ar)  # temp90
VIF(d$salinity_quantile_10, d$salinity_quantile_90, d$temp_quantile_90,     d$ho, d$ar)  # sal10
VIF(d$salinity_quantile_90, d$salinity_quantile_10, d$temp_quantile_90,     d$ho, d$ar)  # sal90
VIF(d$ho, d$salinity_quantile_10, d$salinity_quantile_90, d$temp_quantile_90, d$ar)      # ho
VIF(d$ar, d$salinity_quantile_10, d$salinity_quantile_90, d$temp_quantile_90, d$ho)      # ar

# alternative set retaining temp Q10 (excluding temp Q90), for comparison
VIF(d$temp_quantile_10,     d$salinity_quantile_10, d$salinity_quantile_90, d$ho, d$ar)  # temp10
VIF(d$salinity_quantile_10, d$salinity_quantile_90, d$temp_quantile_10,     d$ho, d$ar)  # sal10
VIF(d$salinity_quantile_90, d$salinity_quantile_10, d$temp_quantile_10,     d$ho, d$ar)  # sal90
VIF(d$ho, d$salinity_quantile_10, d$salinity_quantile_90, d$temp_quantile_10, d$ar)      # ho
VIF(d$ar, d$salinity_quantile_10, d$salinity_quantile_90, d$temp_quantile_10, d$ho)      # ar

```

I think we can move forward with all predictors except for temperature Q10.

### t1 survival, Lewisetta

```{r t1_surv_lew}

# t1 survival, Lewisetta
# temp_quantile_10 excluded (collinearity); all retained VIF < 2
# (1|pop) accounts for 3 bags nested within each of 8 populations —
# predictors are constant within population, so this is where the real
# replication lives. Watch for a singular fit (8 groups is few).
lew_t1 <- surv_mod_gen_env[surv_mod_gen_env$t == 1 & surv_mod_gen_env$bag_site == "Lewisetta", ]

surv_GE1.lew <- lmer(
  surv ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop),
  data = lew_t1
)
isSingular(surv_GE1.lew)          # TRUE = pop variance ~0; fall back to population means
summary(surv_GE1.lew)             # fixed effects; p-values via Satterthwaite (lmerTest)

surv_t1_lew_scaled <- lmer(
  scale(surv) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop),
  data = lew_t1
)
isSingular(surv_t1_lew_scaled)

```

### t2 survival, Lewisetta

```{r t2_surv_lew}

lew_t2 <- surv_mod_gen_env[surv_mod_gen_env$t == 2 & surv_mod_gen_env$bag_site == "Lewisetta", ]

surv_GE2.lew <- lmer(
  surv ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop),
  data = lew_t2
)
isSingular(surv_GE2.lew)
summary(surv_GE2.lew)

surv_t2_lew_scaled <- lmer(
  scale(surv) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop),
  data = lew_t2
)
isSingular(surv_t2_lew_scaled)

```

### t3 survival, Lewisetta

```{r t3_surv_lew}

lew_t3 <- surv_mod_gen_env[surv_mod_gen_env$t == 3 & surv_mod_gen_env$bag_site == "Lewisetta", ]

surv_GE3.lew <- lmer(
  surv ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop),
  data = lew_t3
)
isSingular(surv_GE3.lew)
summary(surv_GE3.lew)

surv_t3_lew_scaled <- lmer(
  scale(surv) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop),
  data = lew_t3
)
isSingular(surv_t3_lew_scaled)

```

### t1 survival, York River

```{r t1_surv_yrk}

yrk_t1 <- surv_mod_gen_env[surv_mod_gen_env$t == 1 & surv_mod_gen_env$bag_site == "YorkRiver", ]

surv_GE1.york <- lmer(
  surv ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop),
  data = yrk_t1
)
isSingular(surv_GE1.york)
summary(surv_GE1.york)

surv_t1_yrk_scaled <- lmer(
  scale(surv) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop),
  data = yrk_t1
)
isSingular(surv_t1_yrk_scaled)

```

### t2 survival, York River

```{r t2_surv_yrk}

yrk_t2 <- surv_mod_gen_env[surv_mod_gen_env$t == 2 & surv_mod_gen_env$bag_site == "YorkRiver", ]

surv_GE2.york <- lmer(
  surv ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop),
  data = yrk_t2
)
isSingular(surv_GE2.york)
summary(surv_GE2.york)

surv_t2_yrk_scaled <- lmer(
  scale(surv) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop),
  data = yrk_t2
)
isSingular(surv_t2_yrk_scaled)

```

### t3 survival, York River

```{r t3_surv_yrk}

yrk_t3 <- surv_mod_gen_env[surv_mod_gen_env$t == 3 & surv_mod_gen_env$bag_site == "YorkRiver", ]

surv_GE3.york <- lmer(
  surv ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop),
  data = yrk_t3
)
isSingular(surv_GE3.york)
summary(surv_GE3.york)

surv_t3_yrk_scaled <- lmer(
  scale(surv) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop),
  data = yrk_t3
)
isSingular(surv_t3_yrk_scaled)

# save everything
sink(paste0("results/juvenile_effects_H2F/models/survival_Q2a_quantiles_results_", Sys.Date(), ".txt"))
cat("\nLewisetta t1 (November 2023) modeling approach 1\n")
cat("singular fit:", isSingular(surv_GE1.lew), "\n")
print(summary(surv_GE1.lew))
cat("\nLewisetta t1 (November 2023) scaled modeling approach 1\n")
print(summary(surv_t1_lew_scaled))
cat("\nLewisetta t2 (May 2024) modeling approach 1\n")
cat("singular fit:", isSingular(surv_GE2.lew), "\n")
print(summary(surv_GE2.lew))
cat("\nLewisetta t2 (May 2024) scaled modeling approach 1\n")
print(summary(surv_t2_lew_scaled))
cat("\nLewisetta t3 (November 2024) modeling approach 1\n")
cat("singular fit:", isSingular(surv_GE3.lew), "\n")
print(summary(surv_GE3.lew))
cat("\nLewisetta t3 (November 2024) scaled modeling approach 1\n")
print(summary(surv_t3_lew_scaled))
cat("\nYork River t1 (November 2023) modeling approach 1\n")
cat("singular fit:", isSingular(surv_GE1.york), "\n")
print(summary(surv_GE1.york))
cat("\nYork River t1 (November 2023) scaled modeling approach 1\n")
print(summary(surv_t1_yrk_scaled))
cat("\nYork River t2 (May 2024) modeling approach 1\n")
cat("singular fit:", isSingular(surv_GE2.york), "\n")
print(summary(surv_GE2.york))
cat("\nYork River t2 (May 2024) scaled modeling approach 1\n")
print(summary(surv_t2_yrk_scaled))
cat("\nYork River t3 (November 2024) modeling approach 1\n")
cat("singular fit:", isSingular(surv_GE3.york), "\n")
print(summary(surv_GE3.york))
cat("\nYork River t3 (November 2024) scaled modeling approach 1\n")
print(summary(surv_t3_yrk_scaled))
sink()

```

## Modeling drivers of differences in length

```{r}
# start by putting together a complete data frame
len_mod_gen     <- merge(lengths_mod_narm_nodead, juv_genetics, by = "pop")
len_mod_gen_env <- merge(len_mod_gen, env_parents, by = "pop")
```

### t1 length, Lewisetta

```{r t1_len_lew}

# t1 length, Lewisetta — individuals within bags within populations.
# temp_quantile_10 excluded (collinearity); all retained VIF < 2.
# (1|pop/bags_label): bag clustering AND population clustering. Predictors vary
# only at the population level (8 pops); watch the pop term for singularity.
len_lew_t1 <- len_mod_gen_env[len_mod_gen_env$t == 1 & len_mod_gen_env$bag_site == "Lewisetta", ]

len_GE1.lew <- lmer(
  length ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop / bags_label),
  data = len_lew_t1
)
isSingular(len_GE1.lew)   # TRUE = pop or bag variance ~0
VarCorr(len_GE1.lew)      # pop vs bag vs residual variance split
summary(len_GE1.lew)      # Satterthwaite df/p-values

len_t1_lew_scaled <- lmer(
  scale(length) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label),
  data = len_lew_t1
)
isSingular(len_t1_lew_scaled)

```

### t2 length, Lewisetta

```{r t2_len_lew}

len_lew_t2 <- len_mod_gen_env[len_mod_gen_env$t == 2 & len_mod_gen_env$bag_site == "Lewisetta", ]

len_GE2.lew <- lmer(
  length ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop / bags_label),
  data = len_lew_t2
)
isSingular(len_GE2.lew)
VarCorr(len_GE2.lew)
summary(len_GE2.lew)

len_t2_lew_scaled <- lmer(
  scale(length) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label),
  data = len_lew_t2
)
isSingular(len_t2_lew_scaled)

```

### t3 length, Lewisetta

```{r t3_len_lew}

len_lew_t3 <- len_mod_gen_env[len_mod_gen_env$t == 3 & len_mod_gen_env$bag_site == "Lewisetta", ]

len_GE3.lew <- lmer(
  length ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop / bags_label),
  data = len_lew_t3
)
isSingular(len_GE3.lew)
VarCorr(len_GE3.lew)
summary(len_GE3.lew)

len_t3_lew_scaled <- lmer(
  scale(length) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label),
  data = len_lew_t3
)
isSingular(len_t3_lew_scaled)

```

### t1 length, York River

```{r t1_len_yrk}

len_yrk_t1 <- len_mod_gen_env[len_mod_gen_env$t == 1 & len_mod_gen_env$bag_site == "YorkRiver", ]

len_GE1.york <- lmer(
  length ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop / bags_label),
  data = len_yrk_t1
)
isSingular(len_GE1.york)
VarCorr(len_GE1.york)
summary(len_GE1.york)

len_t1_yrk_scaled <- lmer(
  scale(length) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label),
  data = len_yrk_t1
)
isSingular(len_t1_yrk_scaled)

```

### t2 length, York River

```{r t2_len_yrk}

len_yrk_t2 <- len_mod_gen_env[len_mod_gen_env$t == 2 & len_mod_gen_env$bag_site == "YorkRiver", ]

len_GE2.york <- lmer(
  length ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop / bags_label),
  data = len_yrk_t2
)
isSingular(len_GE2.york)
VarCorr(len_GE2.york)
summary(len_GE2.york)

len_t2_yrk_scaled <- lmer(
  scale(length) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label),
  data = len_yrk_t2
)
isSingular(len_t2_yrk_scaled)

```

### t3 length, York River

```{r t3_len_yrk}

len_yrk_t3 <- len_mod_gen_env[len_mod_gen_env$t == 3 & len_mod_gen_env$bag_site == "YorkRiver", ]

len_GE3.york <- lmer(
  length ~ ho + ar + salinity_quantile_10 + salinity_quantile_90 + temp_quantile_90 +
    (1 | pop / bags_label),
  data = len_yrk_t3
)
isSingular(len_GE3.york)
VarCorr(len_GE3.york)
summary(len_GE3.york)

len_t3_yrk_scaled <- lmer(
  scale(length) ~ scale(ho) + scale(ar) + scale(salinity_quantile_10) +
    scale(salinity_quantile_90) + scale(temp_quantile_90) + (1 | pop / bags_label),
  data = len_yrk_t3
)
isSingular(len_t3_yrk_scaled)

```

```{r}

# save everything
sink(paste0("results/juvenile_effects_H2F/models/length_Q2a_quantiles_results_", Sys.Date(), ".txt"))
cat("\nLewisetta t1 (November 2023) modeling approach 1\n")
cat("singular fit:", isSingular(len_GE1.lew), "\n")
print(VarCorr(len_GE1.lew))
print(summary(len_GE1.lew))
cat("\nLewisetta t1 (November 2023) scaled modeling approach 1\n")
print(summary(len_t1_lew_scaled))
cat("\nLewisetta t2 (May 2024) modeling approach 1\n")
cat("singular fit:", isSingular(len_GE2.lew), "\n")
print(VarCorr(len_GE2.lew))
print(summary(len_GE2.lew))
cat("\nLewisetta t2 (May 2024) scaled modeling approach 1\n")
print(summary(len_t2_lew_scaled))
cat("\nLewisetta t3 (November 2024) modeling approach 1\n")
cat("singular fit:", isSingular(len_GE3.lew), "\n")
print(VarCorr(len_GE3.lew))
print(summary(len_GE3.lew))
cat("\nLewisetta t3 (November 2024) scaled modeling approach 1\n")
print(summary(len_t3_lew_scaled))
cat("\nYork River t1 (November 2023) modeling approach 1\n")
cat("singular fit:", isSingular(len_GE1.york), "\n")
print(VarCorr(len_GE1.york))
print(summary(len_GE1.york))
cat("\nYork River t1 (November 2023) scaled modeling approach 1\n")
print(summary(len_t1_yrk_scaled))
cat("\nYork River t2 (May 2024) modeling approach 1\n")
cat("singular fit:", isSingular(len_GE2.york), "\n")
print(VarCorr(len_GE2.york))
print(summary(len_GE2.york))
cat("\nYork River t2 (May 2024) scaled modeling approach 1\n")
print(summary(len_t2_yrk_scaled))
cat("\nYork River t3 (November 2024) modeling approach 1\n")
cat("singular fit:", isSingular(len_GE3.york), "\n")
print(VarCorr(len_GE3.york))
print(summary(len_GE3.york))
cat("\nYork River t3 (November 2024) scaled modeling approach 1\n")
print(summary(len_t3_yrk_scaled))
sink()

```

## Figures

```{r}

# helper: tidy a set of scaled models, tag time, and build honest t-based 95% CIs.
# df comes from lmerTest's Satterthwaite approximation (~2 here), so qt(0.975, df)
# gives the correct multiplier (~4.3), NOT 2. Bars now match the Satterthwaite p-values.
build_slope_df <- function(m_t1, m_t2, m_t3) {
  bind_rows(
    tidy(m_t1, effects = "fixed") %>% mutate(t = "t1"),
    tidy(m_t2, effects = "fixed") %>% mutate(t = "t2"),
    tidy(m_t3, effects = "fixed") %>% mutate(t = "t3")
  ) %>%
    filter(term != "(Intercept)") %>%
    mutate(
      tmult = qt(0.975, df),
      lower = estimate - tmult * std.error,
      upper = estimate + tmult * std.error
    ) %>%
    select(explanatory = term, slope = estimate, lower, upper, t)
}

# shared y-axis ordering + labels (all five predictors, present in every cell now)
y_limits <- c("scale(salinity_quantile_10)", "scale(salinity_quantile_90)",
              "scale(temp_quantile_90)", "scale(ho)", "scale(ar)")
y_labels <- c("Salinity Q10", "Salinity Q90", "Temperature Q90",
              "Heterozygosity", "Allelic richness")

time_labels <- c("t1" = "November 2023", "t2" = "May 2024", "t3" = "November 2024")

# Lewisetta survival
lew_surv_df <- build_slope_df(surv_t1_lew_scaled, surv_t2_lew_scaled, surv_t3_lew_scaled)

lew_surv_plot <- lew_surv_df %>%
  ggplot(aes(x = slope, y = explanatory)) +
  geom_point(aes(alpha = t, shape = t), color = "#F06292", size = 3.5,
             position = position_dodge2v(height = .85, reverse = T, preserve = "total")) +
  geom_errorbarh(aes(xmax = upper, xmin = lower, alpha = t),
                 color = "#F06292", height = 0, size = 1.2,
                 position = position_dodgev(height = -.85, preserve = "total")) +
  scale_alpha_manual(name = "Time", na.translate = F,
                     values = c("t1" = .45, "t2" = .7, "t3" = 1), labels = time_labels) +
  scale_shape_manual(name = "Time", na.translate = F,
                     values = c("t1" = 15, "t2" = 16, "t3" = 17), labels = time_labels) +
  coord_cartesian(xlim = c(-1.6, 2)) +
  xlab("Slope on scaled data") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_y_discrete(name = "Explanatory variable", limits = y_limits, labels = y_labels) +
  theme_classic() +
  theme(axis.text = element_text(color = "black", size = 14),
        axis.title = element_text(size = 15),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 15))

lew_surv_plot
ggsave("Fig3A_lew_surv_quantiles.pdf", plot = lew_surv_plot, device = "pdf",
       path = "figures/juvenile_effects_H2F", height = 4, width = 7)

# York survival
yrk_surv_df <- build_slope_df(surv_t1_yrk_scaled, surv_t2_yrk_scaled, surv_t3_yrk_scaled)

yrk_surv_plot <- yrk_surv_df %>%
  ggplot(aes(x = slope, y = explanatory)) +
  geom_point(aes(alpha = t, shape = t), color = "#880E4F", size = 3.5,
             position = position_dodge2v(height = .85, reverse = T, preserve = "total")) +
  geom_errorbarh(aes(xmax = upper, xmin = lower, alpha = t),
                 color = "#880E4F", height = 0, size = 1.2,
                 position = position_dodgev(height = -.85, preserve = "total")) +
  scale_alpha_manual(name = "Time", na.translate = F,
                     values = c("t1" = .4, "t2" = .7, "t3" = 1), labels = time_labels) +
  scale_shape_manual(name = "Time", na.translate = F,
                     values = c("t1" = 15, "t2" = 16, "t3" = 17), labels = time_labels) +
  coord_cartesian(xlim = c(-1.6, 2)) +
  xlab("Slope on scaled data") + ylab("Variable") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_y_discrete(name = "Explanatory variable", limits = y_limits, labels = y_labels) +
  theme_classic() +
  theme(axis.text = element_text(color = "black", size = 14),
        axis.title = element_text(size = 15),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 15))

yrk_surv_plot
ggsave("Fig3C_york_surv_quantiles.pdf", plot = yrk_surv_plot, device = "pdf",
       path = "figures/juvenile_effects_H2F", height = 4, width = 7)

# Lewisetta length
lew_len_df <- build_slope_df(len_t1_lew_scaled, len_t2_lew_scaled, len_t3_lew_scaled)

lew_len_plot <- lew_len_df %>%
  ggplot(aes(x = slope, y = explanatory)) +
  geom_point(aes(alpha = t, shape = t), color = "#F06292", size = 3.5,
             position = position_dodge2v(height = .85, reverse = T, preserve = "total")) +
  geom_errorbarh(aes(xmax = upper, xmin = lower, alpha = t),
                 color = "#F06292", height = 0, size = 1.2,
                 position = position_dodgev(height = -.85, preserve = "total")) +
  scale_alpha_manual(name = "Time", na.translate = F,
                     values = c("t1" = .45, "t2" = .7, "t3" = 1), labels = time_labels) +
  scale_shape_manual(name = "Time", na.translate = F,
                     values = c("t1" = 15, "t2" = 16, "t3" = 17), labels = time_labels) +
  coord_cartesian(xlim = c(-1.6, 2)) +
  xlab("Slope on scaled data") + ylab("Variable") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_y_discrete(name = "Explanatory variable", limits = y_limits, labels = y_labels) +
  theme_classic() +
  theme(axis.text = element_text(color = "black", size = 14),
        axis.title = element_text(size = 15),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 15))

lew_len_plot
ggsave("FigS5A_lew_length_quantiles.pdf", plot = lew_len_plot, device = "pdf",
       path = "figures/juvenile_effects_H2F", height = 4, width = 7)

# York length
yrk_len_df <- build_slope_df(len_t1_yrk_scaled, len_t2_yrk_scaled, len_t3_yrk_scaled)

yrk_len_plot <- yrk_len_df %>%
  ggplot(aes(x = slope, y = explanatory)) +
  geom_point(aes(alpha = t, shape = t), color = "#880E4F", size = 3.5,
             position = position_dodge2v(height = .85, reverse = T, preserve = "total")) +
  geom_errorbarh(aes(xmax = upper, xmin = lower, alpha = t),
                 color = "#880E4F", height = 0, size = 1.2,
                 position = position_dodgev(height = -.85, preserve = "total")) +
  scale_alpha_manual(name = "Time", na.translate = F,
                     values = c("t1" = .45, "t2" = .7, "t3" = 1), labels = time_labels) +
  scale_shape_manual(name = "Time", na.translate = F,
                     values = c("t1" = 15, "t2" = 16, "t3" = 17), labels = time_labels) +
  coord_cartesian(xlim = c(-1.6, 2)) +
  xlab("Slope on scaled data") + ylab("Variable") +
  geom_vline(xintercept = 0, linetype = "dashed") +
  scale_y_discrete(name = "Explanatory variable", limits = y_limits, labels = y_labels) +
  theme_classic() +
  theme(axis.text = element_text(color = "black", size = 14),
        axis.title = element_text(size = 15),
        legend.text = element_text(size = 14),
        legend.title = element_text(size = 15))

yrk_len_plot
ggsave("FigS5C_york_length_quantiles.pdf", plot = yrk_len_plot, device = "pdf",
       path = "figures/juvenile_effects_H2F", height = 4, width = 7)

```

## Check assumptions

```{r}

# collect the raw (unscaled) fitted models
surv_models <- list(
  "surv Lew t1" = surv_GE1.lew, "surv Lew t2" = surv_GE2.lew, "surv Lew t3" = surv_GE3.lew,
  "surv Yrk t1" = surv_GE1.york, "surv Yrk t2" = surv_GE2.york, "surv Yrk t3" = surv_GE3.york
)
len_models <- list(
  "len Lew t1" = len_GE1.lew, "len Lew t2" = len_GE2.lew, "len Lew t3" = len_GE3.lew,
  "len Yrk t1" = len_GE1.york, "len Yrk t2" = len_GE2.york, "len Yrk t3" = len_GE3.york
)
all_models <- c(surv_models, len_models)

# residuals vs fitted + QQ for every model, two per row
op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (nm in names(all_models)) {
  m <- all_models[[nm]]
  plot(fitted(m), resid(m), main = paste(nm, "- resid vs fitted"),
       xlab = "Fitted", ylab = "Residual", pch = 19, col = rgb(0,0,0,0.3))
  abline(h = 0, lty = 2)
  qqnorm(resid(m), main = paste(nm, "- QQ")); qqline(resid(m))
}
par(op)

# quick numeric summary: normality of residuals per model
sapply(all_models, function(m) {
  r <- resid(m)
  c(shapiro_p = tryCatch(shapiro.test(r)$p.value, error = function(e) NA),
    n = length(r))
})

```

```{r}

op <- par(mfrow = c(2, 3), mar = c(4, 4, 2, 1))
for (nm in names(len_models)) {
  qqnorm(resid(len_models[[nm]]), main = nm, pch = 19, col = rgb(0,0,0,0.2))
  qqline(resid(len_models[[nm]]))
}
par(op)

```