---
title: "Q2b_JUV_envrdist_models_field"
author: "Kiran Bajaj"
date: "2026-08-23"
output: pdf_document
---

This file contains analyses that address Q2b: effects of offspring genetic diversity (heterozygosity, allelic richness) and environmental distance on oyster survival and length at a given timepoint (monocultures only).

Modeling is based on code from CR.

260823 KEB: updated these models to include juvenile heterozygosity and allelic richness data which is calculated correctly based on the recoded genotype matrix.

## Load packages

```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = TRUE)
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("tidyverse", "here", "ggplot2", "lme4", "lmerTest", "car", "broom", "broom.mixed", "corrplot", "ggstance", "stringr")
pacman::p_load(char = packages)

r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")

```

```{r}
getwd()
```


## Read in data

```{r}

# survival
surv_mod_ord_rename <- read.csv("data/performance_H2F/field_surv_all_times.csv")[, -1]

# length
lengths_mod_narm_nodead <- read.csv("data/performance_H2F/field_length_all_times.csv")[ ,-1]

# juvenile heterozygosity and allelic richness (recoded genotype matrix)
juv_genetics <- read.csv("results/juvenile_effects_H2F/juvenile_genetic_effects_full.csv")

# pairwise environmental distance matrix (scaled envr quantiles); LEW/YRK cols = distance to each field site
env_dist <- read.csv("results/environment/envr_quantiles_dist_all.csv")

```

Because the H1-HYBRIDMIX and H2-SEEDMIX groups do not have a single environment-of-origin, they are excluded from these analyses.

## Data preparation

```{r}

# genetic data
colnames(juv_genetics) <- c("pop", "ho", "exp_hs", "ar")

# reshape the distance matrix: keep only distance-to-field-site columns
# env_dist$X is the source population; LEW/YRK are distances to each field site.
env_dist_long <- env_dist %>%
  select(pop = X, Lewisetta = LEW, YorkRiver = YRK) %>%
  pivot_longer(cols = c(Lewisetta, YorkRiver),
               names_to = "bag_site", values_to = "dist")

# survival + genetics
surv_mod_gen <- merge(surv_mod_ord_rename, juv_genetics, by = "pop")

# add the site-matched distance (join on pop AND bag_site so each row gets
# the distance from its source pop to the field site it's deployed at)
surv_mod_gen_dist <- merge(surv_mod_gen, env_dist_long, by = c("pop", "bag_site"))

# length + genetics + site-matched distance (per-individual rows kept)
len_mod_gen      <- merge(lengths_mod_narm_nodead, juv_genetics, by = "pop")
len_mod_gen_dist <- merge(len_mod_gen, env_dist_long, by = c("pop", "bag_site"))

# confirm the mixed-origin groups are gone and we have 8 pops per site
unique(surv_mod_gen_dist$pop)
table(surv_mod_gen_dist$pop, surv_mod_gen_dist$bag_site)

```

## Assess collinearity

```{r}
# build a survival + mean-length + genetics + distance frame for the corrplots
length_summarized <- lengths_mod_narm_nodead %>%
  group_by(bag_site, pop, t, bags_label) %>%
  summarise(mean_len = mean(length, na.rm = TRUE), .groups = "drop")

surv_len_gen      <- merge(length_summarized, surv_mod_gen, by = c("pop", "bag_site", "bags_label", "t"))
surv_len_gen_dist <- merge(surv_len_gen, env_dist_long, by = c("pop", "bag_site"))

cor_vars   <- c("dist", "ho", "ar", "surv", "mean_len")
cor_labels <- c("Environmental distance", "Heterozygosity", "Allelic richness", "Survival", "Mean length")

# Lewisetta
lew_mat <- cor(surv_len_gen_dist[surv_len_gen_dist$bag_site == "Lewisetta", cor_vars])
dimnames(lew_mat) <- list(cor_labels, cor_labels)

pdf(file = "figures/juvenile_effects_H2F/FigS2_corrplot_envrdist_lew.pdf")
corrplot(lew_mat, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         title = "Pairwise correlations Q2b — Lewisetta",
         addCoef.col = "black", mar = c(0,0,2,0))
dev.off()

# also show inline
corrplot(lew_mat, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         title = "Pairwise correlations Q2b — Lewisetta",
         addCoef.col = "black", mar = c(0,0,2,0))

# York River
yrk_mat <- cor(surv_len_gen_dist[surv_len_gen_dist$bag_site == "YorkRiver", cor_vars])
dimnames(yrk_mat) <- list(cor_labels, cor_labels)

pdf(file = "figures/juvenile_effects_H2F/FigS2_corrplot_envrdist_yrk.pdf")
corrplot(yrk_mat, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         title = "Pairwise correlations Q2b — York River",
         addCoef.col = "black", mar = c(0,0,2,0))
dev.off()

corrplot(yrk_mat, method = "color", type = "upper", tl.col = "black", tl.srt = 45,
         title = "Pairwise correlations Q2b — York River",
         addCoef.col = "black", mar = c(0,0,2,0))
```

```{r}

# only three predictors here (ho, ar, dist), but check briefly for each site. 
# predictors vary across only 8 source populations.
cor_vars <- c("ho", "ar", "dist")

# Lewisetta
cor(surv_mod_gen_dist[surv_mod_gen_dist$bag_site == "Lewisetta", cor_vars])
# York
cor(surv_mod_gen_dist[surv_mod_gen_dist$bag_site == "YorkRiver", cor_vars])

# quick VIF (3-predictor models)
VIF <- function(a, b, c) 1 / (1 - summary(lm(a ~ b + c))$r.squared)

d_lew <- surv_mod_gen_dist[surv_mod_gen_dist$bag_site == "Lewisetta", ]
VIF(d_lew$ho,  d_lew$ar,  d_lew$dist)  # ho
VIF(d_lew$ar,  d_lew$ho,  d_lew$dist)  # ar
VIF(d_lew$dist, d_lew$ho, d_lew$ar)    # dist

d_yrk <- surv_mod_gen_dist[surv_mod_gen_dist$bag_site == "YorkRiver", ]
VIF(d_yrk$ho,  d_yrk$ar,  d_yrk$dist)  # ho
VIF(d_yrk$ar,  d_yrk$ho,  d_yrk$dist)  # ar
VIF(d_yrk$dist, d_yrk$ho, d_yrk$ar)    # dist
```

## Modeling drivers of differences in survival

### t1 survival, Lewisetta

```{r t1_surv_lew}

# full model, no selection. (1|pop) accounts for bags within the 8 source
# populations. Predictors are population-level.
lew_t1 <- surv_mod_gen_dist[surv_mod_gen_dist$t == 1 & surv_mod_gen_dist$bag_site == "Lewisetta", ]

surv_GE1.lew <- lmer(surv ~ ho + ar + dist + (1 | pop), data = lew_t1)
isSingular(surv_GE1.lew)
summary(surv_GE1.lew)

surv_t1_lew_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop), data = lew_t1)
isSingular(surv_t1_lew_scaled)

```

### t2 survival, Lewisetta

```{r t2_surv_lew}

lew_t2 <- surv_mod_gen_dist[surv_mod_gen_dist$t == 2 & surv_mod_gen_dist$bag_site == "Lewisetta", ]

surv_GE2.lew <- lmer(surv ~ ho + ar + dist + (1 | pop), data = lew_t2)
isSingular(surv_GE2.lew)
summary(surv_GE2.lew)

surv_t2_lew_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop), data = lew_t2)
isSingular(surv_t2_lew_scaled)

```

### t3 survival, Lewisetta

```{r t3_surv_lew}

lew_t3 <- surv_mod_gen_dist[surv_mod_gen_dist$t == 3 & surv_mod_gen_dist$bag_site == "Lewisetta", ]

surv_GE3.lew <- lmer(surv ~ ho + ar + dist + (1 | pop), data = lew_t3)
isSingular(surv_GE3.lew)
summary(surv_GE3.lew)

surv_t3_lew_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop), data = lew_t3)
isSingular(surv_t3_lew_scaled)

```

### t1 survival, York River

```{r t1_surv_yrk}

yrk_t1 <- surv_mod_gen_dist[surv_mod_gen_dist$t == 1 & surv_mod_gen_dist$bag_site == "YorkRiver", ]

surv_GE1.york <- lmer(surv ~ ho + ar + dist + (1 | pop), data = yrk_t1)
isSingular(surv_GE1.york)
summary(surv_GE1.york)

surv_t1_yrk_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop), data = yrk_t1)
isSingular(surv_t1_yrk_scaled)

```

### t2 survival, York River

```{r t2_surv_yrk}

yrk_t2 <- surv_mod_gen_dist[surv_mod_gen_dist$t == 2 & surv_mod_gen_dist$bag_site == "YorkRiver", ]

surv_GE2.york <- lmer(surv ~ ho + ar + dist + (1 | pop), data = yrk_t2)
isSingular(surv_GE2.york)
summary(surv_GE2.york)

surv_t2_yrk_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop), data = yrk_t2)
isSingular(surv_t2_yrk_scaled)

```

### t3 survival, York River

```{r t3_surv_yrk}

yrk_t3 <- surv_mod_gen_dist[surv_mod_gen_dist$t == 3 & surv_mod_gen_dist$bag_site == "YorkRiver", ]

surv_GE3.york <- lmer(surv ~ ho + ar + dist + (1 | pop), data = yrk_t3)
isSingular(surv_GE3.york)
summary(surv_GE3.york)

surv_t3_yrk_scaled <- lmer(scale(surv) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop), data = yrk_t3)
isSingular(surv_t3_yrk_scaled)

# save everything
sink(paste0("results/juvenile_effects_H2F/models/survival_Q2b_envrdist_results_", Sys.Date(), ".txt"))
cat("\nLewisetta t1 (November 2023) modeling approach 2\n")
cat("singular fit:", isSingular(surv_GE1.lew), "\n"); print(summary(surv_GE1.lew))
cat("\nLewisetta t1 (November 2023) scaled modeling approach 2\n"); print(summary(surv_t1_lew_scaled))
cat("\nLewisetta t2 (May 2024) modeling approach 2\n")
cat("singular fit:", isSingular(surv_GE2.lew), "\n"); print(summary(surv_GE2.lew))
cat("\nLewisetta t2 (May 2024) scaled modeling approach 2\n"); print(summary(surv_t2_lew_scaled))
cat("\nLewisetta t3 (November 2024) modeling approach 2\n")
cat("singular fit:", isSingular(surv_GE3.lew), "\n"); print(summary(surv_GE3.lew))
cat("\nLewisetta t3 (November 2024) scaled modeling approach 2\n"); print(summary(surv_t3_lew_scaled))
cat("\nYork River t1 (November 2023) modeling approach 2\n")
cat("singular fit:", isSingular(surv_GE1.york), "\n"); print(summary(surv_GE1.york))
cat("\nYork River t1 (November 2023) scaled modeling approach 2\n"); print(summary(surv_t1_yrk_scaled))
cat("\nYork River t2 (May 2024) modeling approach 2\n")
cat("singular fit:", isSingular(surv_GE2.york), "\n"); print(summary(surv_GE2.york))
cat("\nYork River t2 (May 2024) scaled modeling approach 2\n"); print(summary(surv_t2_yrk_scaled))
cat("\nYork River t3 (November 2024) modeling approach 2\n")
cat("singular fit:", isSingular(surv_GE3.york), "\n"); print(summary(surv_GE3.york))
cat("\nYork River t3 (November 2024) scaled modeling approach 2\n"); print(summary(surv_t3_yrk_scaled))
sink()
```

## Modeling drivers of differences in length

### t1 length, Lewisetta

```{r t1_len_lew}

# individuals within bags within populations; (1|pop/bags_label)
len_lew_t1 <- len_mod_gen_dist[len_mod_gen_dist$t == 1 & len_mod_gen_dist$bag_site == "Lewisetta", ]

len_GE1.lew <- lmer(length ~ ho + ar + dist + (1 | pop / bags_label), data = len_lew_t1)
isSingular(len_GE1.lew); VarCorr(len_GE1.lew); summary(len_GE1.lew)

len_t1_lew_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = len_lew_t1)
isSingular(len_t1_lew_scaled)

```

### t2 length, Lewisetta

```{r t2_len_lew}

len_lew_t2 <- len_mod_gen_dist[len_mod_gen_dist$t == 2 & len_mod_gen_dist$bag_site == "Lewisetta", ]

len_GE2.lew <- lmer(length ~ ho + ar + dist + (1 | pop / bags_label), data = len_lew_t2)
isSingular(len_GE2.lew); VarCorr(len_GE2.lew); summary(len_GE2.lew)

len_t2_lew_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = len_lew_t2)
isSingular(len_t2_lew_scaled)

```

### t3 length, Lewisetta

```{r t3_len_lew}

len_lew_t3 <- len_mod_gen_dist[len_mod_gen_dist$t == 3 & len_mod_gen_dist$bag_site == "Lewisetta", ]

len_GE3.lew <- lmer(length ~ ho + ar + dist + (1 | pop / bags_label), data = len_lew_t3)
isSingular(len_GE3.lew); VarCorr(len_GE3.lew); summary(len_GE3.lew)

len_t3_lew_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = len_lew_t3)
isSingular(len_t3_lew_scaled)

```

### t1 length, York River

```{r t1_len_yrk}

len_yrk_t1 <- len_mod_gen_dist[len_mod_gen_dist$t == 1 & len_mod_gen_dist$bag_site == "YorkRiver", ]

len_GE1.york <- lmer(length ~ ho + ar + dist + (1 | pop / bags_label), data = len_yrk_t1)
isSingular(len_GE1.york); VarCorr(len_GE1.york); summary(len_GE1.york)

len_t1_yrk_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = len_yrk_t1)
isSingular(len_t1_yrk_scaled)

```

### t2 length, York River

```{r t2_len_yrk}

len_yrk_t2 <- len_mod_gen_dist[len_mod_gen_dist$t == 2 & len_mod_gen_dist$bag_site == "YorkRiver", ]

len_GE2.york <- lmer(length ~ ho + ar + dist + (1 | pop / bags_label), data = len_yrk_t2)
isSingular(len_GE2.york); VarCorr(len_GE2.york); summary(len_GE2.york)

len_t2_yrk_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = len_yrk_t2)
isSingular(len_t2_yrk_scaled)

```

### t3 length, York River

```{r t3_len_yrk}

len_yrk_t3 <- len_mod_gen_dist[len_mod_gen_dist$t == 3 & len_mod_gen_dist$bag_site == "YorkRiver", ]

len_GE3.york <- lmer(length ~ ho + ar + dist + (1 | pop / bags_label), data = len_yrk_t3)
isSingular(len_GE3.york); VarCorr(len_GE3.york); summary(len_GE3.york)

len_t3_yrk_scaled <- lmer(scale(length) ~ scale(ho) + scale(ar) + scale(dist) + (1 | pop / bags_label), data = len_yrk_t3)
isSingular(len_t3_yrk_scaled)

# save everything
sink(paste0("results/juvenile_effects_H2F/models/length_Q2b_envrdist_results_", Sys.Date(), ".txt"))
cat("\nLewisetta t1 (November 2023) modeling approach 2\n")
cat("singular fit:", isSingular(len_GE1.lew), "\n"); print(VarCorr(len_GE1.lew)); print(summary(len_GE1.lew))
cat("\nLewisetta t1 (November 2023) scaled modeling approach 2\n"); print(summary(len_t1_lew_scaled))
cat("\nLewisetta t2 (May 2024) modeling approach 2\n")
cat("singular fit:", isSingular(len_GE2.lew), "\n"); print(VarCorr(len_GE2.lew)); print(summary(len_GE2.lew))
cat("\nLewisetta t2 (May 2024) scaled modeling approach 2\n"); print(summary(len_t2_lew_scaled))
cat("\nLewisetta t3 (November 2024) modeling approach 2\n")
cat("singular fit:", isSingular(len_GE3.lew), "\n"); print(VarCorr(len_GE3.lew)); print(summary(len_GE3.lew))
cat("\nLewisetta t3 (November 2024) scaled modeling approach 2\n"); print(summary(len_t3_lew_scaled))
cat("\nYork River t1 (November 2023) modeling approach 2\n")
cat("singular fit:", isSingular(len_GE1.york), "\n"); print(VarCorr(len_GE1.york)); print(summary(len_GE1.york))
cat("\nYork River t1 (November 2023) scaled modeling approach 2\n"); print(summary(len_t1_yrk_scaled))
cat("\nYork River t2 (May 2024) modeling approach 2\n")
cat("singular fit:", isSingular(len_GE2.york), "\n"); print(VarCorr(len_GE2.york)); print(summary(len_GE2.york))
cat("\nYork River t2 (May 2024) scaled modeling approach 2\n"); print(summary(len_t2_yrk_scaled))
cat("\nYork River t3 (November 2024) modeling approach 2\n")
cat("singular fit:", isSingular(len_GE3.york), "\n"); print(VarCorr(len_GE3.york)); print(summary(len_GE3.york))
cat("\nYork River t3 (November 2024) scaled modeling approach 2\n"); print(summary(len_t3_yrk_scaled))
sink()
```

## Figures

```{r}

# helper: tidy scaled models, tag time, build t-based 95% CIs (Satterthwaite df ~2)
build_slope_df <- function(m_t1, m_t2, m_t3) {
  bind_rows(
    tidy(m_t1, effects = "fixed") %>% mutate(t = "t1"),
    tidy(m_t2, effects = "fixed") %>% mutate(t = "t2"),
    tidy(m_t3, effects = "fixed") %>% mutate(t = "t3")
  ) %>%
    filter(term != "(Intercept)") %>%
    mutate(tmult = qt(0.975, df),
           lower = estimate - tmult * std.error,
           upper = estimate + tmult * std.error) %>%
    select(explanatory = term, slope = estimate, lower, upper, t)
}

# three predictors now. Distance is the site-specific term
y_limits <- c("scale(dist)", "scale(ho)", "scale(ar)")
y_labels <- c(str_wrap("Environmental distance (EoO to field site)", width = 20),
              "Heterozygosity", "Allelic richness")
time_labels <- c("t1" = "November 2023", "t2" = "May 2024", "t3" = "November 2024")

plot_slopes <- function(df, point_color) {
  ggplot(df, aes(x = slope, y = explanatory)) +
    geom_point(aes(alpha = t, shape = t), color = point_color, size = 3.5,
               position = position_dodge2v(height = .85, reverse = T, preserve = "total")) +
    geom_errorbarh(aes(xmax = upper, xmin = lower, alpha = t), color = point_color,
                   height = 0, size = 1.2,
                   position = position_dodgev(height = -.85, preserve = "total")) +
    scale_alpha_manual(name = "Time", na.translate = F,
                       values = c("t1" = .45, "t2" = .7, "t3" = 1), labels = time_labels) +
    scale_shape_manual(name = "Time", na.translate = F,
                       values = c("t1" = 15, "t2" = 16, "t3" = 17), labels = time_labels) +
    coord_cartesian(xlim = c(-1.6, 2)) +
    xlab("Slope on scaled data") + ylab("Variable") +
    geom_vline(xintercept = 0, linetype = "dashed") +
    scale_y_discrete(name = "Explanatory variable", limits = y_limits, labels = y_labels) +
    theme_classic() +
    theme(axis.text = element_text(color = "black", size = 14),
          axis.title = element_text(size = 15),
          legend.text = element_text(size = 14),
          legend.title = element_text(size = 15))
}

lew_surv_df <- build_slope_df(surv_t1_lew_scaled, surv_t2_lew_scaled, surv_t3_lew_scaled)
lew_surv_plot <- plot_slopes(lew_surv_df, "#F06292"); lew_surv_plot
ggsave("Fig3B_lew_surv_envrdist.pdf", plot = lew_surv_plot, device = "pdf",
       path = "figures/juvenile_effects_H2F", height = 4, width = 7)

yrk_surv_df <- build_slope_df(surv_t1_yrk_scaled, surv_t2_yrk_scaled, surv_t3_yrk_scaled)
yrk_surv_plot <- plot_slopes(yrk_surv_df, "#880E4F"); yrk_surv_plot
ggsave("Fig3D_york_surv_envrdist.pdf", plot = yrk_surv_plot, device = "pdf",
       path = "figures/juvenile_effects_H2F", height = 4, width = 7)

lew_len_df <- build_slope_df(len_t1_lew_scaled, len_t2_lew_scaled, len_t3_lew_scaled)
lew_len_plot <- plot_slopes(lew_len_df, "#F06292"); lew_len_plot
ggsave("FigS5B_lew_len_envrdist.pdf", plot = lew_len_plot, device = "pdf",
       path = "figures/juvenile_effects_H2F", height = 4, width = 7)

yrk_len_df <- build_slope_df(len_t1_yrk_scaled, len_t2_yrk_scaled, len_t3_yrk_scaled)
yrk_len_plot <- plot_slopes(yrk_len_df, "#880E4F"); yrk_len_plot
ggsave("FigS5D_york_len_envrdist.pdf", plot = yrk_len_plot, device = "pdf",
       path = "figures/juvenile_effects_H2F", height = 4, width = 7)

```

## Check assumptions

```{r}

surv_models <- list(
  "surv Lew t1" = surv_GE1.lew, "surv Lew t2" = surv_GE2.lew, "surv Lew t3" = surv_GE3.lew,
  "surv Yrk t1" = surv_GE1.york, "surv Yrk t2" = surv_GE2.york, "surv Yrk t3" = surv_GE3.york
)
len_models <- list(
  "len Lew t1" = len_GE1.lew, "len Lew t2" = len_GE2.lew, "len Lew t3" = len_GE3.lew,
  "len Yrk t1" = len_GE1.york, "len Yrk t2" = len_GE2.york, "len Yrk t3" = len_GE3.york
)
all_models <- c(surv_models, len_models)

op <- par(mfrow = c(2, 2), mar = c(4, 4, 2, 1))
for (nm in names(all_models)) {
  m <- all_models[[nm]]
  plot(fitted(m), resid(m), main = paste(nm, "- resid vs fitted"),
       xlab = "Fitted", ylab = "Residual", pch = 19, col = rgb(0,0,0,0.3))
  abline(h = 0, lty = 2)
  qqnorm(resid(m), main = paste(nm, "- QQ")); qqline(resid(m))
}
par(op)

sapply(all_models, function(m) {
  r <- resid(m)
  c(shapiro_p = tryCatch(shapiro.test(r)$p.value, error = function(e) NA), n = length(r))
})

```