---
title: "Fstats_juveniles.Rmd"
author: "Kiran Bajaj"
date: "2026-05-29"
output: html_document
---

This file is for calculating pairwise Fst, allelic richness, observed heterozygosity, and running OutFLANK to test for Fst outliers. The analysis uses SNP chip data from the experimental JUVENILES for the H2F paper. MGE generated SNP sets and genotype matrices. 

## Load packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working directory relative to project root
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("here", "ggplot2", "hierfstat", "reshape2", "dplyr", "tidyr", "readr", "adegenet", "viridis", "OutFLANK", "purrr", "tibble")
pacman::p_load(char = packages)

# log versions for reproducibility
r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

## Load data

```{r}

# full SNP set
full_juvs <- readRDS(file.path("data", "juvenile_genotypes", "20250616_FULLSNPs_exp_juvs_labeled.rds"))

# SNP set thinned for LD
thinned_juvs <- readRDS(file.path("data", "juvenile_genotypes", "20250616_THINNEDSNPs_exp_juvs_labeled.rds"))

# SNP info
muts_juvs <- readRDS(file.path("data", "juvenile_genotypes", "20250604_FULLmuts_exp_juvs.rds"))

# sample info
juvs_pca_inds <- read.csv(file.path("data", "juvenile_genotypes", "20251119_juvs_pca_inds_labels.csv"))

```

Need to figure out if all SNPs are accounted for in the muts_juvs dataframe.
260609: Everything is accounted for.

```{r}

# colnames of full_juvs that ARE in Affx.ID
matched <- intersect(colnames(full_juvs), muts_juvs$Affx.ID)
cat("Matched:", length(matched), "\n")

# colnames of full_juvs that are NOT in Affx.ID
unmatched <- setdiff(colnames(full_juvs), muts_juvs$Affx.ID)
cat("Not accounted for:", length(unmatched), "\n")

print(unmatched)

```

```{r}

# rename datasets for processing
full <- full_juvs
gen <- thinned_juvs
muts <- muts_juvs

# check dimensions
dim(full) # full - 267 individuals and 150062 SNPs
dim(gen) # thinned - 267 individuals and 93792 SNPs
dim(muts) # full mutation-level info - 150062 SNPs

```

## Remove mislabeled individuals

From Madeline: There are 3 samples that are potentially mislabeled based on the fact that they cluster with Gulf samples TX/LA but are labeled as Atlantic.

1. VA-JR-DeepWatSh, B10_CviMVPe_PlDNA_11_70433_9fd0250a, PC1 = -0.101961382
2. ME-HogIs, B3_CviMVPe_PlDNA_13_70034_f5a02f77, PC1 = -0.094567742
3. VA-JR-DeepWatSh, B9_CviMVPe_PlDNA_11_70600_195573cd, PC1 = -0.086083214

Then there's also one Gulf sample that groups with the Atlantic.

1. TX-CapBay, B12_CviMVPe_PlDNA_11_70377_42164822, PC1 = 0.055471111

```{r}

head(juvs_pca_inds)

# remove mislabeled individuals
inds_to_remove <- c("B10_CviMVPe_PlDNA_11_70433_9fd0250a", "B3_CviMVPe_PlDNA_13_70034_f5a02f77", "B9_CviMVPe_PlDNA_11_70600_195573cd", "B12_CviMVPe_PlDNA_11_70377_42164822")

juvs_inds_clean <- juvs_pca_inds %>%
  filter(!as.character(clean_ID) %in% inds_to_remove)

dim(juvs_inds_clean) # 267 - 4 = 263 individuals remaining

# now remove them from both genotype matrices too

full_clean <- full %>%
  filter(!rownames(.) %in% inds_to_remove)

dim(full_clean) # 267 - 4 = 263 individuals remaining

gen_clean <- gen %>%
  filter(!rownames(.) %in% inds_to_remove)

dim(gen_clean) # 267 - 4 = 263 individuals remaining
```

## Shared genotype conversion function

This function converts a 0/1/2 genotype matrix to a genind object via string genotypes ("11"/"12"/"22"). This avoids hierfstat misreading raw 0/1/2 as triallelic (see 260609 note). Defined once here and used for both Fst (thinned) and basic stats (full).

```{r}

genotypes_to_genind <- function(numeric_genotypes, pop_labels) {
  geno_numeric <- as.matrix(numeric_genotypes)
  
  geno_strings <- matrix(NA_character_, 
                         nrow = nrow(geno_numeric), 
                         ncol = ncol(geno_numeric))
  geno_strings[geno_numeric == 0] <- "11"
  geno_strings[geno_numeric == 1] <- "12"
  geno_strings[geno_numeric == 2] <- "22"
  
  geno_strings <- as.data.frame(geno_strings)
  colnames(geno_strings) <- colnames(geno_numeric)
  
  df2genind(geno_strings, pop = pop_labels, ploidy = 2, ncode = 1)
}

```

## Shared population recoding

Used for both Fst and basic stats sections to keep labels consistent.

```{r}

recode_pop <- function(labels) {
  dplyr::case_match(labels,
    "TX-CapBay"        ~ "W1-TX",
    "LA-SisLake"       ~ "W2-LA",
    "FL-KingPlan"      ~ "W3-FL",
    "VA-JR-DeepWatSh"  ~ "W4-VA",
    "LOLA"             ~ "S1-LOLA",
    "DEBY"             ~ "S2-DEBY",
    "NH-GrtBay"        ~ "W5-NH",
    "ME-HogIs"         ~ "W6-ME",
    "Poly-NoOutcross"  ~ "SEEDMIX",
    "Poly-WithOutcross" ~ "HYBRIDMIX"
  )
}

population_order <- c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME", "SEEDMIX", "HYBRIDMIX")

```

## Calculating Fst - NOT REREUN FOR JUVENILES

### Data preparation

```{r}

# find SNPs that are NOT all heterozygous (all 1s)
keep_snps_fst <- apply(gen_clean, 2, function(col) !all(col == 1))

# filter genotype matrix for these SNPs
GEN_filtered_fst <- gen_clean[, keep_snps_fst]
dim(GEN_filtered_fst) # 0 columns removed

```

```{r}

# match sample order between genotype matrix and metadata
juvs_inds_clean <- juvs_inds_clean[match(rownames(GEN_filtered_fst), juvs_inds_clean$clean_ID), ]

# verify alignment
identical(rownames(GEN_filtered_fst), as.character(juvs_inds_clean$clean_ID)) # should be TRUE

# build population labels for Fst (mislabeled individuals already removed)
fst_pop_labels <- recode_pop(juvs_inds_clean$SpawnTrt_Label)
table(fst_pop_labels) # check sample counts per population

```

```{r}

# convert thinned genotype matrix to genind -> hierfstat
genind_fst <- genotypes_to_genind(GEN_filtered_fst, pop_labels = fst_pop_labels)

# verify biallelic: should be 2 columns per locus
ncol(genind_fst@tab) / ncol(GEN_filtered_fst) # should be 2

hierfstat_fst <- genind2hierfstat(genind_fst)
dim(hierfstat_fst) # col 1 is pop, rest are loci

```

### Data product: Fst matrix NOT RUN

```{r}

# run Fst calculation - cluster only
#fst_results <- pairwise.WCfst(hierfstat_fst, diploid = TRUE)
#fst_results[is.na(fst_results)] <- 0

# view results
#print(fst_results)

#saveRDS(fst_results, file.path("results", "parental_genetics", "fst_results.rds"))

# load in fst results - after running with cluster
fst_results <- readRDS(file.path("results", "parental_genetics", "20250606_fst_results.rds"))


# convert the matrix to a long format and edit column names
fst_plotting <- melt(fst_results, na.rm = TRUE)
colnames(fst_plotting) <- c("Var1", "Var2", "value")

# reorder the Var1 and Var2 columns based on the defined order
fst_plotting$Var1 <- factor(fst_plotting$Var1, levels = population_order)
fst_plotting$Var2 <- factor(fst_plotting$Var2, levels = population_order)

# view data one more time
fst_plotting

# create heatmap using viridis color palette
fig_fst_heatmap <- ggplot(fst_plotting, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") + 
  labs(x = "Population", y = "Population", fill = "Fst") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig_fst_heatmap

```

## Basic stats with hierfstat

### Data preparation

Same steps as before but with full SNP set. 

```{r}

# prepare the full SNP set
full <- full_juvs

dim(full) # 150062 SNPs

```

```{r}

# filter SNPs that are all heterozygous (all 1s)
keep_snps_bs <- apply(full, 2, function(col) !all(col == 1))

# filter genotype matrix for these SNPs
FULL_filtered <- full[, keep_snps_bs]
dim(FULL_filtered) # 0 columns removed

```

```{r}

# reorder to match metadata
FULL_filtered <- FULL_filtered[match(juvs_pca_inds$clean_ID, rownames(FULL_filtered)), ]

# verify alignment
identical(rownames(FULL_filtered), as.character(juvs_pca_inds$clean_ID)) # should be TRUE

# build population labels for basic stats (uses ALL individuals including mislabeled)
bs_pop_labels <- recode_pop(juvs_pca_inds$SpawnTrt_Label)

# check number of samples per population
table(bs_pop_labels)

# check for any NAs in population labels
sum(is.na(bs_pop_labels)) # should be 0

```

```{r}

# convert full genotype matrix to genind -> hierfstat (uses shared function defined above)
genind_obj_juv <- genotypes_to_genind(FULL_filtered, pop_labels = bs_pop_labels)

# verify biallelic: should be 2 columns per locus
ncol(genind_obj_juv@tab) / ncol(FULL_filtered) # should be 2

```

```{r}

# convert to hierfstat obj
hierfstat_obj_juv <- genind2hierfstat(genind_obj_juv)

# run basic stats
div <- basic.stats(hierfstat_obj_juv)

# quick sanity check on the new values
ho_pop <- colMeans(div$Ho, na.rm = TRUE)
hs_pop <- colMeans(div$Hs, na.rm = TRUE)

data.frame(
  Population = names(ho_pop),
  Ho = round(ho_pop, 3),
  Hs = round(hs_pop, 3),
  FIS_implied = round(1 - ho_pop/hs_pop, 3)
)

```

260609: This makes much more biological sense. 
- Wild Gulf populations (TX, LA, FL) are very similar to each other in Ho, Hs, and FIS. This is consistent with these being relatively connected populations within the Gulf region.
- Range edge populations (NH, ME) have slightly lower Hs than mid-range populations. Populations at range edges can have reduced genetic diversity due to smaller population sizes and reduced gene flow. 
- VA has the highest Hs among wild populations. Chesapeake oysters can show elevated diversity due to mixing of northern and southern lineages around the phylogeographic break.
- LOLA and DEBY have noticeably higher Ho than wild populations. Breeding programs typically maintain outcrossed lines specifically to avoid inbreeding depression, so observed heterozygosity stays high.
- HYBRIDMIX has the highest Ho (0.251). This is expected from an intentional outcross between divergent populations — F1 hybrids show heterozygote excess at loci where parents are fixed for different alleles.
- SEEDMIX has the highest FIS (0.192). This is also expected - when you mix samples from different populations (without outcrossing), the apparent population looks heterozygote-deficient because allele frequencies differ between the source populations.

### Observed and expected heterozygosity

```{r}
# extract values from basic.stats output
Ho <- div$Ho # one value per SNP per population
Hs <- div$Hs # one value per SNP per population
```

#### Sanity checks

```{r}
# range checks: are values within expected bounds?
summary(unlist(Ho))
summary(unlist(Hs))

# any out-of-range values?
cat("Ho > 1:", sum(Ho > 1, na.rm = TRUE), "\n")
cat("Ho < 0:", sum(Ho < 0, na.rm = TRUE), "\n")
cat("Hs > 0.5:", sum(Hs > 0.5, na.rm = TRUE), 
    "(small excess OK due to sample-size correction)\n")
```

```{r}
# population assignments: do they match between input and hierfstat object?
table(hierfstat_obj_juv[, 1])
table(bs_pop_labels)
```

```{r}
# population-level summary of Ho, Hs, and implied FIS
ho_pop <- colMeans(div$Ho, na.rm = TRUE)
hs_pop <- colMeans(div$Hs, na.rm = TRUE)

pop_summary <- data.frame(
  Population = names(ho_pop),
  Ho = round(ho_pop, 4),
  Hs = round(hs_pop, 4),
  FIS_implied = round(1 - ho_pop/hs_pop, 4)
)
pop_summary
```

```{r}
# monomorphic loci per population
monomorphic_per_pop <- apply(div$Hs, 2, function(x) sum(x == 0, na.rm = TRUE))
monomorphic_per_pop
```

```{r}
# convert Ho and Hs matrices to long-format data frames
Ho_long <- as.data.frame(Ho) %>%
  tibble::rownames_to_column("Locus") %>%
  pivot_longer(cols = -Locus, names_to = "Population", values_to = "Ho")

Hs_long <- as.data.frame(Hs) %>%
  tibble::rownames_to_column("Locus") %>%
  pivot_longer(cols = -Locus, names_to = "Population", values_to = "Hs")

# merge into one long-format dataframe
pop_data <- merge(Ho_long, Hs_long, by = c("Locus", "Population"))

# set population factor levels for consistent plotting order
pop_data$Population <- factor(pop_data$Population, levels = population_order)

# also reshape for paired Ho/Hs boxplot
pop_long <- pop_data %>%
  pivot_longer(cols = c("Ho", "Hs"), names_to = "Type", values_to = "Value")
```

#### Plot 1: Ho and Hs distributions by population (boxplot)

```{r}
ggplot(pop_long, aes(x = Population, y = Value, fill = Type)) +
  geom_boxplot() +
  labs(title = "Observed and expected heterozygosity by population",
       x = "Population",
       y = "Heterozygosity") +
  scale_fill_manual(values = c("Ho" = "blue", "Hs" = "orange")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}

# curious what these high dots are

# how many SNPs have Ho > 0.75 in each population?
high_ho_counts <- apply(div$Ho, 2, function(x) sum(x > 0.75, na.rm = TRUE))
high_ho_counts

# what fraction of total SNPs is that?
high_ho_counts / nrow(div$Ho)

```


#### Plot 2: Ho distribution by population (boxplot)

```{r}
ggplot(pop_data, aes(x = Population, y = Ho)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  labs(title = "Observed heterozygosity by population",
       x = "Population",
       y = "Observed heterozygosity (Ho)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Plot 3: Mean Ho and Hs by population (line plot)

```{r}
# calculate population means
avg_het_per_pop <- pop_data %>%
  group_by(Population) %>%
  summarise(
    Ho = mean(Ho, na.rm = TRUE),
    Hs = mean(Hs, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(Ho, Hs), names_to = "Type", values_to = "Value")

ggplot(avg_het_per_pop, aes(x = Population, y = Value, group = Type, color = Type)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Mean Ho and Hs by population",
       x = "Population",
       y = "Mean heterozygosity") +
  scale_color_manual(values = c("Ho" = "blue", "Hs" = "orange")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Allelic richness 

This should be fixed with the new hierfstat object that reads SNPs as biallelic.

```{r}

# calculate allelic richness
ar_results <- allelic.richness(hierfstat_obj_juv, min.n = 48, diploid = TRUE) # min.n = number in smallest population * 2

# extract results
ar_results <- ar_results$Ar # one value per SNP per population

# make column for locus
ar_results <- rownames_to_column(ar_results, var = "Locus")
ar_results <- ar_results[, c("Locus", colnames(ar_results)[-1])] # move to front

# format data for plotting
ar_long <- melt(ar_results, id.vars = "Locus", variable.name = "Population", value.name = "Allelic_Richness")

# calculate means to put on graph
means <- aggregate(Allelic_Richness ~ Population, data = ar_long, mean)

```

#### Plot 1: Mean

```{r}

ggplot(means, aes(x = factor(Population, levels = population_order), 
                  y = Allelic_Richness)) +
  geom_point(size = 4, color = "darkblue") +
  geom_hline(yintercept = 2, linetype = "dashed", color = "grey50") +
  labs(title = "Mean allelic richness by population",
       x = "Population", 
       y = "Mean allelic richness") +
  coord_cartesian(ylim = c(1.5, 2)) +  # zoom in
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#### Plot 2: Proportion of polymorphic SNPs 

```{r}

poly_props <- ar_long %>%
  group_by(Population) %>%
  summarise(prop_polymorphic = mean(Allelic_Richness > 1.5, na.rm = TRUE))

ggplot(poly_props, aes(x = factor(Population, levels = population_order),
                       y = prop_polymorphic)) +
  geom_col(fill = "lightblue") +
  labs(title = "Proportion of polymorphic SNPs by population",
       x = "Population",
       y = "Proportion polymorphic") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}

# make sure allelic richness is never greater than 2
max(ar_long$Allelic_Richness, na.rm = TRUE)  # should be 2

```

```{r}

# calculate average ar per population
avg_ar_per_pop <- ar_long %>%
  group_by(Population) %>%
  summarise(
    avg_allelic_richness = mean(Allelic_Richness, na.rm = TRUE),
  )

print(avg_ar_per_pop)

```

### Data product: AR, Ho, Hs

This will get used in a linear model that includes temperature, salinity, AR, and Ho. It is saved to the "parent_effects_H2F" folder under "data." This is for the JUVENILE data though, not parents.

```{r}

# reshape avg_het_per_pop back to wide format for merging
avg_het_wide <- avg_het_per_pop %>%
  pivot_wider(names_from = Type, values_from = Value) %>%
  rename(avg_observed_ho = Ho, avg_expected_hs = Hs)

# merge with allelic richness data
juv_genetic_effects <- merge(avg_het_wide, avg_ar_per_pop, by = "Population") %>%
  mutate(Population = factor(Population, levels = population_order)) %>%
  arrange(Population)

write_csv(juv_genetic_effects, file.path("results", "parent_effects_H2F", "juvenile_genetic_effects.csv"))

```

## Fst outliers with OutFLANK - NOT RERUN FOR JUVENILES

### Data preparation

```{r}

# rename for processing - using the full filtered matrix and pop labels from basic stats
outflank_data <- as.data.frame(FULL_filtered)
outflank_data$pop <- bs_pop_labels

# move pop column to front
outflank_data <- outflank_data[, c(ncol(outflank_data), 1:(ncol(outflank_data)-1))]

# convert population to factor
outflank_data$pop <- as.factor(outflank_data$pop)

# check dimensions one more time
dim(outflank_data)

```

```{r}

# OutFLANK inputs: snp matrix, locus names, population names
# documentation: https://github.com/whitlock/OutFLANK/blob/master/OutFLANK%20readme.pdf

# make the outflank input data frame
make_outflank_input <- function(df) {
  SNPmat <- as.matrix(df[, -1])         # exclude population column
  SNPmat <- apply(SNPmat, 2, as.numeric)
  SNPmat[is.na(SNPmat)] <- 9            # replace missing with 9
  
  locus_names <- colnames(df)[-1]       # SNP names
  pop_names <- df$pop           # population IDs
  
  FstDataFrame <- MakeDiploidFSTMat(SNPmat, locus_names, pop_names)
  return(FstDataFrame)
}

make_outflank_input(outflank_data)

```

### Data checks

```{r}

# data check: heterozygosity vs. Fst
plot(FstDataFrame$He, FstDataFrame$FST) 

# data check: Fst vs. FstNoCorr
plot(FstDataFrame$FST, FstDataFrame$FSTNoCorr) # uncorrected Fst is always larger

```

### Run OutFLANK

```{r}

# first, run OutFLANK on thinned SNPs
out_trim <- OutFLANK(FstDataFrame[FstDataFrame$LocusName %in% colnames(thinned_snps), ], # using pruned SNPs
                     LeftTrimFraction = 0.05, # defaults
                     RightTrimFraction = 0.05, 
                     Hmin = 0.1, 
                     NumberOfSamples = 8, # number of populations
                     qthreshold = 0.01)

# view results
str(out_trim)
head(out_trim$results)

# check the fit
OutFLANKResultsPlotter(out_trim, withOutliers = TRUE,
                       NoCorr = TRUE, Hmin = 0.1, binwidth = 0.001, Zoom =
                         FALSE, RightZoomFraction = 0.05, titletext = NULL) # looks fine

# zoom in on right tail
OutFLANKResultsPlotter(out_trim , withOutliers = TRUE,
                       NoCorr = TRUE, Hmin = 0.1, binwidth = 0.001, Zoom =
                         TRUE, RightZoomFraction = 0.15, titletext = NULL) # looks fine

# check p value histogram - should look flat (not great)
hist(out_trim$results$pvaluesRightTail)

```

### Calculate p-values

```{r}

# then, calulate p-values on the full set
P1 <- pOutlierFinderChiSqNoCorr(FstDataFrame, Fstbar = out_trim$FSTNoCorrbar, 
                                   dfInferred = out_trim$dfInferred, qthreshold = 0.01, Hmin=0.05)

# view data
head(P1)
tail(P1)

# look for outliers - highlighted in blue
my_out <- P1$OutlierFlag==TRUE
plot(P1$He, P1$FST, pch=19, col=rgb(0,0,0,0.1))
points(P1$He[my_out], P1$FST[my_out], col="blue")

# check the p-value histogram for the full set of data
# if there are outliers, it should look flat with an inflation near 0
hist(P1$pvaluesRightTail)

```

### Data product: OutFLANK results, SNP info

```{r}

# separating by chromosome
a = sprintf("%08d", muts$Position)
muts$chrom_position <- paste(muts$Chromosome, a, sep = ".")

# add chromosome positions to dataframe
P1$Affx.ID <- P1$LocusName
joined_P1 <- left_join(P1, muts, by = c("Affx.ID" = "Affx.ID"))

write_csv(joined_P1, file.path("results", "parental_genetics", "outflank_results.csv"))

```
