---
title: "Envr_Genetics_Length_Survival_Hatchery"
output: pdf_document
date: "2025-02-20"
---

This code analyzes the effect of conditions in the environments-of-origin and parent genetics on offspring oyster survival and shell length in the VIMS ABC hatchery and nursery.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Set working directory
```{r setwd}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Load packages
```{r packages}
library(tidyverse)
library(car)
library(ggplot2)
library(gridExtra)
library(multcompView) #extract significance letters from Tukey-Kramer test
library(cowplot) #arrange ggplots
library(ggpmisc)
library(vegan) #for Mantel tests
library(broom)#save lm results to tibbles
library(outliers)
```

##Read in necessary files
```{r read_csvs}
#this file contains raw lengths of all oysters aged 15-78 days
length <- read.csv(file.path("../../data/performance_H2F/latestage_length.csv"))

#this file has summary statistics on temperature and salinity for the 8 sites (W1-TX, W2-LA, W3-FL, S1-LOLA, S2-DEBY, W4-VA, W5-NH, W6-ME)
envr_summary <-  read.csv("../../data/envr_of_origin/envr_summary_stats.csv")

#this file contains survival rate data for larval oysters up to day 21 post hatching
survival <-  read.csv("../../data/performance_H2F/CViMVP_larvae_survival.csv")

parent_genetics <- read.csv("../../data/parent_effects_H2F/parent_genetic_effects.csv")
```

Extract survival rates from survival data frame, and reorganize data frame
```{r survival_cleaning}
#For survival rates, we only care about survival from day 21 (marked as day_an 15-21), so filter for those day_an only. Reminder that days 15-21 were grouped for analysis because there were constant larval drops occurring during that time as individuals reached the eyed stage, but the number of drops varied by site_name. Also exclude LARMIX, since there is no environmental data for that, given that it represents oysters from multiple sites
survival_day21 <- survival %>% 
  filter(Day_analysis == "15-21") %>% 
  filter(Tank_naming != "MVP-LARMIX")

#Exclude any groups with "small" in their label. The length data only considers the eyed oysters.
small <- grep("small", survival_day21$Group_Day_Label, ignore.case = TRUE)

#make new dataframe WITHOUT any indices saved in small, mean remaining data are everything except small larvae
survival_day21_sub <- survival_day21[-small, ]

#exclude "MVP" from group names and add in W or S labels
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-TX"] <- "W1-TX"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LA"] <- "W2-LA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"]  == "MVP-FL"] <- "W3-FL"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LOLA"] <- "S1-LOLA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-DEBY"] <- "S2-DEBY"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-JR"] <- "W4-VA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-NH"] <- "W5-NH"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-ME"] <- "W6-ME"

#select only tank naming and survival rate columns
survival_filter <- survival_day21_sub[, c("Tank_naming", "Survival_rate_perc")]

#rename columns
colnames(survival_filter) <- c("site_name", "percent_survival")
```

##Length df organizing
```{r length_cleaning_means}
#select columns with site name, day, and shell length in mm
length <- length[,c("site_name", "day_an", "shell_length_mm")]

colnames(length) <- c("site_name", "day_an", "length")

#filter out LARMIX from group
length_filter <- length %>% 
  filter(site_name != "H1-LARMIX") %>% 
  mutate(site_name = fct_relevel(site_name, c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME")))
  

#keep only data from days 15-21
length_21 <- length_filter %>% 
  filter(day_an == "15-21")

length_78 <- length_filter %>% 
  filter(day_an == "78")
```

##Remove extra columns from environmental data frame and add latitude
```{r envr_cleaning}
#remove extra site label, mean annual temp, mean annual salinity, and standard deviations from envr_summary
envr <- subset(envr_summary, select = -c(X, Temperature_st_dev, Salinity_st_dev, Mean_Annual_Temperature_C, Mean_Annual_Salinity_ppt))

summary(envr)

#rename cols
colnames(envr) <- c("site_name", "mean_max_temp", "mean_min_temp", "mean_max_sal", "mean_min_sal")
```

#Parent genetics df cleaning
```{r parent_genetics_cleaning}
#rename column names 
colnames(parent_genetics) <- c("site_name", "ho", "avg_expected_hs", "ar")
```

#Join survival, genetic, and envr data frames
```{r surv_join_dfs}
#make df of just genetic and envr data
envr_gen <- left_join(envr, parent_genetics, by = "site_name")

#merge survival data with genetic and environmental data
surv_merge <- left_join(survival_filter, envr, by = "site_name") %>% 
  left_join(parent_genetics, by = "site_name")

write.csv(surv_merge, "../../data/parent_effects_H2F/H_Survival_Envr_Genetics.csv", row.names = FALSE)
```

Since we want to use the full raw length data, we can't just left join the length, envr, and genetic data frames as they have different numbers of rows. We'll use for loops to subset our data for each site_name and create new columns for allelic richness, heterozygosity, mean annual temp, and mean annual salinity from there.

#Join length day 21, genetic, and envr data frames
```{r len21_join_dfs}
#make list of site names
site_ls <- as.character(unique(length_filter$site_name))
#make empty list for results of joined data for length 21
length21_join_ls <- list()
#use for loop to merge data
for (site in site_ls) {
  #check if site has been processed and placed in the length21_join_ls
	if(!(site %in% names(length21_join_ls))) {
    
	  # subset length, genetics, and envr dataframes for current site
		len21_sub <- length_21 %>%
			filter(site_name == site)
		
		gen_sub <- parent_genetics %>%
			filter(site_name == site)	
		
		envr_sub <- envr %>% 
		  filter(site_name == site)
	  
		#add columns to length subset for allelic richness, average observed heterozygosity, mean annual temp, and mean annual salinity
		len21_sub$ar <- gen_sub$ar
		len21_sub$ho <- gen_sub$ho
		len21_sub$mean_min_temp <- envr_sub$mean_min_temp
		len21_sub$mean_max_temp <- envr_sub$mean_max_temp
		len21_sub$mean_min_sal <- envr_sub$mean_min_sal
		len21_sub$mean_max_sal <- envr_sub$mean_max_sal
		
		#save len21_sub for the given site_name to the list and name it after the current site
		length21_join_ls[[site]] <- len21_sub
	}	
}

#rbind all the separate site's data frames into one data frame and remove row names
len21_envr_gen <- do.call(rbind, length21_join_ls)
rownames(len21_envr_gen) <- NULL

#Sanity check- make sure row numbers match between the original length_21 data frame and the new one with the added columns
nrow(len21_envr_gen) #nrow = 619
nrow(length_21) #nrow = 619

summary(len21_envr_gen)
len21_envr_gen$site_name <- as.character(len21_envr_gen$site_name)

write.csv(len21_envr_gen, "../../data/parent_effects_H2F/H_Length21_Envr_Genetics.csv", row.names = FALSE)
```

#Join length day 78, genetic, and envr data frames
```{r len78_join_dfs}
#make empty list for results of joined data for length 78
length78_join_ls <- list()
#use for loop to merge data
for (site in site_ls) {
  #check if site has been processed and placed in the length78_join_ls
	if(!(site %in% names(length78_join_ls))) {
    
	  # subset length, genetics, and envr dataframes for current site
		len78_sub <- length_78 %>%
			filter(site_name == site)
		
		gen_sub <- parent_genetics %>%
			filter(site_name == site)	
		
		envr_sub <- envr %>% 
		  filter(site_name == site)
	  
		#add columns to length subset for allelic richness, average observed heterozygosity, mean annual temp, and mean annual salinity
		len78_sub$ar <- gen_sub$ar
		len78_sub$ho <- gen_sub$ho
		len78_sub$mean_min_temp <- envr_sub$mean_min_temp
		len78_sub$mean_max_temp <- envr_sub$mean_max_temp
		len78_sub$mean_min_sal <- envr_sub$mean_min_sal
		len78_sub$mean_max_sal <- envr_sub$mean_max_sal
		
		#save len78_sub for the given site_name to the list and name it after the current site
		length78_join_ls[[site]] <- len78_sub
	}	
}

#rbind all the separate site's data frames into one data frame and remove row names
len78_envr_gen <- do.call(rbind, length78_join_ls)
rownames(len78_envr_gen) <- NULL

#Sanity check- make sure row numbers match between the original length_78 data frame and the new one with the added columns
nrow(len78_envr_gen) #nrow = 971
nrow(length_78) #nrow = 971

summary(len78_envr_gen)
len78_envr_gen$site_name <- as.character(len78_envr_gen$site_name)

write.csv(len78_envr_gen, "../../data/parent_effects_H2F/H_Length78_Envr_Genetics.csv", row.names = FALSE)
```

#Check for multicollinearity among explanatory variables
```{r multicollinearity}
cor_matrix <- cor(envr_gen[,c("mean_min_temp", "mean_max_temp", "mean_min_sal", "mean_max_sal", "ho", "ar")])
plot(envr_gen[,c("mean_min_temp", "mean_max_temp", "mean_min_sal", "mean_max_sal", "ho", "ar")])
#Average observed Ho and temperature are borderline collinear (r = 0.68).

#calculate variance inflation factor (VIF)- create VIF function
VIF <- function(a, b, c, d, e, f){
  1/(1-summary(lm(a ~ b + c + d + e + f))$r.squared)
}

#get VIF for mean minimum temperature
VIF(envr_gen$mean_min_temp, envr_gen$mean_max_temp, envr_gen$mean_min_sal, envr_gen$mean_max_sal, envr_gen$ho, envr_gen$ar)
#VIF_min_temp = 10.76

#VIF for mean maximum temp
VIF(envr_gen$mean_max_temp, envr_gen$mean_min_sal, envr_gen$mean_max_sal, envr_gen$ho, envr_gen$ar, envr_gen$mean_min_temp)
#VIF_max_temp = 4.75

#VIF for mean minimum salinity
VIF(envr_gen$mean_min_sal, envr_gen$mean_max_sal, envr_gen$ho, envr_gen$ar, envr_gen$mean_min_temp, envr_gen$mean_max_temp)
#VIF_min_sal = 2.80

#VIF for mean maximum salinity
VIF(envr_gen$mean_max_sal, envr_gen$ho, envr_gen$ar, envr_gen$mean_min_temp, envr_gen$mean_max_temp, envr_gen$mean_min_sal)
#VIF_mean_max = 13.54

#VIF for average observed heterozygosity
VIF(envr_gen$ho, envr_gen$ar, envr_gen$mean_min_temp, envr_gen$mean_max_temp, envr_gen$mean_min_sal, envr_gen$mean_max_sal)
#VIF_ho = 15.04

#VIF for average allelic richness
VIF(envr_gen$ar, envr_gen$mean_min_temp, envr_gen$mean_max_temp, envr_gen$mean_min_sal, envr_gen$mean_max_sal, envr_gen$ho)
#VIF_ar = 1.51

par(mfrow = c(1,1))

#Mean min temperature, mean max salinity, and average observed heterozygosity have somewhat high VIFs (VIF > 10). Run the lm excluding each high-VIF variable individually.

#exclude mean min temperature
no_min_temp_lm <- lm(length ~ mean_max_temp + mean_min_sal + mean_max_sal + ar + ho, data = len21_envr_gen)
no_min_temp_resid <- no_min_temp_lm$residuals
shapiro.test(no_min_temp_resid) #p = 2.7e-7, not normal
hist(no_min_temp_resid) #left-skewed

#exclude mean max salinity
no_max_sal_lm <- lm(length ~ mean_min_temp + mean_max_temp + mean_min_sal + ar + ho, data = len21_envr_gen)
no_max_sal_resid <- no_max_sal_lm$residuals
shapiro.test(no_max_sal_resid) #p = 3.5e-8, not normal
hist(no_max_sal_resid) #left-skewed

#exclude ho
no_ho_lm <- lm(length ~ mean_min_temp + mean_max_temp + mean_min_sal + mean_max_sal + ar, data = len21_envr_gen)
no_ho_resid <- no_ho_lm$residuals
shapiro.test(no_ho_resid) #p = 1.5e-8, not normal
hist(no_ho_resid) #left-skewed

par(mfrow = c(2,2))
plot(no_min_temp_lm) #homoscedasticity looks good, no outliers
plot(no_max_sal_lm) #homoscedasticity slightly better than no min temp, no outliers
plot(no_ho_lm) #homoscedasticity not quite as good as no max sal, but still really even, no outliers
```

#Model selection: outlier exclusion
```{r outlier_exclusion}
#there's one exceptionally low length from the TX group that is throwing off normality, I'm going to exclude it and try the models
len21_envr_gen_sub <- len21_envr_gen %>% 
  filter(length > 0.22)

#exclude mean min temperature
no_min_temp_lm2 <- lm(length ~ mean_max_temp + mean_min_sal + mean_max_sal + ar + ho, data = len21_envr_gen_sub)
no_min_temp_resid2 <- no_min_temp_lm2$residuals
shapiro.test(no_min_temp_resid2) #p = 0.007, not normal, but very improved from full data set
hist(no_min_temp_resid2) #left-skewed

#exclude mean max salinity
no_max_sal_lm2 <- lm(length ~ mean_min_temp + mean_max_temp + mean_min_sal + ar + ho, data = len21_envr_gen_sub)
no_max_sal_resid2 <- no_max_sal_lm2$residuals
shapiro.test(no_max_sal_resid2) #p = 0.001, not normal, but improved from full data set
hist(no_max_sal_resid2) #left-skewed

#exclude ho
no_ho_lm2 <- lm(length ~ mean_min_temp + mean_max_temp + mean_min_sal + mean_max_sal + ar, data = len21_envr_gen_sub)
no_ho_resid2 <- no_ho_lm2$residuals
shapiro.test(no_ho_resid2) #p = 0005, not normal, but improved from full data set
hist(no_ho_resid) #left-skewed

par(mfrow = c(2,2))
plot(no_min_temp_lm2) #homoscedasticity looks good, no outliers, best normality
plot(no_max_sal_lm2) #homoscedasticity looks good, no outliers
plot(no_ho_lm2) #homoscedasticity looks good, no outliers, lowest normality
```

Because the model excluding mean minimum temperature has the best normality out of all the models, I am going to recalculate multicollinearity and VIF excluding that variable.
```{r no_mean_min_temp}
cor_matrix2 <- cor(envr_gen[,c("mean_max_temp", "mean_min_sal", "mean_max_sal", "ho", "ar")])
plot(envr_gen[,c("mean_max_temp", "mean_min_sal", "mean_max_sal", "ho", "ar")])
#Average observed Ho and mean max temperature have highest correlation, but not over 0.7 threshold (r = 0.67).

#calculate variance inflation factor (VIF)- create VIF function
VIF2 <- function(a, b, c, d, e){
  1/(1-summary(lm(a ~ b + c + d + e))$r.squared)
}

#VIF for mean maximum temp
VIF2(envr_gen$mean_max_temp, envr_gen$mean_min_sal, envr_gen$mean_max_sal, envr_gen$ho, envr_gen$ar)
#VIF_max_temp = 4.49

#VIF for mean minimum salinity
VIF2(envr_gen$mean_min_sal, envr_gen$mean_max_sal, envr_gen$ho, envr_gen$ar, envr_gen$mean_max_temp)
#VIF_min_sal = 1.33

#VIF for mean maximum salinity
VIF2(envr_gen$mean_max_sal, envr_gen$ho, envr_gen$ar, envr_gen$mean_max_temp, envr_gen$mean_min_sal)
#VIF_mean_max = 2.64

#VIF for average observed heterozygosity
VIF2(envr_gen$ho, envr_gen$ar, envr_gen$mean_max_temp, envr_gen$mean_min_sal, envr_gen$mean_max_sal)
#VIF_ho = 4.77

#VIF for average allelic richness
VIF2(envr_gen$ar, envr_gen$mean_max_temp, envr_gen$mean_min_sal, envr_gen$mean_max_sal, envr_gen$ho)
#VIF_ar = 1.43
```

The next chunk includes multiple regressions testing the effects of several environmental and genetic parameters on length and survival in the hatchery/nursery. The explanatory variables are 1) mean maximum temperature at the environment-of-origin (mean_max_temp), 2) mean minimum salinity at the environment of origin (mean_min_sal), 3) mean maximum salinity at the environment of origin (mean_max_sal), 4) mean observed heterozygosity in the parent oysters from each site (Ho), and 5) mean allelic richness (AR) in the parent oysters from each site.

#Day 21 Length: Multiple regression
```{r}
#I am using the data excluding the one very low length measurement. I will square the length values to try to normalize the left-skewed residuals
len21_lm <- lm(length^2 ~ mean_max_temp + mean_min_sal + mean_max_sal + ho + ar, data = len21_envr_gen_sub)
len21_resid <- len21_lm$residuals
shapiro.test(len21_resid) #p = 0.12.42, normal, p = 0.0004 when including all data
#plot(len21_lm) #no outliers, homoscedasticity looks good
summary(len21_lm) #everything is significant. Model stats: Adj. R squared = 0.1075, F = 15.87, p = 1.035e-14
```

#Day 78 Length: Multiple regression and model selection 
```{r day78_lm}
#using all data for day 78
#day 78 multiple regression
len78_lm <- lm(length ~ mean_max_temp + mean_min_sal + mean_max_sal + ho + ar, data = len78_envr_gen)
len78_resid <- len78_lm$residuals
shapiro.test(len78_resid) #p = 0.0282 with all data
hist(len78_resid) #these data are right skewed
#plot(len78_lm) #no outliers, homoscedasticty is good

#try log transforming length values to improve normality
loglen78_lm <- lm(log10(length) ~ mean_max_temp + mean_min_sal + mean_max_sal + ho + ar, data = len78_envr_gen)
loglen78_resid <- loglen78_lm$residuals
shapiro.test(loglen78_resid) #p < 2.2e-16 not normal, much worse than raw data

#try square root transforming length values to improve normality
sqrtlen78_lm <- lm(length^(1/2) ~ mean_max_temp + mean_min_sal + mean_max_sal + ho + ar, data = len78_envr_gen)
sqrtlen78_resid <- sqrtlen78_lm$residuals
shapiro.test(sqrtlen78_resid) #p = 0.00012 not normal, worse than raw data

#try cube root transforming length values to improve normality
cu_len78_lm <- lm(length^(1/3) ~ mean_max_temp + mean_min_sal + mean_max_sal + ho + ar, data = len78_envr_gen)
cu_len78_resid <- cu_len78_lm$residuals
shapiro.test(cu_len78_resid) #p = 2.42e-9 not normal, much worse than raw data

#try squaring transforming length values to improve normality
sq_len78_lm <- lm(length^(2) ~ mean_max_temp + mean_min_sal + mean_max_sal + ho + ar, data = len78_envr_gen)
sq_len78_resid <- sq_len78_lm$residuals
shapiro.test(sq_len78_resid) #p < 2.2e-16 not normal, much worse than raw data

#going to stick with the raw length for this day for now, normality isn't that off
summary(len78_lm)
#mean max salinity is not a significant predictor of length at day 78 (p = 0.1178). Everything else is significant
#model stats: Adj. R squared = 0.280, F = 76.46, p < 2.2e-16

#remove mean max salinity from model and rerun
len78_no_max_sal_lm <- lm(length ~ mean_max_temp + mean_min_sal + ho + ar, data = len78_envr_gen)
len78_no_max_sal_resid <- len78_no_max_sal_lm$residuals
shapiro.test(len78_no_max_sal_resid) #p = 0.025, not normal, as expected
#plot(len78_no_max_sal_lm) #no outliers, homoscedasticty is good
summary(len78_no_max_sal_lm) #all variables significant, overall model stats: adj. R squared = 0.279, F = 94.82, p < 2.2e-16
```

#Day 21 Survival: Multiple Regression
```{r day21_surv_lm}
#day 78 multiple regression
surv_lm <- lm(log10(percent_survival) ~ mean_max_temp + mean_min_sal + mean_max_sal + ho + ar, data = surv_merge)
surv_resid <- surv_lm$residuals
shapiro.test(surv_resid) #p = 0.37, normal
#plot(surv_lm) #no outliers, homoscedasticty is good
summary(surv_lm)

#I don't think this model is worth running, too few data points, many outliers, not homoscedastic
```

Below adjusted p-values have not been updated to reflect current model p values 
#adjust p-values
```{r}
#model_p_values <- c(0.06075, 0.89, 0.5623)

#model_p_adj <- p.adjust(model_p_values, method = "BH")

#model_p_adj
```


