---
title: "Envr_Genetics_Length_Survival_Field"
output: pdf_document
date: "2025-02-20"
---

This code analyzes the effect of conditions in the environments-of-origin and parent genetics on offspring oyster survival and shell length in field sites at the York and Lewisetta Rivers

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Set working directory
```{r setwd}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Load packages
```{r packages}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(multcompView) #extract significance letters from Tukey-Kramer test
library(grid)#for ggplot arranging
library(cowplot) #arrange ggplots
library(ggpmisc)
library(vegan) #for Mantel tests
library(broom)#save lm results to tibbles
```

##Read in necessary files
```{r read_csvs}
#this file contains mean shell lengths oysters in each bag at each of the three field sampling dates. There are six bags per broodstock group, with three bags at the York site and three at the Lewisetta site.
length <- read.csv(file.path("../../data/performance_H2F/lengths_bags_022025.csv"))

#this file contains percent survival at each of the three field sampling dates. There are six bags per broodstock group, with three bags at the York site and three at the Lewisetta site.
survival <- read.csv(file.path("../../data/performance_H2F/mortality_bags_022025.csv"))

#this file has summary statistics on temperature and salinity for the 8 sites (W1-TX, W2-LA, W3-FL, S1-LOLA, S2-DEBY, W4-VA, W5-NH, W6-ME)
envr_summary <-  read.csv("../../data/envr_of_origin/envr_summary_stats.csv")

#this file contains average observed heterozygosity, average expected heterozygosity, and average allelic richness across every loci at each site
parent_genetics <- read.csv("../../data/parent_effects_H2F/parent_genetic_effects.csv")
```

Clean survival df
```{r survival_cleaning}
#Exclude any LARMIX and SEEDMIX bags, which do are crosses of oysters from different groups, and do not contain distinct environmental or genetic data
survival <- survival %>% 
  filter(pop != "SEEDMIX" & pop != "LARMIX")

#add in W or S labels to group names
survival["pop"][survival["pop"] == "TX"] <- "W1-TX"
survival["pop"][survival["pop"] == "LA"] <- "W2-LA"
survival["pop"][survival["pop"] == "FL"] <- "W3-FL"
survival["pop"][survival["pop"] == "LOLA"] <- "S1-LOLA"
survival["pop"][survival["pop"] == "DEBY"] <- "S2-DEBY"
survival["pop"][survival["pop"] == "JR"] <- "W4-VA"
survival["pop"][survival["pop"] == "NH"] <- "W5-NH"
survival["pop"][survival["pop"] == "ME"] <- "W6-ME"

#change bag_site York River to York
survival["bag_site"][survival["bag_site"] == "YorkRiver"] <- "York"

#select only site names (labeled "pop" for now), bag labels, bag field site locations, and survival at times t1, t2, and t3 (each of the field sampling sites respectively).
survival_filter <- survival[, c("pop", "bags_label", "bag_site", "t1", "t2", "t3")]

#rename columns
colnames(survival_filter) <- c("site_name", "bag_label", "bag_site", "t1_surv", "t2_surv", "t3_surv")

#reorder df by site_name
survival_filter <- survival_filter %>% 
  mutate(site_name = factor(site_name)) %>% 
  mutate(site_name = fct_relevel(site_name, c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME"))) %>% 
  arrange(site_name)

```

##Length df organizing
```{r length_cleaning_means}
#select columns with bag, monitoring event, length, and field site name
length <- length[,c("bag", "monitoring_event", "length", "site")]
#rename columns
colnames(length) <- c("bag_label", "monitoring_event", "length", "bag_site")

#change bag_site York River to York
length["bag_site"][length["bag_site"] == "York River"] <- "York"

#make string of site names with 6 replicates of each name (for each field bag) in the order they appear in the 
site_name1 <- c(rep("S2-DEBY", 6), rep("W3-FL", 6), rep("W4-VA", 6), rep("W2-LA", 6), rep("H1-LARMIX", 6), rep("S1-LOLA", 6), rep("W6-ME", 6), rep("W5-NH", 6), rep("H2-SEEDMIX", 6), rep("W1-TX", 6))

#replicate site_name1 string 3 times for the three field sampling dates
site_name <- rep(site_name1, 3)
#add site_name column to length df
length$site_name <- site_name

#Now remove any rows with site names H2-SEEDMIX or H1-LARMIX, since neither of these have genetic/environment-of-origin data
length_filter <- length %>% 
  filter(site_name != "H1-LARMIX" & site_name != "H2-SEEDMIX")

#reorder df by site_name
length_filter <- length_filter %>% 
  mutate(site_name = factor(site_name)) %>% 
  mutate(site_name = fct_relevel(site_name, c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME"))) %>% 
  arrange(site_name)

length_filter <- length_filter[, c(5,1,4,2,3)]
```

##Remove extra columns from environmental data frame and add latitude
```{r envr_cleaning}
#keep only site name, mean annual temp, and mean annual salinity
envr <- subset(envr_summary, select = c("site_name", "Mean_Annual_Temperature_C", "Mean_Annual_Salinity_ppt"))

summary(envr)

#add sampling site latitudes. Note that these latitudes ARE NOT the same as the latitudes of where environmental data came from. These latitudes, while similar, were often slightly different, and I am using our actual sampling site latitudes.
envr$lat <- c(28.096000, 29.239925, 30.440030, 37.1501163, 37.98030, 37.249107, 43.053746, 44.01330)
```


```{r parent_genetics_cleaning}
#rename column names 
colnames(parent_genetics) <- c("site_name", "avg_observed_ho", "avg_expected_hs", "avg_allelic_richness")
```

#Join genetics and envr data frames and merge with length and survival
```{r join_dfs}
#merge genetics and envr data
envr_gen <- left_join(envr, parent_genetics, by = "site_name")

#merge envr_gen with survival data and save as csv
surv_envr_gen <- full_join(survival_filter, envr_gen, by = "site_name")
write.csv(surv_envr_gen, "../../data/parent_effects_H2F/Field_Survival_Envr_ParentGenetics.csv", row.names = FALSE)

#merge envr_gen with length data and save as csv
length_envr_gen <- full_join(length_filter, envr_gen, by = "site_name")
write.csv(length_envr_gen, "../../data/parent_effects_H2F/Field_Length_Envr_ParentGenetics.csv", row.names = FALSE)

#split up data based on bag_site
lew_surv_df <- surv_envr_gen %>% 
  filter(bag_site == "Lewisetta")
york_surv_df <- surv_envr_gen %>% 
  filter(bag_site == "York")
lew_length_df <- length_envr_gen %>% 
  filter(bag_site == "Lewisetta")
york_length_df <- length_envr_gen %>% 
  filter(bag_site == "York")

#test correlation between temperature and latitude. If they are highly correlated, one must be excluded from the upcoming multiple regression.
temp_lat_corr <- cor.test(length_envr_gen$Mean_Annual_Temperature_C, length_envr_gen$lat)
temp_lat_corr #significant correlation (r = -0.962, p = 0.00014) between temp and lat, cannot include both in multiple regression
```

The next chunk includes multiple regressions testing the effects of several environmental and genetic parameters on length and survival in the field. Data are split up by bag_site (Lewisetta or York) The explanatory variables are 1) mean annual temperature at the environment-of-origin (mean temp), 2) mean annual salinity at the environment of origin (mean sal), 3) mean observed heterozygosity in the parent oysters from each site (Ho), and 4) mean allelic richness (AR) in the parent oysters from each site.

## Lewisetta multiple regressions
```{r mult_regressions}
#multiple regression testing effect of mean annual temp, mean annual salinity, mean Ho, and mean AR on percent survival at the first field monitoring time. Assess normality of residuals with Shapiro-Wilk test 
lew_t1surv_mr <- lm(t1_surv ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt + avg_observed_ho + avg_allelic_richness, data = lew_surv_df)
lew_t1surv_resid <- lew_t1surv_mr$residuals
shapiro.test(lew_t1surv_resid) #p = 0.7306, data are normal
#plot(lew_t1surv_mr) #homoscedasticity looks fine, no outliers
summary(lew_t1surv_mr) 
#model p-value = 0.0059
#model F-stat = 5.292
#adjusted R squared = 0.45
surv21_tidy <- tidy(surv21_mr) #save coefficients as table
```

#adjust p-values
```{r}
model_p_values <- c(0.06075, 0.89, 0.5623)

model_p_adj <- p.adjust(model_p_values, method = "BH")

model_p_adj
```


