---
title: "Envr_Genetics_Length_Survival_Hatchery"
output: pdf_document
date: "2025-02-20"
---

This code analyzes the effect of conditions in the environments-of-origin and parent genetics on offspring oyster survival and shell length in the VIMS ABC hatchery and nursery.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Set working directory
```{r setwd}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Load packages
```{r packages}
library(tidyverse)
library(car)
library(ggplot2)
library(gridExtra)
library(multcompView) #extract significance letters from Tukey-Kramer test
library(cowplot) #arrange ggplots
library(ggpmisc)
library(vegan) #for Mantel tests
library(broom)#save lm results to tibbles
```

##Read in necessary files
```{r read_csvs}
#this file contains raw lengths of all oysters aged 15-78 days
length <- read.csv(file.path("../../data/performance_H2F/latestage_length.csv"))

#this file has summary statistics on salinity in the environments of origin for the 8 sites (W1-TX, W2-LA, W3-FL, S1-LOLA, S2-DEBY, W4-VA, W5-NH, W6-ME)
sal <-  read.csv("../../data/environment/all_salinity.csv")

#this file has summary statistics on temperature in the environments of origin for the 8 sites (W1-TX, W2-LA, W3-FL, S1-LOLA, S2-DEBY, W4-VA, W5-NH, W6-ME)
temp <-  read.csv("../../data/environment/all_temperature.csv")

#this file contains survival rate data for larval oysters up to day 21 post hatching
survival <-  read.csv("../../data/performance_H2F/CViMVP_larvae_survival.csv")

parent_genetics <- read.csv("../../data/parent_effects_H2F/parent_genetic_effects.csv")
```

Extract survival rates from survival data frame, and reorganize data frame
```{r survival_cleaning}
#For survival rates, we only care about survival from day 21 (marked as day_an 15-21), so filter for those day_an only. Reminder that days 15-21 were grouped for analysis because there were constant larval drops occurring during that time as individuals reached the eyed stage, but the number of drops varied by site_name. Also exclude LARMIX, since there is no environmental data for that, given that it represents oysters from multiple sites
survival_day21 <- survival %>% 
  filter(Day_analysis == "15-21") %>% 
  filter(Tank_naming != "MVP-LARMIX")

#Exclude any groups with "small" in their label. The length data only considers the eyed oysters.
small <- grep("small", survival_day21$Group_Day_Label, ignore.case = TRUE)

#make new dataframe WITHOUT any indices saved in small, mean remaining data are everything except small larvae
survival_day21_sub <- survival_day21[-small, ]

#exclude "MVP" from group names and add in W or S labels
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-TX"] <- "W1-TX"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LA"] <- "W2-LA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"]  == "MVP-FL"] <- "W3-FL"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LOLA"] <- "S1-LOLA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-DEBY"] <- "S2-DEBY"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-JR"] <- "W4-VA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-NH"] <- "W5-NH"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-ME"] <- "W6-ME"

#select only tank naming and survival rate columns
survival_filter <- survival_day21_sub[, c("Tank_naming", "Survival_rate_perc")]

#rename columns
colnames(survival_filter) <- c("site_name", "percent_survival")
```

##Length df organizing
```{r length_cleaning_means}
#select columns with site name, day, and shell length in mm
length <- length[,c("site_name", "day_an", "shell_length_mm")]

colnames(length) <- c("site_name", "day_an", "length")

#filter out LARMIX from group
length_filter <- length %>% 
  filter(site_name != "H1-LARMIX") %>% 
  mutate(site_name = fct_relevel(site_name, c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME")))
  

#keep only data from days 15-21
length_21 <- length_filter %>% 
  filter(day_an == "15-21")

length_78 <- length_filter %>% 
  filter(day_an == "78")
```

##Remove extra columns from environmental data frames and add latitude
```{r envr_cleaning}
#merge temp and sal into one file
envr <- left_join(temp, sal, by = "site_name")
summary(envr)

#remove extra columns
envr_sub <- subset(envr, select = c(site_name, temp_quantile_10, temp_quantile_90, Temperature_st_dev, salinity_quantile_10, salinity_quantile_90, Salinity_st_dev))

#make site names standard format with the rest of the data frames
envr_sub["site_name"][envr_sub["site_name"] == "W1_TX"] <- "W1-TX"
envr_sub["site_name"][envr_sub["site_name"] == "W2_LA"] <- "W2-LA"
envr_sub["site_name"][envr_sub["site_name"] == "W3_FL"] <- "W3-FL"
envr_sub["site_name"][envr_sub["site_name"] == "W4_VA"] <- "W4-VA"
envr_sub["site_name"][envr_sub["site_name"] == "W5_NH"] <- "W5-NH"
envr_sub["site_name"][envr_sub["site_name"] == "W6_ME"] <- "W6-ME"

#remove rows for LEW and YRK - these are field sites not analyzed in this code
envr_sub <- envr_sub %>% 
  filter(site_name != "YRK" & site_name != "LEW")

colnames(envr_sub) <- c("site_name", "temp10", "temp90", "temp_st_dev", "sal10", "sal90", "sal_st_dev")
```

#Parent genetics df cleaning
```{r parent_genetics_cleaning}
#rename column names 
colnames(parent_genetics) <- c("site_name", "ho", "avg_expected_hs", "ar")
```

#Join survival, genetic, and envr data frames
```{r surv_join_dfs}
#make df of just genetic and envr data
envr_gen <- left_join(envr_sub, parent_genetics, by = "site_name")

#merge survival data with genetic and environmental data
surv_merge <- left_join(survival_filter, envr_sub, by = "site_name") %>% 
  left_join(parent_genetics, by = "site_name")

write.csv(surv_merge, "../../data/parent_effects_H2F/H_Survival_Envr_Genetics.csv", row.names = FALSE)
```

Since we want to use the full raw length data, we can't just left join the length, envr, and genetic data frames as they have different numbers of rows. We'll use for loops to subset our data for each site_name and create new columns for allelic richness, heterozygosity, mean annual temp, and mean annual salinity from there.

#Join length day 21, genetic, and envr data frames
```{r len21_join_dfs}
#make list of site names
site_ls <- as.character(unique(length_filter$site_name))
#make empty list for results of joined data for length 21
length21_join_ls <- list()
#use for loop to merge data
for (site in site_ls) {
  #check if site has been processed and placed in the length21_join_ls
	if(!(site %in% names(length21_join_ls))) {
    
	  # subset length, genetics, and envr dataframes for current site
		len21_sub <- length_21 %>%
			filter(site_name == site)
		
		gen_sub <- parent_genetics %>%
			filter(site_name == site)	
		
		envr1 <- envr_sub %>% 
		  filter(site_name == site)
	  
		#add columns to length subset for allelic richness, average observed heterozygosity, mean annual temp, and mean annual salinity
		len21_sub$ar <- gen_sub$ar
		len21_sub$ho <- gen_sub$ho
		len21_sub$temp10 <- envr1$temp10
		len21_sub$temp90 <- envr1$temp90
		len21_sub$sal10 <- envr1$sal10
		len21_sub$sal90 <- envr1$sal90
		
		#save len21_sub for the given site_name to the list and name it after the current site
		length21_join_ls[[site]] <- len21_sub
	}	
}

#rbind all the separate site's data frames into one data frame and remove row names
len21_envr_gen <- do.call(rbind, length21_join_ls)
rownames(len21_envr_gen) <- NULL

#Sanity check- make sure row numbers match between the original length_21 data frame and the new one with the added columns
nrow(len21_envr_gen) #nrow = 619
nrow(length_21) #nrow = 619

summary(len21_envr_gen)
len21_envr_gen$site_name <- as.character(len21_envr_gen$site_name)

write.csv(len21_envr_gen, "../../data/parent_effects_H2F/H_Length21_Envr_Genetics.csv", row.names = FALSE)
```

#Join length day 78, genetic, and envr data frames
```{r len78_join_dfs}
#make empty list for results of joined data for length 78
length78_join_ls <- list()
#use for loop to merge data
for (site in site_ls) {
  #check if site has been processed and placed in the length78_join_ls
	if(!(site %in% names(length78_join_ls))) {
    
	  # subset length, genetics, and envr dataframes for current site
		len78_sub <- length_78 %>%
			filter(site_name == site)
		
		gen_sub <- parent_genetics %>%
			filter(site_name == site)	
		
		envr1 <- envr_sub %>% 
		  filter(site_name == site)
	  
		#add columns to length subset for allelic richness, average observed heterozygosity, mean annual temp, and mean annual salinity
		len78_sub$ar <- gen_sub$ar
		len78_sub$ho <- gen_sub$ho
		len78_sub$temp10 <- envr1$temp10
		len78_sub$temp90 <- envr1$temp90
		len78_sub$sal10 <- envr1$sal10
		len78_sub$sal90 <- envr1$sal90
		
		#save len78_sub for the given site_name to the list and name it after the current site
		length78_join_ls[[site]] <- len78_sub
	}	
}

#rbind all the separate site's data frames into one data frame and remove row names
len78_envr_gen <- do.call(rbind, length78_join_ls)
rownames(len78_envr_gen) <- NULL

#Sanity check- make sure row numbers match between the original length_78 data frame and the new one with the added columns
nrow(len78_envr_gen) #nrow = 971
nrow(length_78) #nrow = 971

summary(len78_envr_gen)
len78_envr_gen$site_name <- as.character(len78_envr_gen$site_name)

write.csv(len78_envr_gen, "../../data/parent_effects_H2F/H_Length78_Envr_Genetics.csv", row.names = FALSE)
```

#Check for multicollinearity among explanatory variables
```{r multicollinearity}
cor_matrix <- cor(envr_gen[,c("temp10", "temp90", "sal10", "sal90", "ho", "ar")])
plot(envr_gen[,c("temp10", "temp90", "sal10", "sal90", "ho", "ar")])
#temp quantile 90 and ho are collinear (r = 0.805). Temp quantile 90 is also collinear with ar (r = 0.75). Everything else looks fine. Remove temp quantile 90 and check VIF.

#calculate variance inflation factor (VIF)- create VIF function
VIF <- function(a, b, c, d, e){
  1/(1-summary(lm(a ~ b + c + d + e))$r.squared)
}

#get VIF for temp quantile 10
VIF(envr_gen$temp10, envr_gen$sal10, envr_gen$sal90,  envr_gen$ho, envr_gen$ar)
#VIF_temp10 = 4.47

#VIF for salinity quantile 10
VIF(envr_gen$sal10, envr_gen$sal90,  envr_gen$ho, envr_gen$ar, envr_gen$temp10)
#VIF_sal10 = 3.47

#VIF for salinity quantile 90
VIF(envr_gen$sal90,  envr_gen$ho, envr_gen$ar, envr_gen$temp10, envr_gen$sal10)
#VIF_min_sal = 2.95

#VIF for ho
VIF(envr_gen$ho, envr_gen$ar, envr_gen$temp10, envr_gen$sal10, envr_gen$sal90)
#VIF_ho = 3.41

#VIF for average allelic richness
VIF(envr_gen$ar, envr_gen$temp10, envr_gen$sal10, envr_gen$sal90, envr_gen$ho)
#VIF_ar = 2.53

par(mfrow = c(1,1))
```

The next chunk includes multiple regressions testing the effects of several environmental and genetic parameters on length and survival in the hatchery/nursery. The explanatory variables are 1) temperature quantile 10, 2) salinity quantile 10, 3) salinity quantile 90, 4) mean observed heterozygosity in the parent oysters from each site (Ho), and 5) mean allelic richness (AR) in the parent oysters from each site.

#Day 21 Length: Multiple regression
```{r}
len21_lm <- lm(length ~ temp10 + sal10+ sal90 + ho + ar, data = len21_envr_gen)
len21_resid <- len21_lm$residuals
shapiro.test(len21_resid) #p = 2.048e-8, data are not normal, 
plot(len21_lm) #no outliers, homoscedasticity looks good
summary(len21_lm) #everything is significant. Model stats: Adj. R squared = 0.2354, F = 39.05, p < 2.2e-16

#there's one exceptionally low length from the TX group that is throwing off normality, I'm going to exclude it and try the models
len21_envr_gen_sub <- len21_envr_gen %>% 
  filter(length > 0.22)

par(mfrow = c(1,1))
#I am using the data excluding the one very low length measurement.
len21_lm2 <- lm(length ~ temp10 + sal10+ sal90 + ho + ar, data = len21_envr_gen_sub)
len21_resid2 <- len21_lm2$residuals
shapiro.test(len21_resid2) #p = 1.27e-5, data are not normal, but improved, multiple regressions are pretty robust to non-normality
#plot(len21_lm2) #no outliers, homoscedasticity looks good
summary(len21_lm2) #everything is significant. Model stats: Adj. R squared = 0.2329, F = 38.46, p < 2.2e-16

#model selection
drop1(len21_lm2) #this is the model with the lowest AIC (-5135.3)
```

#Day 78 Length: Multiple regression and model selection 
```{r day78_lm}
#using all data for day 78
#day 78 multiple regression
len78_lm <- lm(length ~ temp10 + sal10+ sal90 + ho + ar, data = len78_envr_gen)
len78_resid <- len78_lm$residuals
shapiro.test(len78_resid) #p = 0.00086 
hist(len78_resid) #these data are right skewed
#plot(len78_lm) #no outliers, homoscedasticty is good

#try log transforming length values to improve normality
loglen78_lm <- lm(log10(length) ~ temp10 + sal10+ sal90 + ho + ar, data = len78_envr_gen)
loglen78_resid <- loglen78_lm$residuals
shapiro.test(loglen78_resid) #p < 2.2e-16 not normal, much worse than raw data

#try square root transforming length values to improve normality
sqrtlen78_lm <- lm(length^(1/2) ~ temp10 + sal10+ sal90 + ho + ar, data = len78_envr_gen)
sqrtlen78_resid <- sqrtlen78_lm$residuals
shapiro.test(sqrtlen78_resid) #p = 6.541e-5 not normal, worse than raw data

#try cube root transforming length values to improve normality
cu_len78_lm <- lm(length^(1/3) ~ temp10 + sal10+ sal90 + ho + ar, data = len78_envr_gen)
cu_len78_resid <- cu_len78_lm$residuals
shapiro.test(cu_len78_resid) #p = 2.07e-9 not normal, much worse than raw data

#try squaring transforming length values to improve normality
sq_len78_lm <- lm(length^(2) ~ temp10 + sal10+ sal90 + ho + ar, data = len78_envr_gen)
sq_len78_resid <- sq_len78_lm$residuals
shapiro.test(sq_len78_resid) #p < 2.2e-16 not normal, much worse than raw data

#going to stick with the raw length for this day for now, normality isn't that off
summary(len78_lm)
#Everything is significant
#model stats: Adj. R squared = 0.2302, F = 59.01, p < 2.2e-16

#model selection
drop1(len78_lm) #drop nothing, but lowest AIC is 2410.2
```

#Day 21 Survival: Multiple Regression
I don't think this model is worth running, too few data points, many outliers, not homoscedastic, but I'll leave it here for now for reference.
```{r day21_surv_lm}
#day 78 multiple regression
surv_lm <- lm(percent_survival ~ temp10 + sal10+ sal90 + ho + ar, data = surv_merge)
surv_resid <- surv_lm$residuals
shapiro.test(surv_resid) #p = 0.99, normal
plot(surv_lm) #two outliers
summary(surv_lm) #only sal10 is significant

#model selection
drop1(surv_lm) #drop ho

#rerun model without ho
surv_lm2 <- lm(percent_survival ~ temp10 + sal10+ sal90 + ar, data = surv_merge)
drop1(surv_lm2) #dropping nothing at this point results in the lowest AIC, stop model selection now.
summary(surv_lm2) #now everything except sal90 is significant
#model stats: p = 0.001406, adj. R squared = 0.9841, F = 109 on 4 and 3 DF

```

Below adjusted p-values have not been updated to reflect current model p values 
#adjust p-values
```{r}
#model_p_values <- c(0.06075, 0.89, 0.5623)

#model_p_adj <- p.adjust(model_p_values, method = "BH")

#model_p_adj
```


