---
title: "Envr_of_Origin_Length_Survival_Hatchery"
output: pdf_document
date: "2024-10-26"
---

This code analyzes the effect of condition in the environments-of-origin on offspring oyster survival and shell length in the hatchery and nursery at VIMS ABC.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#setwd("~/Desktop/MVP-H2F-HatcheryField")
```

```{r packages}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(multcompView) #extract significance letters from Tukey-Kramer test
library(tidyverse)
library(grid)#for ggplot arranging
library(cowplot) #arrange ggplots
library(ggpmisc)
library(vegan) #for Mantel tests
```

#read in necessary files
```{r}
#this file contains raw lengths of all oysters aged 15-78 days
length <- read.csv(file.path("..", "..", "data", "performance_H2F", "latestage_length.csv"))

#this file has summary statistics on temperature and salinity for the 8 sites (DEBY, LOLA, JR, TX, LA, FL, NH, ME)
envr_summary <-  read.csv("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/NM_envr_for_violinplots.csv")

#this file is a matrix of the environmental distances between each site
envr_dist <- read.csv("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/NM_envr_dist.csv")

#this file contains survival rate data for larval oysters up to day 21 post hatching
survival <-  read.csv("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/CViMVP_larvae_survival.csv")
```

Extract survival rates from survival data frame, and reorganize data frame
```{r}
summary(survival)

#For survival rates, we only care about survival from day 21 (marked as day_an 15-21), so filter for those day_an only. Reminder that days 15-21 were grouped for analysis because there were constant larval drops occurring during that time, but the number of drops varied by site. Also exclude LARMIX, since there is no environmental data for that, given that it represents oysters from multiple sites
survival_day21 <- survival %>% 
  filter(Day_analysis == "15-21") %>% 
  filter(Tank_naming != "MVP-LARMIX")

#Exclude any groups with "small" in their label. The length data only considers the eyed oysters.
#use grep function to make a vector of all indices in latestage_len with the phrase "small" in them, ignore.case=TRUE means query is not case sensitive
small <- grep("small", survival_day21$Group_Day_Label, ignore.case = TRUE)

#make new dataframe WITHOUT any indices saved in small, mean remaining data are everything except small larvae
survival_day21_sub <- survival_day21[-small, ]

#exclude "MVP" from group names to shorten them
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-DEBY"] <- "DEBY"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"]  == "MVP-FL"] <- "FL"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-JR"] <- "JR"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LA"] <- "LA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LOLA"] <- "LOLA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-ME"] <- "ME"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-NH"] <- "NH"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-TX"] <- "TX"

#select only tank naming and survival rate columns
survival_filter <- survival_day21_sub[, c("Tank_naming", "Survival_rate_perc")]

#rename columns
colnames(survival_filter) <- c("site_name", "percent_survival")
```

#length df organizing
```{r}
#select columns with group, day, and shell length in mm
length <- length[,c("group", "day_an", "shell_length_mm")]

#filter out LARMIX from group
length_filter <- length %>% 
  filter(group != "LARMIX")

#keep only data from days 15-21
length_21 <- length_filter %>% 
  filter(day_an == "15-21")

length_78 <- length_filter %>% 
  filter(day_an == "78")

#find mean shell lengths of each group on each day
means_mm_21 <- aggregate(shell_length_mm ~ group, FUN = mean, data = length_21)
colnames(means_mm_21) <- c("site_name", "mean_length_mm_21")

means_mm_78 <- aggregate(shell_length_mm ~ group, FUN = mean, data = length_78)
colnames(means_mm_78) <- c("site_name", "mean_length_mm_78")
```

#remove extra columns from environmental data frame
```{r}
#remove extra site label and standard deviations from envr_summary
envr <- envr_summary[,-c(2,6,10)]

summary(envr)

#rename cols
colnames(envr) <- c("site_name", "Mean_Annual_Temperature_C", "Mean_max_temperature_C", "Mean_min_temperature_C", "Mean_Annual_Salinity_ppt", "Mean_max_Salinity_ppt", "Mean_min_Salinity_ppt")

#add sampling site latitudes. Note that these latitudes ARE NOT the same as the latitudes of where environmental data came from. These latitudes, while similar, were often slightly different, and I am using our actual sampling site latitudes. These latitudes are the same ones as those in table 1 in my thesis.
envr$lat <- c(37.249107, 37.98030, 37.1501163, 28.096000, 29.239925, 30.440030, 43.053746, 44.01330)
```

#Join data frames
```{r}
surv_length_envr <- left_join(envr, means_mm_21, by = "site_name") %>% 
  left_join(means_mm_78, by = "site_name") %>% 
  left_join(survival_filter, by = "site_name")

write.csv(surv_length_envr, "/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/Survival_Length_Envr_Data.csv", row.names = FALSE)
```

```{r}
#compare mean length on day 21 to mean annual temp, mean annual salinity, and latitude

length21_temp_lm <- lm(mean_length_mm_21 ~ Mean_Annual_Temperature_C, data = surv_length_envr)
summary(length21_temp_lm)
#extract p-value (.568)
length21_temp_pval <- summary(length21_temp_lm)$coefficients[,4]
length21_temp_pval <- unname(length21_temp_pval[2])


length21_sal_lm <- lm(mean_length_mm_21 ~ Mean_Annual_Salinity_ppt, data = surv_length_envr)
#extract p-value (.411)
length21_sal_pval <- summary(length21_sal_lm)$coefficients[,4]
length21_sal_pval <- unname(length21_sal_pval[2])

length21_lat_lm <- lm(mean_length_mm_21 ~ lat, data = surv_length_envr)
summary(length21_lat_lm)
#extract p-value (0.646)
length21_lat_pval <- summary(length21_lat_lm)$coefficients[,4]
length21_lat_pval <- unname(length21_lat_pval[2])

#compare percent survival to mean annual temp, mean annual salinity, and envr distance from DEBY
surv_temp_lm <- lm(percent_survival ~ Mean_Annual_Temperature_C, data = surv_length_envr)
summary(surv_temp_lm)
#extract p-value (0.024)
surv_temp_pval <- summary(surv_temp_lm)$coefficients[,4]
surv_temp_pval <- unname(surv_temp_pval[2])

surv_sal_lm <- lm(percent_survival ~ Mean_Annual_Salinity_ppt, data = surv_length_envr)
summary(surv_sal_lm)
#extract p-value (0.172)
surv_sal_pval <- summary(surv_sal_lm)$coefficients[,4]
surv_sal_pval <- unname(surv_sal_pval[2])

surv_lat_lm <- lm(percent_survival ~ lat, data = surv_length_envr)
summary(surv_lat_lm)
#extract p-value (0.0466)
surv_lat_pval <- summary(surv_lat_lm)$coefficients[,4]
surv_lat_pval <- unname(surv_lat_pval[2])

#compare length at day 78 to mean annual temp, mean annual salinity, and envr distance from DEBY
length78_temp_lm <- lm(mean_length_mm_78 ~ Mean_Annual_Temperature_C, data = surv_length_envr)
summary(length78_temp_lm)
#extract p-value (0.543)
length78_temp_pval <- summary(length78_temp_lm)$coefficients[,4]
length78_temp_pval <- unname(length78_temp_pval[2])


length78_sal_lm <- lm(mean_length_mm_78 ~ Mean_Annual_Salinity_ppt, data = surv_length_envr)
#extract p-value (0.983)
length78_sal_pval <- summary(length78_sal_lm)$coefficients[,4]
length78_sal_pval <- unname(length78_sal_pval[2])

length78_lat_lm <- lm(mean_length_mm_78 ~ lat, data = surv_length_envr)
#extract p-value (0.962)
length78_lat_pval <- summary(length78_lat_lm)$coefficients[,4]
length78_lat_pval <- unname(length78_lat_pval[2])
```

```{r}
#test correlation between temperature and latitude. If they are highly correlated, one must be excluded from the upcoming multiple regression.
temp_lat_corr <- cor.test(surv_length_envr$Mean_Annual_Temperature_C, surv_length_envr$lat)
temp_lat_corr #significant correlation between temp and lat, cannot include both in multiple regression

#multiple regression testing effect of mean annual temp and mean annual salinity at the environments-of-origin on percent survival
surv_temp_sal_lm <- lm(percent_survival ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt, data = surv_length_envr)
summary(surv_temp_sal_lm) #both explanatory variables significant, whole model significant

#multiple regression testing effect of mean annual temp and mean annual salinity at the environments-of-origin on mean length at the eyed stage (days 15-21)
length21_temp_sal_lm <- lm(mean_length_mm_21 ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt, data = surv_length_envr)
summary(length21_temp_sal_lm) #neither explanatory variable significant, model not significant

#multiple regression testing effect of mean annual temp and mean annual salinity at the environments-of-origin on mean length at the pre-deployment stage (day 78)
length78_temp_sal_lm <- lm(mean_length_mm_78 ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt, data = surv_length_envr)
summary(length78_temp_sal_lm) #neither explanatory variable significant, model insignificant
#put multiple regression output table in results
```


#adjust p-values
```{r}
p_values <- c(length21_temp_pval, length21_sal_pval, length21_lat_pval, surv_temp_pval, surv_sal_pval, surv_lat_pval, length78_temp_pval ,length78_sal_pval, length78_lat_pval)

p_adj <- p.adjust(p_values, method = "BH")

p_values

p_adj

```
ADJUSTED P-VALUES
- Mean annual temp is not a significant predictor of mean length at day 21 (p = 0.831)
- Mean annual salinity is not a significant predictor of mean length at day 21 (p = 0.831). 
- Latitude is not a significant predictor of mean length at day 21 (p = 0.831). 

- Mean annual temp is no longer a significant predictor of percent survival at day 21 (p = 0.210).
- Mean annual salinity is not a significant predictor of percent survival at day 21 (p = 0.516). 
- Latitude is not a significant predictor of percent survival at day 21 (p = 0.210). 

- Mean annual temp is not a significant predictor of mean length at day 78 (p = 0.831)
- Mean annual salinity is not a significant predictor of mean length at day 78 (p = 0.983). 
- Latitude is not a significant predictor of mean length at day 78 (p = 0.983).

```{r}
#make function to wrap title text
wrapper <- function(x, ...) 
{
  paste(strwrap(x, ...), collapse = "\n")
}

surv_temp_plot <- surv_length_envr %>%
  ggplot(aes(x = Mean_Annual_Temperature_C, y = percent_survival))+
  geom_point()+
  labs(subtitle = "A: p = 0.009, slope = 0.28")+
  ggtitle("Percent Survival vs. Mean Annual Temperature & Mean Annual Salinity")+
  ylab("% Survival")+
  xlab("Mean Annual Temperature (°C)")+
  ylim(c(0,5))+
  theme(axis.title.y = element_text(size = 9), axis.title.x = element_text(size = 9), axis.text.x = element_text(size = 9), plot.subtitle = element_text(size = 10), plot.title = element_text(size = 12))+
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE)

surv_temp_plot

surv_sal_plot <- surv_length_envr %>%
  ggplot(aes(x = Mean_Annual_Salinity_ppt, y = percent_survival))+
  geom_point()+
  labs(subtitle = "B: p = 0.041, slope = -0.08")+
  ylab("% Survival")+
  xlab("Mean Annual Salinity (ppt)")+
  theme(axis.title.x = element_text(size = 9), plot.title = element_blank(), plot.subtitle = element_text(size = 10), axis.title.y = element_text(size = 9))+
  geom_smooth(method = "lm", se = TRUE, fullrange = TRUE)

surv_sal_plot

surv_plots <- plot_grid(surv_temp_plot, surv_sal_plot, ncol = 1, align = "v")

surv_plots

ggsave("surv_plots.png",
       plot = surv_plots,
       device = png,
       path = "/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/results/NM_results/Result_plots")
```


