---
title: "Envr_Genetics_Length_Survival_Hatchery"
output: pdf_document
date: "2025-02-20"
---

This code analyzes the effect of conditions in the environments-of-origin and parent genetics on offspring oyster survival and shell length in the VIMS ABC hatchery and nursery.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

##Set working directory
```{r setwd}
setwd(dirname(rstudioapi::getActiveDocumentContext()$path))
```

## Load packages
```{r packages}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(multcompView) #extract significance letters from Tukey-Kramer test
library(cowplot) #arrange ggplots
library(ggpmisc)
library(vegan) #for Mantel tests
library(broom)#save lm results to tibbles
```

##Read in necessary files
```{r read_csvs}
#this file contains raw lengths of all oysters aged 15-78 days
length <- read.csv(file.path("../../data/performance_H2F/latestage_length.csv"))

#this file has summary statistics on temperature and salinity for the 8 sites (W1-TX, W2-LA, W3-FL, S1-LOLA, S2-DEBY, W4-VA, W5-NH, W6-ME)
envr_summary <-  read.csv("../../data/envr_of_origin/envr_summary_stats.csv")

#this file contains survival rate data for larval oysters up to day 21 post hatching
survival <-  read.csv("../../data/performance_H2F/CViMVP_larvae_survival.csv")

parent_genetics <- read.csv("../../data/parent_effects_H2F/parent_genetic_effects.csv")
```

Extract survival rates from survival data frame, and reorganize data frame
```{r survival_cleaning}
#For survival rates, we only care about survival from day 21 (marked as day_an 15-21), so filter for those day_an only. Reminder that days 15-21 were grouped for analysis because there were constant larval drops occurring during that time as individuals reached the eyed stage, but the number of drops varied by site_name. Also exclude LARMIX, since there is no environmental data for that, given that it represents oysters from multiple sites
survival_day21 <- survival %>% 
  filter(Day_analysis == "15-21") %>% 
  filter(Tank_naming != "MVP-LARMIX")

#Exclude any groups with "small" in their label. The length data only considers the eyed oysters.
small <- grep("small", survival_day21$Group_Day_Label, ignore.case = TRUE)

#make new dataframe WITHOUT any indices saved in small, mean remaining data are everything except small larvae
survival_day21_sub <- survival_day21[-small, ]

#exclude "MVP" from group names and add in W or S labels
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-TX"] <- "W1-TX"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LA"] <- "W2-LA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"]  == "MVP-FL"] <- "W3-FL"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LOLA"] <- "S1-LOLA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-DEBY"] <- "S2-DEBY"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-JR"] <- "W4-VA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-NH"] <- "W5-NH"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-ME"] <- "W6-ME"

#select only tank naming and survival rate columns
survival_filter <- survival_day21_sub[, c("Tank_naming", "Survival_rate_perc")]

#rename columns
colnames(survival_filter) <- c("site_name", "percent_survival")
```

##Length df organizing
```{r length_cleaning_means}
#select columns with site name, day, and shell length in mm
length <- length[,c("site_name", "day_an", "shell_length_mm")]

#filter out LARMIX from group
length_filter <- length %>% 
  filter(site_name != "H1-LARMIX") %>% 
  mutate(site_name = fct_relevel(site_name, c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME")))
  

#keep only data from days 15-21
length_21 <- length_filter %>% 
  filter(day_an == "15-21")

length_78 <- length_filter %>% 
  filter(day_an == "78")
```

##Remove extra columns from environmental data frame and add latitude
```{r envr_cleaning}
#remove extra site label and standard deviations from envr_summary
envr <- subset(envr_summary, select = -c(X, Temperature_st_dev, Salinity_st_dev))

summary(envr)

#rename cols
colnames(envr) <- c("site_name", "Mean_Annual_Temperature_C", "Mean_max_temperature_C", "Mean_min_temperature_C", "Mean_Annual_Salinity_ppt", "Mean_max_Salinity_ppt", "Mean_min_Salinity_ppt")
```

#Parent genetics df cleaning
```{r parent_genetics_cleaning}
#rename column names 
colnames(parent_genetics) <- c("site_name", "avg_observed_ho", "avg_expected_hs", "avg_allelic_richness")
```

#Join survival, genetic, and envr data frames
```{r surv_join_dfs}
#make df of just genetic and envr data
envr_gen <- left_join(envr, parent_genetics, by = "site_name")

#merge survival data with genetic and environmental data
surv_merge <- left_join(survival_filter, envr, by = "site_name") %>% 
  left_join(parent_genetics, by = "site_name")

write.csv(surv_merge, "../../data/parent_effects_H2F/H_Survival_Envr_Genetics.csv", row.names = FALSE)
```

Since we want to use the full raw length data, we can't just left join the length, envr, and genetic data frames as they have different numbers of rows. We'll use for loops to subset our data for each site_name and create new columns for allelic richness, heterozygosity, mean annual temp, and mean annual salinity from there.

#Join length day 21, genetic, and envr data frames
```{r len21_join_dfs}
#make list of site names
site_ls <- as.character(unique(length_filter$site_name))
#make empty list for results of joined data for length 21
length21_join_ls <- list()
#use for loop to merge data
for (site in site_ls) {
  #check if site has been processed and placed in the length21_join_ls
	if(!(site %in% names(length21_join_ls))) {
    
	  # subset length, genetics, and envr dataframes for current site
		len21_sub <- length_21 %>%
			filter(site_name == site)
		
		gen_sub <- parent_genetics %>%
			filter(site_name == site)	
		
		envr_sub <- envr %>% 
		  filter(site_name == site)
	  
		#add columns to length subset for allelic richness, average observed heterozygosity, mean annual temp, and mean annual salinity
		len21_sub$avg_allelic_richness <- gen_sub$avg_allelic_richness
		len21_sub$avg_observed_ho <- gen_sub$avg_observed_ho
		len21_sub$Mean_Annual_Temperature_C <- envr_sub$Mean_Annual_Temperature_C
		len21_sub$Mean_Annual_Salinity_ppt <- envr_sub$Mean_Annual_Salinity_ppt
		
		#save len21_sub for the given site_name to the list and name it after the current site
		length21_join_ls[[site]] <- len21_sub
	}	
}

#rbind all the separate site's data frames into one data frame and remove row names
len21_envr_gen <- do.call(rbind, length21_join_ls)
rownames(len21_envr_gen) <- NULL

#Sanity check- make sure row numbers match between the original length_21 data frame and the new one with the added columns
nrow(len21_envr_gen) #nrow = 619
nrow(length_21) #nrow = 619

write.csv(len21_envr_gen, "../../data/parent_effects_H2F/H_Length21_Envr_Genetics.csv", row.names = FALSE)
```

#Join length day 78, genetic, and envr data frames
```{r len78_join_dfs}
#make empty list for results of joined data for length 78
length78_join_ls <- list()
#use for loop to merge data
for (site in site_ls) {
  #check if site has been processed and placed in the length78_join_ls
	if(!(site %in% names(length78_join_ls))) {
    
	  # subset length, genetics, and envr dataframes for current site
		len78_sub <- length_78 %>%
			filter(site_name == site)
		
		gen_sub <- parent_genetics %>%
			filter(site_name == site)	
		
		envr_sub <- envr %>% 
		  filter(site_name == site)
	  
		#add columns to length subset for allelic richness, average observed heterozygosity, mean annual temp, and mean annual salinity
		len78_sub$avg_allelic_richness <- gen_sub$avg_allelic_richness
		len78_sub$avg_observed_ho <- gen_sub$avg_observed_ho
		len78_sub$Mean_Annual_Temperature_C <- envr_sub$Mean_Annual_Temperature_C
		len78_sub$Mean_Annual_Salinity_ppt <- envr_sub$Mean_Annual_Salinity_ppt
		
		#save len78_sub for the given site_name to the list and name it after the current site
		length78_join_ls[[site]] <- len78_sub
	}	
}

#rbind all the separate site's data frames into one data frame and remove row names
len78_envr_gen <- do.call(rbind, length78_join_ls)
rownames(len78_envr_gen) <- NULL

#Sanity check- make sure row numbers match between the original length_78 data frame and the new one with the added columns
nrow(len78_envr_gen) #nrow = 971
nrow(length_78) #nrow = 971

write.csv(len78_envr_gen, "../../data/parent_effects_H2F/H_Length78_Envr_Genetics.csv", row.names = FALSE)
```

#Check for multicollinearity among explanatory variables

```{r multicollinearity}
cor_matrix2 <- cor(envr_gen[,c("Mean_Annual_Temperature_C", "Mean_Annual_Salinity_ppt", "avg_observed_ho", "avg_allelic_richness")])
plot(envr_gen[,c("Mean_Annual_Temperature_C", "Mean_Annual_Salinity_ppt", "avg_observed_ho", "avg_allelic_richness")])
#Average observed Ho and temperature are borderline collinear (r = 0.68).

#calculate variance inflation factor (VIF)- create VIF function
VIF <- function(a, b, c, d){
  1/(1-summary(lm(a ~ b + c + d))$r.squared)
}

#get VIF for mean annual temperature
VIF(envr_gen$Mean_Annual_Temperature_C, envr_gen$Mean_Annual_Salinity_ppt, envr_gen$avg_observed_ho, envr_gen$avg_allelic_richness)
#VIF_temp = 5.6

#VIF for mean annual salinity
VIF(envr_gen$Mean_Annual_Salinity_ppt, envr_gen$avg_observed_ho, envr_gen$avg_allelic_richness, envr_gen$Mean_Annual_Temperature_C)
#VIF_sal = 3.82

#VIF for average observed ho
VIF(envr_gen$avg_observed_ho, envr_gen$avg_allelic_richness, envr_gen$Mean_Annual_Temperature_C, envr_gen$Mean_Annual_Salinity_ppt)
#VIF_obs_Ho = 6.68

#VIF for average allelic richness
VIF(envr_gen$avg_allelic_richness,envr_gen$Mean_Annual_Temperature_C, envr_gen$Mean_Annual_Salinity_ppt, envr_gen$avg_observed_ho)
#VIF_AR = 2.01

par(mfrow = c(1,1))

len21_envr_gen_test <- len21_envr_gen %>% 
  filter(shell_length_mm >= 0.23)
#Both temperature and average observed heterozygosity have somewhat high VIFs (VIF > 5). Run lm with one of the response variables with each of them without the other. I'll use length at day 21.
#exclude Ho
temp_lm <- lm(shell_length_mm ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt + avg_allelic_richness, data = len21_envr_gen)
temp_resid <- temp_lm$residuals
shapiro.test(temp_resid) #p < 0.01, not normal
hist(temp_resid)
summary(temp_lm)
#exclude temperature
ho_lm <- lm(shell_length_mm ~ Mean_Annual_Salinity_ppt + avg_allelic_richness + avg_observed_ho, data = len21_envr_gen)
ho_resid <- ho_lm$residuals
shapiro.test(ho_resid) #p < 0.01, not normal
summary(ho_lm)

#plot results
par(mfrow=c(2,2))
plot(temp_sal_ar_lm)
plot(sal_ho_ar_lm)
#both models have no outliers and are decently homoscedastic, but normality is off for both. No real difference in how well each model fits the assumptions.

#drop insignigicant predictors and recompare

#retry each lm with transformed length
#exclude Ho
q_temp_lm <- lm((shell_length_mm)^4 ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt + avg_allelic_richness, data = len21_envr_gen)
q_temp_resid <- q_temp_lm$residuals
shapiro.test(q_temp_resid) #p < 0.48, normal
hist(q_temp_resid)

#exclude temperature
q_ho_lm <- lm((shell_length_mm)^4 ~ Mean_Annual_Salinity_ppt + avg_allelic_richness + avg_observed_ho, data = len21_envr_gen)
q_ho_resid <- q_ho_lm$residuals
shapiro.test(q_ho_resid) #p = 0.079


par(mfrow = c(2,2))
plot(q_temp_lm)
plot(q_ho_lm)
#normality is improved, but still no difference in how well each model meets the assumptions. I'm inclined to just use both temp and ho in the model since their VIFs aren't that high
```

#Model selection
```{r}
len21_lm <- lm(shell_length_mm ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt + avg_allelic_richness + avg_observed_ho, data = len21_envr_gen)
len21_resid <- len21_lm$residuals
shapiro.test(len21_resid) #p < 0.01, not normal


qlen21_lm <- lm((shell_length_mm)^4 ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt + avg_allelic_richness + avg_observed_ho, data = len21_envr_gen)
qlen21_resid <- qlen21_lm$residuals
shapiro.test(qlen21_resid) #p = 0.48, normal
plot(qlen21_lm)
summary(qlen21_lm)

#rerun without salinity and avg observed ho (insigificant)
qlen21_lm <- lm((shell_length_mm)^4 ~ Mean_Annual_Temperature_C + avg_allelic_richness, data = len21_envr_gen)
qlen21_resid <- qlen21_lm$residuals
shapiro.test(qlen21_resid) #p = 0.06, normal
plot(qlen21_lm)
summary(qlen21_lm)
```



The next chunk includes multiple regressions testing the effects of several environmental and genetic parameters on length and survival in the hatchery/nursery. The explanatory variables are 1) mean annual temperature at the environment-of-origin (mean temp), 2) mean annual salinity at the environment of origin (mean sal), 3) mean observed heterozygosity in the parent oysters from each site (Ho), and 4) mean allelic richness (AR) in the parent oysters from each site.

## Multiple regressions
```{r mult_regressions}
#multiple regression testing effect of mean annual temp, mean annual salinity, mean Ho, and mean AR on percent survival at day 21 (end of hatchery stage)
surv21_mr <- lm(percent_survival ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt + avg_observed_ho + avg_allelic_richness, data = surv_length_envr_gen)
surv21_resid <- surv21_mr$residuals
shapiro.test(surv21_resid) #p = 0.018, not normal, try log10 transforming survival

logsurv21_mr <- lm(log10(percent_survival) ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt + avg_observed_ho + avg_allelic_richness, data = surv_length_envr_gen)
logsurv21_resid <- logsurv21_mr$residuals
shapiro.test(logsurv21_resid) #p = 0.075, data are normal
plot(logsurv21_mr)


#multiple regression testing effect of mean annual temp, mean annual salinity, mean Ho, and mean AR on percent length at day 21 (end of hatchery stage)
length21_mr <- lm(mean_length_mm_21 ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt + avg_observed_ho + avg_allelic_richness, data = surv_length_envr_gen)
summary(length21_mr) 
#model p-value = 0.89
#model F-stat = 0.2551
#Adjusted R squared = -0.74
length21_tidy <- tidy(length21_mr) #save coefficients in table

#multiple regression testing effect of mean annual temp, mean annual salinity, mean Ho, and mean AR on percent length at day 78 (end of nursery stage)
length78_mr <- lm(mean_length_mm_78 ~ Mean_Annual_Temperature_C + Mean_Annual_Salinity_ppt + avg_observed_ho + avg_allelic_richness, data = surv_length_envr_gen)
summary(length78_mr) 
#model p-value = 0.5623
#model F-stat = 0.8865
#adjusted R squared = -0.06934
length78_tidy <- tidy(length78_mr) #save coefficients as table
```

#adjust p-values
```{r}
model_p_values <- c(0.06075, 0.89, 0.5623)

model_p_adj <- p.adjust(model_p_values, method = "BH")

model_p_adj
```

Adjusted model p-values
Survival at day 21: 0.18225 
Length at day 21: 0.89000 
Length at day 78: 0.84345


