---
title: "Q2_models_hatchery_nursery"
author: "Nicole Mongillo"
output: pdf_document
date: "2025-02-20"
---

This file is for modeling the effects of heterozygosity, allelic richness, and temperature and salinity quantiles of environments-of-origin on oyster survival and length in the VIMS ABC hatchery and nursery.

CR: Updated 2025-11-16 with new envr quantiles file.
KEB: Updated 2026-08-22 with heterozygosity and allelic richness data that was calculated correctly on recoded data.

## Load packages
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# set working directory relative to project root
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("tidyverse", "here", "ggplot2", "stringr", "broom", "ggstance")
pacman::p_load(char = packages)

# log versions for reproducibility
r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

## Load data
```{r}
# this file contains raw lengths of all oysters aged 15-78 days
length <- read.csv("data/performance_H2F/nursery_length.csv")

# this file has 0.1 and 0.9 quantiles for temperature and salinity for all groups including field sites
quantiles <- read.csv("results/environment/envr_quantiles_allsites.csv")

# this file contains survival rate data for larval oysters up to day 21 post hatching
survival <-  read.csv("data/performance_H2F/CViMVP_larvae_survival.csv")

# KEB: 260822 - now calculated correctly with recoded data
parent_genetics <- read.csv("results/parent_effects_H2F/parent_genetic_effects_full.csv")

```

## Data preparation

### Survival

For survival rates, select only survival from day 21 as this is the "final" hatchery survival (marked as day_an 15-21). Reminder that days 15-21 were grouped for analysis because there were constant larval drops occurring during that time as individuals reached the eyed stage, but the number of drops varied by population. Exclude LARMIX, since it has no environmental data. Exclude groups labelled "small" and only keep measurements from eyed larvae. 

```{r}

surv21_sub <- survival %>%
  # keep the 15-21 day window, drop LARMIX
  filter(
    Day_analysis == "15-21",
    Tank_naming != "MVP-LARMIX",
    # drop "small" groups
    !str_detect(Group_Day_Label, regex("small", ignore_case = TRUE))
  )

survival_filter <- surv21_sub %>%
  # rename each population: W1-W6 (wild) or S1-S2 (selection line)
  # NOTE: JR is the VA site, so it maps straight to the wild VA label here
  mutate(site_name = case_match(
    Tank_naming,
    "MVP-TX" ~ "W1-TX",
    "MVP-LA" ~ "W2-LA",
    "MVP-FL" ~ "W3-FL",
    "MVP-JR" ~ "W4-VA",
    "MVP-NH" ~ "W5-NH",
    "MVP-ME" ~ "W6-ME",
    "MVP-LOLA" ~ "S1-LOLA",
    "MVP-DEBY" ~ "S2-DEBY",
    .default = Tank_naming
  )) %>%
  # keep only site name + survival, and rename the survival column
  select(site_name, percent_survival = Survival_rate_perc)

```

### Length

Split into end of hatchery stage (day 21) and end of nursery stage (78).

```{r}

length_filter <- length %>%
  select(site_name, day_an, length = shell_length_mm) %>%
  filter(site_name != "H1-HYBRIDMIX") %>%
  mutate(site_name = fct_relevel(
    site_name,
    "W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME"
  ))

# split out day-21 and day-78 length
length_21 <- length_filter %>% filter(day_an == "15-21")
length_78 <- length_filter %>% filter(day_an == "78")

```

### Environment

Make separate environmental data frames for everything except field sites/only field sites.

```{r}

# rename all columns
colnames(quantiles) <- c("site_name", "sal10", "sal90", "sal_sd", "temp10", "temp90", "temp_sd", "salinity")

# split into everything except YRK/LEW vs. only YRK/LEW
quant_eoo   <- quantiles %>% filter(!site_name %in% c("YRK", "LEW"))
quant_field <- quantiles %>% filter(site_name %in% c("YRK", "LEW"))

```

### Genetic diversity

```{r}

# rename columns 
colnames(parent_genetics) <- c("site_name", "ho", "avg_expected_hs", "ar")

```

### Join datasets

Combine environmental, genetic, survival, and shell-length data into per-analysis data frames, keyed on site_name, for the downstream models.

```{r}

# environment + genetics, then add survival
envr_gen <- left_join(quant_eoo, parent_genetics, by = "site_name")
surv_merge <- left_join(survival_filter, envr_gen, by = "site_name")

# add length data and split by day
len_envr_gen <- left_join(length_filter, envr_gen, by = "site_name")

len21_envr_gen <- len_envr_gen %>%
  filter(day_an == "15-21") %>%
  mutate(site_name = as.character(site_name))
len78_envr_gen <- len_envr_gen %>%
  filter(day_an == "78") %>%
  mutate(site_name = as.character(site_name))

# sanity check: row counts should match the pre-join length frames (619 / 971)
stopifnot(
  nrow(len21_envr_gen) == nrow(length_21),
  nrow(len78_envr_gen) == nrow(length_78)
)

nrow(len21_envr_gen)
nrow(len78_envr_gen)

```

### Multicollinearity

With heterozygosity/allelic richness calculated on the recoded data, nothing is collinear. When you keep all the predictors in, VIF is super inflated.

```{r}

par(mfrow = c(1,1))
cor_matrix <- cor(envr_gen[,c("temp10", "temp90", "sal10", "sal90", "ho", "ar")])

# for supplement, rename variables for a cleaner graph 
colnames(cor_matrix) <- c("Temp 0.1 quantile", "Temp 0.9 quantile", "Salinity 0.1 quantile", "Salinity 0.9 quantile", "Heterozygosity", "Allelic Richness")

rownames(cor_matrix) <- c("Temp 0.1 quantile", "Temp 0.9 quantile", "Salinity 0.1 quantile", "Salinity 0.9 quantile", "Heterozygosity", "Allelic Richness")

#calculate variance inflation factor (VIF)
vif_vars <- c("temp10", "temp90", "sal10", "sal90", "ho", "ar")

vif_all <- sapply(vif_vars, function(v) {
  others <- setdiff(vif_vars, v)
  1 / (1 - summary(lm(reformulate(others, response = v), data = envr_gen))$r.squared)
})
round(vif_all, 2)

```

KEB 260822: ***Not updated from here*** We decided to remove the hatchery/nursery models from the paper because the data is pseudoreplicated as we can't get a random effect.

## Multiple regression and model selection

### Day 21 length
```{r}
len21_lm <- lm(length ~ temp10 + sal10+ sal90 + ho + ar, data = len21_envr_gen)
len21_resid <- len21_lm$residuals
shapiro.test(len21_resid) #p = 1.8e-8, data are not normal, 
#plot(len21_lm) #no outliers, homoscedasticity looks good
summary(len21_lm) #everything is significant. Model stats: Adj. R squared = 0.2354, F = 39.05, p < 2.2e-16

#there's one exceptionally low length from the TX group that is throwing off normality, I'm going to exclude it and try the models
len21_envr_gen_sub <- len21_envr_gen %>% 
  filter(length > 0.22)

#I am using the data excluding the one very low length measurement.
len21_lm2 <- lm(length ~ temp10 + sal10+ sal90 + ho + ar, data = len21_envr_gen_sub)
len21_resid2 <- len21_lm2$residuals
shapiro.test(len21_resid2) #p = 8.64e-5, data are not normal, but improved, multiple regressions are pretty robust to non-normality
#plot(len21_lm2) #no outliers, homoscedasticity looks good
summary(len21_lm2) #everything is significant. Model stats: Adj. R squared = 0.22, F = 45.49 on 4 and 613 DF, p < 2.2e-16

#model selection
drop1(len21_lm2) #this is the model with the lowest AIC (-5137.9)

len21_lm2_scaled <- lm(scale(length) ~ scale(temp10) + scale(sal10) + scale(sal90) + scale(ho) + scale(ar), data = len21_envr_gen_sub)
summary(len21_lm2_scaled)
```

###Day 78 length

We call this day 78 here but these measurements are more likely from day 77 (from SK). 

```{r day78_lm}
#using all data for day 78
#day 78 multiple regression
len78_lm <- lm(length ~ temp10 + sal10+ sal90 + ho + ar, data = len78_envr_gen)
len78_resid <- len78_lm$residuals
shapiro.test(len78_resid) #p = 0.00092, data not normal, but not bad
#plot(len78_lm) #no outliers, homoscedasticty is good
summary(len78_lm) #all significant except sal90
#model stats: Adj. R squared = 0.266, F = 71.22 on 5 and 965 df, p < 2.2e-16

#model selection
drop1(len78_lm) #drop sal90

len78_lm2 <- lm(length ~ temp10 + sal10+ ho + ar, data = len78_envr_gen)
len78_resid2 <- len78_lm2$residuals
shapiro.test(len78_resid2) #p = 0.00098, data not normal, but not bad
#plot(len78_lm2) #no outliers, homoscedasticty is good
summary(len78_lm2) #all significant
#model stats: Adj. R squared = 0.266, F = 88.86 on 4 and 966 df, p < 2.2e-16

len78_lm_scaled <- lm(scale(length) ~ scale(temp10) + scale(sal10)+  scale(ho) + scale(ar), data = len78_envr_gen)
summary(len78_lm_scaled)
```

###Day 21 survival
I don't think this model is worth running, too few data points, many outliers, not homoscedastic, but I'll leave it here for now for reference.
```{r day21_surv_lm}
#day 21 survival multiple regression
surv_lm <- lm(percent_survival ~ temp10 + sal10+ sal90 + ho + ar, data = surv_merge)
surv_resid <- surv_lm$residuals
shapiro.test(surv_resid) #p = 0.4449, normal
#plot(surv_lm) # 3 outliers
summary(surv_lm) #nothing signif

#model selection
drop1(surv_lm) #drop sal90

#rerun model without ho
surv_lm2 <- lm(percent_survival ~ temp10 + sal10+ ho + ar, data = surv_merge)
summary(surv_lm2) #temp10, sal10 signif
#model stats: adj. R squared = 0.8513, F = 11.02 on 4 and 3 DF, p = 0.039
drop1(surv_lm2) #dropping nothing at this point results in the lowest AIC, stop model selection now.

```

```{r}

# save models
sink(paste0("results/parent_effects_H2F/models/hatchery_nursery_model_results_", Sys.Date(), ".txt"))
cat("\nDay 21 length (excluding outlier)\n")
summary(len21_lm2)
cat("\nDay 21 length scaled\n")
summary(len21_lm2_scaled)
cat("\nDay 77 length (final model)\n")
summary(len78_lm2)
cat("\nDay 77 length scaled\n")
summary(len78_lm_scaled)
cat("\nDay 21 survival (final model)\n")
summary(surv_lm2)
sink()

```

## Plot of slopes
```{r}
#save slopes from each model to one data frame
slopes <- rbind(tidy(len21_lm2_scaled), tidy(len78_lm_scaled))
#add column for time point
slopes$t <- c(rep(c("15-21"), 6), rep(c("78"), 5))

#add dummy row for time point lacking variable so that position_dodge v plots each data point in the same time point at consistent vertical positions for each y axis variable
slopes[nrow(slopes) + 1,] = list("scale(sal90)", -.1, .01, NA, NA, "78")

#rename columns
colnames(slopes) <- c("explanatory", "slope", "se", "Fstat", "pval", "t")

#remove rows with Intercept slopes
slopes_df <- slopes %>% 
  filter(explanatory != "(Intercept)")

slopes <- slopes_df %>% 
  ggplot(aes(x = slope, y = explanatory))+
  geom_point(aes(alpha = t, shape = t), color = "black", size = 3.5, position = position_dodge2v(height = .75, reverse = T, preserve = "total"))+
  geom_errorbarh(aes(xmax = slope + 2*se, xmin = slope - 2*se, alpha= t), color = "black", height = 0, size = 1, position = position_dodgev(height = -.75, preserve = "total"))+
  scale_alpha_manual(name = "Time", na.translate = F, values = c("15-21" = .5, "78" = 1), labels = c("15-21" = "End hatchery", "78" = "End nursery"))+
  scale_shape_manual(name = "Time", na.translate = F, values = c("15-21" = 15, "78" = 16), labels = c("15-21" = "End hatchery", "78" = "End nursery"))+
  xlim(c(-1,2))+
  ggtitle(str_wrap(c("Slopes of survival vs. environmental quantiles, heterozygosity, and allelic richness"), width = 50))+
  labs(subtitle = "End of larval stage (days 15-21) and nursery stage (day 77)")+
  xlab("Slope on scaled data")+
  ylab("Variable")+
  geom_vline(xintercept = 0, linetype = "dashed")+
  scale_y_discrete(name = "Explanatory variable", 
                   limits = c("scale(sal10)", "scale(sal90)", "scale(temp10)", "scale(ho)", "scale(ar)"), 
                   labels = c("Salinity Q10", "Salinity Q90", "Temperature Q10",  str_wrap(c("Heterozygosity/ Temperature Q90"), width = 17), str_wrap(c("Allelic richness/ Temperature Q90"), width = 20)))+
  theme_classic()+
    theme(axis.text = element_text(color = "black", size = 12),
        axis.title = element_text(size = 13),
        legend.text = element_text(size = 12),
        legend.title = element_text(size = 13),
        plot.title = element_text(size = 14),
        plot.subtitle = element_text(size = 13))

slopes

ggsave("FigS6_hatchery_nursery_surv_quantiles.pdf",
       plot = slopes,
       device = "pdf",
       path = "figures/parent_effects_H2F",
       height = 4,
       width = 7)

```
