---
title: "exp_structure_analysis"
author: "Nicole Mongillo"
date: "2025-05-23"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Loading packages
```{r}
library(ggplot2)
library(LEA)
library(scatterpie)
library(sf)
library(rnaturalearthdata)
library(rnaturalearth)
library(dplyr)
library(stringr)
```

### Set region colors
```{r}
setwd("~/Desktop/GitHub/MVP-H2F-HatcheryField/src/parental_genetics")

# can update region colors to whatever we want to see here
region_colors <- c(
  "W1-TX" = "#3e4989",
  "W2-LA" = "#31688e", 
  "W3-FL" = "#26828e",
  "W4-VA" = "#1f9e89",
  "S1-LOLA" = "#440154", 
  "S2-DEBY" = "#482878",
  "W5-NH" = "#35b779",
  "W6-ME" = "#6ece58"
  )
```

```{r}
pca_data <- read.csv("../../data/parental_genetics_structure/20250203_pca_data_exp.csv")

#rename values in ID_SiteDate to match naming scheme
pca_data <- pca_data %>%
  mutate(ID_SiteDate = case_match(ID_SiteDate,
                                   "TX-CapBay" ~ "W1-TX",
                                   "LA-SisLake" ~ "W2-LA",
                                   "FL-KingPlan" ~ "W3-FL",
                                   "VA-DeepWatSh" ~ "W4-VA",
                                   "LOLA" ~ "S1-LOLA",
                                   "DEBY" ~ "S2-DEBY",
                                   "NH-GrtBay" ~ "W5-NH",
                                   "ME-HogIs" ~ "W6-ME"
                                   ))

# plot PCA by plate to check for batch effects - just see pl6 and pl7 here
pca_plot <- ggplot(pca_data, aes(x = PC1, y = PC2, color = Plate)) +
  geom_point(size = 1) + 
  labs(title = "PCA - Plate Extractions, exp Dataset", x = "PC1", y = "PC2", color = "Plate") +
  theme_minimal()

pca_plot

# PCs 3 and 4 checking for batch effects
pca_plot2 <- ggplot(pca_data, aes(x = PC3, y = PC4, color = Plate)) +
  geom_point(size = 1) + 
  labs(title = "PCA - Plate Extractions, Whole Dataset", x = "PC3", y = "PC4", color = "Plate") +
  theme_minimal()

pca_plot2

# PCA plot PCs 1 and 2
centroids <- pca_data %>%
  group_by(ID_SiteDate) %>%
  summarize(PC1 = mean(PC1), PC2 = mean(PC2))

pca_plot3 <- ggplot(pca_data, aes(x = PC1, y = PC2, color = ID_SiteDate)) +
  geom_point(size = 1) +
  stat_ellipse(aes(group = ID_SiteDate), level = 0.95) +
  #geom_text_repel(data = centroids, aes(x = PC1, y = PC2, label = ID_SiteDate), 
  # size = 3, fontface = "bold", box.padding = 0.5, point.padding = 0.5) +
  labs(title = "Parental Groups Genetic PCA", x = "PC1", y = "PC2") +
  scale_color_manual(values = region_colors) +
  theme_minimal()+
  theme(legend.position = "none")

pca_plot3

ggsave("parental_pca.pdf",
       plot = pca_plot3,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 4,
       width = 5)
```

```{r}
# now PCs 3 and 4
centroids34 <- pca_data %>%
  group_by(ID_SiteDate) %>%
  summarize(PC3 = mean(PC3), PC4 = mean(PC4))

pca_plot4 <- ggplot(pca_data, aes(x = PC3, y = PC4, color = ID_SiteDate)) +
  geom_point(size = 1) +
  stat_ellipse(aes(group = ID_SiteDate), level = 0.95) +
  #geom_text_repel(data = centroids34, aes(x = PC3, y = PC4, label = ID_SiteDate), 
                  #size = 3, fontface = "bold", box.padding = 0.5, point.padding = 0.5) +
  labs(title = "PCA Plot exp Samples", x = "PC3", y = "PC4", color = "Region") +
  theme_minimal()

pca_plot4
```


# snmf ancestry analysis

# subset the last geno matrix that we made (GEN_filtered) with our thinned SNPs list from the auto_SVD function
gen_filtered_subset <- GEN_imputed[, thinned_snps]

# now save our matrix as a .geno format to run the snmf analysis
write.geno(gen_filtered_subset, "/Users/madelineeppley/Desktop/cvi-diploid-files/20250204geno_imputedfile_exp.geno")


```{r}
# now generate the snmf project for our plate
snmf_exp <- snmf("../../data/parental_genetics_structure/20250204geno_imputedfile_exp.geno", K = 2, repetitions = 10, ploidy = 2, entropy = TRUE, project = "new")

# plot cross-entropy criterion of all runs of the project
plot(snmf_exp, cex = 1.2, col = "#1565C0", pch = 19)

# get the cross-entropy of the 10 runs for k=2
ce = cross.entropy(snmf_exp, K = 2)

# select the run with the lowest cross-entropy for k=2
best = which.min(ce)

# display the q-matrix

my.colors <- c("#81C784", "#1565C0")

# plot the ancestry proportions
bp <- barchart(snmf_exp, K = 2, run = best, 
         border = NA, space = 0, col = my.colors, 
         xlab = "Individuals", ylab = "Ancestry proportions", 
         main = "Ancestry matrix for Experimental Samples",
         sort.by.Q = T)

#read in RDS file
samp_full_subset <- readRDS("../../data/parental_genetics_structure/20250203_samp_full_subset_exp.rds")
samp_full_subset <- samp_full_subset %>%
  mutate(ID_SiteDate = case_match(ID_SiteDate,
                                   "TX-CapBay" ~ "W1-TX",
                                   "LA-SisLake" ~ "W2-LA",
                                   "FL-KingPlan" ~ "W3-FL",
                                   "VA-DeepWatSh" ~ "W4-VA",
                                   "LOLA" ~ "S1-LOLA",
                                   "DEBY" ~ "S2-DEBY",
                                   "NH-GrtBay" ~ "W5-NH",
                                   "ME-HogIs" ~ "W6-ME"
                                   ))
#extract region labels
region_labels <- samp_full_subset$ID_SiteDate

reordered_labels <- region_labels[bp$order]

axis(1, at = 1:length(region_labels), 
     labels = reordered_labels, las = 3, cex.axis = .4)
```

### Create snmf plot results on map 
```{r}
# read in site data
site_data <- read.csv("../../data/performance_H2F/MVP23-FieldBags - spawn_trt2.csv")

site_data <- site_data %>%
  mutate(SpawnTrt_Label = case_match(SpawnTrt_Label,
                                   "TX-CapBay" ~ "W1-TX",
                                   "LA-SisLake" ~ "W2-LA",
                                   "FL-KingPlan" ~ "W3-FL",
                                   "VA-DeepWatSh" ~ "W4-VA",
                                   "LOLA" ~ "S1-LOLA",
                                   "DEBY" ~ "S2-DEBY",
                                   "NH-GrtBay" ~ "W5-NH",
                                   "ME-HogIs" ~ "W6-ME"
                                   ))

# get the Q-matrix from snmf results
geno <- read.geno("../../data/parental_genetics_structure/20250204geno_imputedfile_exp.geno")
snmf_project <- load.snmfProject("../../data/parental_genetics_structure/20250204geno_imputedfile_exp.snmfProject")
best_run <- which.min(cross.entropy(snmf_project, K = 2))  # choose the best run
Q_matrix <- Q(snmf_project, K = 2, run = best_run)  # extract best run into q-matrix

# pull the site and region labels and merge them with the ancestry proportions
region_labels <- samp_full_subset$ID_SiteDate
region_labels <- as.data.frame(region_labels)
colnames(region_labels) <- "ID_SiteDate"
region_labels <- samp_full_subset[, c("ID_SiteDate", "merge_ID_inds")]
region_labels$Q1 <- Q_matrix[, 1]  # q matrix ancestry group 1
region_labels$Q2 <- Q_matrix[, 2]  # q matrix ancestry group 2

colnames(site_data)[2] = "ID_SiteDate"

# Merge ancestry proportions into the full data
merged_data <- merge(site_data, samp_full_subset, by = "ID_SiteDate")
merged_data <- merge(region_labels, merged_data, by = "merge_ID_inds", all.x = TRUE)
```

### Make base map
```{r}
# load US coastline 
coastline <- ne_coastline(scale = "medium", returnclass = "sf")

# now plot on the map - create a base map with our coastline as the outline of the plot
base_map <- ggplot(data = coastline) +
  geom_sf() +
  theme_minimal() +
  labs(
    title = "Ancestry Proportions by Parental Group",
    x = "Longitude", y = "Latitude"
  ) +
  coord_sf(xlim = c(-100, -60), ylim = c(24, 50), expand = FALSE)

base_map
```

```{r}
# now make site averages of our ancestry proportions 
site_data_averaged <- merged_data %>%
  group_by(ID_SiteDate.x) %>%
  summarise(Q1_avg = mean(Q1), Q2_avg = mean(Q2),
            Longitude = first(longitudeDecimal),
            Latitude = first(latitudeDecimal))   

# we need to add some arbitrary coordinates for the VA populations or they will all plot on top of each other. let's put them in order from south to north, Deep Water Shoal, DEBY (York Riv), LOLA (Lew Riv)
site_data_averaged[site_data_averaged$ID_SiteDate.x == "W4-VA", "Latitude"] <- 35.61813
site_data_averaged[site_data_averaged$ID_SiteDate.x == "W4-VA", "Longitude"] <- -74.29945
site_data_averaged[site_data_averaged$ID_SiteDate.x == "S2-DEBY", "Latitude"] <- 37.5
site_data_averaged[site_data_averaged$ID_SiteDate.x == "S2-DEBY", "Longitude"] <- -74.95174
site_data_averaged[site_data_averaged$ID_SiteDate.x == "S1-LOLA", "Latitude"] <- 38.8223
site_data_averaged[site_data_averaged$ID_SiteDate.x == "S1-LOLA", "Longitude"] <- -73.19229

map_with_sites <- base_map + 
  geom_scatterpie(
    data = site_data_averaged,
    aes(x = Longitude, y = Latitude, r = 0.9),  
    cols = c("Q1_avg", "Q2_avg"), 
    color = NA,
    pie_scale = 1.5) +  
  scale_fill_manual(name = "Genetic Cluster", breaks = c("Q1_avg", "Q2_avg"), labels = c("Gulf Cluster", "Atlantic Cluster"), values = c("#1565C0", "#81C784")) +  
  theme(
    legend.position = "inside",
    legend.position.inside = c(.3,.7), 
    legend.text = element_text(size = 10),
    panel.grid = element_blank(), 
    panel.background = element_blank(), 
    axis.line = element_line(color = "black"),
    axis.text = element_text(color = "black")
  )

map_with_sites

ggsave("ancestry_map.pdf",
       plot = map_with_sites,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 4.2,
       width = 6)
```

### Common garden sites inset map
```{r}
# load US coastline 
coastline <- ne_coastline(scale = "medium", returnclass = "sf")

# now plot on the map - create a base map with Chesapeake Bay coastline as the outline of the plot
inset_base <- ggplot(data = coastline) +
  geom_sf() +
  theme_minimal() +
  coord_sf(xlim = c(-78, -74), ylim = c(36.5, 39.5), expand = FALSE)+
  theme(panel.grid = element_blank(),
        axis.title = element_blank(),
        axis.text = element_blank())
inset_base


#Create list of coordinates for Lewisetta and York common garden sites
cg_site <- c("LEW", "YRK")
Latitude <- c(37.98030, 37.249107)
Longitude <- c(-76.46190, -76.497032)

common_garden <- as.data.frame(cbind(cg_site, Latitude, Longitude))

common_garden <- common_garden %>%
  st_as_sf(coords = c('Longitude', 'Latitude'))

# set crs to 4326 for lat/lon
common_garden$geometry <- st_set_crs(common_garden$geometry, 4326)

inset_map <- inset_base +
  geom_point(data = common_garden, aes(x = Longitude, y = Latitude, shape = cg_site, fill = cg_site), size = 4)+
  scale_fill_manual(name = " ", breaks = c("LEW", "YRK"), 
                    values = c("#F06292", "#880E4F"),
                    labels = c("Lewisetta (LEW): low salinity and disease", "York (YRK): higher salinity and disease"))+
  scale_shape_manual(name = " ", breaks = c("LEW", "YRK"), 
                    values = c(23, 24),
                    labels = c("Lewisetta (LEW): low salinity and disease", "York (YRK): higher salinity and disease"))+
  ggtitle("Common Garden Sites")+
  scale_x_continuous(breaks = seq(-78, -74, by =1))+
  scale_y_continuous(breaks = seq(36.5, 39.5, by = 1))+
  theme(legend.position = "right",
        legend.direction = "vertical",
        legend.text = element_text(size = 11),
        plot.title = element_text(size = 11))

inset_map

ggsave("inset_map.pdf",
       plot = inset_map,
       device = "pdf",
       path = "../../figures/envr_of_origin",
       height = 3,
       width = 4)
```

