---
title: "Fstats_experimental.Rmd"
author: "Kiran Bajaj"
output: html_document
date: "2025-02-04"
---

This file is for calculating pairwise Fst, allelic richness, observed heterozygosity, and running OutFLANK to test for Fst outliers. The analysis uses SNP chip data from the experimental PARENTS for the H2F paper. MGE generated SNP sets and genotype matrices. 

## Load packages

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Set working directory relative to project root
knitr::opts_knit$set(root.dir = here::here())

if (!require("pacman")) install.packages("pacman")

packages <- c("here", "ggplot2", "hierfstat", "reshape2", "dplyr", "tidyr", "readr", "adegenet", "viridis", "OutFLANK", "purrr", "tibble")
pacman::p_load(char = packages)

# log versions for reproducibility
r_version <- paste(R.version$major, R.version$minor, sep = ".")
lib_ver <- packages %>%
  map_chr(~ paste0(.x, " (v", packageVersion(.x), ")")) %>%
  paste(collapse = ", ")
```

```{r}
getwd()
```

## Load data

```{r}

# full SNP set
full_snps <- readRDS(file.path("data", "parental_genetics_structure", "20250606_FULLSNPs_exp_labeled.rds"))

# SNP set thinned for LD
thinned_snps <- readRDS(file.path("data", "parental_genetics_structure", "20250606_THINNEDSNPs_exp_labeled.rds"))

# SNP info
muts_maf_filtered <- readRDS(file.path("data", "parental_genetics_structure", "20250604_FULLmuts_exp.rds"))

# sample info
samp_full_subset <- readRDS(file.path("data", "parental_genetics_structure", "20250604_samp_full_subset_exp.rds"))

```

## Data product (BCODMO): full SNPs genotypes in csv format

```{r}

write.csv(full_snps, file.path("data", "parental_genetics_structure", "20250606_FULLSNPs_exp_labeled.csv"), row.names = FALSE)

```

```{r}

# rename datasets for processing
full <- full_snps
gen <- thinned_snps
muts <- muts_maf_filtered

# check dimensions
dim(gen) # thinned - 160 individuals and 105356 SNPs
dim(muts) # full mutation-level info - 154102 SNPs

```

## Generate supplementary SNP files

We want one file that lists the SNPs and another file with the SNP ID and imputed matrix of genotypes, ideally as a vcf file. Let's add a column to the full SNP set that lists the SNPs that indicates whether or not the SNP is in the thinned SNP set.

```{r}

# file 1: SNP info with column indicating if the SNP is in the thinned set
thinned_snp_ids <- colnames(gen)
muts$in_thinned_set <- ifelse(muts$Affx.ID %in% thinned_snp_ids, "T", "F")
table(muts$in_thinned_set)
write.csv(muts, "supp_files/1-supplementary_snp_info.csv", row.names = FALSE)

# file 2: VCF of full genotype matrix
snp_ids <- colnames(full)
snp_info <- muts[match(snp_ids, muts$Affx.ID), ]

# extract REF and ALT from Sequence column (e.g., "...[C/T]...")
snp_info$REF <- gsub(".*\\[([A-Z])/[A-Z]\\].*", "\\1", snp_info$Sequence)
snp_info$ALT <- gsub(".*\\[[A-Z]/([A-Z])\\].*", "\\1", snp_info$Sequence)

# convert 0/1/2 -> GT format
gt_matrix <- apply(full, c(1, 2), function(x) {
  switch(as.character(x), "0" = "0/0", "1" = "0/1", "2" = "1/1", "./.")
})

# build VCF
vcf_df <- data.frame(
  `#CHROM` = snp_info$Chromosome,
  POS      = snp_info$Position,
  ID       = snp_ids,
  REF      = snp_info$REF,
  ALT      = snp_info$ALT,
  QUAL     = ".",
  FILTER   = "PASS",
  INFO     = ".",
  FORMAT   = "GT",
  check.names = FALSE
)

# append genotype columns
gt_df <- as.data.frame(t(gt_matrix))
colnames(gt_df) <- rownames(full)
vcf_df <- cbind(vcf_df, gt_df)

# write VCF
vcf_file <- "supp_files/2-supplementary_genotypes.vcf"
writeLines(c(
  "##fileformat=VCFv4.2",
  '##FORMAT=<ID=GT,Number=1,Type=String,Description="Genotype">',
  '##INFO=<ID=.,Number=.,Type=.,Description="No additional info">',
  '##source=Affymetrix_OysterCVI200K_array'
), vcf_file)
write.table(vcf_df, vcf_file, append = TRUE, sep = "\t",
            row.names = FALSE, quote = FALSE)

```

## Shared genotype conversion function

This function converts a 0/1/2 genotype matrix to a genind object via string genotypes ("11"/"12"/"22"). This avoids hierfstat misreading raw 0/1/2 as triallelic. Defined once here and used for both Fst (thinned) and basic stats (full).

```{r}

genotypes_to_genind <- function(numeric_genotypes, pop_labels) {
  geno_numeric <- as.matrix(numeric_genotypes)
  
  geno_strings <- matrix(NA_character_, 
                         nrow = nrow(geno_numeric), 
                         ncol = ncol(geno_numeric))
  geno_strings[geno_numeric == 0] <- "11"
  geno_strings[geno_numeric == 1] <- "12"
  geno_strings[geno_numeric == 2] <- "22"
  
  geno_strings <- as.data.frame(geno_strings)
  colnames(geno_strings) <- colnames(geno_numeric)
  
  df2genind(geno_strings, pop = pop_labels, ploidy = 2, ncode = 1)
}

```

## Shared population recoding

Used for both Fst and basic stats sections to keep labels consistent.

```{r}

recode_pop <- function(labels) {
  dplyr::case_match(labels,
    "TX-CapBay"    ~ "W1-TX",
    "LA-SisLake"   ~ "W2-LA",
    "FL-KingPlan"  ~ "W3-FL",
    "VA-DeepWatSh" ~ "W4-VA",
    "LOLA"         ~ "S1-LOLA",
    "DEBY"         ~ "S2-DEBY",
    "NH-GrtBay"    ~ "W5-NH",
    "ME-HogIs"     ~ "W6-ME"
  )
}

population_order <- c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME")

```

## Calculating Fst 

### Data preparation

```{r}

# find SNPs that are NOT all heterozygous (all 1s)
keep_snps_fst <- apply(gen, 2, function(col) !all(col == 1))

# filter genotype matrix for these SNPs
GEN_filtered_fst <- gen[, keep_snps_fst]
dim(GEN_filtered_fst) # 0 columns removed

```

```{r}

# match sample order between genotype matrix and metadata
samp_ordered <- samp_full_subset[match(rownames(GEN_filtered_fst), samp_full_subset$clean_ID), ]

# verify alignment
identical(rownames(GEN_filtered_fst), as.character(samp_ordered$clean_ID)) # should be TRUE

# build population labels for Fst
fst_pop_labels <- recode_pop(samp_ordered$ID_SiteDate)
table(fst_pop_labels) # check sample counts per population

```

```{r}

# convert thinned genotype matrix to genind -> hierfstat
genind_fst <- genotypes_to_genind(GEN_filtered_fst, pop_labels = fst_pop_labels)

# verify biallelic: should be 2 columns per locus
ncol(genind_fst@tab) / ncol(GEN_filtered_fst) # should be 2

hierfstat_fst <- genind2hierfstat(genind_fst)
dim(hierfstat_fst) # col 1 is pop, rest are loci

```

### Data product: Fst matrix 

```{r}

# run Fst calculation - cluster only
#fst_results <- pairwise.WCfst(hierfstat_fst, diploid = TRUE)
#fst_results[is.na(fst_results)] <- 0

# view results
#print(fst_results)

#saveRDS(fst_results, file.path("results", "parental_genetics", "fst_results.rds"))

# load in fst results - after running with cluster
fst_results <- readRDS(file.path("results", "parental_genetics", "20250606_fst_results.rds"))


# convert the matrix to a long format and edit column names
fst_plotting <- melt(fst_results, na.rm = TRUE)
colnames(fst_plotting) <- c("Var1", "Var2", "value")

# reorder the Var1 and Var2 columns based on the defined order
fst_plotting$Var1 <- factor(fst_plotting$Var1, levels = population_order)
fst_plotting$Var2 <- factor(fst_plotting$Var2, levels = population_order)

# view data one more time
fst_plotting

# create heatmap using viridis color palette
fig_fst_heatmap <- ggplot(fst_plotting, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") + 
  labs(x = "Population", y = "Population", fill = "Fst") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

fig_fst_heatmap

```

## Basic stats with hierfstat

### Data preparation

Same steps as before but with full SNP set. 

```{r}

# prepare the full SNP set
full <- full_snps

dim(full) # 154102 SNPs

```

```{r}

# filter SNPs that are all heterozygous (all 1s)
keep_snps_bs <- apply(full, 2, function(col) !all(col == 1))

# filter genotype matrix for these SNPs
FULL_filtered <- full[, keep_snps_bs]
dim(FULL_filtered) # 0 columns removed

```

```{r}

# reorder to match metadata
FULL_filtered <- FULL_filtered[match(samp_full_subset$clean_ID, rownames(FULL_filtered)), ]

# verify alignment
identical(rownames(FULL_filtered), as.character(samp_full_subset$clean_ID)) # should be TRUE

# build population labels for basic stats
bs_pop_labels <- recode_pop(samp_full_subset$ID_SiteDate)

# check number of samples per population
table(bs_pop_labels)

# check for any NAs in population labels
sum(is.na(bs_pop_labels)) # should be 0

```

```{r}

# convert full genotype matrix to genind -> hierfstat (uses shared function defined above)
genind_obj <- genotypes_to_genind(FULL_filtered, pop_labels = bs_pop_labels)

# verify biallelic: should be 2 columns per locus
ncol(genind_obj@tab) / ncol(FULL_filtered) # should be 2

```

```{r}

# convert to hierfstat obj
#hierfstat_obj <- genind2hierfstat(genind_obj)

# run basic stats - change here to run on thinned
div <- basic.stats(hierfstat_fst)

# quick sanity check on the new values
ho_pop <- colMeans(div$Ho, na.rm = TRUE)
hs_pop <- colMeans(div$Hs, na.rm = TRUE)

data.frame(
  Population = names(ho_pop),
  Ho = round(ho_pop, 3),
  Hs = round(hs_pop, 3),
  FIS_implied = round(1 - ho_pop/hs_pop, 3)
)

```

### Observed and expected heterozygosity

```{r}
# extract values from basic.stats output
Ho <- div$Ho # one value per SNP per population
Hs <- div$Hs # one value per SNP per population
```

#### Sanity checks

```{r}
# range checks: are values within expected bounds?
summary(unlist(Ho))
summary(unlist(Hs))

# any out-of-range values?
cat("Ho > 1:", sum(Ho > 1, na.rm = TRUE), "\n")
cat("Ho < 0:", sum(Ho < 0, na.rm = TRUE), "\n")
cat("Hs > 0.5:", sum(Hs > 0.5, na.rm = TRUE), 
    "(small excess OK due to sample-size correction)\n")
```

```{r}
# population assignments: do they match between input and hierfstat object?
table(hierfstat_obj[, 1])
table(bs_pop_labels)
```

```{r}
# population-level summary of Ho, Hs, and implied FIS
ho_pop <- colMeans(div$Ho, na.rm = TRUE)
hs_pop <- colMeans(div$Hs, na.rm = TRUE)

pop_summary <- data.frame(
  Population = names(ho_pop),
  Ho = round(ho_pop, 4),
  Hs = round(hs_pop, 4),
  FIS_implied = round(1 - ho_pop/hs_pop, 4)
)
pop_summary
```

```{r}
# monomorphic loci per population
monomorphic_per_pop <- apply(div$Hs, 2, function(x) sum(x == 0, na.rm = TRUE))
monomorphic_per_pop
```

```{r}
# convert Ho and Hs matrices to long-format data frames
Ho_long <- as.data.frame(Ho) %>%
  tibble::rownames_to_column("Locus") %>%
  pivot_longer(cols = -Locus, names_to = "Population", values_to = "Ho")

Hs_long <- as.data.frame(Hs) %>%
  tibble::rownames_to_column("Locus") %>%
  pivot_longer(cols = -Locus, names_to = "Population", values_to = "Hs")

# merge into one long-format dataframe
pop_data <- merge(Ho_long, Hs_long, by = c("Locus", "Population"))

# set population factor levels for consistent plotting order
pop_data$Population <- factor(pop_data$Population, levels = population_order)

# also reshape for paired Ho/Hs boxplot
pop_long <- pop_data %>%
  pivot_longer(cols = c("Ho", "Hs"), names_to = "Type", values_to = "Value")
```

#### Plot 1: Ho and Hs distributions by population (boxplot)

```{r}
ggplot(pop_long, aes(x = Population, y = Value, fill = Type)) +
  geom_boxplot() +
  labs(title = "Observed and expected heterozygosity by population",
       x = "Population",
       y = "Heterozygosity") +
  scale_fill_manual(values = c("Ho" = "blue", "Hs" = "orange")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}
# curious what the high outlier dots are

# how many SNPs have Ho > 0.75 in each population?
high_ho_counts <- apply(div$Ho, 2, function(x) sum(x > 0.75, na.rm = TRUE))
high_ho_counts

# what fraction of total SNPs is that?
high_ho_counts / nrow(div$Ho)
```

#### Plot 2: Ho distribution by population (boxplot)

```{r}
ggplot(pop_data, aes(x = Population, y = Ho)) +
  geom_boxplot(fill = "lightblue", alpha = 0.7) +
  labs(title = "Observed heterozygosity by population",
       x = "Population",
       y = "Observed heterozygosity (Ho)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

#### Plot 3: Mean Ho and Hs by population (line plot)

```{r}
# calculate population means
avg_het_per_pop <- pop_data %>%
  group_by(Population) %>%
  summarise(
    Ho = mean(Ho, na.rm = TRUE),
    Hs = mean(Hs, na.rm = TRUE)
  ) %>%
  pivot_longer(cols = c(Ho, Hs), names_to = "Type", values_to = "Value")

ggplot(avg_het_per_pop, aes(x = Population, y = Value, group = Type, color = Type)) +
  geom_line(size = 1.2) +
  geom_point(size = 3) +
  labs(title = "Mean Ho and Hs by population",
       x = "Population",
       y = "Mean heterozygosity") +
  scale_color_manual(values = c("Ho" = "blue", "Hs" = "orange")) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

### Allelic richness 

This should be fixed with the new hierfstat object that reads SNPs as biallelic.

```{r}

# calculate allelic richness
ar_results <- allelic.richness(hierfstat_obj, min.n = 40, diploid = TRUE) # min.n = number in smallest population * 2

# extract results
ar_results <- ar_results$Ar # one value per SNP per population

# make column for locus
ar_results <- rownames_to_column(ar_results, var = "Locus")
ar_results <- ar_results[, c("Locus", colnames(ar_results)[-1])] # move to front

# format data for plotting
ar_long <- melt(ar_results, id.vars = "Locus", variable.name = "Population", value.name = "Allelic_Richness")

# calculate means to put on graph
means <- aggregate(Allelic_Richness ~ Population, data = ar_long, mean)

```

#### Plot 1: Mean

```{r}

ggplot(means, aes(x = factor(Population, levels = population_order), 
                  y = Allelic_Richness)) +
  geom_point(size = 4, color = "darkblue") +
  geom_hline(yintercept = 2, linetype = "dashed", color = "grey50") +
  labs(title = "Mean allelic richness by population",
       x = "Population", 
       y = "Mean allelic richness") +
  coord_cartesian(ylim = c(1.5, 2)) +  # zoom in
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

#### Plot 2: Proportion of polymorphic SNPs 

```{r}

poly_props <- ar_long %>%
  group_by(Population) %>%
  summarise(prop_polymorphic = mean(Allelic_Richness > 1.5, na.rm = TRUE))

ggplot(poly_props, aes(x = factor(Population, levels = population_order),
                       y = prop_polymorphic)) +
  geom_col(fill = "lightblue") +
  labs(title = "Proportion of polymorphic SNPs by population",
       x = "Population",
       y = "Proportion polymorphic") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

```

```{r}

# make sure allelic richness is never greater than 2
max(ar_long$Allelic_Richness, na.rm = TRUE)  # should be 2

```


```{r}

# calculate average ar per population
avg_ar_per_pop <- ar_long %>%
  group_by(Population) %>%
  summarise(
    avg_allelic_richness = mean(Allelic_Richness, na.rm = TRUE),
  )

print(avg_ar_per_pop)

```

### Data product: AR, Ho, Hs

This will get used in a linear model that includes temperature, salinity, AR, and Ho. It is saved to the "parent_effects_H2F" folder under "data." This is for the JUVENILE data though, not parents.

```{r}

# reshape avg_het_per_pop back to wide format for merging
avg_het_wide <- avg_het_per_pop %>%
  pivot_wider(names_from = Type, values_from = Value) %>%
  rename(avg_observed_ho = Ho, avg_expected_hs = Hs)

# merge with allelic richness data
parent_genetic_effects <- merge(avg_het_wide, avg_ar_per_pop, by = "Population") %>%
  mutate(Population = factor(Population, levels = population_order)) %>%
  arrange(Population)

write_csv(parent_genetic_effects, file.path("results", "parent_effects_H2F", "parent_genetic_effects_thinned.csv"))

```

## Fst outliers with OutFLANK - NOT RERUN FOR JUVENILES

### Data preparation

```{r}

# rename for processing - using the full filtered matrix and pop labels from basic stats
outflank_data <- as.data.frame(FULL_filtered)
outflank_data$pop <- bs_pop_labels

# move pop column to front
outflank_data <- outflank_data[, c(ncol(outflank_data), 1:(ncol(outflank_data)-1))]

# convert population to factor
outflank_data$pop <- as.factor(outflank_data$pop)

# check dimensions one more time
dim(outflank_data)

```

```{r}

# OutFLANK inputs: snp matrix, locus names, population names
# documentation: https://github.com/whitlock/OutFLANK/blob/master/OutFLANK%20readme.pdf

# make the outflank input data frame
make_outflank_input <- function(df) {
  SNPmat <- as.matrix(df[, -1])         # exclude population column
  SNPmat <- apply(SNPmat, 2, as.numeric)
  SNPmat[is.na(SNPmat)] <- 9            # replace missing with 9
  
  locus_names <- colnames(df)[-1]       # SNP names
  pop_names <- df$pop           # population IDs
  
  FstDataFrame <- MakeDiploidFSTMat(SNPmat, locus_names, pop_names)
  return(FstDataFrame)
}

make_outflank_input(outflank_data)

```

### Data checks

```{r}

# data check: heterozygosity vs. Fst
plot(FstDataFrame$He, FstDataFrame$FST) 

# data check: Fst vs. FstNoCorr
plot(FstDataFrame$FST, FstDataFrame$FSTNoCorr) # uncorrected Fst is always larger

```

### Run OutFLANK

```{r}

# first, run OutFLANK on thinned SNPs
out_trim <- OutFLANK(FstDataFrame[FstDataFrame$LocusName %in% colnames(thinned_snps), ], # using pruned SNPs
                     LeftTrimFraction = 0.05, # defaults
                     RightTrimFraction = 0.05, 
                     Hmin = 0.1, 
                     NumberOfSamples = 8, # number of populations
                     qthreshold = 0.01)

# view results
str(out_trim)
head(out_trim$results)

# check the fit
OutFLANKResultsPlotter(out_trim, withOutliers = TRUE,
                       NoCorr = TRUE, Hmin = 0.1, binwidth = 0.001, Zoom =
                         FALSE, RightZoomFraction = 0.05, titletext = NULL) # looks fine

# zoom in on right tail
OutFLANKResultsPlotter(out_trim , withOutliers = TRUE,
                       NoCorr = TRUE, Hmin = 0.1, binwidth = 0.001, Zoom =
                         TRUE, RightZoomFraction = 0.15, titletext = NULL) # looks fine

# check p value histogram - should look flat (not great)
hist(out_trim$results$pvaluesRightTail)

```

### Calculate p-values

```{r}

# then, calulate p-values on the full set
P1 <- pOutlierFinderChiSqNoCorr(FstDataFrame, Fstbar = out_trim$FSTNoCorrbar, 
                                   dfInferred = out_trim$dfInferred, qthreshold = 0.01, Hmin=0.05)

# view data
head(P1)
tail(P1)

# look for outliers - highlighted in blue
my_out <- P1$OutlierFlag==TRUE
plot(P1$He, P1$FST, pch=19, col=rgb(0,0,0,0.1))
points(P1$He[my_out], P1$FST[my_out], col="blue")

# check the p-value histogram for the full set of data
# if there are outliers, it should look flat with an inflation near 0
hist(P1$pvaluesRightTail)

```

### Data product: OutFLANK results, SNP info

```{r}

# separating by chromosome
a = sprintf("%08d", muts$Position)
muts$chrom_position <- paste(muts$Chromosome, a, sep = ".")

# add chromosome positions to dataframe
P1$Affx.ID <- P1$LocusName
joined_P1 <- left_join(P1, muts, by = c("Affx.ID" = "Affx.ID"))

write_csv(joined_P1, file.path("results", "parental_genetics", "outflank_results.csv"))

```
