---
title: "condition_index_analysis"
author: "Zea Segnitz"
output: html_document
date: "2025-04-10"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load in libraries

```{r}
library(fitdistrplus)
library(dplyr)
library(lme4)
library(lmerTest)
library(pbkrtest)
library(emmeans)
library(multcompView)
library(multcomp)
library(tidyverse)
library(lme4) 
library(effects)
library(ggsignif) 
library(renv)
library(readr)
```

### Set up data 

```{r}
# load in dataset 
MVP_phenotyping_final <- read_csv("../data/condition_index_H2F/MVP_phenotyping_final.csv", show_col_types = FALSE)

# put the populations in correct factor order
MVP_phenotyping_final$population <- factor(MVP_phenotyping_final$population, levels = c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME", "H1-LARMIX","H2-SEEDMIX"))

population_colors <- c(
  "W1-TX"     = "#3e4989", "W2-LA"     = "#31688e", "W3-FL"     = "#26828e", "S1-LOLA"   = "#440154","S2-DEBY"   = "#482878", "W4-VA"     = "#1f9e89",
  "W5-NH"     = "#35b779","W6-ME"     = "#6ece58", "H1-LARMIX" = "#fde725","H2-SEEDMIX"= "#b5de2b")

#subset the timepoints 
data202405 = MVP_phenotyping_final %>% filter(samplingTimePoint == "2024-05")
data202411 = MVP_phenotyping_final %>% filter(samplingTimePoint == "2024-11")

#set the levels 
data202405$population <- factor(data202405$population, levels = c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME", "H1-LARMIX", "H2-SEEDMIX"))

data202411$population <- factor(data202411$population, levels = c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME", "H1-LARMIX", "H2-SEEDMIX"))
```

### Plots
This creates boxplots of the condition index by population, with unique colors for LEW and YR. 

```{r}
#May 2024
ggplot(data202405, aes(x = population, y = CI, fill = site)) +
geom_boxplot(position = position_dodge(width = 0.8)) +  
  scale_x_discrete() +  
  labs(x = "Population", y = "Condition Index", title = "Condition Index by Population (May 2024)", fill = "Site") +   scale_fill_manual(values = c("LEW" = "steelblue", "YR" = "tomato"), labels = c("LEW" = "Lewisetta", "YR" = "York River")) +  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# Nov 2024
ggplot(data202411, aes(x = population, y = CI, fill = site)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +  
  scale_x_discrete() +  
  labs(x = "Population", y = "Condition Index", title = "Condition Index by Population (Nov. 2024)", fill = "Site") +   scale_fill_manual(values = c("LEW" = "steelblue", "YR" = "tomato"), labels = c("LEW" = "Lewisetta", "YR" = "York River")) +  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

create linear mixed-effects models 
```{r}
# create the linear mixed-effects model including site, population, and their interaction
mixed.model.2 <- lmer(CI ~ site + population + (1 | bagID), data = data202405)

mixed.model.3 <- lmer(CI ~ site*population + (1 | bagID), data = data202411)
mixed.model.3.site.pop = lmer(CI ~ site.pop + (1 | bagID), data = data202411) # build this as well to be able to compute apirwise differences between every site.pop combination 
```

post-hoc testing. Use emmeans to find and pairwise compare means of every population. 
```{r}
#pairwise comparisons of every site-population combination ex. LEW.TX etc. - only for Nov 2024 because site is not significant at May 2024
mixed.model.2.emmeans = emmeans(mixed.model.2, pairwise ~ population, adjust = "tukey")
mm3.site.pop.emmeans = emmeans(mixed.model.3.site.pop, pairwise ~ site.pop, adjust = "tukey")

# get just the emmeans output 
mm2.emm = mixed.model.2.emmeans$emmeans
mm3.emm = mm3.site.pop.emmeans$emmeans

# Generate compact letter display
mm2.cld_letters <- cld(mm2.emm, adjust = "tukey", Letters = letters)
mm3.cld_letters <- cld(mm3.emm, adjust = "tukey", Letters = letters)

#make sure the letters are in dataframes
as.data.frame(mm2.cld_letters)
as.data.frame(mm3.cld_letters)

```

### Plot every site-population combination
I used this to visualize the letters but ended up manually overlaying them onto a prettier figure 

```{r}
# Plot mm2 - REMOVE THIS 
ggplot(mm2.cld_letters, aes(x = site.pop, y = emmean)) +
  geom_col(fill = "skyblue", color = "black") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_text(aes(label = .group, y = upper.CL + 0.1), vjust = 0, size = 5) +
  labs(
    title = "Estimated Marginal Means by Site.Pop (May 2024)",
    x = "Site.Pop",
    y = "Estimated Mean"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

#plot mm3
ggplot(mm3.cld_letters, aes(x = site.pop, y = emmean)) +
  geom_col(fill = "skyblue", color = "black") +
  geom_errorbar(aes(ymin = lower.CL, ymax = upper.CL), width = 0.2) +
  geom_text(aes(label = .group, y = upper.CL + 0.1), vjust = 0, size = 5) +
  labs(
    title = "Estimated Marginal Means by Site.Pop (Nov. 2024)",
    x = "Site.Pop",
    y = "Estimated Mean"
  ) +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

model testing with time inclusion
```{r}
mixed.model.4 = lmer(CI ~ site*population*samplingTimePoint + (1 | bagID), data = MVP_phenotyping_final %>% filter(samplingTimePoint == "2024-05" | samplingTimePoint == "2024-11"))
anova(mixed.model.4)
step(mixed.model.4)
### confirms that current model is the best one 
```

### Scatter plots of tissue vs shell weights - May 2024

```{r}
# Lewisetta 
ggplot(data202411 %>% filter(site == "LEW"), aes(x = shell_dryweight_g, y = tissue_dryweight_g,
                       color = population_colors, shape = grouping)) +
  geom_point(size = 3, alpha = 0.7, aes(color = population)) + scale_color_manual(values = population_colors) +
  labs(
    x = "Shell Dry Weight (g)", 
    y = "Tissue Dry Weight (g)", 
    title = "Shell vs. Tissue Dry Weights by Population at Lewisetta (May 2024)",
    color = "Population", 
    shape = "Grouping"
  )  +
  theme_minimal() +
  theme(legend.position = "right", axis.text = element_text(size = 12)) +
  geom_smooth(aes(x = shell_dryweight_g, y = tissue_dryweight_g), 
              method = "lm", se = FALSE, color = "blue", inherit.aes = FALSE)

# York River
ggplot(data202411 %>% filter(site == "YR"), aes(x = shell_dryweight_g, y = tissue_dryweight_g,
                       color = population_colors, shape = grouping)) +
  geom_point(size = 3, alpha = 0.7, aes(color = population)) + scale_color_manual(values = population_colors) +
  labs(
    x = "Shell Dry Weight (g)", 
    y = "Tissue Dry Weight (g)", 
    title = "Shell vs. Tissue Dry Weights by Population at York River (May 2024)",
    color = "Population", 
    shape = "Grouping"
  )  +
  theme_minimal() +
  theme(legend.position = "right", axis.text = element_text(size = 12)) +
  geom_smooth(aes(x = shell_dryweight_g, y = tissue_dryweight_g), 
              method = "lm", se = FALSE, color = "blue", inherit.aes = FALSE)
```

### Scatter plots by group and site 
ex. Wild pops at LEW - LEWISETTA

```{r}
# wild at LEW
ggplot(data202411 %>% filter(site == "LEW")  %>%  filter(grouping == "Wild"), aes(x = shell_dryweight_g, y = tissue_dryweight_g, color = population_colors, shape = grouping)) +
    geom_point(size = 3, alpha = 0.7, aes(color = population)) + scale_color_manual(values = population_colors) +
    labs(x = "Shell Dry Weight (g)",  y = "Tissue Dry Weight (g)",  title = "Shell vs. Tissue Dry Weights by Population (Wild) at Lewisetta (May 2024)",color = "Population", shape = "Grouping")  +
    theme_minimal() +
    theme(legend.position = "right", axis.text = element_text(size = 12)) +
    coord_cartesian(xlim = c(0, 55), ylim = c(0, 2.6)) +
    geom_abline(slope = 0.029636, intercept = 0.047439, color = "blue", linetype = "solid", size = 1) 

# selection at LEW
ggplot(data202411 %>% filter(site == "LEW")  %>%  filter(grouping == "Selection"), aes(x = shell_dryweight_g, y = tissue_dryweight_g, color = population_colors, shape = grouping)) +
    geom_point(size = 3, alpha = 0.7, aes(color = population)) + scale_color_manual(values = population_colors) +
    labs(x = "Shell Dry Weight (g)",  y = "Tissue Dry Weight (g)",  title = "Shell vs. Tissue Dry Weights by Population (Selection) at Lewisetta (May 2024)",color = "Population", shape = "Grouping")  +
    theme_minimal() +
    theme(legend.position = "right", axis.text = element_text(size = 12)) +
    coord_cartesian(xlim = c(0, 55), ylim = c(0, 2.6)) +
    geom_abline(slope = 0.029636, intercept = 0.047439, color = "blue", linetype = "solid", size = 1) 

# mixtures at LEW
ggplot(data202411 %>% filter(site == "LEW")  %>%  filter(grouping == "Mixtures"), aes(x = shell_dryweight_g, y = tissue_dryweight_g, color = population_colors, shape = grouping)) +
    geom_point(size = 3, alpha = 0.7, aes(color = population)) + scale_color_manual(values = population_colors) +
    labs(x = "Shell Dry Weight (g)",  y = "Tissue Dry Weight (g)",  title = "Shell vs. Tissue Dry Weights by Population (Mixtures) at Lewisetta (May 2024)",color = "Population", shape = "Grouping")  +
    theme_minimal() +
    theme(legend.position = "right", axis.text = element_text(size = 12)) +
    coord_cartesian(xlim = c(0, 55), ylim = c(0, 2.6)) +
    geom_abline(slope = 0.029636, intercept = 0.047439, color = "blue", linetype = "solid", size = 1) 


#---------------------------------------------------------------------------------------------------------------------------------------------------------------

# wild at YR
ggplot(data202411 %>% filter(site == "YR")  %>%  filter(grouping == "Wild"), aes(x = shell_dryweight_g, y = tissue_dryweight_g, color = population_colors, shape = grouping)) +
    geom_point(size = 3, alpha = 0.7, aes(color = population)) + scale_color_manual(values = population_colors) +
    labs(x = "Shell Dry Weight (g)",  y = "Tissue Dry Weight (g)",  title = "Shell vs. Tissue Dry Weights by Population (Wild) at York River (May 2024)",color = "Population", shape = "Grouping")  +
    theme_minimal() +
    theme(legend.position = "right", axis.text = element_text(size = 12)) +
    coord_cartesian(xlim = c(0, 55), ylim = c(0, 2.6)) +
    geom_abline(slope = 0.022701, intercept = -0.024714, color = "blue", linetype = "solid", size = 1) 

# selection at YR
ggplot(data202411 %>% filter(site == "YR")  %>%  filter(grouping == "Selection"), aes(x = shell_dryweight_g, y = tissue_dryweight_g, color = population_colors, shape = grouping)) +
    geom_point(size = 3, alpha = 0.7, aes(color = population)) + scale_color_manual(values = population_colors) +
    labs(x = "Shell Dry Weight (g)",  y = "Tissue Dry Weight (g)",  title = "Shell vs. Tissue Dry Weights by Population (Selection) at York River (May 2024)",color = "Population", shape = "Grouping")  +
    theme_minimal() +
    theme(legend.position = "right", axis.text = element_text(size = 12)) +
    coord_cartesian(xlim = c(0, 55), ylim = c(0, 2.6)) +
    geom_abline(slope = 0.022701, intercept = -0.024714, color = "blue", linetype = "solid", size = 1) 

# mixtures at YR
ggplot(data202411 %>% filter(site == "YR")  %>%  filter(grouping == "Mixtures"), aes(x = shell_dryweight_g, y = tissue_dryweight_g, color = population_colors, shape = grouping)) +
    geom_point(size = 3, alpha = 0.7, aes(color = population)) + scale_color_manual(values = population_colors) +
    labs(x = "Shell Dry Weight (g)",  y = "Tissue Dry Weight (g)",  title = "Shell vs. Tissue Dry Weights by Population (Mixtures) at York River (May 2024)",color = "Population", shape = "Grouping")  +
    theme_minimal() +
    theme(legend.position = "right", axis.text = element_text(size = 12)) +
    coord_cartesian(xlim = c(0, 55), ylim = c(0, 2.6)) +
    geom_abline(slope = 0.022701, intercept = -0.024714, color = "blue", linetype = "solid", size = 1) 
```

### Figures of shell and tissue weight across populations 
In May and Nov 2024 (code was not working in the markdown but was working when pasted into the console)

```{r{}}
# May 2024, Shell
ggplot(data = data202405, aes(x=population, y = shell_dryweight_g, fill = population)) + geom_boxplot() + theme_minimal_grid() + labs( x = "Population", y = "Dry Shell Weight (g)", title = "Dry Shell Weight by Population (May 2024)") +
  scale_fill_manual(values = population_colors) + theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") 

# May 2024, Tissue 
ggplot(data = data202405, aes(x=population, y = tissue_dryweight_g, fill = population)) + geom_boxplot() + theme_minimal_grid() + labs( x = "Population", y = "Dry Tissue Weight (g)", title = "Dry Tissue Weight by Population (May 2024)") +
  scale_fill_manual(values = population_colors) + theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") 

#Nov. 2024, LEW, shell
ggplot(data = data202411  %>% filter (site == "LEW"), aes(x=population, y = shell_dryweight_g, fill = population)) + geom_boxplot() + theme_minimal_grid() + labs( x = "Population", y = "Dry Shell Weight (g)", title = "Dry Shell Weight by Population at Lewisetta (Nov. 2024)") +
  scale_fill_manual(values = population_colors) + theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") 

#Nov 2024, YR, Shell
ggplot(data = data202411  %>% filter (site == "YR"), aes(x=population, y = shell_dryweight_g, fill = population)) + geom_boxplot() + theme_minimal_grid() + labs( x = "Population", y = "Dry Shell Weight (g)", title = "Dry Shell Weight by Population at York River (Nov. 2024)") +
  scale_fill_manual(values = population_colors) + theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") 

#Nov 2024, LEW, tissue
ggplot(data = data202411 %>% filter (site == "LEW"), aes(x=population, y = tissue_dryweight_g, fill = population)) + geom_boxplot() + theme_minimal_grid() + labs( x = "Population", y = "Dry Tissue Weight (g)", title = "Dry Tissue Weight by Population at Lewisetta (Nov. 2024)") +
  scale_fill_manual(values = population_colors) + theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") 

#Nov 2024, YR, tissue
ggplot(data = data202411 %>% filter (site == "YR"), aes(x=population, y = tissue_dryweight_g, fill = population)) + geom_boxplot() + theme_minimal_grid() + labs( x = "Population", y = "Dry Tissue Weight (g)", title = "Dry Tissue Weight by Population at York River (Nov. 2024)") +
  scale_fill_manual(values = population_colors) + theme(axis.text.x = element_text(angle = 45, hjust = 1), legend.position = "none") 
```

