---
title: "Q1_condition_survival_length_figures"
author: "Nicole Mongillo"
date: "2025-06-02"
output: pdf_document
---
Condition plots created from code developed by Zea Segnitz (condition_index_analysis.Rmd).

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Load libraries
```{r}
library(fitdistrplus)
library(dplyr)
library(lme4)
library(lmerTest)
library(pbkrtest)
library(emmeans)
library(multcompView)
library(multcomp)
library(tidyverse)
library(lme4) 
library(effects)
library(ggsignif) 
library(renv)
library(readr)
library(car)
```

```{r}
#setwd("~/Desktop/GitHub/MVP-H2F-HatcheryField/src/performance_H2F")
```

### Condition Index
Condition index calculations and modeling performed by ZS.
#### Data preparation
```{r}
# load in condition index dataset 
MVP_phenotyping_final <- read_csv("../../data/performance_H2F/MVP_phenotyping_final.csv", show_col_types = FALSE)

MVP_phenotyping_final$population <- as.character(MVP_phenotyping_final$population)

#replace LARMIX with HYBRIDMIX
MVP_phenotyping_final["population"][MVP_phenotyping_final["population"] == "H1-LARMIX"] <- "H1-HYBRIDMIX"

# put the populations in correct factor order
MVP_phenotyping_final$population <- factor(MVP_phenotyping_final$population, levels = c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME", "H1-HYBRIDMIX","H2-SEEDMIX"))

population_colors <- c(
  "W1-TX" = "#3e4989",
  "W2-LA" = "#31688e",
  "W3-FL" = "#26828e",
  "W4-VA" = "#1f9e89",
  "S1-LOLA" = "#440154",
  "S2-DEBY" = "#482878",
  "W5-NH" = "#35b779",
  "W6-ME" = "#6ece58",
  "H1-HYBRIDMIX" = "#fde725",
  "H2-SEEDMIX"= "#b5de2b")

#subset the timepoints 
data202405 = MVP_phenotyping_final %>% filter(samplingTimePoint == "2024-05")
data202411 = MVP_phenotyping_final %>% filter(samplingTimePoint == "2024-11")

#set the levels 
data202405$population <- factor(data202405$population, levels = c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME", "H1-HYBRIDMIX", "H2-SEEDMIX"))

data202411$population <- factor(data202411$population, levels = c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY",  "W5-NH", "W6-ME", "H1-HYBRIDMIX", "H2-SEEDMIX"))
```

#### Create linear mixed-effects models 
```{r}
mixed.model.3 <- lmer(CI ~ site*population + (1 | bagID), data = data202411)
summary(mixed.model.3)
mm3_anova <- Anova(mixed.model.3, type = 3, ddf = "Satterthwaite")

mixed.model.3.site.pop = lmer(CI ~ site.pop + (1 | bagID), data = data202411) # build this as well to be able to compute pairwise differences between every site.pop combination 
```

#### p=Post-hoc testing. 
Use emmeans to find and pairwise compare means of every population. 
```{r}
#pairwise comparisons of every site-population combination ex. LEW.TX etc. - only for Nov 2024 because site is not significant at May 2024
mm3.site.pop.emmeans = emmeans(mixed.model.3.site.pop, pairwise ~ site.pop, adjust = "tukey")

# get just the emmeans output 
mm3.emm = mm3.site.pop.emmeans$emmeans

# Generate compact letter display
mm3.cld_letters <- cld(mm3.emm, adjust = "tukey", Letters = letters, decreasing = T)

#make sure the letters are in dataframes
as.data.frame(mm3.cld_letters)

#add cols for site and parental group separately based on those in data202411
site_pop <- subset(data202411, select = c("site.pop", "site", "population"))
site_pop1 <- site_pop %>% 
  distinct(site.pop, .keep_all = TRUE)

mm3.cld_letters1 <- left_join(mm3.cld_letters, site_pop1, by = "site.pop")
```

#### Model with time inclusion
```{r}
mixed.model.4 = lmer(CI ~ site*population*samplingTimePoint + (1 | bagID), data = MVP_phenotyping_final %>% filter(samplingTimePoint == "2024-05" | samplingTimePoint == "2024-11"))
anova(mixed.model.4)
step(mixed.model.4)
### confirms that current model is the best one 
```

### Figure product: Condition index vs parental group
This creates box plots of the condition index by population, with unique colors for LEW and YR. Plotting by NM.
```{r}
mm3.cld_letters1$population <- as.factor(mm3.cld_letters1$population)

site.pop_ls <- c("LEW.TX", "YR.TX", "LEW.LA", "YR.LA", "LEW.FL", "YR.FL", "LEW.VA", "YR.VA", "LEW.LOLA", "YR.LOLA", "LEW.DEBY", "YR.DEBY", "LEW.NH", "YR.NH", "LEW.ME", "YR.ME", "LEW.LARMIX", "YR.LARMIX", "LEW.SEEDMIX", "YR.SEEDMIX")
x1 <- c(0.7, 1.1, 1.7, 2.1, 2.7, 3.1, 3.7, 4.1, 4.7, 5.1, 5.7, 6.1, 6.7, 7.1, 7.7, 8.1, 8.7, 9.1, 9.7, 10.1)
x2 <- c(0.9, 1.3, 1.9, 2.3, 2.9, 3.3, 3.9, 4.3, 4.9, 5.3, 5.9, 6.3, 6.9, 7.3, 7.9, 8.3, 8.9, 9.3, 9.9, 10.3)

xvals <- as.data.frame(cbind(site.pop_ls, x1, x2))
colnames(xvals) <- c("site.pop", "x1", "x2")

mm3.cld_letters2 <- as.data.frame(left_join(mm3.cld_letters1, xvals, by = "site.pop"))
mm3.cld_letters2$x1 <- as.numeric(mm3.cld_letters2$x1)
mm3.cld_letters2$x2 <- as.numeric(mm3.cld_letters2$x2)

# Nov 2024
field_ci <- ggplot(data202411, aes(x = population, y = CI, fill = site)) +
  geom_boxplot(position = position_dodge(width = 0.8)) +  
  scale_x_discrete() +  
  labs(x = "Treatment group", y = "Condition index", title = "Condition index vs. Treatment group", subtitle = "Final field monitoring event, November 2024", fill = "Site") +   scale_fill_manual(labels = c("LEW" = "Lewisetta", "YR" = "York River"), values = c("LEW" = "#F06292", "YR" = "#880E4F")) +  
  ylim(c(0, 12.2))+
  theme_minimal() +
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 10),
        axis.title = element_text(size = 12),
        axis.text.y = element_text(color = "black", size = 10),
        legend.text = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

field_ci

#make list of significance letters
letters_ls <- c("a", "b", "c", "d", "e", "f")

#make dfs where each parental group with a given significance letter is subsetted out
for(x in letters_ls){
  names <- paste(x, "df", sep = "_")
  
  df <- mm3.cld_letters2 %>% 
  filter(grepl(x, .group) == TRUE)
  
  assign(names, df)
}

#add horizontal line segments for each significance letter based on the dataframes we just made
field_ci_sig <- field_ci + 
  geom_segment(data = a_df, aes(x = x1, xend = x2, y = 12, yend =12), color = "#8A8989") + 
  geom_segment(data = b_df, aes(x = x1, xend = x2, y = 11.8, yend = 11.8), color = "#8A8989")+
  geom_segment(data = c_df, aes(x = x1, xend = x2, y = 11.6, yend =11.6), color = "#8A8989")+
  geom_segment(data = d_df, aes(x = x1, xend = x2, y = 11.4, yend =11.4), color = "#8A8989")+
  geom_segment(data = e_df, aes(x = x1, xend = x2, y = 11.2, yend =11.2), color = "#8A8989")+
  geom_segment(data = f_df, aes(x = x1, xend = x2, y = 11, yend =11), color = "#8A8989")+
  geom_hline(yintercept = 10.5, linetype = "dashed")

field_ci_sig
ggsave("FigS4_field_ci.pdf",
       plot = field_ci_sig,
       device = "pdf",
       path = "../../figures/performance_H2F",
       height = 6,
       width = 8)
```

### Survival
Survival calculations and models conducted by CR in len_surv_modeling_ooters.Rmd. Plotting by NM.
```{r}
#final field time-point survival analyses were conducted in len_surv_modeling_ooters.Rmd
survival <- read.csv("../../data/performance_H2F/field_surv_final_time.csv")[,-c(1, 8)]

#rename columns
colnames(survival) <- c("pop_site", "site", "population", "bag_label", "t", "survival", "letters")

survival$population <- factor(survival$population, levels = c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY",  "W5-NH", "W6-ME", "H1-HYBRIDMIX", "H2-SEEDMIX"))
```

```{r}
surv_t3_plot <- ggplot(survival, aes(x = population, y = survival, fill = site)) + 
  geom_boxplot(position=position_dodge(width = 0.8), linewidth = .4, fatten = 0.5) +
  ggtitle("Survival vs Treatment Group") +
  scale_x_discrete() +  
  ylim(c(0, 1.5))+
  labs(x = "Treatment Group", y = "Survival", subtitle = "Final Field Monitoring Event, Nov. 2024", fill = "Site") +   scale_fill_manual(values = c("Lewisetta" = "#F06292", "YorkRiver" = "#880E4F"), labels = c("Lewisetta" = "Lewisetta", "YorkRiver" = "York River")) + 
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 10),
        axis.title = element_text(size = 12),
        axis.text.y = element_text(color = "black", size = 10),
        legend.text = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())
surv_t3_plot

#make df of each pop_site and significance letters
survival_sig <- survival[seq(1, nrow(survival), 3), ]

surv_ps_ls <- c("Lewisetta:TX", "YorkRiver:TX", "Lewisetta:LA", "YorkRiver:LA", "Lewisetta:FL", "YorkRiver:FL", "Lewisetta:VA", "YorkRiver:VA", "Lewisetta:LOLA", "YorkRiver:LOLA", "Lewisetta:DEBY", "YorkRiver:DEBY", "Lewisetta:NH", "YorkRiver:NH", "Lewisetta:ME", "YorkRiver:ME", "Lewisetta:HYBRIDMIX", "YorkRiver:HYBRIDMIX", "Lewisetta:SEEDMIX", "YorkRiver:SEEDMIX")

site_xvals <- as.data.frame(cbind(surv_ps_ls, x1, x2))
colnames(site_xvals) <- c("pop_site", "x1", "x2")

survival_sig2 <- as.data.frame(left_join(survival_sig, site_xvals, by = "pop_site"))
survival_sig2$x1 <- as.numeric(survival_sig2$x1)
survival_sig2$x2 <- as.numeric(survival_sig2$x2)

#make list of significance letters
s_letters_ls <- c("a", "b", "c", "d", "e", "f", "g", "h", "i")

#make dfs where each parental group with a given significance letter is subsetted out
for(x in s_letters_ls){
  names <- paste(x, "s_df", sep = "_")
  
  df <- survival_sig2 %>% 
  filter(grepl(x, letters) == TRUE)
  
  assign(names, df)
}

#add horizontal line segments for each significance letter based on the dataframes we just made
surv_t3_plot_sig <- surv_t3_plot + 
  geom_segment(data = a_s_df, aes(x = x1, xend = x2, y = 1.44, yend =1.44), color = "#8A8989") + 
  geom_segment(data = b_s_df, aes(x = x1, xend = x2, y = 1.41, yend = 1.41), color = "#8A8989")+
  geom_segment(data = c_s_df, aes(x = x1, xend = x2, y = 1.38, yend =1.38), color = "#8A8989")+
  geom_segment(data = d_s_df, aes(x = x1, xend = x2, y = 1.35, yend =1.35), color = "#8A8989")+
  geom_segment(data = e_s_df, aes(x = x1, xend = x2, y = 1.32, yend =1.32), color = "#8A8989")+
  geom_segment(data = f_s_df, aes(x = x1, xend = x2, y = 1.29, yend =1.29), color = "#8A8989") +
  geom_segment(data = g_s_df, aes(x = x1, xend = x2, y = 1.26, yend =1.26), color = "#8A8989")+
  geom_segment(data = h_s_df, aes(x = x1, xend = x2, y = 1.23, yend =1.23), color = "#8A8989")+
  geom_segment(data = i_s_df, aes(x = x1, xend = x2, y = 1.2, yend =1.2), color = "#8A8989")+
  geom_hline(yintercept = 1.1, linetype = "dashed")

surv_t3_plot_sig
ggsave("field_surv.pdf",
       plot = surv_t3_plot_sig,
       device = "pdf",
       path = "../../figures/performance_H2F",
       height = 4,
       width = 6)
```

```{r}
length <- read.csv("../../data/performance_H2F/field_length_final_time.csv")[, 2:10]

length$pop <- factor(length$pop, levels = c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY",  "W5-NH", "W6-ME", "H1-HYBRIDMIX", "H2-SEEDMIX"))

len_plot <- ggplot(length, aes(x = pop, y = length, fill = bag_site)) + 
  geom_boxplot(position=position_dodge(.8), linewidth = 0.4, fatten = 0.5) +
  ylim(c(0, 170))+
  labs(x = "Treatment Group", y = "Length (mm)", title = "Shell Length vs. Treatment Group", subtitle = "Final Field Monitoring Event, Nov. 2024", fill = "Site")+
  #scale_x_discrete(labels = c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME", "H1-HYBRIDMIX", "H2-SEEDMIX"))+
  scale_fill_manual(values = c("Lewisetta" = "#F06292", "YorkRiver" = "#880E4F"), labels = c("Lewisetta" = "Lewisetta", "YorkRiver" = "York River")) + 
  theme_classic()+
  theme(axis.text.x = element_text(angle = 45, hjust = 1, color = "black", size = 10),
        axis.title = element_text(size = 12),
        axis.text.y = element_text(color = "black", size = 10),
        legend.text = element_text(size = 10),
        panel.grid.major = element_blank(),
        panel.grid.minor = element_blank())

len_plot

colnames(length) <- c("pop_site", "bag_site", "pop", "t", "bags_label", "tags_label", "length", "width", "letters")

length_sig <- as.data.frame(left_join(length, site_xvals, by = "pop_site"))
length_sig$x1 <- as.numeric(length_sig$x1)
length_sig$x2 <- as.numeric(length_sig$x2)

#make list of significance letters
l_letters_ls <- c("a", "b", "c", "d", "e", "f", "g", "h", "i", "j")

#make dfs where each parental group with a given significance letter is subsetted out
for(x in l_letters_ls){
  names <- paste(x, "l_df", sep = "_")
  
  df <- length_sig %>% 
  filter(grepl(x, letters) == TRUE)
  
  assign(names, df)
}

#add horizontal line segments for each significance letter based on the dataframes we just made
len_plot_sig <- len_plot + 
  geom_segment(data = a_l_df, aes(x = x1, xend = x2, y = 170, yend =170), color = "#8A8989") + 
  geom_segment(data = b_l_df, aes(x = x1, xend = x2, y = 166, yend =166), color = "#8A8989")+
  geom_segment(data = c_l_df, aes(x = x1, xend = x2, y = 162, yend =162), color = "#8A8989")+
  geom_segment(data = d_l_df, aes(x = x1, xend = x2, y = 158, yend =158), color = "#8A8989")+
  geom_segment(data = e_l_df, aes(x = x1, xend = x2, y = 154, yend =154), color = "#8A8989")+
  geom_segment(data = f_l_df, aes(x = x1, xend = x2, y = 150, yend =150), color = "#8A8989") +
  geom_segment(data = g_l_df, aes(x = x1, xend = x2, y = 146, yend =146), color = "#8A8989")+
  geom_segment(data = h_l_df, aes(x = x1, xend = x2, y = 142, yend =142), color = "#8A8989")+
  geom_segment(data = i_l_df, aes(x = x1, xend = x2, y = 138, yend =138), color = "#8A8989")+
  geom_segment(data = j_l_df, aes(x = x1, xend = x2, y = 134, yend =134), color = "#8A8989")+
  geom_hline(yintercept = 130, linetype = "dashed")

len_plot_sig
ggsave("field_len.pdf",
       plot = len_plot_sig,
       device = "pdf",
       path = "../../figures/performance_H2F",
       height = 4,
       width = 6)
```

