---
title: "Envr_Performance_Mantel_Tests"
output: pdf_document
date: "2025-01-23"
---

This code analyzes the effect of pairwise environmental distance between conditions at the environments-of-origin and pairwise genetic distance between parent populations on pairwise differences in survival and shell length for juvenile oysters raised in the hatchery and nursery at VIMS ABC.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

### Packages

```{r packages}

library(tidyverse)
library(ggplot2)
library(gridExtra)
library(multcompView) #extract significance letters from Tukey-Kramer test
library(tidyverse)
library(grid)#for ggplot arranging
library(cowplot) #arrange ggplots
library(ggpmisc)
library(vegan) #for Mantel tests
library(reshape2)

```

```{r}
#setwd("~/Desktop/MVP-H2F-HatcheryField")
```

### Load data

```{r}

######## File paths - we should probably store these all in the pairwise analyses data folder

#this file contains raw lengths of all oysters aged 15-78 days
length <- read.csv(file.path("..", "..", "data", "performance_H2F", "latestage_length.csv"))

#this file has summary statistics on temperature and salinity for the 8 groups 
envr_summary <-  read.csv(file.path("..", "..", "data", "envr_of_origin", "envr_quantiles_summary.csv"))

#this file is a matrix of the environmental distances between each group
envr_dist <- readRDS(file.path("..", "..", "data", "envr_of_origin", "envr_quantiles_dist.rds"))

#this file contains survival rate data for larval oysters up to day 21 post hatching
survival <-  read.csv(file.path("..", "..", "data", "performance_H2F", "CViMVP_larvae_survival.csv"))

# this file is a matrix of genetic distances between each group
fst_mat <- readRDS(file.path("..", "..", "data", "pairwise_differences_analyses", "fst_results.rds"))

```

### Data preparation

#### Survival

```{r}
summary(survival)

#For survival rates, we only care about survival from day 21 (marked as day_an 15-21), so filter for those day_an only. Reminder that days 15-21 were grouped for analysis because there were constant larval drops occurring during that time, but the number of drops varied by site. Also exclude LARMIX, since there is no environmental data for that, given that it represents oysters from multiple sites
survival_day21 <- survival %>% 
  filter(Day_analysis == "15-21") %>% 
  filter(Tank_naming != "MVP-LARMIX")

#Exclude any groups with "small" in their label. The length data only considers the eyed oysters.
#use grep function to make a vector of all indices in latestage_len with the phrase "small" in them, ignore.case=TRUE means query is not case sensitive
small <- grep("small", survival_day21$Group_Day_Label, ignore.case = TRUE)

#make new dataframe WITHOUT any indices saved in small, mean remaining data are everything except small larvae
survival_day21_sub <- survival_day21[-small, ]

#exclude "MVP" from group names to shorten them
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-DEBY"] <- "S2-DEBY"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"]  == "MVP-FL"] <- "W3-FL"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-JR"] <- "W4-VA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LA"] <- "W2-LA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LOLA"] <- "S1-LOLA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-ME"] <- "W6-ME"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-NH"] <- "W5-NH"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-TX"] <- "W1-TX"

#select only tank naming and survival rate columns
survival_filter <- survival_day21_sub[, c("Tank_naming", "Survival_rate_perc")]

#rename columns
colnames(survival_filter) <- c("site_name", "percent_survival")
```

#### Length

```{r}
#select columns with group, day, and shell length in mm
length <- length[,c("site_name", "day_an", "shell_length_mm")]

#filter out LARMIX from group
length_filter <- length %>% 
  filter(site_name != "H1-LARMIX")

#keep only data from days 15-21
length_21 <- length_filter %>% 
  filter(day_an == "15-21")

length_78 <- length_filter %>% 
  filter(day_an == "78")

#find mean shell lengths of each group on each day
means_mm_21 <- aggregate(shell_length_mm ~ site_name, FUN = mean, data = length_21)
colnames(means_mm_21) <- c("site_name", "mean_length_mm_21")

means_mm_78 <- aggregate(shell_length_mm ~ site_name, FUN = mean, data = length_78)
colnames(means_mm_78) <- c("site_name", "mean_length_mm_78")
```

#### Environmental
```{r}
#remove extra site label and standard deviations from envr_summary
envr <- envr_summary %>% select(-X)

summary(envr)

#add sampling site latitudes. Note that these latitudes ARE NOT the same as the latitudes of where environmental data came from. These latitudes, while similar, were often slightly different, and I am using our actual sampling site latitudes. These latitudes are the same ones as those in table 1 in my thesis.
envr$lat <- c(37.249107, 37.98030, 37.1501163, 28.096000, 29.239925, 30.440030, 43.053746, 44.01330)
```

#### Join data frames
```{r}
surv_length_envr <- left_join(envr, means_mm_21, by = "site_name") %>% 
  left_join(means_mm_78, by = "site_name") %>% 
  left_join(survival_filter, by = "site_name")

#save csv for further analyses
write.csv(surv_length_envr, file.path("..", "..", "data", "parent_effects_H2F", "Survival_Length_Envr_Data.csv"), row.names = FALSE)
```

Make distance matrices for difference in mean length across groups at days 21 and 78 and difference in survival across groups at day 21.

```{r}
#save vector of row and column names
population_order <- c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME")

# if you pull things out of surv_length_envr, they are already in the right order - Kiran
day21_len <- surv_length_envr %>% select(mean_length_mm_21)

day21_len_dist <- as.matrix(dist(day21_len$mean_length_mm_21))
rownames(day21_len_dist) <- population_order
colnames(day21_len_dist) <- population_order

day78_len <- surv_length_envr %>% select(mean_length_mm_78)

day78_len_dist <- as.matrix(dist(day78_len$mean_length_mm_78))
rownames(day78_len_dist) <- population_order
colnames(day78_len_dist) <- population_order

day21_surv <- surv_length_envr %>% select(percent_survival)

day21_surv_dist <- as.matrix(dist(day21_surv$percent_survival))
rownames(day21_surv_dist) <- population_order
colnames(day21_surv_dist) <- population_order

```

#### Fix Fst matrix

```{r}

# reorder
fst_mat <- fst_mat[population_order, population_order]

```

#### Check order of all matrices

```{r}

check_matrix_order <- function(matrices, population_order) {
  all_good <- TRUE
  for (i in seq_along(matrices)) {
    if (!identical(rownames(matrices[[i]]), population_order)) {
      message(sprintf("Row names do not match in matrix %d", i))
      all_good <- FALSE
    }
    if (!identical(colnames(matrices[[i]]), population_order)) {
      message(sprintf("Column names do not match in matrix %d", i))
      all_good <- FALSE
    }
  }
  if (all_good) {
    message("All matrices have matching row and column names.")
  }
}

check_matrix_order(list(fst_mat, envr_dist, day21_len_dist, day78_len_dist, day21_surv_dist), population_order)

```

### Partial Mantel tests

Comparison between measures of genetic diversity of parents and size and survival in the hatchery of offspring.
Comparison between genetic distance of parent populations and size and survival distance of offspring populations.

To test the effect of pairwise Fst on pairwise length at day 21, length at day 78, and survival at day 21.

#### Assumptions

```{r}

# Linearity

plot(as.vector(as.matrix(fst_mat)), as.vector(as.matrix(envr_dist)))

plot(as.vector(as.matrix(fst_mat)), as.vector(as.matrix(day21_len_dist)))
plot(as.vector(as.matrix(fst_mat)), as.vector(as.matrix(day78_len_dist)))
plot(as.vector(as.matrix(fst_mat)), as.vector(as.matrix(day21_surv_dist)))

plot(as.vector(as.matrix(envr_dist)), as.vector(as.matrix(day21_len_dist)))
plot(as.vector(as.matrix(envr_dist)), as.vector(as.matrix(day78_len_dist)))
plot(as.vector(as.matrix(envr_dist)), as.vector(as.matrix(day21_surv_dist)))

```

#### Partial tests

To test the effect of genetic distance while controlling for environmental distance.

```{r}

# test for the effect of genetic distance on size/survival, controlling for environmental distance

# length on day 21
partial_mantel_gen_len_21 <- mantel.partial(day21_len_dist, fst_mat, envr_dist, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_gen_len_21

# length on day 78
partial_mantel_gen_len_78 <- mantel.partial(day78_len_dist, fst_mat, envr_dist, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_gen_len_78

# survival on day 21
partial_mantel_gen_surv_21 <- mantel.partial(day21_surv_dist, fst_mat, envr_dist, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_gen_surv_21

# collect p-values
partial_mantel_gen_p <- c(0.46275, 0.44756, 0.13309)

# adjust p-vaules to correct for multiple tests
partial_mantel_gen_p_adj <- p.adjust(partial_mantel_gen_p, method = "BH")

# print results
partial_mantel_gen_p_adj

```

To test the effect of environmental distance while controlling for genetic distance.

```{r}

# test for the effect of environmental distance on size/survival, controlling for genetic distance

# length on day 21
partial_mantel_envr_len_21 <- mantel.partial(day21_len_dist, envr_dist, fst_mat, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_envr_len_21

# length on day 78
partial_mantel_envr_len_78 <- mantel.partial(day78_len_dist, envr_dist, fst_mat, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_envr_len_78

# survival on day 21
partial_mantel_envr_surv_21 <- mantel.partial(day21_surv_dist, envr_dist, fst_mat, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_envr_surv_21

# collect p-values
partial_mantel_envr_p <- c(0.082192, 0.92851, 0.25207)

# adjust p-vaules to correct for multiple tests
partial_mantel_envr_p_adj <- p.adjust(partial_mantel_envr_p, method = "BH")

# print results
partial_mantel_envr_p_adj

```

### Heatmaps

#### Data preparation

```{r}

# function to put all matrices in the same format

process_matrix <- function(matrix_data, population_order) {
  # convert the matrix to long format
  long_data <- melt(matrix_data, na.rm = TRUE)
  
  # edit column names
  colnames(long_data) <- c("Var1", "Var2", "value")
  
  # reorder rows and columns
  long_data <- long_data %>%
    mutate(
      Var1 = factor(Var1, levels = population_order),
      Var2 = factor(Var2, levels = population_order)
    ) %>%
    arrange(Var1, Var2)
  
  return(long_data)
}

# create a vector for the desired order of populations
population_order <- c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME")

# gather all matrices
all_data <- list(fst_mat, envr_dist, day21_len_dist, day78_len_dist, day21_surv_dist)

# apply function to each matrix
processed_data <- lapply(all_data, process_matrix, population_order = population_order)

# Access individual processed data frames
plot1_fst <- processed_data[[1]]
plot2_envr <- processed_data[[2]]
plot3_len21 <- processed_data[[3]]
plot3_len78 <- processed_data[[4]]
plot3_surv21 <- processed_data[[5]]

```

#### Heatmaps

```{r}

heatmap_fst <- ggplot(plot1_fst, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Pairwise Fst", x = "Population", y = "Population", fill = "Fst") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heatmap_envr <- ggplot(plot2_envr, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Environmental distance", x = "Population", y = "Population", fill = "Distance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heatmap_len21 <- ggplot(plot3_len21, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Length (day 21)", x = "Population", y = "Population", fill = "Length (mm)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heatmap_len78 <- ggplot(plot3_len78, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Length (day 78)", x = "Population", y = "Population", fill = "Length (mm)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heatmap_surv21 <- ggplot(plot3_surv21, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Survival (day 21)", x = "Population", y = "Population", fill = "Survival (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# arrange figures in a grid
fig_heatmap_grid <- grid.arrange(
  heatmap_fst, heatmap_envr,
  heatmap_len21, heatmap_len78, heatmap_surv21,
  nrow = 2  # 2 rows
)

ggsave("fig_heatmap_grid.png", fig_heatmap_grid, width = 14, height = 8)

```


