---
title: "Envr_Performance_Mantel_Tests"
author: "Kiran Bajaj"
output: pdf_document
date: "2025-01-23"
---

# Overview
This analysis tests whether pairwise environmental distance (at origin sites) and 
pairwise genetic distance (Fst) predict differences in oyster survival and shell length.
Method: Partial Mantel tests with Spearman correlation and 10,000 permutations.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Packages
```{r packages, message=FALSE}
library(tidyverse)
library(vegan)      # Mantel tests
library(patchwork)  # combining plots
```

## Configuration
```{r config}
# Population display order (south to north within wild, then selected)
POP_ORDER <- c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME")

# Site name mapping from raw data to standardized names
SITE_RECODE <- c(
  "MVP-TX" = "W1-TX", "MVP-LA" = "W2-LA", "MVP-FL" = "W3-FL", "MVP-JR" = "W4-VA",
  "MVP-LOLA" = "S1-LOLA", "MVP-DEBY" = "S2-DEBY", "MVP-NH" = "W5-NH", "MVP-ME" = "W6-ME"
)

# Groups to exclude from analysis (no environmental data available)
EXCLUDE_GROUPS <- c("MVP-LARMIX", "MVP-SEEDMIX", "H1-LARMIX", "H1-HYBRIDMIX")

# Paths
DATA_DIR <- file.path("..", "..", "data")
FIG_DIR <- file.path("..", "..", "figures", "pairwise_differences_analyses")
```

## Helper Functions
```{r helpers}
#' Recode site names and filter excluded groups
clean_site_names <- function(df, site_col = "site_name") {
  df %>%
    filter(!.data[[site_col]] %in% EXCLUDE_GROUPS) %>%
    mutate(!!site_col := recode(.data[[site_col]], !!!SITE_RECODE))
}

#' Reorder dataframe by population order
order_by_pop <- function(df, site_col = "site_name") {
  df %>% arrange(factor(.data[[site_col]], levels = POP_ORDER))
}

#' Create distance matrix from a two-column dataframe (site_name, value)
make_dist_matrix <- function(df) {
  df <- df %>% filter(!site_name %in% c("H1-HYBRIDMIX", EXCLUDE_GROUPS))
  mat <- as.matrix(dist(df[[2]]))
  rownames(mat) <- colnames(mat) <- POP_ORDER
  mat
}

#' Summarize field data by monitoring event and site
summarize_field_data <- function(df, value_col, group_col = "bag_site") {
  df %>%
    group_by(monitoring_event, .data[[group_col]], site_name) %>%
    summarise(value = mean(.data[[value_col]], na.rm = TRUE), .groups = "drop")
}

#' Split field summary into named list of dataframes
split_field_data <- function(df, prefix, group_col = "bag_site") {
  df %>%
    group_split(monitoring_event, .data[[group_col]]) %>%
    setNames(
      df %>%
        group_keys(monitoring_event, .data[[group_col]]) %>%
        mutate(name = paste0(prefix, "_event", monitoring_event, "_", 
                            tolower(gsub(" ", "", .data[[group_col]])))) %>%
        pull(name)
    ) %>%
    map(~ .x %>% order_by_pop() %>% select(site_name, value))
}

#' Create a heatmap from a distance matrix
plot_heatmap <- function(mat, fill_label, title = "") {
  as.data.frame(as.table(mat)) %>%
    ggplot(aes(Var1, Var2, fill = Freq)) +
    geom_tile(color = "white") +
    scale_fill_viridis_c(option = "D", na.value = "grey50",
                         guide = guide_colorbar(label.theme = element_text(angle = 45, size = 12))) +
    labs(title = title, x = "Treatment group", y = "Treatment group", fill = fill_label) +
    theme_minimal() +
    theme(
      axis.text.x = element_text(angle = 45, hjust = 1, size = 14),
      axis.text.y = element_text(size = 14),
      axis.title = element_text(size = 14),
      legend.title = element_text(size = 14),
      legend.position = "bottom"
    )
}
```

## Load Data
```{r load-data}
# Performance data
hatchery_length <- read.csv(file.path(DATA_DIR, "performance_H2F", "latestage_length.csv"))
hatchery_survival <- read.csv(file.path(DATA_DIR, "performance_H2F", "CViMVP_larvae_survival.csv"))
field_length <- read.csv(file.path(DATA_DIR, "performance_H2F", "lengths_bags_022025.csv"))
field_survival <- read.csv(file.path(DATA_DIR, "performance_H2F", "mortality_bags_022025.csv"))

# Environmental and genetic distance matrices
envr_summary <- read.csv(file.path(DATA_DIR, "envr_of_origin", "envr_quantiles_summary.csv"))
envr_dist <- readRDS(file.path(DATA_DIR, "envr_of_origin", "envr_quantiles_dist.rds"))
fst_mat <- readRDS(file.path(DATA_DIR, "pairwise_differences_analyses", "20250606_fst_results.rds"))

# Reorder Fst matrix to match population order
fst_mat <- fst_mat[POP_ORDER, POP_ORDER]
```

## Process Hatchery Data
```{r hatchery-data}
# Survival at day 21 (days 15-21 grouped due to variable larval drop timing)
surv_day21 <- hatchery_survival %>%
  filter(Day_analysis == "15-21", !grepl("small", Group_Day_Label, ignore.case = TRUE)) %>%
  clean_site_names("Tank_naming") %>%
  select(site_name = Tank_naming, surv_21 = Survival_rate_perc) %>%
  order_by_pop()

# Shell length at days 21 and 78
hatchery_length_clean <- hatchery_length %>%
  select(site_name, day_an, shell_length_mm) %>%
  filter(!site_name %in% EXCLUDE_GROUPS)

calc_mean_length <- function(df, day) {
  df %>%
    filter(day_an == day) %>%
    group_by(site_name) %>%
    summarise(mean_length = mean(shell_length_mm), .groups = "drop") %>%
    order_by_pop()
}

means_mm_21 <- calc_mean_length(hatchery_length_clean, "15-21")
means_mm_78 <- calc_mean_length(hatchery_length_clean, "78")
```

## Process Field Data
```{r field-data}
# Helper to parse field bag labels
parse_field_bags <- function(df, bag_col) {
  df %>%
    separate(!!bag_col, into = c("site_name", "bag_number"), sep = "-(?=\\d+$)") %>%
    clean_site_names()
}

# Field length: average across bags per group/event/site
field_length_list <- field_length %>%
  parse_field_bags("bag") %>%
  summarize_field_data("length", "site") %>%
  split_field_data("len", "site")

# Field survival: reshape to long format, then summarize
field_survival_list <- field_survival %>%
  parse_field_bags("bags_label") %>%
  pivot_longer(cols = c(t1, t2, t3), names_to = "monitoring_event", values_to = "survival") %>%
  mutate(monitoring_event = str_replace(monitoring_event, "t", "")) %>%
  summarize_field_data("survival", "bag_site") %>%
  split_field_data("surv", "bag_site")
```

## Prepare Environmental Data
```{r envr-data}
# Add sampling site latitudes (from Table 1 of Nicole thesis)
envr <- envr_summary %>%
  select(-X) %>%
  mutate(lat = c(28.096, 29.240, 30.440, 37.150, 37.980, 37.249, 43.054, 44.013))
```

## Export Combined Dataset
```{r export}
surv_length_envr <- envr %>%
  left_join(means_mm_21 %>% rename(mean_length_mm_21 = mean_length), by = "site_name") %>%
  left_join(means_mm_78 %>% rename(mean_length_mm_78 = mean_length), by = "site_name") %>%
  left_join(surv_day21, by = "site_name")

write.csv(surv_length_envr, 
          file.path(DATA_DIR, "parent_effects_H2F", "Survival_Length_Envr_Data.csv"), 
          row.names = FALSE)
```

## Build Distance Matrices
```{r distance-matrices}
# Combine all response variables into named list
response_dfs <- c(
  list(surv_day21 = surv_day21 %>% select(site_name, value = surv_21),
       len_21 = means_mm_21 %>% rename(value = mean_length),
       len_78 = means_mm_78 %>% rename(value = mean_length)),
  field_length_list,
  field_survival_list
)

# Convert each to distance matrix
dist_matrices <- map(response_dfs, make_dist_matrix)

# Verify all matrices have correct ordering
stopifnot(all(map_lgl(dist_matrices, ~ identical(rownames(.x), POP_ORDER))))
message("All matrices have correct row/column ordering.")
```

## Partial Mantel Tests
```{r mantel-tests}
# Test genetic distance effect (controlling for environment)
genetic_results <- map(dist_matrices, ~ mantel.partial(
  .x, fst_mat, envr_dist, method = "spearman", permutations = 10000, na.rm = TRUE
))

# Test environmental distance effect (controlling for genetics)
envr_results <- map(dist_matrices, ~ mantel.partial(
  .x, envr_dist, fst_mat, method = "spearman", permutations = 10000, na.rm = TRUE
))

# Extract p-values and apply FDR correction
genetic_p <- map_dbl(genetic_results, "signif")
envr_p <- map_dbl(envr_results, "signif")
all_p_adj <- p.adjust(c(genetic_p, envr_p), method = "BH")

# Compile results table
mantel_results <- tibble(
  test = rep(names(dist_matrices), 2),
  comparison = rep(c("Genetic|Environment", "Environment|Genetic"), each = length(dist_matrices)),
  statistic = c(map_dbl(genetic_results, "statistic"), map_dbl(envr_results, "statistic")),
  p = c(genetic_p, envr_p),
  p_adj = all_p_adj
)

print(mantel_results)
```

## Assumption Check: Linearity
```{r assumptions, fig.height=4, fig.width=8}
# Visual check of linearity between predictor and response distances
for (name in names(dist_matrices)) {
  par(mfrow = c(1, 2))
  plot(as.vector(fst_mat), as.vector(dist_matrices[[name]]),
       main = paste("Fst vs", name), xlab = "Fst", ylab = "Response distance")
  plot(as.vector(envr_dist), as.vector(dist_matrices[[name]]),
       main = paste("Environment vs", name), xlab = "Environmental distance", ylab = "Response distance")
}
```

## Heatmaps
```{r heatmaps, fig.height=10, fig.width=12}
# Predictor distance heatmaps
heatmap_fst <- plot_heatmap(fst_mat, expression("Pairwise " * F[ST]))
heatmap_envr <- plot_heatmap(envr_dist, "Standardized Euclidean\nenvironmental distance")

# Combined figure
stacked_heatmap <- heatmap_fst / heatmap_envr + plot_annotation(tag_levels = "A")
ggsave(file.path(FIG_DIR, "fig_stacked_envr_fst_heatmap.pdf"), stacked_heatmap, width = 12, height = 10)

# Response variable heatmaps (event 3 as example)
response_heatmaps <- list(
  len_lew = plot_heatmap(dist_matrices$len_event3_lewisetta, "Euclidean distance\nin shell length"),
  len_york = plot_heatmap(dist_matrices$len_event3_yorkriver, "Euclidean distance\nin shell length"),
  surv_lew = plot_heatmap(dist_matrices$surv_event3_lewisetta, "Euclidean distance\nin survival"),
  surv_york = plot_heatmap(dist_matrices$surv_event3_yorkriver, "Euclidean distance\nin survival")
)

# Save response heatmaps
iwalk(response_heatmaps, ~ ggsave(
  file.path(FIG_DIR, paste0("fig_", .y, "_event3_heatmap.pdf")), .x, width = 12, height = 10
))

# Display
heatmap_fst
heatmap_envr
```