---
title: "Envr_Performance_Mantel_Tests"
author: "Kiran Bajaj"
output: pdf_document
date: "2025-01-23"
---

This code analyzes the effect of pairwise environmental distance between conditions at the environments-of-origin and pairwise genetic distance between parent populations on pairwise differences in survival and shell length.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#setwd("~/Desktop/MVP-H2F-HatcheryField")
```

### Packages

```{r packages}

library(tidyverse)
library(ggplot2)
library(grid)
library(reshape2)
library(vegan) # for Mantel tests
library(patchwork)

```

### Load data

```{r}

# raw lengths of all oysters aged 15-78 days
hatchery_length <- read.csv(file.path("..", "..", "data", "performance_H2F", "latestage_length.csv"))

# survival rate data for larval oysters up to day 21 post hatching
hatchery_survival <-  read.csv(file.path("..", "..", "data", "performance_H2F", "CViMVP_larvae_survival.csv"))

# lengths of all oysters averaged by bag for field monitoring events 1-3
field_length <- read.csv(file.path("..", "..", "data", "performance_H2F", "lengths_bags_022025.csv"))

# survival of all oysters averaged by bag for field monitoring events 1-3
field_survival <- read.csv(file.path("..", "..", "data", "performance_H2F", "mortality_bags_022025.csv"))

# summary statistics - temperature and salinity quantiles for the 8 groups 
envr_summary <-  read.csv(file.path("..", "..", "data", "envr_of_origin", "envr_quantiles_summary.csv"))

# matrix of euclidean environmental distances using SCALED data
envr_dist <- readRDS(file.path("..", "..", "data", "envr_of_origin", "envr_quantiles_dist.rds"))

# matrix of genetic distances between each group
fst_mat <- readRDS(file.path("..", "..", "data", "pairwise_differences_analyses", "20250606_fst_results.rds"))

```

### Global variables

Population order

```{r}

# save vector of population order
population_order <- c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME")

```

### Data preparation

#### Hatchery survival

For hatchery survival rate, we used survival from day 21, so filter for those values only. Days 15-21 were grouped for analysis because there were constant larval drops occurring during that time, but the number of drops varied by site. Also exclude LARMIX, since there is no environmental data for group, given that it represents oysters from multiple sites. 

```{r}

# filter for day 21, exclude LARMIX, and recode site names
survival_day21 <- hatchery_survival %>% 
  filter(Day_analysis == "15-21") %>% 
  filter(Tank_naming != "MVP-LARMIX") %>%
  mutate(Tank_naming = recode(Tank_naming,
                           "MVP-DEBY" = "S2-DEBY",
                           "MVP-FL" = "W3-FL",
                           "MVP-LA" = "W2-LA",
                           "MVP-LOLA" = "S1-LOLA",
                           "MVP-ME" = "W6-ME",
                           "MVP-NH" = "W5-NH",
                           "MVP-TX" = "W1-TX",
                           "MVP-JR" = "W4-VA"))

# exclude any groups with "small" in their label, we only want to analyze eyed oysters
surv_day21 <- survival_day21 %>%
  filter(!grepl("small", Group_Day_Label, ignore.case = TRUE)) %>%
  select(Tank_naming, Survival_rate_perc) %>%
  arrange(factor(Tank_naming, levels = population_order)) # reorder

# rename columns
colnames(surv_day21) <- c("site_name", "surv_21")

```

#### Hatchery length

```{r}

# exclude LARMIX
hatchery_length_filter <- hatchery_length %>%
  select(site_name, day_an, shell_length_mm)  %>% 
  filter(site_name != "H1-LARMIX") # filter out LARMIX

# data frame for days 15-21
length_21 <- hatchery_length_filter %>% 
  filter(day_an == "15-21")

# data frame for day 78
length_78 <- hatchery_length_filter %>% 
  filter(day_an == "78")

# calc mean shell lengths of each group on each day
means_mm_21 <- aggregate(shell_length_mm ~ site_name, FUN = mean, data = length_21)
colnames(means_mm_21) <- c("site_name", "mean_length_mm_21")
means_mm_21 <- means_mm_21 %>% arrange(factor(site_name, levels = population_order)) # reorder

means_mm_78 <- aggregate(shell_length_mm ~ site_name, FUN = mean, data = length_78)
colnames(means_mm_78) <- c("site_name", "mean_length_mm_78")
means_mm_78 <- means_mm_78 %>% arrange(factor(site_name, levels = population_order)) # reorder

```

#### Field length

The goal is to create 6 data frames: one for each monitoring event (3) and each field site (2) with
the average length across all bags for each group.

```{r}

# separate bag labels, recode sites, and average by bag and field site
field_length_clean <- field_length %>%
  separate(bag, into = c("site_name", "bag_number"), sep = "-(?=\\d+$)") %>%
  filter(!site_name %in% c("MVP-SEEDMIX", "MVP-LARMIX")) %>% # remove LARMIX and SEEDMIX
  mutate(site_name = recode(site_name,
                           "MVP-DEBY" = "S2-DEBY",
                           "MVP-FL" = "W3-FL",
                           "MVP-LA" = "W2-LA",
                           "MVP-LOLA" = "S1-LOLA",
                           "MVP-ME" = "W6-ME",
                           "MVP-NH" = "W5-NH",
                           "MVP-TX" = "W1-TX",
                           "MVP-JR" = "W4-VA")) 

# average lengths across all bags per group 
# within-bag lengths are are already averaged
# average separately for each monitoring event and field site
field_length_summary <- field_length_clean %>%
  group_by(monitoring_event, site, site_name) %>%
  summarise(mean_length = mean(length, na.rm = TRUE), .groups = "drop")

# split the data frame by monitoring event and site
split_list <- field_length_summary %>%
  group_split(monitoring_event, site)

# give each data frame a name according to monitoring event and field site
split_names <- field_length_summary %>%
  group_keys(monitoring_event, site) %>%
  mutate(name = paste0("length_event", monitoring_event, "_", gsub(" ", "", tolower(site)))) %>%
  pull(name)

split_list <- lapply(split_list, function(df) {
  df %>%
    arrange(factor(site_name, levels = population_order)) %>% # reorder
    select(site_name, mean_length) # remove columns except for population and length
  })

# assign each data frame to a variable
names(split_list) <- split_names
list2env(split_list, envir = .GlobalEnv)

```

#### Field survival

The goal is to create 6 data frames: one for each monitoring event (3) and each field site (2) with
the average survival across all bags for each group.

```{r}

# separate bag labels, recode sites, and average by bag and field site
field_survival_clean <- field_survival %>%
  separate(bags_label, into = c("site_name", "bag_number"), sep = "-(?=\\d+$)") %>%
  filter(!site_name %in% c("MVP-SEEDMIX", "MVP-LARMIX")) %>% # remove LARMIX and SEEDMIX
  mutate(site_name = recode(site_name,
                           "MVP-DEBY" = "S2-DEBY",
                           "MVP-FL" = "W3-FL",
                           "MVP-LA" = "W2-LA",
                           "MVP-LOLA" = "S1-LOLA",
                           "MVP-ME" = "W6-ME",
                           "MVP-NH" = "W5-NH",
                           "MVP-TX" = "W1-TX",
                           "MVP-JR" = "W4-VA")) %>%
  select(site_name, bag_number, bag_site, t1, t2, t3)

# pivot so there is one row per site name, bag, field site, and monitoring event
field_survival_long <- field_survival_clean %>%
  pivot_longer(
    cols = starts_with("t"),  # columns t1, t2, t3
    names_to = "monitoring_event", 
    values_to = "survival"
  ) %>%
  mutate(
    monitoring_event = recode(monitoring_event, 
                              t1 = "1", 
                              t2 = "2", 
                              t3 = "3")
  )

# now can use code from above to make one data frame per monitoring event per field site

# average lengths across all bags per group 
# within-bag lengths are are already averaged
# average separately for each monitoring event and field site
field_survival_summary <- field_survival_long %>%
  group_by(monitoring_event, bag_site, site_name) %>%
  summarise(mean_survival = mean(survival, na.rm = TRUE), .groups = "drop")

# split the data frame by monitoring event and field site
split_list <- field_survival_summary %>%
  group_split(monitoring_event, bag_site)

# give each data frame a name according to monitoring event and field site
split_names <- field_survival_summary %>%
  group_keys(monitoring_event, bag_site) %>%
  mutate(name = paste0("survival_event", monitoring_event, "_", gsub(" ", "", tolower(bag_site)))) %>%
  pull(name)

split_list <- lapply(split_list, function(df) {
  df %>%
    arrange(factor(site_name, levels = population_order)) %>% # reorder
    select(site_name, mean_survival) # remove columns except for population and length
  })

# assign each data frame to a variable
names(split_list) <- split_names
list2env(split_list, envir = .GlobalEnv)

```

#### Environment-of-origin

```{r}

# remove extra column from environmental data summary table
envr <- envr_summary %>% select(-X)

# add sampling site latitudes. from table 1 of Nicole thesis

envr$lat <- c(28.096000, 29.239925, 30.440030, 37.1501163, 37.98030, 37.249107, 43.053746, 44.01330)

```

#### Data product: hatchery survival, hatchery length, environment-of-origin

For further analysis in src/parent_effects_H2F. 

```{r}

surv_length_envr <- left_join(envr, means_mm_21, by = "site_name") %>% 
  left_join(means_mm_78, by = "site_name") %>% 
  left_join(surv_day21, by = "site_name")

# save csv for further analyses
write.csv(surv_length_envr, file.path("..", "..", "data", "parent_effects_H2F", "Survival_Length_Envr_Data.csv"), row.names = FALSE)

```

#### Fst

```{r}

# reorder
fst_mat <- fst_mat[population_order, population_order]

```

### Distance matrices

Make distance matrices for all survival and length variables

```{r}

# list of all the data frames that need to be made into distance matrices
df_list <- list(
  surv_day21 = surv_day21,
  len_21 = means_mm_21,
  len_78 = means_mm_78,
  len_1_lew = length_event1_lewisetta,
  len_1_york = length_event1_yorkriver,
  len_2_lew = length_event2_lewisetta,
  len_2_york = length_event2_yorkriver,
  len_3_lew = length_event3_lewisetta,
  len_3_york = length_event3_yorkriver,
  surv_1_lew = survival_event1_lewisetta,
  surv_1_york = survival_event1_yorkriver,
  surv_2_lew = survival_event2_lewisetta,
  surv_2_york = survival_event2_yorkriver,
  surv_3_lew = survival_event3_lewisetta,
  surv_3_york = survival_event3_yorkriver
)

# empty list to store distance matrices
dist_matrices <- list()

# loop through each data frame in the list
for (name in names(df_list)) {
  df <- df_list[[name]]
  
  # remove polyculture
  df <- df[!df$site_name %in% "H1-HYBRIDMIX", ]
  
  # set the second column as values (first column is population names)
  values <- df[[2]]

  # create distance matrix
  dist_matrix <- as.matrix(dist(values))
  
  # set row and col names
  rownames(dist_matrix) <- population_order
  colnames(dist_matrix) <- population_order
  
  # save into the list
  dist_matrices[[name]] <- dist_matrix
}

```

#### Data check

```{r}

check_matrix_order <- function(matrices, population_order) {
  all_good <- TRUE
  for (i in seq_along(matrices)) {
    if (!identical(rownames(matrices[[i]]), population_order)) {
      message(sprintf("Row names do not match in matrix %d", i))
      all_good <- FALSE
    }
    if (!identical(colnames(matrices[[i]]), population_order)) {
      message(sprintf("Column names do not match in matrix %d", i))
      all_good <- FALSE
    }
  }
  if (all_good) {
    message("All matrices have matching row and column names.")
  }
}

check_matrix_order(dist_matrices, population_order)

```

### Partial Mantel tests

#### Assumptions

```{r}

# Linearity

for (name in names(dist_matrices)) {
  mat <- dist_matrices[[name]]

  # Plot against fst_mat
  plot(as.vector(as.matrix(fst_mat)), as.vector(as.matrix(mat)),
    main = paste("FST vs", name),
    xlab = "FST",
    ylab = paste(name, "Response variable distance")
  )

  # plot against envr_dist
  plot(
    as.vector(as.matrix(envr_dist)), as.vector(as.matrix(mat)),
    main = paste("Environment vs", name),
    xlab = "Environmental distance",
    ylab = paste(name, "Response variable distance")
  )
}

```

#### Partial tests

To test the effect of genetic distance while controlling for environmental distance and vice versa. 

```{r}

# genetic distance controlling for environmental distance
genetic_dist_results <- lapply(dist_matrices, function(dist_mat) {
  mantel.partial(dist_mat, fst_mat, envr_dist, method = "spearman",
                 permutations = 10000, na.rm = TRUE)
})
genetic_p <- sapply(genetic_dist_results, function(res) res$signif)

# environmental distance controlling for genetic distance
envr_dist_results <- lapply(dist_matrices, function(dist_mat) {
  mantel.partial(dist_mat, envr_dist, fst_mat, method = "spearman",
                 permutations = 10000, na.rm = TRUE)
})
envr_p <- sapply(envr_dist_results, function(res) res$signif)

# adjust all p-values together
all_p <- c(genetic_p, envr_p)
all_p_adj <- p.adjust(all_p, method = "BH")

# split adjusted p-values back
genetic_p_adj <- all_p_adj[1:length(genetic_p)]
envr_p_adj <- all_p_adj[(length(genetic_p) + 1):length(all_p)]

# create results data frame
mantel_results <- data.frame(
  test = rep(names(dist_matrices), times = 2),
  comparison = rep(c("Genetic|Environment", "Environment|Genetic"), each = length(dist_matrices)),
  p = c(genetic_p, envr_p),
  p_adj = c(genetic_p_adj, envr_p_adj)
)

print(mantel_results)

```

### Heatmaps

```{r}

fst_long <- as.data.frame(as.table(fst_mat))

heatmap_fst <- ggplot(fst_long, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
scale_fill_viridis_c(
    option = "D", 
    na.value = "grey50", 
    guide = guide_colorbar(label.theme = element_text(angle = 45, size = 12))  # legend tick labels
  ) +
  labs(title = "", x = "Treatment group", y = "Treatment group", fill = expression("Pairwise " * F[ST])) + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),  # x tick labels
    axis.text.y = element_text(size = 14),                        # y tick labels
    axis.title.x = element_text(size = 14),                       # x-axis label
    axis.title.y = element_text(size = 14),                       # y-axis label
    legend.title = element_text(size = 14),                       # legend title
    text = element_text(color = "black"),
    legend.position = "bottom"
  )

ggsave(file.path("..", "..", "figures", "pairwise_differences_analyses", "fig_fst_heatmap.pdf"), heatmap_fst, width = 12, height = 10, dpi = 300)

heatmap_fst

envr_long <- as.data.frame(as.table(envr_dist))

heatmap_envr <- ggplot(envr_long, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
scale_fill_viridis_c(
    option = "D", 
    na.value = "grey50", 
    guide = guide_colorbar(label.theme = element_text(angle = 45, size = 12))  # legend tick labels
  ) +
  labs(title = "", 
       x = "Treatment group", y = "Treatment group",   fill = "Standardized Euclidean\nenvironmental distance") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),  # x tick labels
    axis.text.y = element_text(size = 14),                        # y tick labels
    axis.title.x = element_text(size = 14),                       # x-axis label
    axis.title.y = element_text(size = 14),                       # y-axis label
    legend.title = element_text(size = 14),                       # legend title
    text = element_text(color = "black"),
    legend.position = "bottom"
  )

ggsave(file.path("..", "..", "figures", "pairwise_differences_analyses", "fig_envr_dist_heatmap.pdf"), heatmap_envr, width = 12, height = 10, dpi = 300)

heatmap_envr

stacked_envr_fst_heatmap <- heatmap_fst / heatmap_envr +
  plot_annotation(tag_levels = "A")

ggsave(file.path("..", "..", "figures", "pairwise_differences_analyses", "fig_stacked_envr_fst_heatmap.pdf"), stacked_envr_fst_heatmap, width = 12, height = 10, dpi = 300)

```

```{r}

len_3_lew_long <- as.data.frame(as.table(dist_matrices$len_3_lew))
surv_3_lew_long <- as.data.frame(as.table(dist_matrices$surv_3_lew))
len_3_york_long <- as.data.frame(as.table(dist_matrices$len_3_york))
surv_3_york_long <- as.data.frame(as.table(dist_matrices$surv_3_york))

heatmap_len_3_lew <- ggplot(len_3_lew_long, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", 
    na.value = "grey50", 
    guide = guide_colorbar(label.theme = element_text(angle = 45, size = 12))  # legend tick labels
  ) +
  labs(title = "", x = "Treatment group", y = "Treatment group", fill = "Euclidean distance\nin shell length") + 
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),  # x tick labels
    axis.text.y = element_text(size = 14),                        # y tick labels
    axis.title.x = element_text(size = 14),                       # x-axis label
    axis.title.y = element_text(size = 14),                       # y-axis label
    text = element_text(color = "black"),
    legend.position = "bottom"
  )

heatmap_surv_3_lew <- ggplot(surv_3_lew_long, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", 
    na.value = "grey50", 
    guide = guide_colorbar(label.theme = element_text(angle = 45, size = 12))  # legend tick labels
  ) +
  labs(title = "", 
       x = "Treatment group", y = "Treatment group",   fill = "Euclidean distance\nin survival") +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),  # x tick labels
    axis.text.y = element_text(size = 14),                        # y tick labels
    axis.title.x = element_text(size = 14),                       # x-axis label
    axis.title.y = element_text(size = 14),                       # y-axis label
    text = element_text(color = "black"),
    legend.position = "bottom"
  )

heatmap_len_3_york <- ggplot(len_3_york_long, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", 
    na.value = "grey50", 
    guide = guide_colorbar(label.theme = element_text(angle = 45, size = 12))  # legend tick labels
  ) +
  labs(title = "", x = "Treatment group", y = "Treatment group", fill = "Euclidean distance\nin shell length") +  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),  # x tick labels
    axis.text.y = element_text(size = 14),                        # y tick labels
    axis.title.x = element_text(size = 14),                       # x-axis label
    axis.title.y = element_text(size = 14),                       # y-axis label
    text = element_text(color = "black"),
    legend.position = "bottom"
  )

heatmap_surv_3_york <- ggplot(surv_3_york_long, aes(Var1, Var2, fill = Freq)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", 
    na.value = "grey50", 
    guide = guide_colorbar(label.theme = element_text(angle = 45, size = 12))  # legend tick labels
  ) +
  labs(title = "", 
       x = "Treatment group", y = "Treatment group",   fill = "Euclidean distance\nin survival") +  
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 14),  # x tick labels
    axis.text.y = element_text(size = 14),                        # y tick labels
    axis.title.x = element_text(size = 14),                       # x-axis label
    axis.title.y = element_text(size = 14),                       # y-axis label
    text = element_text(color = "black"),
    legend.position = "bottom"
  )


ggsave(file.path("..", "..", "figures", "pairwise_differences_analyses", "fig_len_lewisetta_heatmap.pdf"), heatmap_len_3_lew, width = 12, height = 10, dpi = 300)

ggsave(file.path("..", "..", "figures", "pairwise_differences_analyses", "fig_len_york_heatmap.pdf"), heatmap_len_3_york, width = 12, height = 10, dpi = 300)

ggsave(file.path("..", "..", "figures", "pairwise_differences_analyses", "fig_surv_lewisetta_heatmap.pdf"), heatmap_surv_3_lew, width = 12, height = 10, dpi = 300)

ggsave(file.path("..", "..", "figures", "pairwise_differences_analyses", "fig_surv_york_heatmap.pdf"), heatmap_surv_3_york, width = 12, height = 10, dpi = 300)


```


