---
title: "Envr_Performance_Mantel_Tests"
output: pdf_document
date: "2025-01-23"
---

This code analyzes the effect of pairwise environmental distance between conditions at the environments-of-origin on pairwise differences in survival and shell length for oysters in the hatchery and nursery at VIMS ABC.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r packages}
library(tidyverse)
library(ggplot2)
library(gridExtra)
library(multcompView) #extract significance letters from Tukey-Kramer test
library(tidyverse)
library(grid)#for ggplot arranging
library(cowplot) #arrange ggplots
library(ggpmisc)
library(vegan) #for Mantel tests
```

#read in necessary files
```{r}
#this file contains raw lengths of all oysters aged 15-78 days
length <- read.csv("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/latestage_length.csv")

#this file has summary statistics on temperature and salinity for the 8 sites (DEBY, LOLA, JR, TX, LA, FL, NH, ME)
envr_summary <-  read.csv("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/NM_envr_for_violinplots.csv")

#this file is a matrix of the environmental distances between each site
envr_dist <- read.csv("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/envr_raw_data/NM_envr_dist.csv")

#this file contains survival rate data for larval oysters up to day 21 post hatching
survival <-  read.csv("/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/CViMVP_larvae_survival.csv")
```

Extract survival rates from survival data frame, and reorganize data frame
```{r}
summary(survival)

#For survival rates, we only care about survival from day 21 (marked as day_an 15-21), so filter for those day_an only. Reminder that days 15-21 were grouped for analysis because there were constant larval drops occurring during that time, but the number of drops varied by site. Also exclude LARMIX, since there is no environmental data for that, given that it represents oysters from multiple sites
survival_day21 <- survival %>% 
  filter(Day_analysis == "15-21") %>% 
  filter(Tank_naming != "MVP-LARMIX")

#Exclude any groups with "small" in their label. The length data only considers the eyed oysters.
#use grep function to make a vector of all indices in latestage_len with the phrase "small" in them, ignore.case=TRUE means query is not case sensitive
small <- grep("small", survival_day21$Group_Day_Label, ignore.case = TRUE)

#make new dataframe WITHOUT any indices saved in small, mean remaining data are everything except small larvae
survival_day21_sub <- survival_day21[-small, ]

#exclude "MVP" from group names to shorten them
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-DEBY"] <- "DEBY"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"]  == "MVP-FL"] <- "FL"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-JR"] <- "JR"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LA"] <- "LA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-LOLA"] <- "LOLA"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-ME"] <- "ME"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-NH"] <- "NH"
survival_day21_sub["Tank_naming"][survival_day21_sub["Tank_naming"] == "MVP-TX"] <- "TX"

#select only tank naming and survival rate columns
survival_filter <- survival_day21_sub[, c("Tank_naming", "Survival_rate_perc")]

#rename columns
colnames(survival_filter) <- c("site_name", "percent_survival")
```

#length df organizing
```{r}
#select columns with group, day, and shell length in mm
length <- length[,c("group", "day_an", "shell_length_mm")]

#filter out LARMIX from group
length_filter <- length %>% 
  filter(group != "LARMIX")

#keep only data from days 15-21
length_21 <- length_filter %>% 
  filter(day_an == "15-21")

length_78 <- length_filter %>% 
  filter(day_an == "78")

#find mean shell lengths of each group on each day
means_mm_21 <- aggregate(shell_length_mm ~ group, FUN = mean, data = length_21)
colnames(means_mm_21) <- c("site_name", "mean_length_mm_21")

means_mm_78 <- aggregate(shell_length_mm ~ group, FUN = mean, data = length_78)
colnames(means_mm_78) <- c("site_name", "mean_length_mm_78")
```

#remove extra columns from environmental data frame
```{r}
#remove extra site label and standard deviations from envr_summary
envr <- envr_summary[,-c(2,6,10)]

summary(envr)

#rename cols
colnames(envr) <- c("site_name", "Mean_Annual_Temperature_C", "Mean_max_temperature_C", "Mean_min_temperature_C", "Mean_Annual_Salinity_ppt", "Mean_max_Salinity_ppt", "Mean_min_Salinity_ppt")

#add sampling site latitudes. Note that these latitudes ARE NOT the same as the latitudes of where environmental data came from. These latitudes, while similar, were often slightly different, and I am using our actual sampling site latitudes. These latitudes are the same ones as those in table 1 in my thesis.
envr$lat <- c(37.249107, 37.98030, 37.1501163, 28.096000, 29.239925, 30.440030, 43.053746, 44.01330)
```

#Join data frames
```{r}
surv_length_envr <- left_join(envr, means_mm_21, by = "site_name") %>% 
  left_join(means_mm_78, by = "site_name") %>% 
  left_join(survival_filter, by = "site_name")

write.csv(surv_length_envr, "/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/data/Survival_Length_Envr_Data.csv", row.names = FALSE)
```

Make distance matrices for difference in mean length across groups at days 21 and 78 and difference in survival across groups at day 21
```{r}
#save vector of row and column names
rownames <- c("DEBY", "LOLA", "JR", "TX", "LA", "FL", "NH", "ME")
colnames <- c("DEBY", "LOLA", "JR", "TX", "LA", "FL", "NH", "ME")

means_mm_21 <- subset(means_mm_21, select = -c(site_name))

day21_len_dist <- as.matrix(dist(means_mm_21$mean_length_mm_21))
rownames(day21_len_dist) <- rownames
colnames(day21_len_dist) <- colnames

day78_len_dist <- as.matrix(dist(means_mm_78$mean_length_mm_78))
rownames(day78_len_dist) <- rownames
colnames(day78_len_dist) <- colnames

day21_surv_dist <- as.matrix(dist(survival_filter$percent_survival))
rownames(day21_surv_dist) <- rownames
colnames(day21_surv_dist) <- colnames

#remove site_name column from envr_dist, and make it a matrix
envr_dist <- subset(envr_dist, select = -c(X))
envr_dist_mat <- as.matrix(envr_dist)
```

I will run Mantel tests comparing days 21 and 78 length distance and day 21 survival distance to environmental distance. The goal is to see if more environmentally distant pairs of groups also have greater differences in length and survival. 

H0: There is no correlation between the environmental distance matrix values and the length or survival matrix values. Environmental distance and length/survival do not vary with each other in any predictable way

#Mantel test
```{r}
length21_envr = mantel(day21_len_dist, envr_dist_mat, method = "spearman", permutations = 10000, na.rm = TRUE)
length21_envr
#mantel r = 0.3054, p = 0.078492. As environmental distance between a pair of sights increases, so does the distance between their mean shell length at day 21. This is not significant.

length78_envr = mantel(day78_len_dist, envr_dist_mat, method = "spearman", permutations = 10000, na.rm = TRUE)
length78_envr 
#mantel r = -0.1653. p = 0.73843. As environmental distance between a pair of sights increases, the distance between their mean shell length at day 78 decreases. This is not significant.

surv21_envr = mantel(day21_surv_dist, envr_dist_mat, method = "spearman", permutations = 10000, na.rm = TRUE)
surv21_envr
#mantel r = -0.2299, p = 0.50545
```

#Correct Mantel p-values
```{r}
mantel_p <- c(0.078492, 0.73843, 0.50545)

mantel_p_adj <- p.adjust(mantel_p, method = "BH")

mantel_p_adj

#length21_envr corrected p = 0.235476, length78_envr corrected p = 0.738430, surv21_envr corrected p = 0.738430
```

```{r}
#make function to wrap title text
wrapper <- function(x, ...) 
{
  paste(strwrap(x, ...), collapse = "\n")
}

```

#Plot Mantel test results
```{r}
envr_pairwise_dist <- as.vector(envr_dist_mat)

day21_pairwise_len_dist <- as.vector(day21_len_dist)

day78_pairwise_len_dist <- as.vector(day78_len_dist)

day21_pairwise_surv_dist <- as.vector(day21_surv_dist)

pairwise_dist <- as.data.frame(cbind(envr_pairwise_dist, day21_pairwise_len_dist, day78_pairwise_len_dist, day21_pairwise_surv_dist))

length21_envr_dist <- pairwise_dist %>%
  ggplot(aes(x = envr_pairwise_dist, y = day21_pairwise_len_dist))+
  geom_point()+
  labs(subtitle = "A: p = 0.078, corrected p = 0.235, r = 0.31")+
  ylab(wrapper("Difference in Day 21 Length (mm)", width = 20))+
  theme(axis.title.y = element_text(size = 9), axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(), plot.subtitle = element_text(size = 10), plot.title= element_blank())

length21_envr_dist

length78_envr_dist <- pairwise_dist %>%
  ggplot(aes(x = envr_pairwise_dist, y = day78_pairwise_len_dist))+
  geom_point()+
  labs(subtitle = "B: p = 0.738, corrected p = 0.738, r = -0.17")+
  ylab(wrapper("Difference in Day 78 Length (mm)", width = 20))+
  theme(axis.title.y = element_text(size = 9), axis.title.x = element_blank(), axis.text.x = element_blank(), axis.ticks.x = element_blank(), plot.subtitle = element_text(size = 10))

length78_envr_dist

surv_envr_dist <- pairwise_dist %>%
  ggplot(aes(x = envr_pairwise_dist, y = day21_pairwise_surv_dist))+
  geom_point()+
  labs(subtitle = "C: p = 0.505, corrected p = 0.738, r = -0.23")+
  ylab(wrapper("Difference in Survival (%)", width = 20))+
  xlab("Environmental Distance")+
  theme(axis.title.y = element_text(size = 9), axis.title.x = element_text(size = 9), plot.subtitle = element_text(size = 10))

surv_envr_dist

mantel_plots <- plot_grid(length21_envr_dist, length78_envr_dist, surv_envr_dist, ncol = 1, align = "v")

mantel_plots

ggsave("mantel_plots.png",
       plot = mantel_plots,
       device = png,
       path = "/Users/nicolemongillo/Desktop/GitHub/MVP_Chesapeake_VIMS_hatchery/results/NM_results/Result_plots")
```
