---
title: "Envr_Performance_Mantel_Tests"
author: "Kiran Bajaj"
output: pdf_document
date: "2025-01-23"
---

This code analyzes the effect of pairwise environmental distance between conditions at the environments-of-origin and pairwise genetic distance between parent populations on pairwise differences in survival and shell length.

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
#setwd("~/Desktop/MVP-H2F-HatcheryField")
```

### Packages

```{r packages}

library(tidyverse)
library(ggplot2)
library(grid)
library(reshape2)

```

### Load data

```{r}

# raw lengths of all oysters aged 15-78 days
hatchery_length <- read.csv(file.path("..", "..", "data", "performance_H2F", "latestage_length.csv"))

# survival rate data for larval oysters up to day 21 post hatching
hatchery_survival <-  read.csv(file.path("..", "..", "data", "performance_H2F", "CViMVP_larvae_survival.csv"))

# lengths of all oysters averaged by bag for field monitoring events 1-3
field_length <- read.csv(file.path("..", "..", "data", "performance_H2F", "lengths_bags_022025.csv"))

# summary statistics - temperature and salinity quantiles for the 8 groups 
envr_summary <-  read.csv(file.path("..", "..", "data", "envr_of_origin", "envr_quantiles_summary.csv"))

# matrix of euclidean environmental distances using SCALED data
envr_dist <- readRDS(file.path("..", "..", "data", "envr_of_origin", "envr_quantiles_dist.rds"))

# matrix of genetic distances between each group
fst_mat <- readRDS(file.path("..", "..", "data", "pairwise_differences_analyses", "fst_results.rds"))

```

### Global variable

Population order

```{r}

# save vector of population order
population_order <- c("W1-TX", "W2-LA", "W3-FL", "W4-VA", "S1-LOLA", "S2-DEBY", "W5-NH", "W6-ME")

```

### Data preparation

#### Hatchery survival

For hatchery survival rate, we used survival from day 21, so filter for those values only. Days 15-21 were grouped for analysis because there were constant larval drops occurring during that time, but the number of drops varied by site. Also exclude LARMIX, since there is no environmental data for group, given that it represents oysters from multiple sites. 

```{r}

# filter for day 21, exclude LARMIX, and recode site names
survival_day21 <- hatchery_survival %>% 
  filter(Day_analysis == "15-21") %>% 
  filter(Tank_naming != "MVP-LARMIX") %>%
  mutate(Tank_naming = recode(Tank_naming,
                           "MVP-DEBY" = "S2-DEBY",
                           "MVP-FL" = "W3-FL",
                           "MVP-LA" = "W2-LA",
                           "MVP-LOLA" = "S1-LOLA",
                           "MVP-ME" = "W6-ME",
                           "MVP-NH" = "W5-NH",
                           "MVP-TX" = "W1-TX",
                           "MVP-JR" = "W4-VA"))

# exclude any groups with "small" in their label, we only want to analyze eyed oysters
surv_day21 <- survival_day21 %>%
  filter(!grepl("small", Group_Day_Label, ignore.case = TRUE)) %>%
  select(Tank_naming, Survival_rate_perc) %>%
  arrange(factor(Tank_naming, levels = population_order)) # reorder

# rename columns
colnames(surv_day21) <- c("site_name", "surv_21")
```

#### Hatchery length

```{r}

# exclude LARMIX
hatchery_length_filter <- hatchery_length %>%
  select(site_name, day_an, shell_length_mm)  %>% 
  filter(site_name != "H1-LARMIX") # filter out LARMIX

# data frame for days 15-21
length_21 <- hatchery_length_filter %>% 
  filter(day_an == "15-21")

# data frame for day 78
length_78 <- hatchery_length_filter %>% 
  filter(day_an == "78")

# calc mean shell lengths of each group on each day
means_mm_21 <- aggregate(shell_length_mm ~ site_name, FUN = mean, data = length_21)
colnames(means_mm_21) <- c("site_name", "mean_length_mm_21")
means_mm_21 <- means_mm_21 %>% arrange(factor(site_name, levels = population_order)) # reorder

means_mm_78 <- aggregate(shell_length_mm ~ site_name, FUN = mean, data = length_78)
colnames(means_mm_78) <- c("site_name", "mean_length_mm_78")
means_mm_78 <- means_mm_78 %>% arrange(factor(site_name, levels = population_order)) # reorder

```

#### Field length

The goal is to create 6 data frames: one for each monitoring event (3) and each field site (2) with
the average length across all bags for each group.

```{r}

# separate bag labels, recode sites, and average by bag and field site
field_length_clean <- field_length %>%
  separate(bag, into = c("site_name", "bag_number"), sep = "-(?=\\d+$)") %>%
  filter(!site_name %in% c("MVP-SEEDMIX", "MVP-LARMIX")) %>% # remove LARMIX and SEEDMIX
  mutate(site_name = recode(site_name,
                           "MVP-DEBY" = "S2-DEBY",
                           "MVP-FL" = "W3-FL",
                           "MVP-LA" = "W2-LA",
                           "MVP-LOLA" = "S1-LOLA",
                           "MVP-ME" = "W6-ME",
                           "MVP-NH" = "W5-NH",
                           "MVP-TX" = "W1-TX",
                           "MVP-JR" = "W4-VA")) 

# average lengths across all bags per group 
# within-bag lengths are are already averaged
# average separately for each monitoring event and field site
field_length_summary <- field_length_clean %>%
  group_by(monitoring_event, site, site_name) %>%
  summarise(mean_length = mean(length, na.rm = TRUE), .groups = "drop")

# split the data frame by monitoring event and site
split_list <- field_length_summary %>%
  group_split(monitoring_event, site)

# give each data frame a name according to monitoring event and field site
split_names <- field_length_summary %>%
  group_keys(monitoring_event, site) %>%
  mutate(name = paste0("length_event", monitoring_event, "_", gsub(" ", "", tolower(site)))) %>%
  pull(name)

split_list <- lapply(split_list, function(df) {
  df %>%
    arrange(factor(site_name, levels = population_order)) %>% # reorder
    select(site_name, mean_length) # remove columns except for population and length
  })

# assign each data frame to a variable
names(split_list) <- split_names
list2env(split_list, envir = .GlobalEnv)

```

#### Environment-of-origin

```{r}

# remove extra column from environmental data summary table
envr <- envr_summary %>% select(-X)

# add sampling site latitudes. from table 1 of Nicole thesis

envr$lat <- c(28.096000, 29.239925, 30.440030, 37.1501163, 37.98030, 37.249107, 43.053746, 44.01330)

```

#### Data product: hatchery survival, hatchery length, environment-of-origin

```{r}

surv_length_envr <- left_join(envr, means_mm_21, by = "site_name") %>% 
  left_join(means_mm_78, by = "site_name") %>% 
  left_join(survival_day21_filter, by = "site_name")

# save csv for further analyses
write.csv(surv_length_envr, file.path("..", "..", "data", "parent_effects_H2F", "Survival_Length_Envr_Data.csv"), row.names = FALSE)

```

#### Fst

```{r}

# reorder
fst_mat <- fst_mat[population_order, population_order]

```

### Distance matrices

Make distance matrices for all survival and length variables

```{r}

# list of all the data frames that need to be made into distance matrices
df_list <- list(
  surv_day21 = surv_day21,
  len_21 = means_mm_21,
  len78 = means_mm_78,
  len_1_lew = length_event1_lewisetta,
  len_1_york = length_event1_yorkriver,
  len_2_lew = length_event2_lewisetta,
  len_2_york = length_event2_yorkriver,
  len_3_lew = length_event3_lewisetta,
  len_3_york = length_event3_yorkriver
)

# empty list to store distance matrices
dist_matrices <- list()

# loop through each data frame in the list
for (name in names(df_list)) {
  df <- df_list[[name]]
  
  # set the second column as values (first column is population names)
  values <- df[[2]]

  # create distance matrix
  dist_matrix <- as.matrix(dist(values))
  
  # set row and col names
  rownames(dist_matrix) <- population_order
  colnames(dist_matrix) <- population_order
  
  # save into the list
  dist_matrices[[name]] <- dist_matrix
}

```

#### Data check

```{r}

check_matrix_order <- function(matrices, population_order) {
  all_good <- TRUE
  for (i in seq_along(matrices)) {
    if (!identical(rownames(matrices[[i]]), population_order)) {
      message(sprintf("Row names do not match in matrix %d", i))
      all_good <- FALSE
    }
    if (!identical(colnames(matrices[[i]]), population_order)) {
      message(sprintf("Column names do not match in matrix %d", i))
      all_good <- FALSE
    }
  }
  if (all_good) {
    message("All matrices have matching row and column names.")
  }
}

check_matrix_order(dist_matrices, population_order)

```

### Partial Mantel tests

Effect of pairwise environmental distance when controlling for pairwise genetic difference, and vice certa. 

#### Assumptions

```{r}

# Linearity

for (name in names(dist_matrices)) {
  mat <- dist_matrices[[name]]

  # Plot against fst_mat
  plot(as.vector(as.matrix(fst_mat)), as.vector(as.matrix(mat)),
    main = paste("FST vs", name),
    xlab = "FST",
    ylab = paste(name, "Response variable distance")
  )

  # Plot against envr_dist
  plot(
    as.vector(as.matrix(envr_dist)), as.vector(as.matrix(mat)),
    main = paste("Environment vs", name),
    xlab = "Environmental distance",
    ylab = paste(name, "Response variable distance")
  )
}

```

#### Partial tests

Kiran - the data is all prepped. Just need to figure out a way to run these with loops and adjusting for multiple tests. Should be okay.

To test the effect of genetic distance while controlling for environmental distance.

```{r}

# test for the effect of genetic distance on size/survival, controlling for environmental distance

# length on day 21
partial_mantel_gen_len_21 <- mantel.partial(day21_len_dist, fst_mat, envr_dist, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_gen_len_21

# length on day 78
partial_mantel_gen_len_78 <- mantel.partial(day78_len_dist, fst_mat, envr_dist, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_gen_len_78

# survival on day 21
partial_mantel_gen_surv_21 <- mantel.partial(day21_surv_dist, fst_mat, envr_dist, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_gen_surv_21

# collect p-values
partial_mantel_gen_p <- c(0.46275, 0.44756, 0.13309)

# adjust p-vaules to correct for multiple tests
partial_mantel_gen_p_adj <- p.adjust(partial_mantel_gen_p, method = "BH")

# print results
partial_mantel_gen_p_adj

```

To test the effect of environmental distance while controlling for genetic distance.

```{r}

# test for the effect of environmental distance on size/survival, controlling for genetic distance

# length on day 21
partial_mantel_envr_len_21 <- mantel.partial(day21_len_dist, envr_dist, fst_mat, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_envr_len_21

# length on day 78
partial_mantel_envr_len_78 <- mantel.partial(day78_len_dist, envr_dist, fst_mat, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_envr_len_78

# survival on day 21
partial_mantel_envr_surv_21 <- mantel.partial(day21_surv_dist, envr_dist, fst_mat, method = "spearman", permutations = 10000, na.rm = TRUE)
partial_mantel_envr_surv_21

# collect p-values
partial_mantel_envr_p <- c(0.082192, 0.92851, 0.25207)

# adjust p-vaules to correct for multiple tests
partial_mantel_envr_p_adj <- p.adjust(partial_mantel_envr_p, method = "BH")

# print results
partial_mantel_envr_p_adj

```

### Heatmaps

#### Data preparation

```{r}

# function to put all matrices in the same format

process_matrix <- function(matrix_data, population_order) {
  # convert the matrix to long format
  long_data <- melt(matrix_data, na.rm = TRUE)
  
  # edit column names
  colnames(long_data) <- c("Var1", "Var2", "value")
  
  # reorder rows and columns
  long_data <- long_data %>%
    mutate(
      Var1 = factor(Var1, levels = population_order),
      Var2 = factor(Var2, levels = population_order)
    ) %>%
    arrange(Var1, Var2)
  
  return(long_data)
}

# create a vector for the desired order of populations
population_order <- c("W1-TX", "W2-LA", "W3-FL", "S1-LOLA", "S2-DEBY", "W4-VA", "W5-NH", "W6-ME")

# gather all matrices
all_data <- list(fst_mat, envr_dist, day21_len_dist, day78_len_dist, day21_surv_dist)

# apply function to each matrix
processed_data <- lapply(all_data, process_matrix, population_order = population_order)

# Access individual processed data frames
plot1_fst <- processed_data[[1]]
plot2_envr <- processed_data[[2]]
plot3_len21 <- processed_data[[3]]
plot3_len78 <- processed_data[[4]]
plot3_surv21 <- processed_data[[5]]

```

#### Heatmaps

```{r}

heatmap_fst <- ggplot(plot1_fst, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Pairwise Fst", x = "Population", y = "Population", fill = "Fst") + 
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heatmap_envr <- ggplot(plot2_envr, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Environmental distance", x = "Population", y = "Population", fill = "Distance") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heatmap_len21 <- ggplot(plot3_len21, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Length (day 21)", x = "Population", y = "Population", fill = "Length (mm)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heatmap_len78 <- ggplot(plot3_len78, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Length (day 78)", x = "Population", y = "Population", fill = "Length (mm)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

heatmap_surv21 <- ggplot(plot3_surv21, aes(Var1, Var2, fill = value)) +
  geom_tile(color = "white") +
  #geom_text(aes(label = round(value, 3)), size = 2, color = "black") +
  scale_fill_viridis_c(option = "D", na.value = "grey50") +  # Use viridis color scale
  labs(title = "Survival (day 21)", x = "Population", y = "Population", fill = "Survival (%)") +
  theme_minimal() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

# arrange figures in a grid
fig_heatmap_grid <- grid.arrange(
  heatmap_fst, heatmap_envr,
  heatmap_len21, heatmap_len78, heatmap_surv21,
  nrow = 2  # 2 rows
)

ggsave("fig_heatmap_grid.png", fig_heatmap_grid, width = 14, height = 8)

```


